---
title: "effort-risk"
output: html_document
---

```{r}
library(here)
library(tidyverse)
library(tidyboot)
library(lme4)

theme_set(theme_classic(base_size = 15))
```

## effort

```{r}
d.effort <- read_csv(here("data/effort/main_trials.csv")) %>% 
  pivot_longer(
    cols = starts_with("action_"),
    names_to = "action",
    names_prefix = "action_",
    values_to = "likert_rating"
  ) %>%
  mutate(likert_rating = likert_rating + 1, action = as.integer(action))  # ensure action is numeric (0, 1, 2, 3)
```

```{r}
d.effort.summary <- d.effort %>%
  group_by(scenario_label, action) %>%
  tidyboot_mean(likert_rating, na.rm = T)
```

```{r}
write_csv(d.effort.summary, here("data/effort/effort_summary.csv"))
```


## plot by scenario

```{r fig.width=8, fig.height=8}
ggplot(d.effort.summary, aes(x = action, y = mean, color = scenario_label)) +
  geom_line() +
  geom_point() +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = scenario_label), alpha = 0.2, color = NA) +
  scale_x_continuous(breaks = 0:3) +
  scale_y_continuous(breaks = 1:7, limits = c(1, 7)) +
  labs(
    title = "Effort ratings by scenario",
    x = "Action (0 = least risky, 3 = most risky)",
    y = "How effortful?",
    color = "Scenario",
    fill = "Scenario"
  ) + 
  facet_wrap(~ scenario_label) +
  theme(legend.position = "none")
```

## risk 

```{r}
d.risk <- read_csv(here("data/risk/main_trials.csv")) %>%
  pivot_longer(
    cols = starts_with("action_"),
    names_to = "action",
    names_prefix = "action_",
    values_to = "likert_rating"
  ) %>%
  mutate(likert_rating = likert_rating + 1, action = as.integer(action))  
```

```{r}
d.risk.summary <- d.risk %>%
  group_by(scenario_label, action) %>%
  tidyboot_mean(likert_rating)
```

```{r}
write_csv(d.risk.summary, here("data/risk/risk_summary.csv"))
```

```{r fig.width=8, fig.height=8}
ggplot(d.risk.summary, aes(x = action, y = mean, color = scenario_label)) +
  geom_line() +
  geom_point() +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = scenario_label), alpha = 0.2, color = NA) +
  scale_x_continuous(breaks = 0:3) +
  scale_y_continuous(breaks = 1:7, limits = c(1, 7)) +
  labs(
    title = "Risk ratings by scenario",
    x = "Action (0 = least risky, 3 = most risky)",
    y = "How risky?",
    color = "Scenario",
    fill = "Scenario"
  ) + 
  facet_wrap(~ scenario_label) +
  theme(legend.position = "none")
```


## how related are effort and risk? 

For each scenario and action, plot effort vs. risk, shape (or color) is scenario? 

```{r}
d.effort.risk <- d.effort.summary %>%
  rename(effort_mean = mean, effort_ci_lower = ci_lower, effort_ci_upper = ci_upper) %>%
  inner_join(
    d.risk.summary %>%
      rename(risk_mean = mean, risk_ci_lower = ci_lower, risk_ci_upper = ci_upper),
    by = c("scenario_label", "action")
  )
```

```{r fig.width=10, fig.height=8}
ggplot(d.effort.risk, aes(x = risk_mean, y = effort_mean, color = scenario_label, shape = as.factor(action), group = scenario_label)) +
  geom_point(size = 4) +
  geom_errorbarh(aes(xmin = risk_ci_lower, xmax = risk_ci_upper), height = 0.1) +
  geom_errorbar(aes(ymin = effort_ci_lower, ymax = effort_ci_upper), width = 0.1) +
  geom_line() +
  labs(
    x = "Risk rating",
    y = "Effort rating",
    color = "Scenario",
    shape = "Action"
  ) +
  scale_x_continuous(breaks = 1:7, limits = c(1, 7)) +
  scale_y_continuous(breaks = 1:7, limits = c(1, 7)) + 
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  facet_wrap(~ scenario_label) 
```

```{r fig.width=4, fig.height=4}
ggplot(d.effort.risk, aes(x = effort_mean, y = risk_mean, color = scenario_label, shape = as.factor(action), group = scenario_label)) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = risk_ci_lower, ymax = risk_ci_upper), width = 0.1) +
  geom_errorbarh(aes(xmin = effort_ci_lower, xmax = effort_ci_upper), height = 0.1) +
  labs(
    x = "Mean effort rating",
    y = "Mean risk rating",
    color = "Scenario",
    shape = "Action"
  ) +
  scale_x_continuous(breaks = 1:7, limits = c(1, 7)) +
  scale_y_continuous(breaks = 1:7, limits = c(1, 7)) + 
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  theme(legend.position = "none")
```

## Discomfort

```{r}
d.discomfort <- read_csv(here("data/discomfort/main_trials.csv")) %>%
  pivot_longer(
    cols = starts_with("action_"),
    names_to = "action",
    names_prefix = "action_",
    values_to = "likert_rating"
  ) %>%
  mutate(likert_rating = likert_rating + 1, action = as.integer(action), 
         closeness_condition = factor(closeness_condition, levels = c("not_close", "somewhat_close", "close", "extremely_close")), 
         closeness = case_when(
           closeness_condition == "not_close" ~ 0,
           closeness_condition == "somewhat_close" ~ 1,
           closeness_condition == "close" ~ 2,
           closeness_condition == "extremely_close" ~ 3
         )) %>%
  relocate(closeness, .after = closeness_condition)
```
```{r}
d.discomfort.summary.scenario <- d.discomfort %>%
  group_by(scenario_label, closeness_condition, closeness, action) %>%
  tidyboot_mean(likert_rating)
```

```{r}
write_csv(d.discomfort.summary.scenario, here("data/discomfort/discomfort_summary.csv"))
```


Plot

```{r fig.width=8, fig.height=8}
ggplot(d.discomfort.summary.scenario, aes(x = action, y = mean, color = closeness_condition)) +
  geom_line() +
  geom_point() +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = closeness_condition), alpha = 0.2, color = NA) +
  scale_x_continuous(breaks = 0:3) +
  scale_y_continuous(breaks = 1:7, limits = c(1, 7)) +
  labs(
    title = "Discomfort ratings by scenario and closeness condition",
    x = "Action (0 = least risky, 3 = most risky)",
    y = "How uncomfortable?",
    color = "Closeness",
    fill = "Closeness"
  ) + 
  scale_fill_viridis_d(begin = 0, end = 0.85) +
  scale_color_viridis_d(begin = 0, end = 0.85) +
  facet_wrap(~ scenario_label) +
  theme(legend.position = "top")
```


```{r}
d.discomfort.summary <- d.discomfort %>%
  group_by(closeness_condition, action) %>%
  tidyboot_mean(likert_rating)
```

```{r}
ggplot(d.discomfort.summary, aes(x = action, y = mean, color = closeness_condition)) +
  geom_jitter(data = d.discomfort, aes(x = action, y = likert_rating), size = 2.5, shape = 16, alpha = 0.21) +
  geom_line() +
  geom_point(size = 3) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = closeness_condition), alpha = 0.2, color = NA) +
  scale_x_continuous(breaks = 0:3) +
  scale_y_continuous(breaks = 1:7, limits = c(1, 7)) +
  labs(
    title = "Discomfort ratings by closeness condition",
    x = "Action (0 = least risky, 3 = most risky)",
    y = "How uncomfortable?",
    color = "Closeness",
    fill = "Closeness"
  ) + 
  scale_fill_viridis_d(begin = 0, end = 0.85) +
  scale_color_viridis_d(begin = 0, end = 0.85) +
  theme(legend.position = "top")
```

On x axis, plot empirical risk (mean risk rating for that action in that scenario)

```{r}
d.discomfort.risk <- d.discomfort.summary.scenario %>%
  rename(discomfort_mean = empirical_stat, discomfort_ci_lower = ci_lower, discomfort_ci_upper = ci_upper) %>%
  inner_join(
    d.risk.summary %>%
      rename(risk_mean = empirical_stat, risk_ci_lower = ci_lower, risk_ci_upper = ci_upper),
    by = c("scenario_label", "action")
  )
```

Plot risk on x axis, discomfort on y axis

```{r}
ggplot(d.discomfort.risk, aes(x = risk_mean, y = discomfort_mean, color = closeness_condition, shape = as.factor(action), group = closeness_condition)) +
  geom_point(size = 4) +
  geom_errorbarh(aes(xmin = risk_ci_lower, xmax = risk_ci_upper), height = 0.1) +
  geom_errorbar(aes(ymin = discomfort_ci_lower, ymax = discomfort_ci_upper), width = 0.1) +
  geom_line() +
  labs(
    x = "Risk rating",
    y = "Discomfort rating",
    color = "Closeness",
    shape = "Action"
  ) +
  scale_x_continuous(breaks = 1:7, limits = c(1, 7)) +
  scale_y_continuous(breaks = 1:7, limits = c(1, 7)) + 
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  scale_fill_viridis_d(begin = 0, end = 0.85) +
  scale_color_viridis_d(begin = 0, end = 0.85) +
  facet_wrap(~ scenario_label)
```

```{r}
ggplot(d.discomfort.risk, aes(x = risk_mean, y = discomfort_mean, color = closeness_condition, group = as.factor(closeness_condition))) +
  geom_point(aes(shape = factor(action)), size = 4, alpha = 0.3) +
  geom_errorbar(aes(ymin = discomfort_ci_lower, ymax = discomfort_ci_upper), width = 0.1, alpha = 0.3) +
  geom_errorbarh(aes(xmin = risk_ci_lower, xmax = risk_ci_upper), height = 0.1, alpha = 0.3) +
  geom_smooth(method = "gam", alpha = 0.5, fill = "lightgray") +
  labs(
    x = "Risk rating",
    y = "Discomfort rating",
    color = "Closeness",
    shape = "Action"
  ) +
  scale_x_continuous(breaks = 1:7, limits = c(1, 7)) +
  scale_y_continuous(breaks = 1:7, limits = c(1, 7)) + 
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  scale_fill_viridis_d(begin = 0, end = 0.85) +
  scale_color_viridis_d(begin = 0, end = 0.85)
```

Plot effort rating on x axis, discomfort rating on y axis

```{r}
d.discomfort.effort <- d.discomfort.summary.scenario %>%
  rename(discomfort_mean = empirical_stat, discomfort_ci_lower = ci_lower, discomfort_ci_upper = ci_upper) %>%
  inner_join(
    d.effort.summary %>%
      rename(effort_mean = empirical_stat, effort_ci_lower = ci_lower, effort_ci_upper = ci_upper),
    by = c("scenario_label", "action")
  )
```

```{r}
ggplot(d.discomfort.effort, aes(x = effort_mean, y = discomfort_mean, color = closeness_condition, shape = as.factor(action), group = closeness_condition)) +
  geom_point(size = 4) +
  geom_errorbarh(aes(xmin = effort_ci_lower, xmax = effort_ci_upper), height = 0.1) +
  geom_errorbar(aes(ymin = discomfort_ci_lower, ymax = discomfort_ci_upper), width = 0.1) +
  geom_line() +
  labs(
    x = "Effort rating",
    y = "Discomfort rating",
    color = "Closeness",
    shape = "Action"
  ) +
  scale_x_continuous(breaks = 1:7, limits = c(1, 7)) +
  scale_y_continuous(breaks = 1:7, limits = c(1, 7)) + 
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  scale_fill_viridis_d(begin = 0, end = 0.85) +
  scale_color_viridis_d(begin = 0, end = 0.85) +
  facet_wrap(~ scenario_label)
```
```{r}
ggplot(d.discomfort.effort, aes(x = effort_mean, y = discomfort_mean, color = closeness_condition, group = as.factor(closeness_condition))) +
  geom_point(aes(shape = factor(action)), size = 4, alpha = 0.3) +
  geom_errorbar(aes(ymin = discomfort_ci_lower, ymax = discomfort_ci_upper), width = 0.1, alpha = 0.3) +
  geom_errorbarh(aes(xmin = effort_ci_lower, xmax = effort_ci_upper), height = 0.1, alpha = 0.3) +
  geom_smooth(method = "gam", alpha = 0.5, fill = "lightgray") +
  labs(
    x = "Effort rating",
    y = "Discomfort rating",
    color = "Closeness",
    shape = "Action"
  ) +
  scale_x_continuous(breaks = 1:7, limits = c(1, 7)) +
  scale_y_continuous(breaks = 1:7, limits = c(1, 7)) + 
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  scale_fill_viridis_d(begin = 0, end = 0.85) +
  scale_color_viridis_d(begin = 0, end = 0.85)
```

one reason this might look like this is because effort also correlates with discomfort in the data.

So we want to see, how much does effort explain discomfort ratings, above and beyond risk?

```{r}
d.discomfort.effort.risk <- d.discomfort.effort %>% 
  inner_join(
    d.risk.summary %>%
      rename(risk_mean = empirical_stat, risk_ci_lower = ci_lower, risk_ci_upper = ci_upper),
    by = c("scenario_label", "action")
  ) %>% mutate(
    closeness = case_when(
      closeness_condition == "not_close" ~ 0,
      closeness_condition == "somewhat_close" ~ 1,
      closeness_condition == "close" ~ 2,
      closeness_condition == "extremely_close" ~ 3
    )
  )
```

Null hypothesis: discomfort is based on risk of saliva sharing action

```{r}
m0 <- lm(discomfort_mean ~ risk_mean, data = d.discomfort.effort.risk)
summary(m0)
```
Alternative hypothesis: discomfort is based on risk of saliva sharing action modulated by closeness of relationship

```{r}
m1 <- lm(discomfort_mean ~ risk_mean * closeness, data = d.discomfort.effort.risk)
summary(m1)
```
which one is better? 

```{r}
anova(m0, m1)
```
does adding effort help? nope... 

```{r}
m2 <- lm(discomfort_mean ~ risk_mean * closeness * effort_mean, data = d.discomfort.effort.risk)
summary(m2)
anova(m1, m2)
```


## PLANNING STUDY!!

```{r}
d.planning <- read_csv(here("data/planning-1/main_trials.csv")) %>%
  pivot_longer(
    cols = starts_with("action_"),
    names_to = "action",
    names_prefix = "action_",
    values_to = "likert_rating"
  ) %>%
  mutate(likert_rating = likert_rating + 1, action = as.integer(action), 
         closeness_condition = factor(closeness_condition, levels = c("not_close", "somewhat_close", "close", "extremely_close")), 
         closeness = case_when(
           closeness_condition == "not_close" ~ 0,
           closeness_condition == "somewhat_close" ~ 1,
           closeness_condition == "close" ~ 2,
           closeness_condition == "extremely_close" ~ 3
         )) %>% 
  relocate(closeness, .after = closeness_condition) %>%
  group_by(subject_id, scenario_label) %>%
  mutate(p_action = likert_rating / sum(likert_rating)) %>%
  ungroup()

write_csv(d.planning, here("data/planning-1/main_trials_tidy.csv"))
```

```{r}
d.planning.summary.scenario <- d.planning %>%
  group_by(scenario_label, closeness_condition, action) %>%
  tidyboot_mean(likert_rating)
```

```{r}
d.planning.summary <- d.planning %>%
  group_by(closeness_condition, action) %>%
  tidyboot_mean(likert_rating)
```

```{r}
ggplot(d.planning.summary, aes(x = action, y = mean, color = closeness_condition)) +
  geom_jitter(data = d.planning, aes(x = action, y = likert_rating), size = 2.5, shape = 16, alpha = 0.1) +
  geom_line() +
  geom_point(size = 3) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = closeness_condition), alpha = 0.2, color = NA) +
  scale_x_continuous(breaks = 0:3) +
  scale_y_continuous(breaks = 1:7, limits = c(1, 7)) +
  labs(
    title = "No communication",
    x = "Action (0 = least risky, 3 = most risky)",
    y = "How likely?",
    color = "Closeness",
    fill = "Closeness"
  ) + 
  scale_fill_viridis_d(begin = 0, end = 0.85) +
  scale_color_viridis_d(begin = 0, end = 0.85) +
  theme(legend.position = "top")
```
Put empirical measured risk on x axis

```{r}
d.planning.risk <- d.planning.summary.scenario %>%
  rename(planning_mean = empirical_stat, planning_ci_lower = ci_lower, planning_ci_upper = ci_upper) %>%
  inner_join(
    d.risk.summary %>%
      rename(risk_mean = empirical_stat, risk_ci_lower = ci_lower, risk_ci_upper = ci_upper),
    by = c("scenario_label", "action")
  )
```

```{r}
ggplot(d.planning.risk, aes(x = risk_mean, y = planning_mean, color = closeness_condition, group = closeness_condition)) +
  geom_point(size = 4, aes(shape = as.factor(action)), alpha = 0.3) +
  geom_errorbarh(aes(xmin = risk_ci_lower, xmax = risk_ci_upper), height = 0.1, alpha = 0.3) +
  geom_errorbar(aes(ymin = planning_ci_lower, ymax = planning_ci_upper), width = 0.1, alpha = 0.3) +
  geom_smooth(method = "gam", alpha = 0.5, fill = "lightgray") +
  labs(
    x = "Risk rating",
    y = "How likely to choose action?",
    title = "No communication",
    color = "Closeness",
    shape = "Action"
  ) +
  scale_x_continuous(breaks = 1:7, limits = c(1, 7)) +
  scale_y_continuous(breaks = 1:7, limits = c(1, 7)) + 
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  scale_fill_viridis_d(begin = 0, end = 0.85) +
  scale_color_viridis_d(begin = 0, end = 0.85) 
```

look at scenario wide

```{r}
ggplot(d.planning.risk, aes(x = risk_mean, y = planning_mean, color = closeness_condition, group = closeness_condition)) +
  geom_point(size = 4, aes(shape = as.factor(action)), alpha = 0.3) +
  geom_errorbarh(aes(xmin = risk_ci_lower, xmax = risk_ci_upper), height = 0.1, alpha = 0.3) +
  geom_errorbar(aes(ymin = planning_ci_lower, ymax = planning_ci_upper), width = 0.1, alpha = 0.3) +
  geom_line()+
  labs(
    x = "Risk rating",
    y = "How likely to choose action?",
    title = "No communication",
    color = "Closeness",
    shape = "Action"
  ) +
  scale_x_continuous(breaks = 1:7, limits = c(1, 7)) +
  scale_y_continuous(breaks = 1:7, limits = c(1, 7)) + 
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  scale_fill_viridis_d(begin = 0, end = 0.85) +
  scale_color_viridis_d(begin = 0, end = 0.85) + 
  facet_wrap(~ scenario_label)
```

## What are people's priors for actions? 

```{r}
d.priors <- read_csv(here("data/planning_priors/main_trials.csv")) %>%
  pivot_longer(
    cols = starts_with("action_"),
    names_to = "action",
    names_prefix = "action_",
    values_to = "likert_rating"
  ) %>%
  mutate(likert_rating = likert_rating + 1, action = as.integer(action)) %>% 
  group_by(subject_id, scenario_label) %>%
  mutate(p_action = likert_rating / sum(likert_rating)) %>%
  ungroup()
```

```{r}
d.priors.summary <- d.priors %>%
  group_by(scenario_label, action) %>%
  tidyboot_mean(p_action, na.rm = T)
```

```{r}
write_csv(d.priors.summary, here("data/planning_priors/priors_summary.csv"))
```

Plot priors

```{r}
ggplot(d.priors.summary, aes(x = action, y = empirical_stat, color = scenario_label)) +
  geom_line() +
  geom_point(size = 3) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = scenario_label), alpha = 0.2, color = NA) +
  scale_x_continuous(breaks = 0:3) +
  scale_y_continuous(limits = c(0, 0.7)) +
  labs(
    title = "Priors for actions",
    x = "Action (0 = least risky, 3 = most risky)",
    y = "How likely?",
    color = "Scenario",
    fill = "Scenario"
  ) + 
  theme(legend.position = "none") +
  facet_wrap(~ scenario_label)
```



## View model predictions

```{r}
d.with.preds <- read_csv(here("data/planning-1/main_trials_tidy_with_preds.csv")) %>% 
  mutate(closeness_condition = factor(closeness_condition, levels = c("not_close", "somewhat_close", "close", "extremely_close"))) %>% 
  left_join(d.risk.summary, by = c("scenario_label", "action")) %>%
  rename(risk_rating = empirical_stat, ci_lower_risk = ci_lower, ci_upper_risk = ci_upper)
```

Aggregate over all risk levels for each action setting

```{r}
d.data.summary <- d.with.preds %>%
  group_by(closeness_condition, action) %>%
  tidyboot_mean(p_action)

d.vanilla.summary <- d.with.preds %>%
  group_by(closeness_condition, action) %>%
  tidyboot_mean(vanilla_pred)

d.relationship.summary <- d.with.preds %>%
  group_by(closeness_condition, action) %>%
  tidyboot_mean(relationship_pred)
```

Scenario-specific

```{r}
d.data.summary.scenario <- d.with.preds %>%
  group_by(scenario_label, closeness_condition, action, risk_rating, ci_lower_risk, ci_upper_risk) %>%
  tidyboot_mean(p_action)

d.vanilla.summary.scenario <- d.with.preds %>%
  group_by(scenario_label, closeness_condition, action, risk_rating, ci_lower_risk, ci_upper_risk) %>%
  tidyboot_mean(vanilla_pred)
```

```{r}
d.relationship.summary.scenario <- d.with.preds %>%
  group_by(scenario_label, closeness_condition, action, risk_rating, ci_lower_risk, ci_upper_risk) %>%
  tidyboot_mean(relationship_pred)
```

### Big plots 

```{r}
ggplot(d.data.summary.scenario, aes(x = risk_rating, y = empirical_stat, color = closeness_condition, group = closeness_condition)) +
  geom_point(size = 4, aes(shape = as.factor(action)), alpha = 0.3, position = position_dodge(width = 0.1)) +
  geom_errorbarh(aes(xmin = ci_lower_risk, xmax = ci_upper_risk), height = 0.015, alpha = 0.3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.1, alpha = 0.3) +
  geom_smooth(method = "gam", alpha = 0.5, fill = "lightgray") +
  labs(
    x = "Risk rating",
    y = "How likely to choose action?",
    title = "Data",
    color = "Closeness",
    shape = "Action"
  ) +
  scale_x_continuous(breaks = 1:7, limits = c(0.9, 6.4)) +
  scale_y_continuous(limits = c(0, 0.65)) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  scale_fill_viridis_d(begin = 0, end = 0.85) +
  scale_color_viridis_d(begin = 0, end = 0.85) 
```

```{r}
ggplot(d.vanilla.summary.scenario, aes(x = risk_rating, y = empirical_stat, color = closeness_condition, group = closeness_condition)) +
  geom_point(size = 4, aes(shape = as.factor(action)), alpha = 0.3, position = position_dodge(width = 0.1)) +
  geom_errorbarh(aes(xmin = ci_lower_risk, xmax = ci_upper_risk), height = 0.015, alpha = 0.3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.1, alpha = 0.3) +
  geom_smooth(method = "gam", alpha = 0.5, fill = "lightgray", color = "black") +
  labs(
    x = "Risk rating",
    y = "How likely to choose action?",
    title = "Model: Vanilla",
    color = "Closeness",
    shape = "Action"
  ) +
  scale_x_continuous(breaks = 1:7, limits = c(0.9, 6.4)) +
  scale_y_continuous(limits = c(0, 0.65)) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  scale_fill_viridis_d(begin = 0, end = 0.85) +
  scale_color_viridis_d(begin = 0, end = 0.85) 
```

```{r}
ggplot(d.relationship.summary.scenario, aes(x = risk_rating, y = empirical_stat, color = closeness_condition, group = closeness_condition)) +
  geom_point(size = 4, aes(shape = as.factor(action)), alpha = 0.3, position = position_dodge(width = 0.1)) +
  geom_errorbarh(aes(xmin = ci_lower_risk, xmax = ci_upper_risk), height = 0.015, alpha = 0.3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.1, alpha = 0.3) +
  geom_smooth(method = "gam", alpha = 0.5, fill = "lightgray") +
  labs(
    x = "Risk rating",
    y = "How likely to choose action?",
    title = "Model: Relationship",
    color = "Closeness",
    shape = "Action"
  ) +
  scale_x_continuous(breaks = 1:7, limits = c(0.9, 6.4)) +
  scale_y_continuous(limits = c(0, 0.65)) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  scale_fill_viridis_d(begin = 0, end = 0.85) +
  scale_color_viridis_d(begin = 0, end = 0.85) 
```

### Scatter plots

Join dataframes

```{r}
d.combined <- d.data.summary.scenario %>%
  rename(data_mean = empirical_stat, data_ci_lower = ci_lower, data_ci_upper = ci_upper) %>%
  inner_join(
    d.vanilla.summary.scenario %>%
      rename(vanilla_mean = empirical_stat, vanilla_ci_lower = ci_lower, vanilla_ci_upper = ci_upper),
    by = c("scenario_label", "closeness_condition", "action")
  ) %>%
  inner_join(
    d.relationship.summary.scenario %>%
      rename(relationship_mean = empirical_stat, relationship_ci_lower = ci_lower, relationship_ci_upper = ci_upper),
    by = c("scenario_label", "closeness_condition", "action")
  ) 
```

Plot: 
- data vs. vanilla
- data vs. relationship


```{r}
ggplot(d.combined, aes(x = vanilla_mean, y = data_mean, color = closeness_condition, shape = as.factor(action), group = closeness_condition)) +
  geom_point(size = 4, alpha = 0.3) +
  geom_errorbar(aes(ymin = data_ci_lower, ymax = data_ci_upper), width = 0.02, alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(
    x = "Model",
    y = "Humans",
    title = "Vanilla",
    color = "Closeness",
    shape = "Action"
  ) +
  scale_x_continuous(breaks = seq(0, 1, by = 0.1), limits = c(0, 0.65)) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.1), limits = c(0, 0.65)) + 
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  scale_fill_viridis_d(begin = 0, end = 0.85) +
  scale_color_viridis_d(begin = 0, end = 0.85) +
  coord_equal()
```

```{r}
ggplot(d.combined, aes(x = relationship_mean, y = data_mean, color = closeness_condition, shape = as.factor(action), group = closeness_condition)) +
  geom_point(size = 4, alpha = 0.3) +
  geom_errorbar(aes(ymin = data_ci_lower, ymax = data_ci_upper), width = 0.02, alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(
    x = "Model",
    y = "Humans",
    title = "Relationship",
    color = "Closeness",
    shape = "Action"
  ) +
  scale_x_continuous(breaks = seq(0, 1, by = 0.1), limits = c(0, 0.65)) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.1), limits = c(0, 0.65)) + 
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  scale_fill_viridis_d(begin = 0, end = 0.85) +
  scale_color_viridis_d(begin = 0, end = 0.85) +
  coord_equal()
```

```{r}
cor.test(d.combined$data_mean, d.combined$vanilla_mean)
cor.test(d.combined$data_mean, d.combined$relationship_mean)
```

What are the scenarios where the model under predicts? 

```{r}
d.combined.resid <- d.combined %>%
  mutate(vanilla_resid = data_mean - vanilla_mean,
         relationship_resid = data_mean - relationship_mean) %>% 
  select(scenario_label, closeness_condition, action, vanilla_resid, relationship_resid) 

ggplot(d.combined.resid, aes(x = scenario_label, y = vanilla_resid, fill = closeness_condition)) +
  geom_bar(stat = "identity", position = position_dodge()) 
```

Plot residuals

```{r}
ggplot(d.combined.resid, aes(x = closeness_condition, y = relationship_resid, color = closeness_condition, shape = as.factor(action))) +
  geom_point(size = 4, alpha = 0.8, position = position_dodge(width = 0.8)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Closeness",
    y = "Residual (data - model)",
    title = "Relationship model residuals",
  ) +
  scale_fill_viridis_d(begin = 0, end = 0.85) +
  scale_color_viridis_d(begin = 0, end = 0.85) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~ scenario_label)
```

Vanilla model residuals

```{r}
ggplot(d.combined.resid, aes(x = closeness_condition, y = vanilla_resid, color = closeness_condition, shape = as.factor(action))) +
  geom_point(size = 4, alpha = 0.8, position = position_dodge(width = 0.8)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Closeness",
    y = "Residual (data - model)",
    title = "Vanilla model residuals",
  ) +
  scale_fill_viridis_d(begin = 0, end = 0.85) +
  scale_color_viridis_d(begin = 0, end = 0.85) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~ scenario_label)
```



# old stuff

DOES NOT RUN RIGHT NOW... todo adapt to previous section with all the different models. 

Scenario specific

```{r}
ggplot(d.data.summary.scenario, aes(x = risk_rating, y = empirical_stat, color = closeness_condition, group = closeness_condition)) +
  geom_point(size = 4, aes(shape = as.factor(action)), alpha = 0.3) +
  geom_errorbarh(aes(xmin = ci_lower_risk, xmax = ci_upper_risk), height = 0.015, alpha = 0.3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.1, alpha = 0.3) +
  geom_smooth(method = "gam", alpha = 0.5, fill = "lightgray") +
  labs(
    x = "Risk rating",
    y = "How likely to choose action?",
    title = "Data",
    color = "Closeness",
    shape = "Action"
  ) +
  scale_x_continuous(breaks = 1:7, limits = c(0.9, 6.4)) +
  scale_y_continuous(limits = c(0, 0.65)) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  scale_fill_viridis_d(begin = 0, end = 0.85) +
  scale_color_viridis_d(begin = 0, end = 0.85) 
```

```{r}
ggplot(d.h0.summary.scenario, aes(x = risk_rating, y = empirical_stat, color = closeness_condition, group = closeness_condition)) +
  geom_point(size = 4, aes(shape = as.factor(action)), alpha = 0.3, position = position_dodge(width = 0.1)) +
  geom_errorbarh(aes(xmin = ci_lower_risk, xmax = ci_upper_risk), height = 0.015, alpha = 0.3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.1, alpha = 0.3) +
  geom_smooth(method = "gam", alpha = 0.5, fill = "lightgray", color = "black") +
  labs(
    x = "Risk rating",
    y = "How likely to choose action?",
    title = "Model: Effort and risk",
    color = "Closeness",
    shape = "Action"
  ) +
  scale_x_continuous(breaks = 1:7, limits = c(0.9, 6.4)) +
  scale_y_continuous(limits = c(0, 0.65)) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  scale_fill_viridis_d(begin = 0, end = 0.85) +
  scale_color_viridis_d(begin = 0, end = 0.85) 
```

```{r}
ggplot(d.h1.summary.scenario, aes(x = risk_rating, y = empirical_stat, color = closeness_condition, group = closeness_condition)) +
  geom_point(size = 4, aes(shape = as.factor(action)), alpha = 0.3) +
  geom_errorbarh(aes(xmin = ci_lower_risk, xmax = ci_upper_risk), height = 0.015, alpha = 0.3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.1, alpha = 0.3) +
  geom_smooth(method = "gam", alpha = 0.5, fill = "lightgray") +
  labs(
    x = "Risk rating",
    y = "How likely to choose action?",
    title = "Model: Effort, risk, and discomfort",
    color = "Closeness",
    shape = "Action"
  ) +
  scale_x_continuous(breaks = 1:7, limits = c(0.9, 6.4)) +
  scale_y_continuous(limits = c(0, 0.65)) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  scale_fill_viridis_d(begin = 0, end = 0.85) +
  scale_color_viridis_d(begin = 0, end = 0.85) 
```

See if model captures scenario-specific patterns

```{r}
ggplot(d.data.summary.scenario, aes(x = risk_rating, y = empirical_stat, color = closeness_condition, group = closeness_condition)) +
  geom_point(size = 4, aes(shape = as.factor(action)), alpha = 0.8) +
  geom_line() +
  geom_errorbarh(aes(xmin = ci_lower_risk, xmax = ci_upper_risk), height = 0.015, alpha = 0.8) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.1, alpha = 0.8) +
  labs(
    x = "Risk rating",
    y = "How likely to choose action?",
    title = "Data",
    color = "Closeness",
    shape = "Action"
  ) +
  scale_x_continuous(breaks = 1:7, limits = c(0.9, 6.4)) +
  scale_y_continuous(limits = c(0, 0.65)) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  scale_fill_viridis_d(begin = 0, end = 0.85) +
  scale_color_viridis_d(begin = 0, end = 0.85) + 
  facet_wrap(~ scenario_label)
```
```{r}
ggplot(d.h0.summary.scenario, aes(x = risk_rating, y = empirical_stat, color = closeness_condition, group = closeness_condition)) +
  geom_point(size = 4, aes(shape = as.factor(action)), alpha = 0.8, position = position_dodge2(width = 0.3)) +
  geom_line(position = position_dodge2(width = 0.3)) +
  geom_errorbarh(aes(xmin = ci_lower_risk, xmax = ci_upper_risk), height = 0.015, alpha = 0.8) +
  labs(
    x = "Risk rating",
    y = "How likely to choose action?",
    title = "Model: effort and risk",
    color = "Closeness",
    shape = "Action"
  ) +
  scale_x_continuous(breaks = 1:7, limits = c(0.9, 6.4)) +
  scale_y_continuous(limits = c(0, 0.65)) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  scale_fill_viridis_d(begin = 0, end = 0.85) +
  scale_color_viridis_d(begin = 0, end = 0.85) + 
  facet_wrap(~ scenario_label)
```
```{r}
ggplot(d.h1.summary.scenario, aes(x = risk_rating, y = empirical_stat, color = closeness_condition, group = closeness_condition)) +
  geom_point(size = 4, aes(shape = as.factor(action)), alpha = 0.8, position = position_dodge2(width = 0.3)) +
  geom_line(position = position_dodge2(width = 0.3)) +
  geom_errorbarh(aes(xmin = ci_lower_risk, xmax = ci_upper_risk), height = 0.015, alpha = 0.8) +
  labs(
    x = "Risk rating",
    y = "How likely to choose action?",
    title = "Model: effort, risk, and discomfort",
    color = "Closeness",
    shape = "Action"
  ) +
  scale_x_continuous(breaks = 1:7, limits = c(0.9, 6.4)) +
  scale_y_continuous(limits = c(0, 0.65)) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  scale_fill_viridis_d(begin = 0, end = 0.85) +
  scale_color_viridis_d(begin = 0, end = 0.85) + 
  facet_wrap(~ scenario_label)
```

Plot scenario specific

```{r}
ggplot(d.data.summary.scenario, aes(x = action, y = empirical_stat, color = closeness_condition)) +
  geom_line() +
  geom_point(size = 3) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = closeness_condition), alpha = 0.2, color = NA) +
  scale_x_continuous(breaks = 0:3) +
  scale_y_continuous(limits = c(0, 0.7)) +
  labs(
    title = "Data: How likely to choose actions?",
    x = "Action (0 = least risky, 3 = most risky)",
    y = "How likely?",
    color = "Closeness",
    fill = "Closeness"
  ) + 
  scale_fill_viridis_d(begin = 0, end = 0.85) +
  scale_color_viridis_d(begin = 0, end = 0.85) +
  facet_wrap(~ scenario_label) +
  theme(legend.position = "top")
```

```{r}
ggplot(d.h0.summary.scenario, aes(x = action, y = empirical_stat, color = closeness_condition)) +
  geom_line(position = position_dodge(width = 0.2)) +
  geom_point(size = 3, position = position_dodge(width = 0.2)) +
  scale_x_continuous(breaks = 0:3) +
  scale_y_continuous(limits = c(0, 0.7)) +
  labs(
    title = "Model H0: Effort and risk",
    x = "Action (0 = least risky, 3 = most risky)",
    y = "How likely?",
    color = "Closeness",
    fill = "Closeness"
  ) + 
  scale_fill_viridis_d(begin = 0, end = 0.85) +
  scale_color_viridis_d(begin = 0, end = 0.85) +
  facet_wrap(~ scenario_label) +
  theme(legend.position = "top")
```

```{r}
ggplot(d.h1.summary.scenario, aes(x = action, y = empirical_stat, color = closeness_condition)) +
  geom_line() +
  geom_point(size = 3) +
  scale_x_continuous(breaks = 0:3) +
  scale_y_continuous(limits = c(0, 0.7)) +
  labs(
    title = "Model H1: Effort, risk, and discomfort",
    x = "Action (0 = least risky, 3 = most risky)",
    y = "How likely?",
    color = "Closeness",
    fill = "Closeness"
  ) + 
  scale_fill_viridis_d(begin = 0, end = 0.85) +
  scale_color_viridis_d(begin = 0, end = 0.85) +
  facet_wrap(~ scenario_label) +
  theme(legend.position = "top")
```


# Communication study


```{r}
d.planning_comm <- read_csv(here("data/planning_comm/main_trials.csv")) %>%
  pivot_longer(
    cols = starts_with("action_"),
    names_to = "action",
    names_prefix = "action_",
    values_to = "likert_rating"
  ) %>%
  mutate(likert_rating = likert_rating + 1, action = as.integer(action), 
         closeness_condition = factor(closeness_condition, levels = c("not_close", "somewhat_close", "close", "extremely_close")), 
         closeness = case_when(
           closeness_condition == "not_close" ~ 0,
           closeness_condition == "somewhat_close" ~ 1,
           closeness_condition == "close" ~ 2,
           closeness_condition == "extremely_close" ~ 3
         )) %>% 
  relocate(closeness, .after = closeness_condition) %>%
  group_by(subject_id, scenario_label) %>%
  mutate(p_action = likert_rating / sum(likert_rating)) %>%
  ungroup()

write_csv(d.planning, here("data/planning_comm/main_trials_tidy.csv"))
```

```{r}
d.planning.comm.summary.scenario <- d.planning_comm %>%
  group_by(scenario_label, closeness_condition, action) %>%
  tidyboot_mean(likert_rating)
```

```{r}
d.planning.comm.summary <- d.planning_comm %>%
  group_by(closeness_condition, action) %>%
  tidyboot_mean(likert_rating)
```

```{r}
ggplot(d.planning.comm.summary, aes(x = action, y = mean, color = closeness_condition)) +
  geom_jitter(data = d.planning, aes(x = action, y = likert_rating), size = 2.5, shape = 16, alpha = 0.1) +
  geom_line() +
  geom_point(size = 3) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = closeness_condition), alpha = 0.2, color = NA) +
  scale_x_continuous(breaks = 0:3) +
  scale_y_continuous(breaks = 1:7, limits = c(1, 7)) +
  labs(
    title = "Communicative goal",
    x = "Action (0 = least risky, 3 = most risky)",
    y = "How likely?",
    color = "Closeness",
    fill = "Closeness"
  ) + 
  scale_fill_viridis_d(begin = 0, end = 0.85) +
  scale_color_viridis_d(begin = 0, end = 0.85) +
  theme(legend.position = "top")
```


Put empirical measured risk on x axis

```{r}
d.planning_comm.risk <- d.planning.comm.summary.scenario %>%
  rename(planning_mean = empirical_stat, planning_ci_lower = ci_lower, planning_ci_upper = ci_upper) %>%
  inner_join(
    d.risk.summary %>%
      rename(risk_mean = empirical_stat, risk_ci_lower = ci_lower, risk_ci_upper = ci_upper),
    by = c("scenario_label", "action")
  )
```

```{r}
ggplot(d.planning_comm.risk, aes(x = risk_mean, y = planning_mean, color = closeness_condition, group = closeness_condition)) +
  geom_point(size = 4, aes(shape = as.factor(action)), alpha = 0.3) +
  geom_errorbarh(aes(xmin = risk_ci_lower, xmax = risk_ci_upper), height = 0.1, alpha = 0.3) +
  geom_errorbar(aes(ymin = planning_ci_lower, ymax = planning_ci_upper), width = 0.1, alpha = 0.3) +
  geom_smooth(method = "gam", alpha = 0.5, fill = "lightgray") +
  labs(
    x = "Risk rating",
    y = "How likely to choose action?",
    title = "Maintain, establish, or communicate about relationship",
    color = "Closeness",
    shape = "Action"
  ) +
  scale_x_continuous(breaks = 1:7, limits = c(1, 7)) +
  scale_y_continuous(breaks = 1:7, limits = c(1, 7)) + 
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  scale_fill_viridis_d(begin = 0, end = 0.85) +
  scale_color_viridis_d(begin = 0, end = 0.85) 
```




