---
title: "2025-10_pilots"
output: html_document
---


```{r}
library(here)
library(tidyverse)
library(tidyboot)
library(lme4)

theme_set(theme_minimal(base_size = 15) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  ))
```

## forw_plan pilot 1

This experiment manipulates how much two people want the food (`low` vs. `high` reward), and the relationship between the two people (0, 50, 75, or 100 on a 100 point scale between 'intimate' and 'formal'). 

```{r}
df_planning <- read_csv(here("data", "forw_plan", "main_trials.csv")) |>
  rename(
    intimacy = intimacy_condition,
    reward = reward_condition
  ) |>
  mutate(
    reward = factor(reward, levels = c("low", "high")),
    model = "humans"
  ) |>
  pivot_longer(
    cols = starts_with("action_"),
    names_to = "action",
    names_prefix = "action_",
    values_to = "p_action"
  )
```
### Look at summary across scenarios

Calculate means

```{r}
df_planning_summary <- df_planning |>
  group_by(intimacy, reward, action) |>
  tidyboot_mean(p_action)
```

Plot

```{r}
ggplot(df_planning_summary, aes(x = action, y = empirical_stat, color = intimacy, group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), size = 1, width = 0.3
  ) +
  geom_line(position = position_dodge(width = 0.2), size = 1) +
  scale_color_viridis_c(begin = 0, end = 0.85) +
  labs(y = "P(action)", x = "Action", color = "Intimacy") +
  theme(legend.position = "bottom") +
  facet_wrap(~reward, labeller = label_both)
```

## Look at toy model predictions

```{r}
df_model_forward <- read_csv(here("model", "sandbox", "actor_preds.csv")) |>
  mutate(
    reward = factor(reward, levels = c("low", "high")),
    action = factor(action)
  )
```

Stack the dataframes

```{r}
df_combined <- bind_rows(df_planning, df_model_forward) |>
  mutate(
    model = case_when(
      model == "humans" ~ "Humans",
      model == "discomfort_only" ~ "Discomfort only",
      model == "vanilla_inv_plan" ~ "Vanilla planning",
      model == "full_model" ~ "Full model"
    )
  ) |>
  mutate(model = factor(model, levels = c(
    "Discomfort only", "Vanilla planning", "Full model", "Humans"
  )))
```

### Summarize combined

```{r}
df_combined_summary <- df_combined |>
  group_by(intimacy, reward, model, action) |>
  tidyboot_mean(p_action)
```
Plot

```{r}
ggplot(df_combined_summary, aes(x = action, y = empirical_stat, color = intimacy, group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), size = 1, width = 0.3
  ) +
  geom_line(position = position_dodge(width = 0.2), size = 1) +
  scale_color_viridis_c(begin = 0, end = 0.85) +
  labs(y = "P(action)", x = "Action", color = "Intimacy", title = "Forward planning") +
  facet_grid(reward ~ model) +
  theme(legend.position = "bottom")
```


# Inverse planning

## Inferring intimacy of relationship, given information about the reward and the action taken

Load data and add belief updates

```{r}
df_inv_intimacy <- read_csv(here("data", "inv_plan_intimacy", "main_trials.csv")) |>
  rename(reward = reward_condition, action = action_condition) |>
  mutate(
    action = str_replace(action, "action_", ""),
    reward = factor(reward, levels = c("low", "high")),
    stage = factor(stage, levels = c("prior", "posterior")),
    model = "humans"
  )

df_inv_intimacy <- df_inv_intimacy |>
  group_by(subject_id, scenario_label) |>
  mutate(belief_update = ifelse(stage == "posterior", intimacy_rating[stage == "posterior"] - intimacy_rating[stage == "prior"], NA)) |>
  ungroup()
```
Plot distributions

```{r}
ggplot(df_inv_intimacy, aes(x = intimacy_rating, fill = action)) +
  geom_density(alpha = 0.5) +
  scale_fill_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  labs(y = "Density", x = "Intimacy rating", fill = "Action") +
  theme(legend.position = "bottom") +
  facet_grid(stage ~ reward, labeller = labeller(reward = c("low" = "motivation: low", "high" = "motivation: high")))
```

Calculate aggregate summary

```{r}
df_inv_intimacy_summary <- df_inv_intimacy |>
  group_by(reward, action, stage, model) |>
  tidyboot_mean(intimacy_rating)
```

Plot priors vs. posteriors for each action and reward condition


```{r}
ggplot() +
  geom_violin(
    data = df_inv_intimacy,
    aes(x = stage, y = intimacy_rating, fill = action),
    position = position_dodge(width = 0.8),
    alpha = 0.5,
    width = 0.7
  ) +
  geom_point(
    data = df_inv_intimacy_summary,
    aes(x = stage, y = empirical_stat, color = action, group = action),
    position = position_dodge(width = 0.8),
    size = 3
  ) +
  geom_errorbar(
    data = df_inv_intimacy_summary,
    aes(x = stage, ymin = ci_lower, ymax = ci_upper, color = action, group = action),
    position = position_dodge(width = 0.8),
    width = 0.1,
    size = 1
  ) +
  geom_line(
    data = df_inv_intimacy_summary,
    aes(x = stage, y = empirical_stat, color = action, group = action),
    position = position_dodge(width = 0.8),
    size = 1
  ) +
  facet_grid(reward ~ action, labeller = label_both) +
  scale_fill_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  scale_color_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  labs(
    title = "Intimacy ratings before and after observing action",
    y = "Intimacy rating",
    x = "Stage",
    fill = "Observed action",
    color = "Observed action"
  )
```


Calculate belief update summary

```{r}
df_inv_intimacy_belief_update_summary <- df_inv_intimacy |>
  filter(stage == "posterior") |>
  group_by(reward, action, model) |>
  tidyboot_mean(belief_update)
```
Plot belief updates

```{r}
ggplot(df_inv_intimacy_belief_update_summary, aes(x = action, y = empirical_stat, group = reward, fill = reward)) +
  geom_col(position = position_dodge(width = 0.9), alpha = 0.9) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.8), size = 1, width = 0.1, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Change in intimacy rating after observing action", y = "Belief update", x = "Observed action", fill = "Reward of sharing food") +
  theme(legend.position = "bottom")
```

Model predictions

```{r}
df_model_inv_intimacy <- read_csv(here("model", "sandbox", "observer_intimacy_preds.csv")) |>
  mutate(
    reward = factor(reward, levels = c("low", "high")),
    action = factor(action),
    intimacy = 100 * intimacy,
    model = factor(model, levels = c("discomfort_only", "vanilla_inv_plan", "full_model"))
  ) # rescale intimacy to be 0-100
```

see basic model predictions

TODO: right now, we don't visualize the model's priors, just the posteriors. But later we should load in the priors (empirically measured using the data, too)?

```{r}
ggplot(df_model_inv_intimacy, aes(x = intimacy, y = density, color = action, fill = action)) +
  geom_area(size = 1, alpha = 0.2, position = "identity") +
  scale_color_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  scale_fill_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  labs(y = "Density", x = "Intimacy", title = "Inverse planning model predictions (uniform prior)") +
  theme(legend.position = "bottom") +
  facet_grid(reward ~ model, labeller = labeller(reward = label_both, model = label_both))
```

Calculate expectation in each `{action, reward}` condition. 

```{r}
df_model_inv_intimacy_summary <- df_model_inv_intimacy |>
  group_by(model, reward, action) |>
  summarize(expected_intimacy = sum(intimacy * density)) |>
  mutate(belief_update = expected_intimacy - 50) # assuming uniform prior with mean 50
```


Combine data and model belief updates


```{r}
df_belief_update_combined <- bind_rows(
  df_inv_intimacy_belief_update_summary |> mutate(model = "Humans", belief_update = empirical_stat),
  df_model_inv_intimacy_summary |> mutate(
    model = case_when(
      model == "humans" ~ "Humans",
      model == "discomfort_only" ~ "Discomfort only",
      model == "vanilla_inv_plan" ~ "Vanilla inv planning",
      model == "full_model" ~ "Full model"
    )
  ) |> select(-expected_intimacy)
) |>
  mutate(model = factor(model, levels = c(
    "Discomfort only", "Vanilla inv planning", "Full model", "Humans"
  )))
```

Plot combined belief updates

```{r}
ggplot(df_belief_update_combined, aes(x = action, y = belief_update, group = interaction(model, reward), fill = reward)) +
  geom_col(position = position_dodge(width = 0.9), alpha = 0.9) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.8), size = 1, width = 0.1, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Change in intimacy rating after observing action", y = "Belief update", x = "Observed action", fill = "Motivational state") +
  theme(legend.position = "bottom") +
  facet_wrap(~model, nrow = 1)
```

### Inferring reward (whether its low reward or high reward) given information about the relationship and the action taken

Load data and calculate belief updates

```{r}
df_inv_reward <- read_csv(here("data", "inv_plan_reward", "main_trials.csv")) |>
  mutate(response = response / 100) |> # convert to probability
  rename(
    intimacy = intimacy_condition, action = action_condition,
    p_high_reward = response
  ) |>
  mutate(
    action = str_replace(action, "action_", ""),
    stage = factor(stage, levels = c("prior", "posterior")),
    model = "humans"
  )

df_inv_reward <- df_inv_reward |>
  group_by(subject_id, scenario_label) |>
  mutate(belief_update = ifelse(stage == "posterior", p_high_reward[stage == "posterior"] - p_high_reward[stage == "prior"], NA)) |>
  ungroup()
```

Plot distributions

```{r}
ggplot(df_inv_reward, aes(x = p_high_reward, fill = action)) +
  geom_density(alpha = 0.5) +
  scale_fill_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  labs(y = "Density", x = "P(high motivation)", fill = "Action") +
  theme(legend.position = "bottom") +
  facet_grid(stage ~ intimacy, labeller = label_both)
```
Calculate aggregate summary

```{r}
df_inv_reward_summary <- df_inv_reward |>
  group_by(intimacy, action, stage, model) |>
  tidyboot_mean(p_high_reward)
```

```{r}
ggplot() +
  geom_violin(
    data = df_inv_reward,
    aes(x = stage, y = p_high_reward, fill = action),
    position = position_dodge(width = 0.8),
    alpha = 0.5,
    width = 0.7
  ) +
  geom_point(
    data = df_inv_reward_summary,
    aes(x = stage, y = empirical_stat, color = action, group = action),
    position = position_dodge(width = 0.8),
    size = 3
  ) +
  geom_errorbar(
    data = df_inv_reward_summary,
    aes(x = stage, ymin = ci_lower, ymax = ci_upper, color = action, group = action),
    position = position_dodge(width = 0.8),
    width = 0.1,
    size = 1
  ) +
  geom_line(
    data = df_inv_reward_summary,
    aes(x = stage, y = empirical_stat, color = action, group = action),
    position = position_dodge(width = 0.8),
    size = 1
  ) +
  facet_grid(intimacy ~ action, labeller = label_both) +
  scale_fill_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  scale_color_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  labs(
    title = "P(high motivation) before and after observing action",
    y = "P(high motivation)",
    x = "Stage",
    fill = "Observed action",
    color = "Observed action"
  )
```
Calculate belief update summary in data

```{r}
df_inv_reward_belief_update_summary <- df_inv_reward |>
  filter(stage == "posterior") |>
  group_by(intimacy, action, model) |>
  tidyboot_mean(belief_update)
```

Plot belief updates

```{r}
ggplot(df_inv_reward_belief_update_summary, aes(x = action, y = empirical_stat, group = intimacy, fill = intimacy)) +
  geom_col(position = position_dodge(width = 0.9), alpha = 0.9) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.8), size = 1, width = 0.1, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_viridis_c(begin = 0, end = 0.85) +
  labs(title = "Change in P(high motivation) after observing action", y = "Belief update", x = "Observed action", fill = "Intimacy") +
  theme(legend.position = "bottom")
```

Model predictions


```{r}
df_model_inverse_reward <- read_csv(here("model", "sandbox", "observer_reward_preds.csv")) |>
  mutate(
    reward = factor(reward, levels = c("low", "high")),
    action = factor(action),
    model = factor(model, levels = c("discomfort_only", "vanilla_inv_plan", "full_model"))
  )
```

Is it that observing more risky actions doesn't necessarily give you more information about reward??? 
(i think yes beause its like the fact that you did a food sharing action versus no food sharing action in the first place...)

Basic model predictions

```{r}
ggplot(df_model_inverse_reward |> filter(reward == "high"), aes(x = action, y = p_reward, color = intimacy, group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 3) +
  geom_line(position = position_dodge(width = 0.2), size = 1) +
  labs(y = "P(high motivation)", x = "Observed action", title = "Inverse planning model predictions") +
  scale_color_viridis_c(begin = 0, end = 0.85) +
  facet_grid(~model, labeller = label_both)
```

Calculate belief updates

```{r}
df_model_inv_reward_summary <- df_model_inverse_reward |>
  filter(reward == "high") |>
  mutate(belief_update = p_reward - 0.5) # assuming uniform prior with mean 0.5
```

Combine data and model belief updates

```{r}
df_reward_belief_update_combined <- bind_rows(
  df_inv_reward_belief_update_summary |> mutate(model = "Humans", belief_update = empirical_stat),
  df_model_inv_reward_summary |> mutate(
    model = case_when(
      model == "humans" ~ "Humans",
      model == "discomfort_only" ~ "Discomfort only",
      model == "vanilla_inv_plan" ~ "Vanilla inv planning",
      model == "full_model" ~ "Full model"
    )
  ) |> select(-p_reward)
) |>
  mutate(model = factor(model, levels = c(
    "Discomfort only", "Vanilla inv planning", "Full model", "Humans"
  )))
```

Plot combined belief updates

```{r}
ggplot(df_reward_belief_update_combined, aes(x = action, y = belief_update, group = interaction(model, intimacy), fill = intimacy)) +
  geom_col(position = position_dodge(width = 0.9), alpha = 0.9) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.8), size = 1, width = 0, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_viridis_c(begin = 0, end = 0.85) +
  labs(title = "Change in P(high motivation) after observing action", y = "Belief update", x = "Observed action", fill = "Intimacy") +
  theme(legend.position = "bottom") +
  facet_wrap(~model, nrow = 1)
```





