---
title: "2025-10_pilots"
output: html_document
---


```{r}
library(here)
library(tidyverse)
library(tidyboot)
library(lme4)

theme_set(theme_classic(base_size = 15))
```

## planning-2 pilot 1

This experiment manipulates how much two people want the food (`low` vs. `high` reward), and the relationship between the two people (0, 50, 75, or 100 on a 100 point scale between 'intimate' and 'formal'). 

```{r}
df_planning <- read_csv(here("data", "planning-2", "main_trials.csv")) |> 
  rename(intimacy = intimacy_condition, 
         reward = reward_condition) |> 
  mutate(reward = factor(reward, levels = c("low", "high")), 
         model = "humans") |>
  pivot_longer(
    cols = starts_with("action_"),
    names_to = "action",
    names_prefix = "action_",
    values_to = "p_action")
```
### Look at summary across scenarios

Calculate means

```{r}
df_planning_summary <- df_planning |> 
  group_by(intimacy, reward, action) |> 
  tidyboot_mean(p_action)
```

Plot

```{r}
ggplot(df_planning_summary, aes(x = action, y = empirical_stat, color = intimacy, group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), 
                position = position_dodge(width = 0.2), size = 1, width = 0.3) + 
  geom_line(position = position_dodge(width = 0.2), size = 1) +
  scale_color_viridis_c(begin = 0, end = 0.85) +
  labs(y = "P(action)", x = "Action", color = "Intimacy") +
  theme(legend.position = "bottom") +
  facet_wrap(~reward, labeller = label_both)
```

## Look at toy model predictions

```{r}
df_model_forward <- read_csv(here("model", "model_preds_forward_planning.csv")) |> 
  mutate(reward = factor(reward, levels = c("low", "high")), 
         action = factor(action)) 
```

Stack the dataframes

```{r}
df_combined <- bind_rows(df_planning, df_model) |> 
  mutate(
    model = case_when(
    model == "humans" ~ "Humans", 
    model == "vanilla" ~ "Vanilla model", 
    model == "relationship" ~ "Relationship model"
  )) |> 
  mutate(model = factor(model, levels = c("Vanilla model", "Relationship model", "Humans")))
```

### Summarize combined

```{r}
df_combined_summary <- df_combined |> 
  group_by(intimacy, reward, model, action) |> 
  tidyboot_mean(p_action)
```

Plot

```{r}
ggplot(df_combined_summary, aes(x = action, y = empirical_stat, color = intimacy, group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), 
                position = position_dodge(width = 0.2), size = 1, width = 0.3) + 
  geom_line(position = position_dodge(width = 0.2), size = 1) +
  scale_color_viridis_c(begin = 0, end = 0.85) +
  labs(y = "P(action)", x = "Action", color = "Intimacy", title = "Forward planning") +
  theme(legend.position = "bottom") +
  facet_grid(reward ~ model)
```


# Inverse planning

```{r}
df_model_inverse <- read_csv(here("model", "model_preds_inverse_planning.csv")) |> 
  mutate(reward = factor(reward, levels = c("low", "high")), 
         action = factor(action))
```

```{r}
ggplot(df_model_inverse, aes(x = intimacy, y = density, color = action, fill = action)) +
  geom_area(size = 1, alpha = 0.2, position = "identity") +
  scale_color_viridis_d(option = "plasma") +
  scale_fill_viridis_d(option = "plasma") +
  labs(y = "Density", x = "Intimacy", title = "Inverse planning model predictions") +
  theme(legend.position = "bottom") +
  facet_grid(model~reward, labeller = label_both)
```

Check if `action=0` cases are always the same as the `action=1` cases

```{r}
df_model_inverse_check <- df_model_inverse |> 
  select(model, reward, intimacy, action, density) |> 
  pivot_wider(names_from = action, values_from = density, names_prefix = "action_") |> 
  mutate(diff = action_0 - action_1)
```

```{r}
df_model_inverse_check |> 
  filter(abs(diff) > 1e-6)
```

### Inferring reward (whether its low reward or high reward) given information about the relationship and the action taken

```{r}
df_model_inverse_reward <- read_csv(here("model", "model_preds_inverse_planning_reward.csv")) |> 
  mutate(reward = factor(reward, levels = c("low", "high")), 
         action = factor(action))
```
Is it that observing more risky actions doesn't necessarily give you more information about reward??? 
(i think yes beause its like the fact that you did a food sharing action versus no food sharing action in the first place...)


```{r}
ggplot(df_model_inverse_reward, aes(x = intimacy, y = p_reward, color = reward, fill = reward, group = reward)) +
  geom_point(position = position_dodge(width = 0.2), size = 3) +
  geom_line(position = position_dodge(width = 0.2), size = 1) +
  labs(y = "P(reward)", x = "Intimacy", title = "Inverse planning model predictions") +
  facet_grid(model~action, labeller = label_both) 
```









