---
title: "2025-10_pilots"
output: html_document
---


```{r}
library(here)
library(tidyverse)
library(tidyboot)
library(lme4)

theme_set(theme_minimal(base_size = 15) +
     theme(
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    ))
```

## planning-2 pilot 1

This experiment manipulates how much two people want the food (`low` vs. `high` reward), and the relationship between the two people (0, 50, 75, or 100 on a 100 point scale between 'intimate' and 'formal'). 

```{r}
df_planning <- read_csv(here("data", "planning-2", "main_trials.csv")) |> 
  rename(intimacy = intimacy_condition, 
         reward = reward_condition) |> 
  mutate(reward = factor(reward, levels = c("low", "high")), 
         model = "humans") |>
  pivot_longer(
    cols = starts_with("action_"),
    names_to = "action",
    names_prefix = "action_",
    values_to = "p_action")
```
### Look at summary across scenarios

Calculate means

```{r}
df_planning_summary <- df_planning |> 
  group_by(intimacy, reward, action) |> 
  tidyboot_mean(p_action)
```

Plot

```{r}
ggplot(df_planning_summary, aes(x = action, y = empirical_stat, color = intimacy, group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), 
                position = position_dodge(width = 0.2), size = 1, width = 0.3) + 
  geom_line(position = position_dodge(width = 0.2), size = 1) +
  scale_color_viridis_c(begin = 0, end = 0.85) +
  labs(y = "P(action)", x = "Action", color = "Intimacy") +
  theme(legend.position = "bottom") +
  facet_wrap(~reward, labeller = label_both)
```

## Look at toy model predictions

```{r}
df_model_forward <- read_csv(here("model", "model_preds_forward_planning.csv")) |> 
  mutate(reward = factor(reward, levels = c("low", "high")), 
         action = factor(action)) 
```

Stack the dataframes

```{r}
df_combined <- bind_rows(df_planning, df_model_forward) |> 
  mutate(
    model = case_when(
    model == "humans" ~ "Humans", 
    model == "vanilla" ~ "Vanilla model", 
    model == "relationship" ~ "Relationship model"
  )) |> 
  mutate(model = factor(model, levels = c("Vanilla model", "Relationship model", "Humans")))
```

### Summarize combined

```{r}
df_combined_summary <- df_combined |> 
  group_by(intimacy, reward, model, action) |> 
  tidyboot_mean(p_action)
```
Plot

```{r}
ggplot(df_combined_summary, aes(x = action, y = empirical_stat, color = intimacy, group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), 
                position = position_dodge(width = 0.2), size = 1, width = 0.3) + 
  geom_line(position = position_dodge(width = 0.2), size = 1) +
  scale_color_viridis_c(begin = 0, end = 0.85) +
  labs(y = "P(action)", x = "Action", color = "Intimacy", title = "Forward planning") +
  theme(legend.position = "bottom") +
  facet_grid(reward ~ model)
```


# Inverse planning

## Inferring intimacy of relationship, given information about the reward and the action taken

Load data and add belief updates

```{r}
df_inv_intimacy <- read_csv(here("data", "inv_plan_intimacy", "main_trials.csv")) |> 
  rename(reward = reward_condition, action = action_condition) |> 
  mutate(action = str_replace(action, "action_", ""), 
         reward = factor(reward, levels = c("low", "high")), 
         stage = factor(stage, levels = c("prior", "posterior")),
         model = "humans")

df_inv_intimacy <- df_inv_intimacy |> 
  group_by(subject_id, scenario_label) |> 
  mutate(belief_update = ifelse(stage == "posterior", intimacy_rating[stage == "posterior"] - intimacy_rating[stage == "prior"], NA)) |>
  ungroup()
```
Plot distributions

```{r}
ggplot(df_inv_intimacy, aes(x = intimacy_rating, fill = action)) +
  geom_density(alpha = 0.5) +
  scale_fill_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  labs(y = "Density", x = "Intimacy rating", fill = "Action") +
  theme(legend.position = "bottom") +
  facet_grid(stage~reward, labeller = label_both)
```

Calculate aggregate summary

```{r}
df_inv_intimacy_summary <- df_inv_intimacy |> 
  group_by(reward, action, stage, model) |> 
  tidyboot_mean(intimacy_rating)
```

Plot priors vs. posteriors for each action and reward condition


```{r}
ggplot() +
  geom_violin(
    data = df_inv_intimacy,
    aes(x = stage, y = intimacy_rating, fill = action),
    position = position_dodge(width = 0.8),
    alpha = 0.5,
    width = 0.7
  ) +
  geom_point(
    data = df_inv_intimacy_summary,
    aes(x = stage, y = empirical_stat, color = action, group = action),
    position = position_dodge(width = 0.8),
    size = 3
  ) +
  geom_errorbar(
    data = df_inv_intimacy_summary,
    aes(x = stage, ymin = ci_lower, ymax = ci_upper, color = action, group = action),
    position = position_dodge(width = 0.8),
    width = 0.1,
    size = 1 
  ) +
  geom_line(
    data = df_inv_intimacy_summary,
    aes(x = stage, y = empirical_stat, color = action, group = action),
    position = position_dodge(width = 0.8),
    size = 1
  ) +
  facet_grid(reward ~ action, labeller = label_both) +
  scale_fill_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  scale_color_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  labs(
    title = "Intimacy ratings before and after observing action",
    y = "Intimacy rating",
    x = "Stage",
    fill = "Observed action",
    color = "Observed action"
  )

```


Calculate belief update summary

```{r}
df_inv_intimacy_belief_update_summary <- df_inv_intimacy |> 
  filter(stage == "posterior") |> 
  group_by(reward, action, model) |> 
  tidyboot_mean(belief_update)
```
Plot belief updates

```{r}
ggplot(df_inv_intimacy_belief_update_summary, aes(x = action, y = empirical_stat, group = reward, fill = reward)) +
  geom_col(position = position_dodge(width = 0.9), alpha = 0.9) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), 
                position = position_dodge(width = 0.8), size = 1, width = 0.1, color = "black", alpha = 0.8) + 
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Change in intimacy rating after observing action", y = "Belief update", x = "Observed action", fill = "Reward of sharing food") +
  theme(legend.position = "bottom") 
```

Model predictions

```{r}
df_model_inv_intimacy <- read_csv(here("model", "model_preds_inverse_planning_relationship.csv")) |> 
  mutate(reward = factor(reward, levels = c("low", "high")), 
         action = factor(action))
```

see basic model predictions

TODO: right now, we don't visualize the model's priors, just the posteriors. But later we should load in the priors (empirically measured using the data, too)?

```{r}
ggplot(df_model_inv_intimacy, aes(x = intimacy, y = density, color = action, fill = action)) +
  geom_area(size = 1, alpha = 0.2, position = "identity") +
  scale_color_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  scale_fill_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  labs(y = "Density", x = "Intimacy", title = "Inverse planning model predictions (uniform prior)") +
  theme(legend.position = "bottom") +
  facet_grid(reward~model, labeller = labeller(reward = label_both, model = label_both))
```

Calculate expectation in each `{action, reward}` condition. 

```{r}
df_model_inv_intimacy_summary <- df_model_inv_intimacy |> 
  group_by(model, reward, action) |> 
  summarize(expected_intimacy = sum(intimacy * density)) |> 
  mutate(belief_update = expected_intimacy - 50) # assuming uniform prior with mean 50
```


Combine data and model belief updates

```{r}
df_belief_update_combined <- bind_rows(
  df_inv_intimacy_belief_update_summary |> mutate(model = "Humans", belief_update = empirical_stat),
  df_model_inv_intimacy_summary |> mutate(
    model = case_when(
      model == "vanilla" ~ "Vanilla model",
      model == "relationship" ~ "Relationship model"
    )
  ) |> select(-expected_intimacy)
) |>
  mutate(model = factor(model, levels = c(
    "Vanilla model", "Relationship model", "Humans"
  )))
```

Plot combined belief updates

```{r}
ggplot(df_belief_update_combined, aes(x = action, y = belief_update, group = interaction(model, reward), fill = reward)) +
  geom_col(position = position_dodge(width = 0.9), alpha = 0.9) +
  geom_errorbar( 
                aes(ymin = ci_lower, ymax = ci_upper), 
                position = position_dodge(width = 0.8), size = 1, width = 0.1, color = "black", alpha = 0.8) + 
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Change in intimacy rating after observing action", y = "Belief update", x = "Observed action", fill = "Reward of sharing food") +
  theme(legend.position = "bottom") +
  facet_wrap(~model)
```

Check if `action=0` cases are always the same as the `action=1` cases...

```{r}
df_model_inverse_check <- df_model_inverse |> 
  select(model, reward, intimacy, action, density) |> 
  pivot_wider(names_from = action, values_from = density, names_prefix = "action_") |> 
  mutate(diff = action_0 - action_1)
```

```{r}
df_model_inverse_check |> 
  filter(abs(diff) > 1e-6)
```

### Inferring reward (whether its low reward or high reward) given information about the relationship and the action taken

```{r}
df_model_inverse_reward <- read_csv(here("model", "model_preds_inverse_planning_reward.csv")) |> 
  mutate(reward = factor(reward, levels = c("low", "high")), 
         action = factor(action))
```
Is it that observing more risky actions doesn't necessarily give you more information about reward??? 
(i think yes beause its like the fact that you did a food sharing action versus no food sharing action in the first place...)


```{r}
ggplot(df_model_inverse_reward, aes(x = intimacy, y = p_reward, color = reward, fill = reward, group = reward)) +
  geom_point(position = position_dodge(width = 0.2), size = 3) +
  geom_line(position = position_dodge(width = 0.2), size = 1) +
  labs(y = "P(reward)", x = "Intimacy", title = "Inverse planning model predictions") +
  facet_grid(model~action, labeller = label_both) 
```


# Reward study

```{r}
df_inv_reward <- read_csv(here("data", "inv_plan_reward", "main_trials.csv")) |> 
  rename(intimacy = intimacy_condition, action = action_condition, 
         p_high_reward = response) |> 
  mutate(action = str_replace(action, "action_", ""), 
         stage = factor(stage, levels = c("prior", "posterior")),
         model = "humans")
```

```{r}
df_inv_reward_summary <- df_inv_reward |> 
  group_by(intimacy, action, stage, model) |> 
  tidyboot_mean(p_high_reward)
```

```{r}
ggplot(df_inv_reward_summary, aes(x = intimacy, y = empirical_stat, color = stage, group = stage)) +
  geom_point(position = position_dodge(width = 0.2), size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), 
                position = position_dodge(width = 0.2), size = 1, width = 0.3) + 
  geom_line(position = position_dodge(width = 0.2), size = 1) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "grey") +
  labs(y = "P(high reward)", x = "Intimacy", color = "Stage") +
  theme(legend.position = "bottom") +
  facet_grid(stage~action, labeller = label_both)
```

Priors aon reward are based on the relationship



