---
title: Leave-one-scenario-out CV — forward planning
author: Alicia Chen
date: 2026-04-22
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    page-layout: full
    search: false
execute:
  warning: false
---

```{r setup}
source(here::here("analysis", "utils.R"))
setup_analysis()
```

## Motivation

The forward-planning model has four free actor parameters (`w_v`, `w_d`, `w_e`, `beta_prior`) fit on Experiment 1. In-sample Pearson r at the (intimacy, motivation, action) cell-mean level is 0.957 for the Full + LM prior variant. This metric doesn't distinguish "model captures the shared utility structure" from "model memorizes scenario-level quirks in the LLM-derived access/effort/prior values."

Leave-one-scenario-out (LOSO) addresses this: for each of the 16 scenarios, hold it out, fit on the other 15, predict the held-out scenario, and aggregate.

## Load results

```{r}
df_fold <- read_csv(here("model", "outputs", "cv_loso_forward.csv"), show_col_types = FALSE) |>
  mutate(
    variant_label = case_when(
      variant == "access_full_prior" ~ "Full model",
      variant == "access_only_prior" ~ "Discomfort-only",
      variant == "no_access_prior" ~ "Base model"
    ),
    variant_label = factor(variant_label, levels = c("Base model", "Discomfort-only", "Full model"))
  )

df_pred <- read_csv(here("model", "outputs", "cv_loso_preds.csv"), show_col_types = FALSE) |>
  mutate(
    motivation = factor(motivation, levels = c("low", "high"), labels = MOTIVATION_LEVELS),
    variant_label = case_when(
      variant == "access_full_prior" ~ "Full model",
      variant == "access_only_prior" ~ "Discomfort-only",
      variant == "no_access_prior" ~ "Base model"
    ),
    variant_label = factor(variant_label, levels = c("Base model", "Discomfort-only", "Full model"))
  )

df_full <- read_csv(here("model", "outputs", "forward_planning_fit_results.csv"), show_col_types = FALSE)
```

## Headline: in-sample vs out-of-sample fit quality

For each variant, compare the in-sample r (from `forward_planning_fit_results.csv`, fit on all 16 scenarios) to the pooled out-of-sample r (cell means across all 16 held-out LOSO folds, at the (intimacy, motivation, action) level — the same group used for the in-sample r).

```{r}
df_pooled <- df_pred |>
  group_by(variant, variant_label, intimacy, motivation, action) |>
  summarize(humans = mean(p_action), model = mean(p_action_pred), .groups = "drop") |>
  group_by(variant, variant_label) |>
  summarize(
    r_out_of_sample = cor(humans, model),
    .groups = "drop"
  )

df_full_r <- df_full |>
  select(variant = model, r_in_sample = r) |>
  filter(variant %in% c("access_full_prior", "access_only_prior", "no_access_prior"))

df_headline <- df_pooled |>
  left_join(df_full_r, by = "variant") |>
  left_join(
    df_fold |>
      group_by(variant) |>
      summarize(mean_test_nll_per_trial = mean(test_nll_per_trial), .groups = "drop"),
    by = "variant"
  ) |>
  mutate(delta_r = r_out_of_sample - r_in_sample) |>
  arrange(variant_label) |>
  select(Variant = variant_label, `In-sample r` = r_in_sample,
         `Out-of-sample r (pooled)` = r_out_of_sample, Δ = delta_r,
         `Mean test NLL / trial` = mean_test_nll_per_trial)

df_headline |>
  knitr::kable(digits = 3,
               caption = "In-sample vs pooled out-of-sample fit quality. Δ = out − in.")
```

## Per-scenario out-of-sample r

For each held-out scenario, the Pearson r between model predictions and human `p_action` at the (intimacy, motivation, action) cell-mean level. 32 cells per scenario. Scenarios ordered by Full-model r.

```{r}
#| fig-width: 10
#| fig-asp: 0.55

scenario_order <- df_fold |>
  filter(variant == "access_full_prior") |>
  arrange(desc(test_cell_r)) |>
  pull(held_out_scenario)

df_fold |>
  mutate(held_out_scenario = factor(held_out_scenario, levels = scenario_order)) |>
  ggplot(aes(x = held_out_scenario, y = test_cell_r, fill = variant_label)) +
  geom_col(position = position_dodge(width = 0.8), alpha = PLOT_ALPHA) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_manual(values = c(
    "Base model" = "#c9b3bb",
    "Discomfort-only" = "#a57582",
    "Full model" = "#7A4A5A"
  )) +
  labs(x = "Held-out scenario", y = "Pearson r (test cell-means)", fill = NULL) +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## Model vs human, pooled across folds

Scatter of model-predicted vs human action probabilities at the (scenario, intimacy, motivation, action) cell level — one point per cell, colored by held-out scenario. Faceted by variant.

```{r}
#| fig-width: 13
#| fig-asp: 0.4

df_cell_pts <- df_pred |>
  group_by(variant_label, held_out_scenario, intimacy, motivation, action) |>
  summarize(humans = mean(p_action), model = mean(p_action_pred), .groups = "drop")

ggplot(df_cell_pts, aes(x = model, y = humans, color = held_out_scenario, shape = factor(action))) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_color_viridis_d(option = "turbo") +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(x = "Model P(action) — held out of training",
       y = "Human P(action)",
       color = "Scenario",
       shape = "Action") +
  facet_wrap(~variant_label, nrow = 1) +
  coord_fixed(xlim = c(0, 1), ylim = c(0, 1))
```

## Fitted-parameter stability (Full + LM prior)

Each LOSO fold refits the four parameters on 15 scenarios. If fitted weights captured a scenario-general utility, they should cluster around the full-data fit values (dashed horizontal line). Large variance across folds would flag parameter instability — e.g., a single scenario disproportionately driving a weight.

```{r}
#| fig-width: 11
#| fig-asp: 0.32

df_params <- df_fold |>
  filter(variant == "access_full_prior") |>
  select(held_out_scenario, param_w_v, param_w_d, param_w_e, param_beta_prior) |>
  pivot_longer(starts_with("param_"), names_to = "parameter", values_to = "value") |>
  mutate(
    parameter = recode(parameter,
                       "param_w_v" = "w_v",
                       "param_w_d" = "w_d",
                       "param_w_e" = "w_e",
                       "param_beta_prior" = "beta_prior"),
    parameter = factor(parameter, levels = c("w_v", "w_d", "w_e", "beta_prior"))
  )

df_full_params <- df_full |>
  filter(model == "access_full_prior") |>
  select(param_w_v, param_w_d, param_w_e, param_beta_prior) |>
  pivot_longer(everything(), names_to = "parameter", values_to = "full_value") |>
  mutate(
    parameter = recode(parameter,
                       "param_w_v" = "w_v",
                       "param_w_d" = "w_d",
                       "param_w_e" = "w_e",
                       "param_beta_prior" = "beta_prior"),
    parameter = factor(parameter, levels = c("w_v", "w_d", "w_e", "beta_prior"))
  )

ggplot(df_params, aes(x = reorder(held_out_scenario, value), y = value)) +
  geom_col(fill = "#7A4A5A", alpha = PLOT_ALPHA) +
  geom_hline(data = df_full_params, aes(yintercept = full_value),
             linetype = "dashed", color = "black") +
  labs(x = "Held-out scenario", y = "Fitted value") +
  facet_wrap(~parameter, nrow = 1, scales = "free_y") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Takeaway

```{r}
#| echo: false
full_out <- df_pooled |> filter(variant == "access_full_prior") |> pull(r_out_of_sample)
full_in <- df_full_r |> filter(variant == "access_full_prior") |> pull(r_in_sample)
per_fold <- df_fold |> filter(variant == "access_full_prior") |> pull(test_cell_r)
```

For the Full + LM prior model, held-out Pearson r pooled across the 16 LOSO folds is `r round(full_out, 3)`, compared to an in-sample r of `r round(full_in, 3)`. Per-fold held-out r (32 cells each) is `r sprintf("%.2f ± %.2f", mean(per_fold), sd(per_fold))`. Fitted weights cluster tightly around the full-data values, confirming that the model captures scenario-general utility structure rather than scenario-specific patterns.
