---
title: Experiment 2b (inverse planning - reward inference) data analysis
author: Alicia Chen
date: 2026-01-15
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    page-layout: full
    search: false
knitr:
  opts_chunk:
    dev: "ragg_png"
execute:
  warning: false
---

```{r setup}
source(here::here("analysis", "utils.R"))
setup_analysis()
```

## Demographics

```{r}
df_exit <- read_csv(here("data", "inv_plan_reward", "exit_survey.csv"), show_col_types = FALSE)

n_total <- nrow(df_exit)
n_passed <- df_exit |>
  filter(attention_passed == TRUE, memory_correct_count > 0) |>
  nrow()

cat("Total participants recruited:", n_total, "\n")
cat("Passed attention checks:", n_passed, "\n")
cat("Mean age:", round(mean(df_exit$age, na.rm = TRUE), 1),
    "SD age:", round(sd(df_exit$age, na.rm = TRUE), 1))
cat("\nGender:\n")
print(table(df_exit$gender))
```

## Results

This experiment presents participants with scenarios where they observe an action taken by one person toward another. Participants rate their belief about the probability that the actors have high motivation (want the food) before (prior) and after (posterior) observing the action. The experiment manipulates:
- **Action**: Which action was taken (action 0-3, varying in saliva-transfer risk)
- **Intimacy**: The relationship closeness between the actors (0, 50, 75, or 100)

## Load human data

```{r}
# main_trials_long.csv is generated by analysis/json_to_csv.py
# It excludes participants who failed attention check or got 0 on memory check
df <- read_csv(here("data", "inv_plan_reward", "main_trials_long.csv")) |>
  mutate(
    action = str_replace(action_condition, "action_", ""),
    intimacy = factor(intimacy),
    p_high_reward = response / 100,  # Convert to probability scale
    stage = factor(stage, levels = c("prior", "posterior")),
    model = "Humans"
  ) |>
  select(-action_condition, -response)

# Calculate belief updates
df <- calculate_belief_update(df, "p_high_reward")
```

## Human data: Prior vs posterior ratings

### Distribution of ratings

```{r}
#| fig-width: 12
#| fig-asp: 0.45

ggplot(df, aes(x = p_high_reward, fill = action)) +
  geom_density(alpha = 0.5) +
  scale_fill_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  labs(y = "Density", x = "P(high motivation)", fill = "Action") +
  theme(legend.position = "right") +
  facet_grid(stage ~ intimacy, labeller = labeller(intimacy = label_both, stage = label_value))
```

### Summary statistics

```{r}
df_summary <- df |>
  group_by(intimacy, action, stage, model) |>
  tidyboot_mean(p_high_reward)
```

### Prior vs posterior by action and intimacy

```{r}
#| fig-width: 12
#| fig-asp: 0.7

ggplot() +
  geom_violin(
    data = df,
    aes(x = stage, y = p_high_reward, fill = action),
    position = position_dodge(width = 0.8),
    alpha = 0.5,
    width = 0.7
  ) +
  geom_point(
    data = df_summary,
    aes(x = stage, y = empirical_stat, color = action, group = action),
    position = position_dodge(width = 0.8),
    size = 3,
    alpha = PLOT_ALPHA
  ) +
  geom_errorbar(
    data = df_summary,
    aes(x = stage, ymin = ci_lower, ymax = ci_upper, color = action, group = action),
    position = position_dodge(width = 0.8),
    width = 0.1,
    size = 1
  ) +
  geom_line(
    data = df_summary,
    aes(x = stage, y = empirical_stat, color = action, group = action),
    position = position_dodge(width = 0.8),
    size = 1
  ) +
  facet_grid(intimacy ~ action, labeller = label_both) +
  scale_fill_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  scale_color_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  labs(
    title = "P(high motivation) before and after observing action",
    y = "P(high motivation)",
    x = "Stage",
    fill = "Observed action",
    color = "Observed action"
  )
```

### Belief updates

```{r}
df_belief_update_summary <- df |>
  filter(stage == "posterior") |>
  group_by(intimacy, action, model) |>
  tidyboot_mean(belief_update)
```

## Model comparison

### Model description

We compare three models that perform inverse planning to infer reward/motivation from observed actions:

**Full model (Social)**: Inverts a forward model where actors consider reward (scaled by intimacy and motivation), discomfort (scaled by intimacy), and effort costs.

**Vanilla model**: Inverts a forward model with reward, discomfort, and effort, but without intimacy scaling these utilities.

**Discomfort-only model**: Inverts a forward model where actors only consider discomfort from saliva transfer (scaled by intimacy).

**Full modified model**: Like the full model, but with reward scaled by $(1 + \beta \cdot \text{intimacy})$. This captures observers' assumption about how much actors value intimacy when sharing: $\beta=1$ is the pre-registered model, $\beta=0$ means observers don't think intimacy affects reward.

All models assume a uniform prior over reward (P(high motivation) = 0.5) and compute posteriors via Bayesian inference. The observer inverse temperature ($\alpha_{\text{observer}}$) is fitted to human data. For the modified model, we also fit $\beta$ (reward-intimacy scaling strength).

### Fitted parameters

```{r}
df_fit_results <- read_csv(here("model", "outputs", "inverse_planning_fit_results.csv")) |>
  filter(experiment == "reward") |>
  mutate(
    model_label = case_when(
      model == "full" ~ "Social (pre-reg)",
      model == "vanilla" ~ "Vanilla",
      model == "discomfort_only" ~ "Discomfort-only",
      model == "full_modified" ~ "Social"
    )
  ) |>
  select(model_label, alpha_observer, beta, nll, n_params)

df_fit_results |>
  knitr::kable(
    digits = 3,
    caption = "Fitted observer parameters for reward inference",
    col.names = c("Model", "α_observer", "beta", "NLL", "n_params")
  )
```

Note: The $\beta$ parameter captures how much observers think actors' reward scales with intimacy. $\beta < 1$ indicates observers believe actors value sharing less with closer friends than the pre-registered model assumes.

### AIC comparison

Following Burnham and Anderson (2002), ΔAIC values of 0–2 indicate essentially equivalent models, 4–7 indicate considerably less support for the higher-AIC model, and values greater than 10 indicate essentially no support for the higher-AIC model.

```{r}
# Calculate AIC
df_fit_results <- df_fit_results |>
  mutate(AIC = 2 * n_params + 2 * nll)

# Find best model
best_aic <- min(df_fit_results$AIC)

df_fit_results |>
  mutate(delta_AIC = AIC - best_aic) |>
  select(model_label, nll, n_params, AIC, delta_AIC) |>
  arrange(AIC) |>
  knitr::kable(
    digits = 2,
    caption = "Model comparison (lower AIC is better)",
    col.names = c("Model", "NLL", "n_params", "AIC", "ΔAIC")
  )
```

### Load model predictions

```{r}
df_model <- read_csv(here("model", "outputs", "inv_plan_reward_preds_summary.csv")) |>
  rename(intimacy = intimacy_condition) |>
  mutate(
    action = as.character(action),
    intimacy = factor(intimacy),
    # Model predictions are P(high reward) in percentage
    p_high_reward = p_high_reward / 100,
    # Belief update from uniform prior (50%)
    belief_update = p_high_reward - 0.5,
    model = case_when(
      model == "full" ~ "Social (pre-reg)",
      model == "vanilla" ~ "Vanilla",
      model == "discomfort_only" ~ "Discomfort-only",
      model == "full_modified" ~ "Social"
    ),
    param_source = "Stipulated"
  )
```

### Combine human and model data

```{r}
df_combined <- bind_rows(
  df_belief_update_summary |>
    mutate(belief_update = empirical_stat) |>
    select(intimacy, action, model, belief_update, ci_lower, ci_upper),
  df_model |>
    group_by(intimacy, action, model) |>
    summarize(belief_update = mean(belief_update), .groups = "drop") |>
    mutate(ci_lower = NA_real_, ci_upper = NA_real_)
) |>
  mutate(
    model = factor(model, levels = c("Vanilla", "Discomfort-only", "Social", "Social (pre-reg)", "Humans")),
    action = factor(action)
  )
```

### Belief updates: Model vs human comparison

Main comparison (excludes preregistered Social):

```{r}
#| fig-width: 12
#| fig-asp: 0.3

df_combined |>
  filter(model != "Social (pre-reg)") |>
  ggplot(aes(x = action, y = belief_update, group = interaction(model, intimacy), fill = intimacy)) +
  geom_crossbar(
    aes(y = 0, ymin = -0.015, ymax = 0.015),
    position = position_dodge(width = 0.9),
    width = 0.9,
    color = NA
  ) +
  geom_col(position = position_dodge(width = 0.9), alpha = PLOT_ALPHA) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.9), size = 1, width = 0, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_intimacy() +
  labs(y = "Belief update", x = "Observed action", fill = "Intimacy") +
  theme(legend.position = "right") +
  facet_wrap(~model, nrow = 1)

ggsave(here("figures", "exp2b_belief_main.pdf"), device = cairo_pdf)
```

Preregistered vs modified model comparison:

```{r}
#| fig-width: 10 
#| fig-asp: 0.35

df_combined |>
  filter(model %in% c("Social (pre-reg)", "Social", "Humans")) |>
  mutate(model = factor(model, levels = c("Social (pre-reg)", "Social", "Humans"))) |>
  ggplot(aes(x = action, y = belief_update, group = interaction(model, intimacy), fill = intimacy)) +
  geom_crossbar(
    aes(y = 0, ymin = -0.015, ymax = 0.015),
    position = position_dodge(width = 0.9),
    width = 0.9,
    color = NA
  ) +
  geom_col(position = position_dodge(width = 0.9), alpha = PLOT_ALPHA) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.9), size = 1, width = 0, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_intimacy() +
  labs(y = "Belief update", x = "Observed action", fill = "Intimacy") +
  theme(legend.position = "right") +
  facet_wrap(~model, nrow = 1)

ggsave(here("figures", "exp2b_belief_comparison.pdf"), device = cairo_pdf)
```

## Scenario-level

### Human belief updates by scenario

```{r}
df_belief_update_scenario <- df |>
  filter(stage == "posterior") |>
  group_by(scenario_label, intimacy, action, model) |>
  tidyboot_mean(belief_update)
```
