---
title: Combined Experiment 2a and 2b correlation plots
author: Alicia Chen
date: today
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
knitr:
  opts_chunk:
    dev: "ragg_png"
execute:
  warning: false
---

```{r setup}
source(here::here("analysis", "utils.R"))
setup_analysis()
library(ggnewscale)

# Define condition sets for each experiment
EXP2A_CONDITIONS <- c("Low desire", "High desire")
EXP2B_CONDITIONS <- c("Intimacy: 0", "Intimacy: 50", "Intimacy: 75", "Intimacy: 100")

# Create intimacy colors with correct names for this analysis
EXP2B_COLORS <- INTIMACY_COLORS
names(EXP2B_COLORS) <- paste0("Intimacy: ", names(INTIMACY_COLORS))
```

This notebook aggregates model-human correlations across both inverse planning experiments:
- **Exp 2a**: Intimacy inference (infer relationship closeness from observed action)
- **Exp 2b**: Reward inference (infer motivation from observed action)

## Load Exp 2a (Intimacy Inference) Data

### Human data

```{r}
df_2a <- read_csv(here("data", "inv_plan_intimacy", "main_trials_long.csv")) |>
  mutate(
    action = str_replace(action_condition, "action_", ""),
    motivation = factor(motivation, levels = c("low", "high"), labels = MOTIVATION_LEVELS),
    stage = factor(stage, levels = c("prior", "posterior"))
  ) |>
  group_by(subject_id, scenario_label) |>
  mutate(
    belief_update = ifelse(
      stage == "posterior",
      intimacy_rating[stage == "posterior"] - intimacy_rating[stage == "prior"],
      NA
    )
  ) |>
  ungroup()

df_2a_summary <- df_2a |>
  filter(stage == "posterior") |>
  group_by(motivation, action) |>
  tidyboot_mean(belief_update) |>
  rename(human_belief_update = empirical_stat)
```

### Model predictions

```{r}
df_model_2a <- read_csv(here("model", "outputs", "inv_plan_intimacy_preds_summary.csv")) |>
  rename(motivation = reward_condition) |>
  filter(param_source == "stipulated") |>
  mutate(
    action = as.character(action),
    motivation = factor(motivation, levels = c("low", "high"), labels = MOTIVATION_LEVELS),
    model_belief_update = expected_intimacy - 50,
    model = case_when(
      model == "full" ~ "Social (pre-reg)",
      model == "vanilla" ~ "Vanilla",
      model == "discomfort_only" ~ "Discomfort-only",
      model == "full_modified" ~ "Social"
    )
  ) |>
  group_by(motivation, action, model) |>
  summarize(model_belief_update = mean(model_belief_update), .groups = "drop")
```

## Load Exp 2b (Reward Inference) Data

### Human data

```{r}
df_2b <- read_csv(here("data", "inv_plan_reward", "main_trials_long.csv")) |>
  mutate(
    action = str_replace(action_condition, "action_", ""),
    intimacy = factor(intimacy),
    p_high_reward = response / 100,
    stage = factor(stage, levels = c("prior", "posterior"))
  ) |>
  group_by(subject_id, scenario_label) |>
  mutate(
    belief_update = ifelse(
      stage == "posterior",
      p_high_reward[stage == "posterior"] - p_high_reward[stage == "prior"],
      NA
    )
  ) |>
  ungroup()

df_2b_summary <- df_2b |>
  filter(stage == "posterior") |>
  group_by(intimacy, action) |>
  tidyboot_mean(belief_update) |>
  rename(human_belief_update = empirical_stat)
```

### Model predictions

```{r}
df_model_2b <- read_csv(here("model", "outputs", "inv_plan_reward_preds_summary.csv")) |>
  rename(intimacy = intimacy_condition) |>
  filter(param_source == "stipulated") |>
  mutate(
    action = as.character(action),
    intimacy = factor(intimacy),
    model_belief_update = (p_high_reward / 100) - 0.5,
    model = case_when(
      model == "full_modified" ~ "Social",
      model == "full" ~ "Social (pre-reg)",
      model == "vanilla" ~ "Vanilla",
      model == "discomfort_only" ~ "Discomfort-only"
    )
  ) |>
  group_by(intimacy, action, model) |>
  summarize(model_belief_update = mean(model_belief_update), .groups = "drop")
```

## Combine and Rescale

We rescale exp 2a belief updates by dividing by 100 to put both experiments on a -1 to 1 scale:
- Exp 2a: intimacy ratings (0-100), so belief updates are on ~-50 to +50 scale → ÷100 gives -0.5 to +0.5
- Exp 2b: probabilities (0-1), so belief updates are already on -0.5 to +0.5 scale

```{r}
# Combine human and model for exp 2a (rescale from 0-100 to 0-1)
df_corr_2a <- df_2a_summary |>
  left_join(df_model_2a, by = c("motivation", "action")) |>
  mutate(
    experiment = "Exp 2a: Intimacy",
    # Rescale from percentage scale to proportion scale
    human_belief_update = human_belief_update / 100,
    ci_lower = ci_lower / 100,
    ci_upper = ci_upper / 100,
    model_belief_update = model_belief_update / 100
  )

# Combine human and model for exp 2b (already on 0-1 scale)
df_corr_2b <- df_2b_summary |>
  left_join(df_model_2b, by = c("intimacy", "action")) |>
  mutate(experiment = "Exp 2b: Reward")

# Combine both experiments
df_combined <- bind_rows(
  df_corr_2a |>
    mutate(condition = as.character(motivation)) |>
    select(action, model, experiment, condition, human_belief_update, ci_lower, ci_upper, model_belief_update),
  df_corr_2b |>
    mutate(condition = paste0("Intimacy: ", intimacy)) |>
    select(action, model, experiment, condition, human_belief_update, ci_lower, ci_upper, model_belief_update)
) |>
  mutate(
    model = factor(model, levels = c("Vanilla", "Discomfort-only", "Social", "Social (pre-reg)")),
    action = factor(action),
    condition = factor(condition, levels = c(
      "Low desire", "High desire",
      "Intimacy: 0", "Intimacy: 50", "Intimacy: 75", "Intimacy: 100"
    ))
  )
```

## Compute Correlations

```{r}
correlations_combined <- df_combined |>
  group_by(model) |>
  summarize(
    boot_result = list(boot_cor(model_belief_update, human_belief_update)),
    .groups = "drop"
  ) |>
  mutate(
    r = sapply(boot_result, function(x) x$r),
    ci_lower = sapply(boot_result, function(x) x$ci_lower),
    ci_upper = sapply(boot_result, function(x) x$ci_upper),
    label = paste0("r = ", round(r, 2), " (", round(ci_lower, 2), ", ", round(ci_upper, 2), ")")
  ) |>
  select(-boot_result)
```

## Aggregated Correlation Plot

Main comparison (excludes preregistered Social):

```{r}
#| fig-width: 11
#| fig-asp: 0.4

df_plot <- df_combined |> filter(model != "Social (pre-reg)")

ggplot(df_plot, aes(x = model_belief_update, y = human_belief_update, shape = action)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
  # Exp 2a points (motivation)
  geom_errorbar(
    data = \(d) filter(d, condition %in% EXP2A_CONDITIONS),
    aes(ymin = ci_lower, ymax = ci_upper, color = condition), width = 0.02
  ) +
  geom_point(
    data = \(d) filter(d, condition %in% EXP2A_CONDITIONS),
    aes(color = condition), size = 3, alpha = PLOT_ALPHA
  ) +
  scale_color_manual(name = "Exp 2a", values = MOTIVATION_COLORS, guide = guide_legend(order = 2)) +
  new_scale_color() +
  # Exp 2b points (intimacy)
  geom_errorbar(
    data = \(d) filter(d, condition %in% EXP2B_CONDITIONS),
    aes(ymin = ci_lower, ymax = ci_upper, color = condition), width = 0.02
  ) +
  geom_point(
    data = \(d) filter(d, condition %in% EXP2B_CONDITIONS),
    aes(color = condition), size = 3, alpha = PLOT_ALPHA
  ) +
  scale_color_manual(
    name = "Exp 2b",
    values = EXP2B_COLORS,
    labels = \(x) gsub("Intimacy: (\\d+)", "\\1 intimacy", x),
    guide = guide_legend(order = 3)
  ) +
  # Correlation labels
  geom_label(
    data = correlations_combined |> filter(model != "Social (pre-reg)"),
    aes(x = -Inf, y = Inf, label = label),
    inherit.aes = FALSE, hjust = -0.08, vjust = 1.5, size = 4.5,
    fill = "white", alpha = 0.5, label.size = 0
  ) +
  scale_shape_manual(values = c(16, 17, 15, 18), guide = guide_legend(order = 1)) +
  labs(x = "Model belief update", y = "Human belief update", shape = "Action") +

  theme(
    legend.position = "right",
    legend.spacing.y = unit(0, "lines"),
  ) +
  facet_wrap(~model, nrow = 1) +
  coord_fixed_symmetric(df_plot$model_belief_update, df_plot$human_belief_update)

ggsave(here("figures", "exp2_combined_correlation_main.pdf"), device = cairo_pdf)
```

Preregistered vs modified model comparison:

```{r}
#| fig-width: 10
#| fig-asp: 0.4

df_plot_comp <- df_combined |>
  filter(model %in% c("Social (pre-reg)", "Social")) |>
  mutate(model = factor(model, levels = c("Social (pre-reg)", "Social")))

ggplot(df_plot_comp, aes(x = model_belief_update, y = human_belief_update, shape = action)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
  # Exp 2a points (motivation)
  geom_errorbar(
    data = \(d) filter(d, condition %in% EXP2A_CONDITIONS),
    aes(ymin = ci_lower, ymax = ci_upper, color = condition), width = 0.02
  ) +
  geom_point(
    data = \(d) filter(d, condition %in% EXP2A_CONDITIONS),
    aes(color = condition), size = 3, alpha = PLOT_ALPHA
  ) +
  scale_color_manual(name = "Exp 2a", values = MOTIVATION_COLORS, guide = guide_legend(order = 2)) +
  new_scale_color() +
  # Exp 2b points (intimacy)
  geom_errorbar(
    data = \(d) filter(d, condition %in% EXP2B_CONDITIONS),
    aes(ymin = ci_lower, ymax = ci_upper, color = condition), width = 0.02
  ) +
  geom_point(
    data = \(d) filter(d, condition %in% EXP2B_CONDITIONS),
    aes(color = condition), size = 3, alpha = PLOT_ALPHA
  ) +
  scale_color_manual(
    name = "Exp 2b",
    values = EXP2B_COLORS,
    labels = \(x) gsub("Intimacy: (\\d+)", "\\1 intimacy", x),
    guide = guide_legend(order = 3)
  ) +
  # Correlation labels
  geom_label(
    data = correlations_combined |> filter(model %in% c("Social (pre-reg)", "Social")),
    aes(x = -Inf, y = Inf, label = label),
    inherit.aes = FALSE, hjust = -0.08, vjust = 1.5, size = 4.5,
    fill = "white", alpha = 0.5, label.size = 0
  ) +
  scale_shape_manual(values = c(16, 17, 15, 18), guide = guide_legend(order = 1)) +
  labs(x = "Model belief update", y = "Human belief update", shape = "Action") +
  theme(
    legend.position = "right",
    legend.spacing.y = unit(0.0, "lines"),
  ) +
  facet_wrap(~model, nrow = 1) +
  coord_fixed_symmetric(df_plot_comp$model_belief_update, df_plot_comp$human_belief_update)

ggsave(here("figures", "exp2_combined_correlation_comparison.pdf"), device = cairo_pdf)
```

## Summary Table

```{r}
correlations_combined |>
  select(model, r, ci_lower, ci_upper) |>
  knitr::kable(
    digits = 3,
    caption = "Aggregated model-human correlations (Exp 2a + 2b)",
    col.names = c("Model", "r", "CI lower", "CI upper")
  )
```
