---
title: Experiment 1 (forward planning) data analysis
author: Alicia Chen
date: 2026-01-03
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    page-layout: full
    search: false
knitr:
  opts_chunk:
    dev: "ragg_png"
execute:
  warning: false 
---

```{r setup}
source(here::here("analysis", "utils.R"))
setup_analysis()
```

This experiment manipulates how much two people want the food (`low` vs. `high` reward), and the relationship between the two people (0, 50, 75, or 100 on a 100 point scale between 'intimate' and 'formal'). 

```{r}
# main_trials_long.csv is generated by analysis/json_to_csv.py
# It excludes participants who failed attention check or got 0 on memory check
df <- read_csv(here("data", "forw_plan", "main_trials_long.csv")) |>
  mutate(
    motivation = factor(motivation, levels = c("low", "high")),
    model = "humans"
  )
```

Calculate means

```{r}
df_summary <- df |>
  group_by(intimacy, motivation, action) |>
  tidyboot_mean(p_action)
```

Plot

```{r}
ggplot(df_summary, aes(x = action, y = empirical_stat, color = intimacy, group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), width = 0.2
  ) +
  geom_line(position = position_dodge(width = 0.2), size = 1) +
  scale_color_intimacy() +
  labs(y = "P(action)", x = "Action", color = "Intimacy") +
  theme(legend.position = "right") +
  facet_wrap(~motivation, labeller = label_both)
```

Plot each scenario separately
```{r}
df_summary_scenario <- df |>
  group_by(scenario_label, intimacy, motivation, action) |>
  tidyboot_mean(p_action)
```

```{r}
ggplot(df_summary_scenario, aes(x = action, y = empirical_stat, color = intimacy, group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), width = 0.2
  ) +
  geom_line(position = position_dodge(width = 0.2), size = 1) +
  scale_color_intimacy() +
  labs(y = "P(action)", x = "Action", color = "Intimacy") +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid(motivation~scenario_label, labeller = labeller(motivation = label_both, scenario_label = label_wrap_gen(15)))
```



## Model comparison

### Model description

We fit three models to the human data. In all models, a joint actor chooses an action proportional to its total utility, following a softmax choice rule with inverse temperature $\alpha$.

**Full model**: The total utility of an action $a$ given motivational state $s$ and intimacy $I$ is:

$$U_{tot}(a|s, I) = w_r \cdot r(a|s, I) - w_d \cdot d(a|I) - w_c \cdot c(a)$$

where:

- $r(a|s, I)$ is the reward, scaled by the motivational state $s$ and the intimacy $I$. The more the actors want to eat the food, and the more intimate the relationship, the higher the reward of eating the food together (for Action 0, the reward is 0).
- $d(a|I) = (1 - I) \cdot \rho(a)$ is the discomfort from saliva transfer, where $\rho(a) = [0, 0, 1, 2]$ is the saliva-transfer risk for actions 0-3. Higher intimacy mitigates discomfort.
- $c(a) = [0, 1, 1, 1]$ is the sharing cost, capturing the baseline cost of coordination.

**Vanilla model**: Includes all utility terms but does NOT consider how intimacy scales either reward or cost. Reward and discomfort are not modulated by relationship closeness.

**Discomfort-only model**: Only considers how intimacy mitigates the discomfort of taking a risky action ($U = -w_d \cdot d(a|I)$), without a representation of reward or sharing cost.

### Load model predictions

```{r}
df_preds <- read_csv(here("model", "forward_planning_fits.csv")) |>
  select(subject_id, scenario_label, intimacy, motivation, action, p_action,
         pred_full, pred_vanilla, pred_discomfort_only) |>
  pivot_longer(
    cols = c(p_action, pred_full, pred_vanilla, pred_discomfort_only),
    names_to = "model",
    values_to = "p_action"
  ) |>
  mutate(
    model = case_when(
      model == "p_action" ~ "Humans",
      model == "pred_full" ~ "Social planning",
      model == "pred_vanilla" ~ "Vanilla planning",
      model == "pred_discomfort_only" ~ "Discomfort only"
    ),
    model = factor(model, levels = c(
      "Vanilla planning", "Discomfort only", "Social planning", "Humans"
    )),
    motivation = factor(motivation, levels = c("low", "high"))
  )

df_preds_summary <- df_preds |>
  group_by(intimacy, motivation, action, model) |>
  tidyboot_mean(p_action) |>
  ungroup()
```

### Model predictions vs human data

Stipulated parameter models:
```{r fig.width=12, fig.height=5.2}
p_model_comparison <- df_preds_summary |>
  filter(model %in% c("Vanilla planning", "Discomfort only", "Social planning", "Humans")) |>
  mutate(model = factor(model, levels = c("Vanilla planning", "Discomfort only", "Social planning", "Humans"))) |>
  ggplot(aes(x = action, y = empirical_stat, color = intimacy, shape = factor(action), group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), width = 0.4
  ) +
  geom_line(position = position_dodge(width = 0.2)) +
  scale_color_intimacy() +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(y = "P(action)", x = "Action", color = "Intimacy", shape = "Action") +
  facet_grid(motivation ~ model, labeller = labeller(motivation = label_both))
p_model_comparison
save_figure(p_model_comparison, "exp1_model_comparison.pdf", width = FIG_WIDTH_LARGE, height = FIG_HEIGHT_TALL)
```

### Model-human correlation

```{r}
# Bootstrap function for correlation CI (handles NAs)
boot_cor <- function(x, y, n_boot = 1000) {
  # Remove NAs
  complete <- complete.cases(x, y)
  x <- x[complete]
  y <- y[complete]
  n <- length(x)
  if (n < 3) return(list(r = NA_real_, ci_lower = NA_real_, ci_upper = NA_real_))
  boot_rs <- replicate(n_boot, {
    idx <- sample(n, replace = TRUE)
    cor(x[idx], y[idx])
  })
  list(
    r = cor(x, y),
    ci_lower = as.numeric(quantile(boot_rs, 0.025, na.rm = TRUE)),
    ci_upper = as.numeric(quantile(boot_rs, 0.975, na.rm = TRUE))
  )
}

# Get human data for joining
df_humans <- df_preds_summary |>
  filter(model == "Humans") |>
  select(intimacy, motivation, action,
         Humans = empirical_stat,
         Humans_ci_lower = ci_lower,
         Humans_ci_upper = ci_upper)

# Create correlation data by joining model predictions with human data
df_corr_all <- df_preds_summary |>
  filter(model != "Humans") |>
  select(intimacy, motivation, action, model, model_pred = empirical_stat) |>
  left_join(df_humans, by = c("intimacy", "motivation", "action")) |>
  filter(!is.na(model_pred), !is.na(Humans)) |>
  mutate(action = factor(action))

# Debug: check what models exist
cat("Models in df_corr_all:", unique(as.character(df_corr_all$model)), "\n")

# Use all models for correlation
df_corr <- df_corr_all |>
  mutate(model = droplevels(model))

cat("Rows in df_corr:", nrow(df_corr), "\n")
```

```{r fig.width=12, fig.height=4}
# Calculate correlations
correlations <- df_corr |>
  group_by(model) |>
  summarize(
    boot_result = list(boot_cor(model_pred, Humans)),
    .groups = "drop"
  ) |>
  mutate(
    r = as.numeric(sapply(boot_result, function(x) x$r)),
    ci_lower = as.numeric(sapply(boot_result, function(x) x$ci_lower)),
    ci_upper = as.numeric(sapply(boot_result, function(x) x$ci_upper)),
    label = paste0("r = ", round(r, 2), " (", round(ci_lower, 2), ", ", round(ci_upper, 2), ")")
  ) |>
  select(-boot_result)

p_correlation <- ggplot(df_corr, aes(x = model_pred, y = Humans, color = intimacy, shape = action)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
  geom_errorbar(aes(ymin = Humans_ci_lower, ymax = Humans_ci_upper), width = 0.02) +
  geom_point(size = 3) +
  geom_text(data = correlations, aes(x = -Inf, y = Inf, label = label),
            inherit.aes = FALSE, hjust = -0.1, vjust = 1.5, size = 5) +
  scale_color_intimacy() +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(x = "Model", y = "Humans", color = "Intimacy", shape = "Action") +
  facet_wrap(~model, nrow = 1) +
  coord_fixed_symmetric(df_corr$model_pred, df_corr$Humans)
p_correlation
save_figure(p_correlation, "exp1_correlation.pdf", width = FIG_WIDTH_STANDARD, height = FIG_HEIGHT_SHORT)
```

Plot motivation on separate panels:

```{r fig.width=12, fig.height=6}
correlations_motiv <- df_corr |>
  group_by(model, motivation) |>
  summarize(
    boot_result = list(boot_cor(model_pred, Humans)),
    .groups = "drop"
  ) |>
  mutate(
    r = as.numeric(sapply(boot_result, function(x) x$r)),
    ci_lower = as.numeric(sapply(boot_result, function(x) x$ci_lower)),
    ci_upper = as.numeric(sapply(boot_result, function(x) x$ci_upper)),
    label = paste0("r = ", round(r, 2), " (", round(ci_lower, 2), ", ", round(ci_upper, 2), ")")
  ) |>
  select(-boot_result)

ggplot(df_corr, aes(x = model_pred, y = Humans, color = intimacy, shape = action)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
  geom_errorbar(aes(ymin = Humans_ci_lower, ymax = Humans_ci_upper), width = 0.02) +
  geom_point(size = 3) +
  geom_text(data = correlations_motiv, aes(x = -Inf, y = Inf, label = label),
            inherit.aes = FALSE, hjust = -0.1, vjust = 1.5, size = 5) +
  scale_color_intimacy() +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(x = "Model", y = "Humans", color = "Intimacy", shape = "Action") +
  facet_grid(motivation~model, labeller = labeller(motivation = label_both)) +
  coord_fixed_symmetric(df_corr$model_pred, df_corr$Humans)
```

### Model fit results

```{r}
fit_results <- read_csv(here("model", "forward_planning_fit_results.csv"))
fit_results |>
  select(model, nll, n_params, starts_with("param_")) |>
  knitr::kable(digits = 3, caption = "Model fitting results")
```

The full model provides the best fit (lowest NLL), followed by the vanilla model and then the discomfort-only model.

### AIC/BIC comparison

```{r}
n_obs <- nrow(df)  # number of observations

fit_results <- fit_results |>
  mutate(
    AIC = 2 * n_params + 2 * nll,
    BIC = n_params * log(n_obs) + 2 * nll
  )

fit_results |>
  select(model, nll, n_params, AIC, BIC) |>
  arrange(AIC) |>
  knitr::kable(digits = 2, caption = "Model comparison (lower AIC/BIC is better)")
```

```{r}
# Compute differences from best model (full)
full_aic <- fit_results$AIC[fit_results$model == "full"]
full_bic <- fit_results$BIC[fit_results$model == "full"]

fit_results |>
  mutate(
    delta_AIC = AIC - full_aic,
    delta_BIC = BIC - full_bic
  ) |>
  select(model, delta_AIC, delta_BIC) |>
  knitr::kable(digits = 2, caption = "Difference from best model (Full)")
```

**Interpretation**: The full (social planning) model with intimacy scaling has the lowest AIC/BIC, indicating that intimacy modulates both reward and discomfort.