---
title: Experiment 2a (inverse planning - intimacy inference) data analysis
author: Alicia Chen
date: 2026-01-15
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    page-layout: full
    search: false
execute:
  warning: false
---

```{r setup}
source(here::here("analysis", "utils.R"))
setup_analysis()
```

This experiment presents participants with scenarios where they observe an action taken by one person toward another. Participants rate their belief about the intimacy of the relationship before (prior) and after (posterior) observing the action. The experiment manipulates:

- **Action**: Which action was taken (action 0-3, varying in saliva-transfer risk)
- **Motivation**: Whether both people want the food (high vs low)

## Load human data

```{r}
# main_trials_long.csv is generated by analysis/json_to_csv.py
# It excludes participants who failed attention check or got 0 on memory check
df <- read_csv(here("data", "inv_plan_intimacy", "main_trials_long.csv")) |>
  mutate(
    action = str_replace(action_condition, "action_", ""),
    motivation = factor(motivation, levels = c("low", "high")),
    stage = factor(stage, levels = c("prior", "posterior")),
    model = "Humans"
  ) |>
  select(-action_condition)

# Calculate belief updates
df <- calculate_belief_update(df, "intimacy_rating")
```

## Human data: Prior vs posterior ratings

### Distribution of ratings

```{r}
ggplot(df, aes(x = intimacy_rating, fill = action)) +
  geom_density(alpha = 0.5) +
  scale_fill_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  labs(y = "Density", x = "Intimacy rating", fill = "Action") +
  theme(legend.position = "bottom") +
  facet_grid(
    stage ~ motivation,
    labeller = labeller(motivation = c("low" = "motivation: low", "high" = "motivation: high"))
  )
```

### Summary statistics

```{r}
df_summary <- df |>
  group_by(motivation, action, stage, model) |>
  tidyboot_mean(intimacy_rating)
```

### Prior vs posterior by action and motivation

```{r fig.width=12, fig.height=6}
ggplot() +
  geom_violin(
    data = df,
    aes(x = stage, y = intimacy_rating, fill = action),
    position = position_dodge(width = 0.8),
    alpha = 0.5,
    width = 0.7
  ) +
  geom_point(
    data = df_summary,
    aes(x = stage, y = empirical_stat, color = action, group = action),
    position = position_dodge(width = 0.8),
    size = 3
  ) +
  geom_errorbar(
    data = df_summary,
    aes(x = stage, ymin = ci_lower, ymax = ci_upper, color = action, group = action),
    position = position_dodge(width = 0.8),
    width = 0.1,
    size = 1
  ) +
  geom_line(
    data = df_summary,
    aes(x = stage, y = empirical_stat, color = action, group = action),
    position = position_dodge(width = 0.8),
    size = 1
  ) +
  facet_grid(motivation ~ action, labeller = label_both) +
  scale_fill_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  scale_color_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  labs(
    title = "Intimacy ratings before and after observing action",
    y = "Intimacy rating",
    x = "Stage",
    fill = "Observed action",
    color = "Observed action"
  )
```

### Belief updates

```{r}
df_belief_update_summary <- df |>
  filter(stage == "posterior") |>
  group_by(motivation, action, model) |>
  tidyboot_mean(belief_update)
```

```{r}
ggplot(df_belief_update_summary, aes(x = action, y = empirical_stat, group = motivation, fill = motivation)) +
  geom_col(position = position_dodge(width = 0.9), alpha = 0.9) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.9), size = 1, width = 0.1, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Change in intimacy rating after observing action",
    y = "Belief update",
    x = "Observed action",
    fill = "Motivation"
  ) +
  theme(legend.position = "bottom")
```

## Model comparison

### Model description

We compare three models that perform inverse planning to infer intimacy from observed actions:

**Full model (Social inv plan)**: Inverts a forward model where actors consider reward (scaled by intimacy and motivation), discomfort (scaled by intimacy), and effort costs.

**Vanilla model**: Inverts a forward model with reward, discomfort, and effort, but without intimacy scaling these utilities.

**Discomfort-only model**: Inverts a forward model where actors only consider discomfort from saliva transfer (scaled by intimacy).

**Full modified model**: A hybrid model that uses vanilla-style reward (no intimacy scaling) with full-model discomfort (scales with 1 - intimacy). This captures the intuition that observers don't think actors value sharing more with closer friends, but do recognize that discomfort decreases with intimacy.

All models assume a uniform prior over intimacy (0-100 scale) and compute posteriors via Bayesian inference. The observer inverse temperature ($\alpha_{\text{observer}}$) is fitted to human data.

### Fitted parameters

```{r}
df_fit_results <- read_csv(here("model", "inverse_planning_fit_results.csv")) |>
  filter(experiment == "intimacy") |>
  mutate(
    model_label = case_when(
      model == "full" ~ "Social inv plan",
      model == "vanilla" ~ "Vanilla inv plan",
      model == "discomfort_only" ~ "Discomfort only",
      model == "full_lm" ~ "Social inv plan (LM)",
      model == "vanilla_lm" ~ "Vanilla inv plan (LM)",
      model == "discomfort_only_lm" ~ "Discomfort only (LM)",
      model == "full_modified" ~ "Social inv plan (modified)",
      model == "full_lm_modified" ~ "Social inv plan (LM, modified)"
    )
  ) |>
  select(model_label, alpha_observer, beta, nll, n_params)

df_fit_results |>
  knitr::kable(
    digits = 3,
    caption = "Fitted observer parameters for intimacy inference",
    col.names = c("Model", "α_observer", "β", "NLL", "n_params")
  )
```

### Load model predictions

```{r}
df_model <- read_csv(here("model", "inv_plan_intimacy_preds_summary.csv")) |>
  rename(motivation = reward_condition) |>
  mutate(
    action = as.character(action),
    motivation = factor(motivation, levels = c("low", "high")),
    # Belief update from uniform prior (mean = 50)
    belief_update = expected_intimacy - 50,
    model = case_when(
      model == "full" ~ "Social inv plan",
      model == "vanilla" ~ "Vanilla inv plan",
      model == "discomfort_only" ~ "Discomfort only",
      model == "full_modified" ~ "Social inv plan (modified)"
    ),
    param_source = case_when(
      param_source == "stipulated" ~ "Stipulated",
      param_source == "lm" ~ "LM"
    )
  )

# Filter to stipulated parameters only for main comparison
df_model_stipulated <- df_model |>
  filter(param_source == "Stipulated")
```

### Combine human and model data

```{r}
df_combined <- bind_rows(
  df_belief_update_summary |>
    mutate(belief_update = empirical_stat) |>
    select(motivation, action, model, belief_update, ci_lower, ci_upper),
  df_model_stipulated |>
    group_by(motivation, action, model) |>
    summarize(belief_update = mean(belief_update), .groups = "drop") |>
    mutate(ci_lower = NA_real_, ci_upper = NA_real_)
) |>
  mutate(
    model = factor(model, levels = c("Vanilla inv plan", "Discomfort only", "Social inv plan", "Social inv plan (modified)", "Humans"))
  )
```

### Belief updates: Model vs human comparison

```{r fig.width=18, fig.height=5}
ggplot(df_combined, aes(x = action, y = belief_update, group = interaction(model, motivation), fill = motivation)) +
  geom_col(position = position_dodge(width = 0.9), alpha = 0.9) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.9), size = 1, width = 0.1, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Change in intimacy rating after observing action",
    y = "Belief update",
    x = "Observed action",
    fill = "Motivation"
  ) +
  theme(legend.position = "bottom") +
  facet_wrap(~model, nrow = 1)
```

### Model-human correlation

```{r}
# Prepare data for correlation
df_corr <- df_combined |>
  select(motivation, action, model, belief_update) |>
  pivot_wider(names_from = model, values_from = belief_update)
```

```{r}
# Reshape for correlation plots
df_corr_long <- df_corr |>
  pivot_longer(
    cols = c(`Vanilla inv plan`, `Discomfort only`, `Social inv plan`, `Social inv plan (modified)`),
    names_to = "model",
    values_to = "model_pred"
  ) |>
  rename(human_pred = Humans) |>
  filter(!is.na(model_pred), !is.na(human_pred)) |>
  mutate(
    model = factor(model, levels = c("Vanilla inv plan", "Discomfort only", "Social inv plan", "Social inv plan (modified)")),
    action = factor(action)
  )

# Calculate correlations
correlations <- df_corr_long |>
  group_by(model) |>
  summarize(
    boot_result = list(boot_cor(model_pred, human_pred)),
    .groups = "drop"
  ) |>
  mutate(
    r = sapply(boot_result, function(x) x$r),
    ci_lower = sapply(boot_result, function(x) x$ci_lower),
    ci_upper = sapply(boot_result, function(x) x$ci_upper),
    label = paste0("r = ", round(r, 2), " [", round(ci_lower, 2), ", ", round(ci_upper, 2), "]")
  ) |>
  select(-boot_result)
```

```{r fig.width=16, fig.height=4}
ggplot(df_corr_long, aes(x = model_pred, y = human_pred, color = motivation, shape = action)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
  geom_point(size = 4) +
  geom_text(
    data = correlations,
    aes(x = -15, y = 25, label = label),
    inherit.aes = FALSE, hjust = 0, size = 4
  ) +
  scale_color_brewer(palette = "Set2") +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(x = "Model prediction", y = "Human data", color = "Motivation", shape = "Action") +
  facet_wrap(~model, nrow = 1) +
  coord_fixed(xlim = c(-20, 30), ylim = c(-20, 30))
```

### Correlation summary

```{r}
correlations |>
  select(model, r, ci_lower, ci_upper) |>
  knitr::kable(
    digits = 3,
    caption = "Model-human correlation (belief updates)",
    col.names = c("Model", "r", "CI lower", "CI upper")
  )
```

## LM-derived parameter models

We also compare models using LLM-derived scenario-specific parameters.

```{r}
df_model_lm <- df_model |>
  filter(param_source == "LM") |>
  mutate(
    model = case_when(
      model == "Social inv plan (modified)" ~ "Social inv plan (LM, modified)",
      TRUE ~ paste0(model, " (LM)")
    )
  )

df_combined_lm <- bind_rows(
  df_belief_update_summary |>
    mutate(belief_update = empirical_stat) |>
    select(motivation, action, model, belief_update, ci_lower, ci_upper),
  df_model_lm |>
    group_by(motivation, action, model) |>
    summarize(belief_update = mean(belief_update), .groups = "drop") |>
    mutate(ci_lower = NA_real_, ci_upper = NA_real_)
) |>
  mutate(
    model = factor(model, levels = c(
      "Vanilla inv plan (LM)", "Discomfort only (LM)", "Social inv plan (LM)", "Social inv plan (LM, modified)", "Humans"
    ))
  )
```

```{r fig.width=18, fig.height=5}
ggplot(df_combined_lm, aes(x = action, y = belief_update, group = interaction(model, motivation), fill = motivation)) +
  geom_col(position = position_dodge(width = 0.9), alpha = 0.9) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.9), size = 1, width = 0.1, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Change in intimacy rating after observing action (LM parameters)",
    y = "Belief update",
    x = "Observed action",
    fill = "Motivation"
  ) +
  theme(legend.position = "bottom") +
  facet_wrap(~model, nrow = 1)
```

### LM model correlations

```{r}
df_corr_lm <- df_combined_lm |>
  select(motivation, action, model, belief_update) |>
  pivot_wider(names_from = model, values_from = belief_update)

df_corr_lm_long <- df_corr_lm |>
  pivot_longer(
    cols = c(`Vanilla inv plan (LM)`, `Discomfort only (LM)`, `Social inv plan (LM)`, `Social inv plan (LM, modified)`),
    names_to = "model",
    values_to = "model_pred"
  ) |>
  rename(human_pred = Humans) |>
  filter(!is.na(model_pred), !is.na(human_pred)) |>
  mutate(
    model = factor(model, levels = c("Vanilla inv plan (LM)", "Discomfort only (LM)", "Social inv plan (LM)", "Social inv plan (LM, modified)")),
    action = factor(action)
  )

correlations_lm <- df_corr_lm_long |>
  group_by(model) |>
  summarize(
    boot_result = list(boot_cor(model_pred, human_pred)),
    .groups = "drop"
  ) |>
  mutate(
    r = sapply(boot_result, function(x) x$r),
    ci_lower = sapply(boot_result, function(x) x$ci_lower),
    ci_upper = sapply(boot_result, function(x) x$ci_upper),
    label = paste0("r = ", round(r, 2), " [", round(ci_lower, 2), ", ", round(ci_upper, 2), "]")
  ) |>
  select(-boot_result)
```

```{r fig.width=16, fig.height=4}
ggplot(df_corr_lm_long, aes(x = model_pred, y = human_pred, color = motivation, shape = action)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
  geom_point(size = 4) +
  geom_text(
    data = correlations_lm,
    aes(x = -15, y = 25, label = label),
    inherit.aes = FALSE, hjust = 0, size = 4
  ) +
  scale_color_brewer(palette = "Set2") +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(x = "Model prediction", y = "Human data", color = "Motivation", shape = "Action") +
  facet_wrap(~model, nrow = 1) +
  coord_fixed(xlim = c(-20, 30), ylim = c(-20, 30))
```

### Summary: All model correlations

```{r}
bind_rows(
  correlations |> mutate(param_source = "Stipulated"),
  correlations_lm |> mutate(param_source = "LM")
) |>
  select(param_source, model, r, ci_lower, ci_upper) |>
  arrange(param_source, model) |>
  knitr::kable(
    digits = 3,
    caption = "Model-human correlations",
    col.names = c("Parameters", "Model", "r", "CI lower", "CI upper")
  )
```

## Scenario-level analysis

### Human belief updates by scenario

```{r}
df_belief_update_scenario <- df |>
  filter(stage == "posterior") |>
  group_by(scenario_label, motivation, action, model) |>
  tidyboot_mean(belief_update)
```

```{r fig.width=14, fig.height=10}
ggplot(df_belief_update_scenario, aes(x = action, y = empirical_stat, fill = motivation)) +
  geom_col(position = position_dodge(width = 0.9), alpha = 0.9) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.9), width = 0.1, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Belief updates by scenario",
    y = "Belief update",
    x = "Observed action",
    fill = "Motivation"
  ) +
  theme(legend.position = "bottom") +
  facet_wrap(~scenario_label, ncol = 4)
```

