---
title: Experiment 1 (forward planning) data analysis
author: Alicia Chen
date: 2026-01-03
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    page-layout: full
    search: false
execute:
  warning: false 
---

```{r setup}
library(here)
library(tidyverse)
library(tidyboot)
library(lme4)

theme_set(theme_classic(base_size = 18))
set.seed(67)
```

This experiment manipulates how much two people want the food (`low` vs. `high` reward), and the relationship between the two people (0, 50, 75, or 100 on a 100 point scale between 'intimate' and 'formal'). 

```{r}
# main_trials_long.csv is generated by analysis/json_to_csv.py
# It excludes participants who failed attention check or got 0 on memory check
df <- read_csv(here("data", "forw_plan", "main_trials_long.csv")) |>
  mutate(
    motivation = factor(motivation, levels = c("low", "high")),
    model = "humans"
  )
```

Calculate means

```{r}
df_summary <- df |>
  group_by(intimacy, motivation, action) |>
  tidyboot_mean(p_action)
```

Plot

```{r}
ggplot(df_summary, aes(x = action, y = empirical_stat, color = intimacy, group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), width = 0.2
  ) +
  geom_line(position = position_dodge(width = 0.2), size = 1) +
  scale_color_viridis_c(begin = 0, end = 0.85) +
  labs(y = "P(action)", x = "Action", color = "Intimacy") +
  theme(legend.position = "bottom") +
  facet_wrap(~motivation, labeller = label_both)
```

Plot each scenario separately
```{r}
df_summary_scenario <- df |>
  group_by(scenario_label, intimacy, motivation, action) |>
  tidyboot_mean(p_action)
```

```{r}
ggplot(df_summary_scenario, aes(x = action, y = empirical_stat, color = intimacy, group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), width = 0.2
  ) +
  geom_line(position = position_dodge(width = 0.2), size = 1) +
  scale_color_viridis_c(begin = 0, end = 0.85) +
  labs(y = "P(action)", x = "Action", color = "Intimacy") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid(motivation~scenario_label, labeller = labeller(motivation = label_both, scenario_label = label_wrap_gen(15)))
```



## Model comparison

### Model description

We fit three models to the human data. In all models, a joint actor chooses an action proportional to its total utility, following a softmax choice rule with inverse temperature $\alpha$.

**Full model**: The total utility of an action $a$ given motivational state $s$ and intimacy $I$ is:

$$U_{tot}(a|s, I) = w_r \cdot r(a|s, I) - w_d \cdot d(a|I) - w_c \cdot c(a)$$

where:

- $r(a|s, I)$ is the reward, scaled by the motivational state $s$ and the intimacy $I$. The more the actors want to eat the food, and the more intimate the relationship, the higher the reward of eating the food together (for Action 0, the reward is 0).
- $d(a|I) = (1 - I) \cdot \rho(a)$ is the discomfort from saliva transfer, where $\rho(a) = [0, 0, 1, 2]$ is the saliva-transfer risk for actions 0-3. Higher intimacy mitigates discomfort.
- $c(a) = [0, 1, 1, 1]$ is the sharing cost, capturing the baseline cost of coordination.

**Vanilla model**: Includes all utility terms but does NOT consider how intimacy scales either reward or cost. Reward and discomfort are not modulated by relationship closeness.

**Discomfort-only model**: Only considers how intimacy mitigates the discomfort of taking a risky action ($U = -w_d \cdot d(a|I)$), without a representation of reward or sharing cost.

**LLM-based models**: We also fit versions of all three models (full, vanilla, discomfort-only) using LLM-derived scenario-specific parameters instead of stipulated constants. The LLM (Llama-3.3-70B-Instruct) was prompted to rate each scenario's actions on scales of saliva-transfer risk, coordination effort, and sharing reward.

**LLM Direct**: Direct LLM predictions without any cognitive model structure. The LLM (Llama-3.3-70B-Instruct) was presented with the exact experiment instructions and stimuli, and asked to predict the probability distribution over actions for each scenario and condition. This tests whether an LLM can directly simulate human behavior without explicit utility-based reasoning.

### Load model predictions

```{r}
# Load LLM direct predictions and join with model fits
llm_direct <- read_csv(here("model", "llm_direct_predictions.csv")) |>
  select(scenario_label, intimacy, motivation, action, pred_llm = pred_llm_mean)

df_preds <- read_csv(here("model", "forward_planning_fits.csv")) |>
  left_join(llm_direct, by = c("scenario_label", "intimacy", "motivation", "action")) |>
  select(subject_id, scenario_label, intimacy, motivation, action, p_action,
         pred_full, pred_vanilla, pred_discomfort_only,
         pred_full_lm, pred_vanilla_lm, pred_discomfort_only_lm, pred_llm) |>
  pivot_longer(
    cols = c(p_action, pred_full, pred_vanilla, pred_discomfort_only,
             pred_full_lm, pred_vanilla_lm, pred_discomfort_only_lm, pred_llm),
    names_to = "model",
    values_to = "p_action"
  ) |>
  mutate(
    model = case_when(
      model == "p_action" ~ "Humans",
      model == "pred_full" ~ "Social planning",
      model == "pred_vanilla" ~ "Vanilla planning",
      model == "pred_discomfort_only" ~ "Discomfort only",
      model == "pred_full_lm" ~ "Social planning (LM)",
      model == "pred_vanilla_lm" ~ "Vanilla planning (LM)",
      model == "pred_discomfort_only_lm" ~ "Discomfort only (LM)",
      model == "pred_llm" ~ "LLM only"
    ),
    # Create param_source variable for grouping
    param_source = case_when(
      model %in% c("Vanilla planning", "Discomfort only", "Social planning") ~ "Stipulated",
      model %in% c("Vanilla planning (LM)", "Discomfort only (LM)", "Social planning (LM)") ~ "LM",
      model == "LLM only" ~ "LLM Direct",
      model == "Humans" ~ "Humans"
    ),
    # Model type without param source suffix (for parallel plotting)
    model_type = case_when(
      model %in% c("Vanilla planning", "Vanilla planning (LM)") ~ "Vanilla",
      model %in% c("Discomfort only", "Discomfort only (LM)") ~ "Discomfort only",
      model %in% c("Social planning", "Social planning (LM)") ~ "Social planning",
      model == "LLM only" ~ "LLM only",
      model == "Humans" ~ "Humans"
    ),
    model = factor(model, levels = c(
      "Vanilla planning", "Discomfort only", "Social planning",
      "Vanilla planning (LM)", "Discomfort only (LM)", "Social planning (LM)",
      "LLM only", "Humans"
    )),
    model_type = factor(model_type, levels = c("Vanilla", "Discomfort only", "Social planning", "LLM only", "Humans")),
    motivation = factor(motivation, levels = c("low", "high"))
  )

df_preds_summary <- df_preds |>
  group_by(intimacy, motivation, action, model, param_source, model_type) |>
  tidyboot_mean(p_action) |>
  ungroup()
```

### Model predictions vs human data

Stipulated parameter models:
```{r fig.width=12, fig.height=5.2}
df_preds_summary |>
  filter(model %in% c("Vanilla planning", "Discomfort only", "Social planning", "Humans")) |>
  mutate(model = factor(model, levels = c("Vanilla planning", "Discomfort only", "Social planning", "Humans"))) |>
  ggplot(aes(x = action, y = empirical_stat, color = intimacy, shape = factor(action), group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), width = 0.4
  ) +
  geom_line(position = position_dodge(width = 0.2)) +
  scale_color_viridis_c(begin = 0, end = 0.85) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(y = "P(action)", x = "Action", color = "Intimacy", shape = "Action",
       title = "Stipulated Parameters") +
  facet_grid(motivation ~ model, labeller = labeller(motivation = label_both))
```

LM-derived parameter models:
```{r fig.width=12, fig.height=5.2}
df_preds_summary |>
  filter(model %in% c("Vanilla planning (LM)", "Discomfort only (LM)", "Social planning (LM)", "Humans")) |>
  mutate(model = factor(model, levels = c("Vanilla planning (LM)", "Discomfort only (LM)", "Social planning (LM)", "Humans"))) |>
  ggplot(aes(x = action, y = empirical_stat, color = intimacy, shape = factor(action), group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), width = 0.4
  ) +
  geom_line(position = position_dodge(width = 0.2)) +
  scale_color_viridis_c(begin = 0, end = 0.85) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(y = "P(action)", x = "Action", color = "Intimacy", shape = "Action",
       title = "LM-Derived Parameters") +
  facet_grid(motivation ~ model, labeller = labeller(motivation = label_both))
```

LLM Direct predictions:
```{r fig.width=6, fig.height=5.2}
df_preds_summary |>
  filter(model %in% c("LLM only", "Humans")) |>
  ggplot(aes(x = action, y = empirical_stat, color = intimacy, shape = factor(action), group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), width = 0.4
  ) +
  geom_line(position = position_dodge(width = 0.2)) +
  scale_color_viridis_c(begin = 0, end = 0.85) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(y = "P(action)", x = "Action", color = "Intimacy", shape = "Action") +
  facet_grid(motivation ~ model, labeller = labeller(motivation = label_both))
```

### Model-human correlation

```{r}
# Bootstrap function for correlation CI (handles NAs)
boot_cor <- function(x, y, n_boot = 1000) {
  # Remove NAs
  complete <- complete.cases(x, y)
  x <- x[complete]
  y <- y[complete]
  n <- length(x)
  if (n < 3) return(list(r = NA_real_, ci_lower = NA_real_, ci_upper = NA_real_))
  boot_rs <- replicate(n_boot, {
    idx <- sample(n, replace = TRUE)
    cor(x[idx], y[idx])
  })
  list(
    r = cor(x, y),
    ci_lower = as.numeric(quantile(boot_rs, 0.025, na.rm = TRUE)),
    ci_upper = as.numeric(quantile(boot_rs, 0.975, na.rm = TRUE))
  )
}

# Get human data for joining
df_humans <- df_preds_summary |>
  filter(model == "Humans") |>
  select(intimacy, motivation, action,
         Humans = empirical_stat,
         Humans_ci_lower = ci_lower,
         Humans_ci_upper = ci_upper)

# Create correlation data by joining model predictions with human data
df_corr_all <- df_preds_summary |>
  filter(model != "Humans") |>
  select(intimacy, motivation, action, model, model_pred = empirical_stat) |>
  left_join(df_humans, by = c("intimacy", "motivation", "action")) |>
  filter(!is.na(model_pred), !is.na(Humans)) |>
  mutate(action = factor(action))

# Debug: check what models exist
cat("Models in df_corr_all:", unique(as.character(df_corr_all$model)), "\n")

# Split by model type
df_corr_stipulated <- df_corr_all |>
  filter(model %in% c("Vanilla planning", "Discomfort only", "Social planning")) |>
  mutate(model = droplevels(model))

df_corr_lm <- df_corr_all |>
  filter(model %in% c("Vanilla planning (LM)", "Discomfort only (LM)", "Social planning (LM)")) |>
  mutate(model = droplevels(model))

df_corr_llm_direct <- df_corr_all |>
  filter(model == "LLM only") |>
  mutate(model = droplevels(model))

cat("Rows in df_corr_stipulated:", nrow(df_corr_stipulated), "\n")
cat("Rows in df_corr_lm:", nrow(df_corr_lm), "\n")
cat("Rows in df_corr_llm_direct:", nrow(df_corr_llm_direct), "\n")
```

Stipulated parameter models:
```{r fig.width=12, fig.height=4}
if (nrow(df_corr_stipulated) > 0) {
  # Calculate correlations for stipulated models
  correlations_stipulated <- df_corr_stipulated |>
    group_by(model) |>
    summarize(
      boot_result = list(boot_cor(model_pred, Humans)),
      .groups = "drop"
    ) |>
    mutate(
      r = as.numeric(sapply(boot_result, function(x) x$r)),
      ci_lower = as.numeric(sapply(boot_result, function(x) x$ci_lower)),
      ci_upper = as.numeric(sapply(boot_result, function(x) x$ci_upper)),
      label = paste0("r = ", round(r, 2), " [", round(ci_lower, 2), ", ", round(ci_upper, 2), "]")
    ) |>
    select(-boot_result)

  print(ggplot(df_corr_stipulated, aes(x = model_pred, y = Humans, color = intimacy, shape = action)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
    geom_errorbar(aes(ymin = Humans_ci_lower, ymax = Humans_ci_upper), width = 0.02) +
    geom_point(size = 3) +
    geom_text(data = correlations_stipulated, aes(x = 0.05, y = 0.9, label = label),
              inherit.aes = FALSE, hjust = 0, size = 5) +
    scale_color_viridis_c(begin = 0, end = 0.85) +
    scale_shape_manual(values = c(16, 17, 15, 18)) +
    labs(x = "Model", y = "Humans", color = "Intimacy", shape = "Action",
         title = "Stipulated Parameters") +
    facet_wrap(~model, nrow = 1) +
    coord_fixed(xlim = c(0, 1), ylim = c(0, 1)))
} else {
  cat("No data for stipulated models correlation plot\n")
}
```

LM-derived parameter models:
```{r fig.width=12, fig.height=4}
if (nrow(df_corr_lm) > 0) {
  # Calculate correlations for LM models
  correlations_lm <- df_corr_lm |>
    group_by(model) |>
    summarize(
      boot_result = list(boot_cor(model_pred, Humans)),
      .groups = "drop"
    ) |>
    mutate(
      r = as.numeric(sapply(boot_result, function(x) x$r)),
      ci_lower = as.numeric(sapply(boot_result, function(x) x$ci_lower)),
      ci_upper = as.numeric(sapply(boot_result, function(x) x$ci_upper)),
      label = paste0("r = ", round(r, 2), " [", round(ci_lower, 2), ", ", round(ci_upper, 2), "]")
    ) |>
    select(-boot_result)

  print(ggplot(df_corr_lm, aes(x = model_pred, y = Humans, color = intimacy, shape = action)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
    geom_errorbar(aes(ymin = Humans_ci_lower, ymax = Humans_ci_upper), width = 0.02) +
    geom_point(size = 3) +
    geom_text(data = correlations_lm, aes(x = 0.05, y = 0.9, label = label),
              inherit.aes = FALSE, hjust = 0, size = 5) +
    scale_color_viridis_c(begin = 0, end = 0.85) +
    scale_shape_manual(values = c(16, 17, 15, 18)) +
    labs(x = "Model", y = "Humans", color = "Intimacy", shape = "Action",
         title = "LM-Derived Parameters") +
    facet_wrap(~model, nrow = 1) +
    coord_fixed(xlim = c(0, 1), ylim = c(0, 1)))
} else {
  cat("No data for LM models correlation plot\n")
}
```

LLM Direct predictions:
```{r fig.width=5, fig.height=4}
if (nrow(df_corr_llm_direct) > 0) {
  # Calculate correlations for LLM Direct
  correlations_llm_direct <- df_corr_llm_direct |>
    group_by(model) |>
    summarize(
      boot_result = list(boot_cor(model_pred, Humans)),
      .groups = "drop"
    ) |>
    mutate(
      r = as.numeric(sapply(boot_result, function(x) x$r)),
      ci_lower = as.numeric(sapply(boot_result, function(x) x$ci_lower)),
      ci_upper = as.numeric(sapply(boot_result, function(x) x$ci_upper)),
      label = paste0("r = ", round(r, 2), " [", round(ci_lower, 2), ", ", round(ci_upper, 2), "]")
    ) |>
    select(-boot_result)

  print(ggplot(df_corr_llm_direct, aes(x = model_pred, y = Humans, color = intimacy, shape = action)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
    geom_errorbar(aes(ymin = Humans_ci_lower, ymax = Humans_ci_upper), width = 0.02) +
    geom_point(size = 3) +
    geom_text(data = correlations_llm_direct, aes(x = 0.05, y = 0.9, label = label),
              inherit.aes = FALSE, hjust = 0, size = 5) +
    scale_color_viridis_c(begin = 0, end = 0.85) +
    scale_shape_manual(values = c(16, 17, 15, 18)) +
    labs(x = "Model", y = "Humans", color = "Intimacy", shape = "Action") +
    facet_wrap(~model, nrow = 1) +
    coord_fixed(xlim = c(0, 1), ylim = c(0, 1)))
} else {
  cat("No data for LLM Direct correlation plot\n")
}
```

Plot motivation on separate panels

Stipulated parameter models:
```{r fig.width=12, fig.height=6}
if (nrow(df_corr_stipulated) > 0) {
  correlations_motiv_stipulated <- df_corr_stipulated |>
    group_by(model, motivation) |>
    summarize(
      boot_result = list(boot_cor(model_pred, Humans)),
      .groups = "drop"
    ) |>
    mutate(
      r = as.numeric(sapply(boot_result, function(x) x$r)),
      ci_lower = as.numeric(sapply(boot_result, function(x) x$ci_lower)),
      ci_upper = as.numeric(sapply(boot_result, function(x) x$ci_upper)),
      label = paste0("r = ", round(r, 2), " [", round(ci_lower, 2), ", ", round(ci_upper, 2), "]")
    ) |>
    select(-boot_result)

  print(ggplot(df_corr_stipulated, aes(x = model_pred, y = Humans, color = intimacy, shape = action)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
    geom_errorbar(aes(ymin = Humans_ci_lower, ymax = Humans_ci_upper), width = 0.02) +
    geom_point(size = 3) +
    geom_text(data = correlations_motiv_stipulated, aes(x = 0.05, y = 0.9, label = label),
              inherit.aes = FALSE, hjust = 0, size = 5) +
    scale_color_viridis_c(begin = 0, end = 0.85) +
    scale_shape_manual(values = c(16, 17, 15, 18)) +
    labs(x = "Model", y = "Humans", color = "Intimacy", shape = "Action",
         title = "Stipulated Parameters") +
    facet_grid(motivation~model, labeller = labeller(motivation = label_both)) +
    coord_fixed(xlim = c(0, 1), ylim = c(0, 1)))
}
```

LM-derived parameter models:
```{r fig.width=12, fig.height=6}
if (nrow(df_corr_lm) > 0) {
  correlations_motiv_lm <- df_corr_lm |>
    group_by(model, motivation) |>
    summarize(
      boot_result = list(boot_cor(model_pred, Humans)),
      .groups = "drop"
    ) |>
    mutate(
      r = as.numeric(sapply(boot_result, function(x) x$r)),
      ci_lower = as.numeric(sapply(boot_result, function(x) x$ci_lower)),
      ci_upper = as.numeric(sapply(boot_result, function(x) x$ci_upper)),
      label = paste0("r = ", round(r, 2), " [", round(ci_lower, 2), ", ", round(ci_upper, 2), "]")
    ) |>
    select(-boot_result)

  print(ggplot(df_corr_lm, aes(x = model_pred, y = Humans, color = intimacy, shape = action)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
    geom_errorbar(aes(ymin = Humans_ci_lower, ymax = Humans_ci_upper), width = 0.02) +
    geom_point(size = 3) +
    geom_text(data = correlations_motiv_lm, aes(x = 0.05, y = 0.9, label = label),
              inherit.aes = FALSE, hjust = 0, size = 5) +
    scale_color_viridis_c(begin = 0, end = 0.85) +
    scale_shape_manual(values = c(16, 17, 15, 18)) +
    labs(x = "Model", y = "Humans", color = "Intimacy", shape = "Action",
         title = "LM-Derived Parameters") +
    facet_grid(motivation~model, labeller = labeller(motivation = label_both)) +
    coord_fixed(xlim = c(0, 1), ylim = c(0, 1)))
}
```

### Model fit results

```{r}
fit_results <- read_csv(here("model", "forward_planning_fit_results.csv"))
fit_results |>
  select(model, nll, n_params, starts_with("param_")) |>
  knitr::kable(digits = 3, caption = "Model fitting results")
```

The full model provides the best fit (lowest NLL), followed by the vanilla model and then the discomfort-only model.

### AIC/BIC comparison

```{r}
n_obs <- nrow(df)  # number of observations

fit_results <- fit_results |>
  mutate(
    AIC = 2 * n_params + 2 * nll,
    BIC = n_params * log(n_obs) + 2 * nll
  )

fit_results |>
  select(model, nll, n_params, AIC, BIC) |>
  arrange(AIC) |>
  knitr::kable(digits = 2, caption = "Model comparison (lower AIC/BIC is better)")
```

```{r}
# Compute differences from best model (full)
full_aic <- fit_results$AIC[fit_results$model == "full"]
full_bic <- fit_results$BIC[fit_results$model == "full"]

fit_results |>
  mutate(
    delta_AIC = AIC - full_aic,
    delta_BIC = BIC - full_bic
  ) |>
  select(model, delta_AIC, delta_BIC) |>
  knitr::kable(digits = 2, caption = "Difference from best model (Full)")
```

**Interpretation**:

*Stipulated vs LM parameters*: We compare stipulated models (hand-crafted risk = [0,0,1,2], effort = [0,1,1,1]) against models using LLM-derived scenario-specific values for risk, effort, and reward.

```{r}
# Compare stipulated vs LM for each model type
comparison_df <- fit_results |>
  mutate(
    model_type = case_when(
      model %in% c("full", "full_lm") ~ "Social planning",
      model %in% c("vanilla", "vanilla_lm") ~ "Vanilla",
      model %in% c("discomfort_only", "discomfort_only_lm") ~ "Discomfort only"
    ),
    param_source = ifelse(grepl("_lm$", model), "LM", "Stipulated")
  ) |>
  select(model_type, param_source, AIC, BIC) |>
  pivot_wider(names_from = param_source, values_from = c(AIC, BIC)) |>
  mutate(
    delta_AIC = AIC_LM - AIC_Stipulated,
    delta_BIC = BIC_LM - BIC_Stipulated
  )

comparison_df |>
  select(model_type, delta_AIC, delta_BIC) |>
  knitr::kable(digits = 2, caption = "LM - Stipulated (negative = LM is better)",
               col.names = c("Model Type", "ΔAIC", "ΔBIC"))
```

*Model comparison*: The full (social planning) model with intimacy scaling has the lowest AIC/BIC among stipulated models, indicating that intimacy modulates both reward and discomfort. Among LM models, the pattern is similar—the full model outperforms vanilla and discomfort-only variants.

## LM Models: Scenario-by-Scenario Predictions

We can examine how the LM-based models capture variation across scenarios, since they use scenario-specific parameters derived from the LLM.

### LM model predictions by scenario

```{r}
df_preds_scenario <- df_preds |>
  filter(model %in% c("Humans", "Vanilla planning (LM)", "Discomfort only (LM)", "Social planning (LM)", "LLM only")) |>
  mutate(model = factor(model, levels = c("Vanilla planning (LM)", "Discomfort only (LM)", "Social planning (LM)", "LLM only", "Humans"))) |>
  group_by(scenario_label, intimacy, motivation, action, model) |>
  tidyboot_mean(p_action) |>
  ungroup()
```

```{r fig.width=18, fig.height=30}
ggplot(df_preds_scenario, aes(x = action, y = empirical_stat, color = intimacy, group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 2) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), width = 0.2
  ) +
  geom_line(position = position_dodge(width = 0.2)) +
  scale_color_viridis_c(begin = 0, end = 0.85) +
  labs(y = "P(action)", x = "Action", color = "Intimacy") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid(scenario_label ~ model + motivation,
             labeller = labeller(motivation = label_both))
```

### Model-human correlation by scenario (LM models)

```{r}
# Compare LM models' correlations per scenario
df_corr_scenario <- df_preds_scenario |>
  filter(model != "Humans") |>
  select(scenario_label, intimacy, motivation, action, model, model_pred = empirical_stat) |>
  left_join(
    df_preds_scenario |>
      filter(model == "Humans") |>
      select(scenario_label, intimacy, motivation, action,
             human_pred = empirical_stat, human_ci_lower = ci_lower, human_ci_upper = ci_upper),
    by = c("scenario_label", "intimacy", "motivation", "action")
  )

# Calculate correlations per scenario and model
scenario_correlations <- df_corr_scenario |>
  group_by(scenario_label, model) |>
  summarize(
    r = cor(model_pred, human_pred),
    .groups = "drop"
  ) |>
  mutate(label = paste0("r = ", round(r, 2)))

ggplot(df_corr_scenario, aes(x = model_pred, y = human_pred, color = intimacy)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
  geom_errorbar(aes(ymin = human_ci_lower, ymax = human_ci_upper), width = 0.02) +
  geom_point(size = 2, alpha = 0.7) +
  geom_text(data = scenario_correlations, aes(x = 0.1, y = 0.95, label = label),
            inherit.aes = FALSE, hjust = 0, size = 3) +
  scale_color_viridis_c(begin = 0, end = 0.85) +
  labs(x = "Model", y = "Humans", color = "Intimacy") +
  facet_grid(model ~ scenario_label) +
  coord_fixed(xlim = c(0, 1), ylim = c(0, 1)) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

### Summary: Correlation comparison across scenarios

```{r}
scenario_correlations |>
  pivot_wider(names_from = model, values_from = c(r, label)) |>
  select(scenario_label,
         `r_Vanilla planning (LM)`, `r_Discomfort only (LM)`,
         `r_Social planning (LM)`, `r_LLM only`) |>
  rename(
    Scenario = scenario_label,
    `Vanilla (LM)` = `r_Vanilla planning (LM)`,
    `Discomfort (LM)` = `r_Discomfort only (LM)`,
    `Social (LM)` = `r_Social planning (LM)`,
    `LLM Direct` = `r_LLM only`
  ) |>
  arrange(desc(`Social (LM)`)) |>
  knitr::kable(digits = 3, caption = "Correlation with human data by scenario (LM models)")
```

### Estimate stability: Bootstrap CIs on per-scenario correlations

With ~15 observations per cell, per-scenario correlations have substantial uncertainty. Here we bootstrap CIs for each scenario's model-human correlation and visualize them as a forest plot.

```{r}
# Bootstrap correlations per scenario and model
scenario_corr_boot <- df_corr_scenario |>
  group_by(scenario_label, model) |>
  summarize(
    boot_result = list(boot_cor(model_pred, human_pred, n_boot = 1000)),
    .groups = "drop"
  ) |>
  mutate(
    r = as.numeric(sapply(boot_result, function(x) x$r)),
    ci_lower = as.numeric(sapply(boot_result, function(x) x$ci_lower)),
    ci_upper = as.numeric(sapply(boot_result, function(x) x$ci_upper))
  ) |>
  select(-boot_result)

# Add overall (pooled) correlation for comparison
overall_corr <- df_corr_scenario |>
  group_by(model) |>
  summarize(
    boot_result = list(boot_cor(model_pred, human_pred, n_boot = 1000)),
    .groups = "drop"
  ) |>
  mutate(
    scenario_label = "OVERALL",
    r = as.numeric(sapply(boot_result, function(x) x$r)),
    ci_lower = as.numeric(sapply(boot_result, function(x) x$ci_lower)),
    ci_upper = as.numeric(sapply(boot_result, function(x) x$ci_upper))
  ) |>
  select(-boot_result)

scenario_corr_boot <- bind_rows(scenario_corr_boot, overall_corr) |>
  mutate(
    scenario_label = factor(scenario_label,
                            levels = c(setdiff(unique(scenario_label), "OVERALL"), "OVERALL")),
    is_overall = scenario_label == "OVERALL"
  )
```

```{r fig.width=14, fig.height=6}
# Forest plot
ggplot(scenario_corr_boot, aes(x = r, y = scenario_label, color = is_overall)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray70") +
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper), height = 0.3) +
  geom_point(aes(size = is_overall)) +
  scale_color_manual(values = c("FALSE" = "steelblue", "TRUE" = "firebrick"), guide = "none") +
  scale_size_manual(values = c("FALSE" = 2, "TRUE" = 4), guide = "none") +
  labs(x = "Correlation (r) with human data", y = "Scenario",
       title = "Per-scenario model-human correlations with 95% bootstrap CIs") +
  facet_wrap(~model, nrow = 1) +
  theme(axis.text.y = element_text(size = 8))
```

```{r}
# Summary statistics on stability
stability_summary <- scenario_corr_boot |>
  filter(!is_overall) |>
  group_by(model) |>
  summarize(
    mean_r = mean(r),
    sd_r = sd(r),
    min_r = min(r),
    max_r = max(r),
    mean_ci_width = mean(ci_upper - ci_lower),
    .groups = "drop"
  )

stability_summary |>
  knitr::kable(digits = 3,
               caption = "Stability of per-scenario correlations (LM models)",
               col.names = c("Model", "Mean r", "SD of r", "Min r", "Max r", "Mean CI width"))
```

**Interpretation**:
- **Mean CI width** indicates average uncertainty per scenario. With 15 obs/cell and ~32 points per scenario (4 intimacy × 2 motivation × 4 action), CIs are often wide.
- **SD of r** shows how much correlations vary across scenarios. If SD is large relative to mean CI width, there may be true scenario heterogeneity. If SD is similar to or smaller than what you'd expect from sampling noise alone, the variation is likely just noise.
- If most scenario CIs overlap substantially with the overall estimate (red), more data probably wouldn't change conclusions much. If some scenarios are clear outliers, that's worth investigating.