---
title: Experiment 1 (forward planning)
author: Alicia Chen
date: 2026-01-03
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    page-layout: full
    search: false
knitr:
  opts_chunk:
    dev: "ragg_png"
execute:
  warning: false 
---

```{r setup}
source(here::here("analysis", "utils.R"))
setup_analysis()
```

## Demographics

```{r}
df_exit <- read_csv(here("data", "forw_plan", "exit_survey.csv"), show_col_types = FALSE)

n_total <- nrow(df_exit)
n_passed <- df_exit |>
  filter(attention_passed == TRUE, memory_correct_count > 0) |>
  nrow()

cat("Total participants recruited:", n_total, "\n")
cat("Passed attention checks:", n_passed, "\n")
cat("Mean age:", round(mean(df_exit$age, na.rm = TRUE), 1),
    "SD age:", round(sd(df_exit$age, na.rm = TRUE), 1),
    "Min age:", min(df_exit$age, na.rm = TRUE),
    "Max age:", max(df_exit$age, na.rm = TRUE))
cat("\nGender:\n")
print(table(df_exit$gender))
```

## Results

This experiment manipulates how much two people want the food (`low` vs. `high` desire), and the relationship between the two people (0, 50, 75, or 100 on a 100 point scale between 'maximally intimate' and 'maximally formal'). 

Load csv, calculate means

```{r}
# main_trials_long.csv is generated by analysis/json_to_csv.py
# It excludes participants who failed attention check or got 0 on memory check
df <- read_csv(here("data", "forw_plan", "main_trials_long.csv")) |>
  mutate(
    intimacy = factor(intimacy),
    motivation = factor(motivation, levels = c("low", "high"), labels = MOTIVATION_LEVELS),
    model = "humans"
  )

df_summary <- df |>
  group_by(intimacy, motivation, action) |>
  tidyboot_mean(p_action)
```

## Model comparison

### Model description

We fit three ablations of the canonical access utility to the human data. In all models, a joint actor chooses an action proportional to its total utility, following a softmax choice rule with inverse temperature $\alpha$. Per-scenario values of $\text{access}(a)$ and $\text{effort}(a)$ are generated by an LLM (see `model/lm_scenario_params.py`); the reward $V(a|s)$ is stipulated as a binary goal-satisfaction gate ($V=1$ iff the action satisfies the active goal — sharing under high motivation, or not-sharing under low motivation).

**Full model**: The full canonical utility is:

$$U(a|s, I) = w_v \cdot V(a|s) - w_d \cdot \text{access}(a) \cdot (1 - I) - w_e \cdot \text{effort}(a)$$

where $V(a|s)$ is the intrinsic food reward, $\text{access}(a)$ captures graded bodily/spatial openness, and $\text{effort}(a)$ is the physical effort of executing the action. Intimacy $I$ scales the access-discomfort term: at high intimacy the $-w_d \cdot \text{access} \cdot (1 - I)$ penalty shrinks toward zero, so higher-access actions become relatively more attractive.

**Access only**: Drops the food-reward and effort terms, keeping only $-w_d \cdot \text{access}(a) \cdot (1 - I)$.

**Base (no access)**: Drops the access term: $U = w_v \cdot V(a|s) - w_e \cdot \text{effort}(a)$.

### Load model predictions

```{r}
df_preds <- read_csv(here("model", "outputs", "forward_planning_fits.csv")) |>
  select(subject_id, scenario_label, intimacy, motivation, action, p_action,
         pred_access_full, pred_access_only, pred_no_access) |>
  pivot_longer(
    cols = c(p_action, pred_access_full, pred_access_only, pred_no_access),
    names_to = "model",
    values_to = "p_action"
  ) |>
  mutate(
    intimacy = factor(intimacy),
    model = case_when(
      model == "p_action" ~ "Humans",
      model == "pred_access_full" ~ "Full model",
      model == "pred_access_only" ~ "Access only",
      model == "pred_no_access" ~ "Base"
    ),
    model = factor(model, levels = c(
      "Base", "Access only", "Full model", "Humans"
    )),
    motivation = factor(motivation, levels = c("low", "high"), labels = MOTIVATION_LEVELS)
  )

df_preds_summary <- df_preds |>
  group_by(intimacy, motivation, action, model) |>
  tidyboot_mean(p_action) |>
  ungroup()
```

### Model predictions vs human data

```{r}
#| fig-width: 12
#| fig-asp: 0.43

df_preds_summary |>
  filter(model %in% c("Base", "Access only", "Full model", "Humans")) |>
  mutate(model = factor(model, levels = c("Base", "Access only", "Full model", "Humans"))) |>
  ggplot(aes(x = action, y = empirical_stat, color = intimacy, shape = factor(action), group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 4, alpha = PLOT_ALPHA) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), size = 1.2, width = 0.5
  ) +
  geom_line(position = position_dodge(width = 0.2), linewidth = 1.2) +
  scale_color_intimacy() +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(y = "P(Action)", x = "Action", color = "Intimacy", shape = "Action") +
  facet_grid(motivation ~ model)

ggsave(here("figures", "exp1_comparison.pdf"),
         device = cairo_pdf)
```

### Model-human correlation

```{r}
# Bootstrap function for correlation CI (handles NAs)
boot_cor <- function(x, y, n_boot = 1000) {
  # Remove NAs
  complete <- complete.cases(x, y)
  x <- x[complete]
  y <- y[complete]
  n <- length(x)
  if (n < 3) return(list(r = NA_real_, ci_lower = NA_real_, ci_upper = NA_real_))
  boot_rs <- replicate(n_boot, {
    idx <- sample(n, replace = TRUE)
    cor(x[idx], y[idx])
  })
  list(
    r = cor(x, y),
    ci_lower = as.numeric(quantile(boot_rs, 0.025, na.rm = TRUE)),
    ci_upper = as.numeric(quantile(boot_rs, 0.975, na.rm = TRUE))
  )
}

# Get human data for joining
df_humans <- df_preds_summary |>
  filter(model == "Humans") |>
  select(intimacy, motivation, action,
         Humans = empirical_stat,
         Humans_ci_lower = ci_lower,
         Humans_ci_upper = ci_upper)

# Create correlation data by joining model predictions with human data
df_corr_all <- df_preds_summary |>
  filter(model != "Humans") |>
  select(intimacy, motivation, action, model, model_pred = empirical_stat) |>
  left_join(df_humans, by = c("intimacy", "motivation", "action")) |>
  filter(!is.na(model_pred), !is.na(Humans)) |>
  mutate(action = factor(action))

# Debug: check what models exist
cat("Models in df_corr_all:", unique(as.character(df_corr_all$model)), "\n")

# Use all models for correlation
df_corr <- df_corr_all |>
  mutate(model = droplevels(model))

cat("Rows in df_corr:", nrow(df_corr), "\n")
```

```{r}
#| label: fig-exp1-correlation
#| fig-width: 10
#| fig-asp: 0.4


# Calculate correlations
correlations <- df_corr |>
  group_by(model) |>
  summarize(boot_result = list(boot_cor(model_pred, Humans)), .groups = "drop") |>
  mutate(
    r = as.numeric(sapply(boot_result, function(x)
      x$r)),
    ci_lower = as.numeric(sapply(boot_result, function(x)
      x$ci_lower)),
    ci_upper = as.numeric(sapply(boot_result, function(x)
      x$ci_upper)),
    label = paste0(
      "r = ",
      round(r, 2),
      " (",
      round(ci_lower, 2),
      ", ",
      round(ci_upper, 2),
      ")"
    )
  ) |>
  select(-boot_result)

ggplot(df_corr,
                        aes(
                          x = model_pred,
                          y = Humans,
                          color = intimacy,
                          shape = action
                        )) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    color = "lightgray"
  ) +
  geom_errorbar(aes(ymin = Humans_ci_lower, ymax = Humans_ci_upper), width = 0.02) +
  geom_point(size = 3, alpha = PLOT_ALPHA) +
  geom_label(
    data = correlations,
    aes(x = -Inf, y = Inf, label = label),
    inherit.aes = FALSE,
    hjust = -0.08,
    vjust = 1.5,
    size = 4.5,
    fill = "white",
    alpha = 0.5,
    label.size = 0
  ) +
  scale_color_intimacy() +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(
    x = "Model",
    y = "Humans",
    color = "Intimacy",
    shape = "Action"
  ) +
  facet_wrap( ~ model, nrow = 1) +
  coord_fixed_symmetric(df_corr$model_pred, df_corr$Humans)

ggsave(here("figures", "exp1_correlation.pdf"),
         device = cairo_pdf)

# p_correlation
# 
# save_figure(p_correlation, "exp1_correlation.pdf")
```

Plot motivation on separate panels:

```{r fig.width=12, fig.height=6}
correlations_motiv <- df_corr |>
  group_by(model, motivation) |>
  summarize(
    boot_result = list(boot_cor(model_pred, Humans)),
    .groups = "drop"
  ) |>
  mutate(
    r = as.numeric(sapply(boot_result, function(x) x$r)),
    ci_lower = as.numeric(sapply(boot_result, function(x) x$ci_lower)),
    ci_upper = as.numeric(sapply(boot_result, function(x) x$ci_upper)),
    label = paste0("r = ", round(r, 2), " (", round(ci_lower, 2), ", ", round(ci_upper, 2), ")")
  ) |>
  select(-boot_result)

ggplot(df_corr, aes(x = model_pred, y = Humans, color = intimacy, shape = action)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
  geom_errorbar(aes(ymin = Humans_ci_lower, ymax = Humans_ci_upper), width = 0.02) +
  geom_point(size = 3, alpha = PLOT_ALPHA) +
  geom_label(
    data = correlations,
    aes(x = -Inf, y = Inf, label = label),
    inherit.aes = FALSE,
    hjust = -0.08,
    vjust = 1.5,
    size = 4.5,
    fill = "white",
    alpha = 0.5,
    label.size = 0
  ) +
  scale_color_intimacy() +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(x = "Model", y = "Humans", color = "Intimacy", shape = "Action") +
  facet_grid(motivation~model) +
  coord_fixed_symmetric(df_corr$model_pred, df_corr$Humans)
```


Plot each scenario separately


```{r}
#| label: fig-exp1-scenario-summary
#| fig-width: 23 
#| fig-asp: 0.18


df_summary_scenario <- df |>
  group_by(scenario_label, intimacy, motivation, action) |>
  tidyboot_mean(p_action)

ggplot(df_summary_scenario, aes(x = action, y = empirical_stat, color = intimacy, group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 3, alpha = PLOT_ALPHA) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), width = 0.4
  ) +
  geom_line(position = position_dodge(width = 0.2), linewidth = 1) +
  scale_color_intimacy() +
  labs(y = "P(action)", x = "Action", color = "Intimacy") +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid(motivation~scenario_label, labeller = labeller(scenario_label = label_wrap_gen(15)))
```

Model predictions per scenario

```{r}
#| label: fig-exp1-scenario-model-preds
#| fig-width: 23
#| fig-asp: 0.55

df_preds_summary_scenario <- df_preds |>
  filter(model != "Humans") |>
  group_by(scenario_label, intimacy, motivation, action, model) |>
  tidyboot_mean(p_action) |>
  ungroup()

ggplot(df_preds_summary_scenario,
       aes(x = action, y = empirical_stat, color = intimacy, group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 3, alpha = PLOT_ALPHA) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), width = 0.4
  ) +
  geom_line(position = position_dodge(width = 0.2), linewidth = 1) +
  scale_color_intimacy() +
  labs(y = "P(action)", x = "Action", color = "Intimacy") +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid(model + motivation ~ scenario_label,
             labeller = labeller(scenario_label = label_wrap_gen(15)))
```

### Model fit results

```{r}
fit_results <- read_csv(here("model", "outputs", "forward_planning_fit_results.csv"))
fit_results |>
  select(model, nll, n_params, starts_with("param_")) |>
  knitr::kable(digits = 3, caption = "Model fitting results")
```

The full access model provides the best fit (lowest NLL), followed by the access-only ablation and the base model.

### AIC/BIC comparison

```{r}
n_obs <- nrow(df)  # number of observations

fit_results <- fit_results |>
  mutate(
    AIC = 2 * n_params + 2 * nll,
    BIC = n_params * log(n_obs) + 2 * nll
  )

fit_results |>
  select(model, nll, n_params, AIC, BIC) |>
  arrange(AIC) |>
  knitr::kable(digits = 2, caption = "Model comparison (lower AIC/BIC is better)")
```

```{r}
# Compute differences from best model (access_full)
best_aic <- fit_results$AIC[fit_results$model == "access_full"]
best_bic <- fit_results$BIC[fit_results$model == "access_full"]

fit_results |>
  mutate(
    delta_AIC = AIC - best_aic,
    delta_BIC = BIC - best_bic
  ) |>
  select(model, delta_AIC, delta_BIC) |>
  knitr::kable(digits = 2, caption = "Difference from best model (Full)")
```
