---
title: Experiment 2b (inverse planning - reward inference) data analysis
author: Alicia Chen
date: 2026-01-15
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    page-layout: full
    search: false
execute:
  warning: false
---

```{r setup}
source(here::here("analysis", "utils.R"))
setup_analysis()
```

This experiment presents participants with scenarios where they observe an action taken by one person toward another. Participants rate their belief about the probability that the actors have high motivation (want the food) before (prior) and after (posterior) observing the action. The experiment manipulates:
- **Action**: Which action was taken (action 0-3, varying in saliva-transfer risk)
- **Intimacy**: The relationship closeness between the actors (0, 50, 75, or 100)

## Load human data

```{r}
# main_trials_long.csv is generated by analysis/json_to_csv.py
# It excludes participants who failed attention check or got 0 on memory check
df <- read_csv(here("data", "inv_plan_reward", "main_trials_long.csv")) |>
  mutate(
    action = str_replace(action_condition, "action_", ""),
    p_high_reward = response / 100,  # Convert to probability scale
    stage = factor(stage, levels = c("prior", "posterior")),
    model = "Humans"
  ) |>
  select(-action_condition, -response)

# Calculate belief updates
df <- calculate_belief_update(df, "p_high_reward")
```

## Human data: Prior vs posterior ratings

### Distribution of ratings

```{r}
ggplot(df, aes(x = p_high_reward, fill = action)) +
  geom_density(alpha = 0.5) +
  scale_fill_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  labs(y = "Density", x = "P(high motivation)", fill = "Action") +
  theme(legend.position = "bottom") +
  facet_grid(stage ~ intimacy, labeller = label_both)
```

### Summary statistics

```{r}
df_summary <- df |>
  group_by(intimacy, action, stage, model) |>
  tidyboot_mean(p_high_reward)
```

### Prior vs posterior by action and intimacy

```{r fig.width=14, fig.height=8}
ggplot() +
  geom_violin(
    data = df,
    aes(x = stage, y = p_high_reward, fill = action),
    position = position_dodge(width = 0.8),
    alpha = 0.5,
    width = 0.7
  ) +
  geom_point(
    data = df_summary,
    aes(x = stage, y = empirical_stat, color = action, group = action),
    position = position_dodge(width = 0.8),
    size = 3
  ) +
  geom_errorbar(
    data = df_summary,
    aes(x = stage, ymin = ci_lower, ymax = ci_upper, color = action, group = action),
    position = position_dodge(width = 0.8),
    width = 0.1,
    size = 1
  ) +
  geom_line(
    data = df_summary,
    aes(x = stage, y = empirical_stat, color = action, group = action),
    position = position_dodge(width = 0.8),
    size = 1
  ) +
  facet_grid(intimacy ~ action, labeller = label_both) +
  scale_fill_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  scale_color_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  labs(
    title = "P(high motivation) before and after observing action",
    y = "P(high motivation)",
    x = "Stage",
    fill = "Observed action",
    color = "Observed action"
  )
```

### Belief updates

```{r}
df_belief_update_summary <- df |>
  filter(stage == "posterior") |>
  group_by(intimacy, action, model) |>
  tidyboot_mean(belief_update)
```

```{r}
ggplot(df_belief_update_summary, aes(x = action, y = empirical_stat, group = intimacy, fill = intimacy)) +
  geom_col(position = position_dodge(width = 0.9), alpha = 0.9) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.9), size = 1, width = 0.1, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_viridis_c(begin = 0, end = 0.85) +
  labs(
    title = "Change in P(high motivation) after observing action",
    y = "Belief update",
    x = "Observed action",
    fill = "Intimacy"
  ) +
  theme(legend.position = "bottom")
```

## Model comparison

### Model description

We compare three models that perform inverse planning to infer reward/motivation from observed actions:

**Full model (Social inv plan)**: Inverts a forward model where actors consider reward (scaled by intimacy and motivation), discomfort (scaled by intimacy), and effort costs.

**Vanilla model**: Inverts a forward model with reward, discomfort, and effort, but without intimacy scaling these utilities.

**Discomfort-only model**: Inverts a forward model where actors only consider discomfort from saliva transfer (scaled by intimacy).

**Full modified model**: Like the full model, but with reward scaled by $(1 + \beta \cdot \text{intimacy})$. This captures observers' assumption about how much actors value intimacy when sharing: $\beta=1$ is the pre-registered model, $\beta=0$ means observers don't think intimacy affects reward.

All models assume a uniform prior over reward (P(high motivation) = 0.5) and compute posteriors via Bayesian inference. The observer inverse temperature ($\alpha_{\text{observer}}$) is fitted to human data. For the modified model, we also fit $\beta$ (reward-intimacy scaling strength).

### Fitted parameters

```{r}
df_fit_results <- read_csv(here("model", "inverse_planning_fit_results.csv")) |>
  filter(experiment == "reward") |>
  mutate(
    model_label = case_when(
      model == "full" ~ "Social inv plan",
      model == "vanilla" ~ "Vanilla inv plan",
      model == "discomfort_only" ~ "Discomfort only",
      model == "full_lm" ~ "Social inv plan (LM)",
      model == "vanilla_lm" ~ "Vanilla inv plan (LM)",
      model == "discomfort_only_lm" ~ "Discomfort only (LM)",
      model == "full_modified" ~ "Social inv plan (modified)",
      model == "full_lm_modified" ~ "Social inv plan (LM, modified)"
    )
  ) |>
  select(model_label, alpha_observer, beta, nll, n_params)

df_fit_results |>
  knitr::kable(
    digits = 3,
    caption = "Fitted observer parameters for reward inference",
    col.names = c("Model", "α_observer", "δ", "NLL", "n_params")
  )
```

Note: The δ parameter captures how much observers think actors' reward scales with intimacy. δ < 1 indicates observers believe actors value sharing less with closer friends than the pre-registered model assumes.

### Load model predictions

```{r}
df_model <- read_csv(here("model", "inv_plan_reward_preds_summary.csv")) |>
  rename(intimacy = intimacy_condition) |>
  mutate(
    action = as.character(action),
    # Model predictions are P(high reward) in percentage
    p_high_reward = p_high_reward / 100,
    # Belief update from uniform prior (50%)
    belief_update = p_high_reward - 0.5,
    model = case_when(
      model == "full" ~ "Social inv plan",
      model == "vanilla" ~ "Vanilla inv plan",
      model == "discomfort_only" ~ "Discomfort only",
      model == "full_modified" ~ "Social inv plan (modified)"
    ),
    param_source = case_when(
      param_source == "stipulated" ~ "Stipulated",
      param_source == "lm" ~ "LM"
    )
  )

# Filter to stipulated parameters only for main comparison
df_model_stipulated <- df_model |>
  filter(param_source == "Stipulated")
```

### Combine human and model data

```{r}
df_combined <- bind_rows(
  df_belief_update_summary |>
    mutate(belief_update = empirical_stat) |>
    select(intimacy, action, model, belief_update, ci_lower, ci_upper),
  df_model_stipulated |>
    group_by(intimacy, action, model) |>
    summarize(belief_update = mean(belief_update), .groups = "drop") |>
    mutate(ci_lower = NA_real_, ci_upper = NA_real_)
) |>
  mutate(
    model = factor(model, levels = c("Vanilla inv plan", "Discomfort only", "Social inv plan", "Social inv plan (modified)", "Humans"))
  )
```

### Belief updates: Model vs human comparison

```{r fig.width=18, fig.height=5}
ggplot(df_combined, aes(x = action, y = belief_update, group = interaction(model, intimacy), fill = intimacy)) +
  geom_col(position = position_dodge(width = 0.9), alpha = 0.9) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.9), size = 1, width = 0, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_viridis_c(begin = 0, end = 0.85) +
  labs(
    title = "Change in P(high motivation) after observing action",
    y = "Belief update",
    x = "Observed action",
    fill = "Intimacy"
  ) +
  theme(legend.position = "bottom") +
  facet_wrap(~model, nrow = 1)
```

### Model-human correlation

```{r}
# Prepare data for correlation
df_corr <- df_combined |>
  select(intimacy, action, model, belief_update) |>
  pivot_wider(names_from = model, values_from = belief_update)
```

```{r}
# Reshape for correlation plots
df_corr_long <- df_corr |>
  pivot_longer(
    cols = c(`Vanilla inv plan`, `Discomfort only`, `Social inv plan`, `Social inv plan (modified)`),
    names_to = "model",
    values_to = "model_pred"
  ) |>
  rename(human_pred = Humans) |>
  filter(!is.na(model_pred), !is.na(human_pred)) |>
  mutate(
    model = factor(model, levels = c("Vanilla inv plan", "Discomfort only", "Social inv plan", "Social inv plan (modified)")),
    action = factor(action)
  )

# Calculate correlations
correlations <- df_corr_long |>
  group_by(model) |>
  summarize(
    boot_result = list(boot_cor(model_pred, human_pred)),
    .groups = "drop"
  ) |>
  mutate(
    r = sapply(boot_result, function(x) x$r),
    ci_lower = sapply(boot_result, function(x) x$ci_lower),
    ci_upper = sapply(boot_result, function(x) x$ci_upper),
    label = paste0("r = ", round(r, 2), " [", round(ci_lower, 2), ", ", round(ci_upper, 2), "]")
  ) |>
  select(-boot_result)
```

```{r fig.width=16, fig.height=4}
ggplot(df_corr_long, aes(x = model_pred, y = human_pred, color = intimacy, shape = action)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
  geom_point(size = 4) +
  geom_text(
    data = correlations,
    aes(x = -0.3, y = 0.35, label = label),
    inherit.aes = FALSE, hjust = 0, size = 4
  ) +
  scale_color_viridis_c(begin = 0, end = 0.85) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(x = "Model prediction", y = "Human data", color = "Intimacy", shape = "Action") +
  facet_wrap(~model, nrow = 1) +
  coord_fixed(xlim = c(-0.35, 0.45), ylim = c(-0.35, 0.45))
```

### Correlation summary

```{r}
correlations |>
  select(model, r, ci_lower, ci_upper) |>
  knitr::kable(
    digits = 3,
    caption = "Model-human correlation (belief updates)",
    col.names = c("Model", "r", "CI lower", "CI upper")
  )
```

## LM-derived parameter models

We also compare models using LLM-derived scenario-specific parameters.

```{r}
df_model_lm <- df_model |>
  filter(param_source == "LM") |>
  mutate(
    model = case_when(
      model == "Social inv plan (modified)" ~ "Social inv plan (LM, modified)",
      TRUE ~ paste0(model, " (LM)")
    )
  )

df_combined_lm <- bind_rows(
  df_belief_update_summary |>
    mutate(belief_update = empirical_stat) |>
    select(intimacy, action, model, belief_update, ci_lower, ci_upper),
  df_model_lm |>
    group_by(intimacy, action, model) |>
    summarize(belief_update = mean(belief_update), .groups = "drop") |>
    mutate(ci_lower = NA_real_, ci_upper = NA_real_)
) |>
  mutate(
    model = factor(model, levels = c(
      "Vanilla inv plan (LM)", "Discomfort only (LM)", "Social inv plan (LM)", "Social inv plan (LM, modified)", "Humans"
    ))
  )
```

```{r fig.width=18, fig.height=5}
ggplot(df_combined_lm, aes(x = action, y = belief_update, group = interaction(model, intimacy), fill = intimacy)) +
  geom_col(position = position_dodge(width = 0.9), alpha = 0.9) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.9), size = 1, width = 0, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_viridis_c(begin = 0, end = 0.85) +
  labs(
    title = "Change in P(high motivation) after observing action (LM parameters)",
    y = "Belief update",
    x = "Observed action",
    fill = "Intimacy"
  ) +
  theme(legend.position = "bottom") +
  facet_wrap(~model, nrow = 1)
```

### LM model correlations

```{r}
df_corr_lm <- df_combined_lm |>
  select(intimacy, action, model, belief_update) |>
  pivot_wider(names_from = model, values_from = belief_update)

df_corr_lm_long <- df_corr_lm |>
  pivot_longer(
    cols = c(`Vanilla inv plan (LM)`, `Discomfort only (LM)`, `Social inv plan (LM)`, `Social inv plan (LM, modified)`),
    names_to = "model",
    values_to = "model_pred"
  ) |>
  rename(human_pred = Humans) |>
  filter(!is.na(model_pred), !is.na(human_pred)) |>
  mutate(
    model = factor(model, levels = c("Vanilla inv plan (LM)", "Discomfort only (LM)", "Social inv plan (LM)", "Social inv plan (LM, modified)")),
    action = factor(action)
  )

correlations_lm <- df_corr_lm_long |>
  group_by(model) |>
  summarize(
    boot_result = list(boot_cor(model_pred, human_pred)),
    .groups = "drop"
  ) |>
  mutate(
    r = sapply(boot_result, function(x) x$r),
    ci_lower = sapply(boot_result, function(x) x$ci_lower),
    ci_upper = sapply(boot_result, function(x) x$ci_upper),
    label = paste0("r = ", round(r, 2), " [", round(ci_lower, 2), ", ", round(ci_upper, 2), "]")
  ) |>
  select(-boot_result)
```

```{r fig.width=16, fig.height=4}
ggplot(df_corr_lm_long, aes(x = model_pred, y = human_pred, color = intimacy, shape = action)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
  geom_point(size = 4) +
  geom_text(
    data = correlations_lm,
    aes(x = -0.3, y = 0.35, label = label),
    inherit.aes = FALSE, hjust = 0, size = 4
  ) +
  scale_color_viridis_c(begin = 0, end = 0.85) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(x = "Model prediction", y = "Human data", color = "Intimacy", shape = "Action") +
  facet_wrap(~model, nrow = 1) +
  coord_fixed(xlim = c(-0.35, 0.45), ylim = c(-0.35, 0.45))
```

### Summary: All model correlations

```{r}
bind_rows(
  correlations |> mutate(param_source = "Stipulated"),
  correlations_lm |> mutate(param_source = "LM")
) |>
  select(param_source, model, r, ci_lower, ci_upper) |>
  arrange(param_source, model) |>
  knitr::kable(
    digits = 3,
    caption = "Model-human correlations",
    col.names = c("Parameters", "Model", "r", "CI lower", "CI upper")
  )
```

## Scenario-level analysis

### Human belief updates by scenario

```{r}
df_belief_update_scenario <- df |>
  filter(stage == "posterior") |>
  group_by(scenario_label, intimacy, action, model) |>
  tidyboot_mean(belief_update)
```


