---
title: Experiment 2a (inverse planning - intimacy inference) data analysis
author: Alicia Chen
date: 2026-01-15
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    page-layout: full
    search: false
execute:
  warning: false
---

```{r setup}
library(here)
library(tidyverse)
library(tidyboot)

theme_set(theme_classic(base_size = 18))
set.seed(67)
```

This experiment presents participants with scenarios where they observe an action taken by one person toward another. Participants rate their belief about the intimacy of the relationship before (prior) and after (posterior) observing the action. The experiment manipulates:

- **Action**: Which action was taken (action 0-3, varying in saliva-transfer risk)
- **Motivation**: Whether both people want the food (high vs low)

## Load human data

```{r}
# main_trials_long.csv is generated by analysis/json_to_csv.py
# It excludes participants who failed attention check or got 0 on memory check
df <- read_csv(here("data", "inv_plan_intimacy", "main_trials_long.csv")) |>
  mutate(
    action = str_replace(action_condition, "action_", ""),
    motivation = factor(motivation, levels = c("low", "high")),
    stage = factor(stage, levels = c("prior", "posterior")),
    model = "Humans"
  ) |>
  select(-action_condition)

# Calculate belief updates
df <- df |>
  group_by(subject_id, scenario_label) |>
  mutate(
    belief_update = ifelse(
      stage == "posterior",
      intimacy_rating[stage == "posterior"] - intimacy_rating[stage == "prior"],
      NA
    )
  ) |>
  ungroup()
```

## Human data: Prior vs posterior ratings

### Distribution of ratings

```{r}
ggplot(df, aes(x = intimacy_rating, fill = action)) +
  geom_density(alpha = 0.5) +
  scale_fill_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  labs(y = "Density", x = "Intimacy rating", fill = "Action") +
  theme(legend.position = "bottom") +
  facet_grid(
    stage ~ motivation,
    labeller = labeller(motivation = c("low" = "motivation: low", "high" = "motivation: high"))
  )
```

### Summary statistics

```{r}
df_summary <- df |>
  group_by(motivation, action, stage, model) |>
  tidyboot_mean(intimacy_rating)
```

### Prior vs posterior by action and motivation

```{r fig.width=12, fig.height=6}
ggplot() +
  geom_violin(
    data = df,
    aes(x = stage, y = intimacy_rating, fill = action),
    position = position_dodge(width = 0.8),
    alpha = 0.5,
    width = 0.7
  ) +
  geom_point(
    data = df_summary,
    aes(x = stage, y = empirical_stat, color = action, group = action),
    position = position_dodge(width = 0.8),
    size = 3
  ) +
  geom_errorbar(
    data = df_summary,
    aes(x = stage, ymin = ci_lower, ymax = ci_upper, color = action, group = action),
    position = position_dodge(width = 0.8),
    width = 0.1,
    size = 1
  ) +
  geom_line(
    data = df_summary,
    aes(x = stage, y = empirical_stat, color = action, group = action),
    position = position_dodge(width = 0.8),
    size = 1
  ) +
  facet_grid(motivation ~ action, labeller = label_both) +
  scale_fill_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  scale_color_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  labs(
    title = "Intimacy ratings before and after observing action",
    y = "Intimacy rating",
    x = "Stage",
    fill = "Observed action",
    color = "Observed action"
  )
```

### Belief updates

```{r}
df_belief_update_summary <- df |>
  filter(stage == "posterior") |>
  group_by(motivation, action, model) |>
  tidyboot_mean(belief_update)
```

```{r}
ggplot(df_belief_update_summary, aes(x = action, y = empirical_stat, group = motivation, fill = motivation)) +
  geom_col(position = position_dodge(width = 0.9), alpha = 0.9) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.9), size = 1, width = 0.1, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Change in intimacy rating after observing action",
    y = "Belief update",
    x = "Observed action",
    fill = "Motivation"
  ) +
  theme(legend.position = "bottom")
```

## Model comparison

### Model description

We compare three models that perform inverse planning to infer intimacy from observed actions:

**Full model (Social planning)**: Inverts a forward model where actors consider reward (scaled by intimacy and motivation), discomfort (scaled by intimacy), and effort costs.

**Vanilla model**: Inverts a forward model with reward, discomfort, and effort, but without intimacy scaling these utilities.

**Discomfort-only model**: Inverts a forward model where actors only consider discomfort from saliva transfer (scaled by intimacy).

All models assume a uniform prior over intimacy (0-100 scale) and compute posteriors via Bayesian inference.

### Load model predictions

```{r}
df_model <- read_csv(here("model", "inv_plan_intimacy_preds_summary.csv")) |>
  rename(motivation = reward_condition) |>
  mutate(
    action = as.character(action),
    motivation = factor(motivation, levels = c("low", "high")),
    # Belief update from uniform prior (mean = 50)
    belief_update = expected_intimacy - 50,
    model = case_when(
      model == "full" ~ "Social planning",
      model == "vanilla" ~ "Vanilla planning",
      model == "discomfort_only" ~ "Discomfort only"
    ),
    param_source = case_when(
      param_source == "stipulated" ~ "Stipulated",
      param_source == "lm" ~ "LM"
    )
  )

# Filter to stipulated parameters only for main comparison
df_model_stipulated <- df_model |>
  filter(param_source == "Stipulated")
```

### Combine human and model data

```{r}
df_combined <- bind_rows(
  df_belief_update_summary |>
    mutate(belief_update = empirical_stat) |>
    select(motivation, action, model, belief_update, ci_lower, ci_upper),
  df_model_stipulated |>
    group_by(motivation, action, model) |>
    summarize(belief_update = mean(belief_update), .groups = "drop") |>
    mutate(ci_lower = NA_real_, ci_upper = NA_real_)
) |>
  mutate(
    model = factor(model, levels = c("Vanilla planning", "Discomfort only", "Social planning", "Humans"))
  )
```

### Belief updates: Model vs human comparison

```{r fig.width=14, fig.height=5}
ggplot(df_combined, aes(x = action, y = belief_update, group = interaction(model, motivation), fill = motivation)) +
  geom_col(position = position_dodge(width = 0.9), alpha = 0.9) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.9), size = 1, width = 0.1, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Change in intimacy rating after observing action",
    y = "Belief update",
    x = "Observed action",
    fill = "Motivation"
  ) +
  theme(legend.position = "bottom") +
  facet_wrap(~model, nrow = 1)
```

### Model-human correlation

```{r}
# Prepare data for correlation
df_corr <- df_combined |>
  select(motivation, action, model, belief_update) |>
  pivot_wider(names_from = model, values_from = belief_update)

# Bootstrap function for correlation CI
boot_cor <- function(x, y, n_boot = 1000) {
  complete <- complete.cases(x, y)
  x <- x[complete]
  y <- y[complete]
  n <- length(x)
  if (n < 3) return(list(r = NA_real_, ci_lower = NA_real_, ci_upper = NA_real_))
  boot_rs <- replicate(n_boot, {
    idx <- sample(n, replace = TRUE)
    cor(x[idx], y[idx])
  })
  list(
    r = cor(x, y),
    ci_lower = as.numeric(quantile(boot_rs, 0.025, na.rm = TRUE)),
    ci_upper = as.numeric(quantile(boot_rs, 0.975, na.rm = TRUE))
  )
}
```

```{r}
# Reshape for correlation plots
df_corr_long <- df_corr |>
  pivot_longer(
    cols = c(`Vanilla planning`, `Discomfort only`, `Social planning`),
    names_to = "model",
    values_to = "model_pred"
  ) |>
  rename(human_pred = Humans) |>
  filter(!is.na(model_pred), !is.na(human_pred)) |>
  mutate(
    model = factor(model, levels = c("Vanilla planning", "Discomfort only", "Social planning")),
    action = factor(action)
  )

# Calculate correlations
correlations <- df_corr_long |>
  group_by(model) |>
  summarize(
    boot_result = list(boot_cor(model_pred, human_pred)),
    .groups = "drop"
  ) |>
  mutate(
    r = sapply(boot_result, function(x) x$r),
    ci_lower = sapply(boot_result, function(x) x$ci_lower),
    ci_upper = sapply(boot_result, function(x) x$ci_upper),
    label = paste0("r = ", round(r, 2), " [", round(ci_lower, 2), ", ", round(ci_upper, 2), "]")
  ) |>
  select(-boot_result)
```

```{r fig.width=12, fig.height=4}
ggplot(df_corr_long, aes(x = model_pred, y = human_pred, color = motivation, shape = action)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
  geom_point(size = 4) +
  geom_text(
    data = correlations,
    aes(x = -15, y = 25, label = label),
    inherit.aes = FALSE, hjust = 0, size = 5
  ) +
  scale_color_brewer(palette = "Set2") +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(x = "Model prediction", y = "Human data", color = "Motivation", shape = "Action") +
  facet_wrap(~model, nrow = 1) +
  coord_fixed(xlim = c(-20, 30), ylim = c(-20, 30))
```

### Correlation summary

```{r}
correlations |>
  select(model, r, ci_lower, ci_upper) |>
  knitr::kable(
    digits = 3,
    caption = "Model-human correlation (belief updates)",
    col.names = c("Model", "r", "CI lower", "CI upper")
  )
```

## LM-derived parameter models

We also compare models using LLM-derived scenario-specific parameters.

```{r}
df_model_lm <- df_model |>
  filter(param_source == "LM") |>
  mutate(
    model = paste0(model, " (LM)")
  )

df_combined_lm <- bind_rows(
  df_belief_update_summary |>
    mutate(belief_update = empirical_stat) |>
    select(motivation, action, model, belief_update, ci_lower, ci_upper),
  df_model_lm |>
    group_by(motivation, action, model) |>
    summarize(belief_update = mean(belief_update), .groups = "drop") |>
    mutate(ci_lower = NA_real_, ci_upper = NA_real_)
) |>
  mutate(
    model = factor(model, levels = c(
      "Vanilla planning (LM)", "Discomfort only (LM)", "Social planning (LM)", "Humans"
    ))
  )
```

```{r fig.width=14, fig.height=5}
ggplot(df_combined_lm, aes(x = action, y = belief_update, group = interaction(model, motivation), fill = motivation)) +
  geom_col(position = position_dodge(width = 0.9), alpha = 0.9) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.9), size = 1, width = 0.1, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Change in intimacy rating after observing action (LM parameters)",
    y = "Belief update",
    x = "Observed action",
    fill = "Motivation"
  ) +
  theme(legend.position = "bottom") +
  facet_wrap(~model, nrow = 1)
```

### LM model correlations

```{r}
df_corr_lm <- df_combined_lm |>
  select(motivation, action, model, belief_update) |>
  pivot_wider(names_from = model, values_from = belief_update)

df_corr_lm_long <- df_corr_lm |>
  pivot_longer(
    cols = c(`Vanilla planning (LM)`, `Discomfort only (LM)`, `Social planning (LM)`),
    names_to = "model",
    values_to = "model_pred"
  ) |>
  rename(human_pred = Humans) |>
  filter(!is.na(model_pred), !is.na(human_pred)) |>
  mutate(
    model = factor(model, levels = c("Vanilla planning (LM)", "Discomfort only (LM)", "Social planning (LM)")),
    action = factor(action)
  )

correlations_lm <- df_corr_lm_long |>
  group_by(model) |>
  summarize(
    boot_result = list(boot_cor(model_pred, human_pred)),
    .groups = "drop"
  ) |>
  mutate(
    r = sapply(boot_result, function(x) x$r),
    ci_lower = sapply(boot_result, function(x) x$ci_lower),
    ci_upper = sapply(boot_result, function(x) x$ci_upper),
    label = paste0("r = ", round(r, 2), " [", round(ci_lower, 2), ", ", round(ci_upper, 2), "]")
  ) |>
  select(-boot_result)
```

```{r fig.width=12, fig.height=4}
ggplot(df_corr_lm_long, aes(x = model_pred, y = human_pred, color = motivation, shape = action)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
  geom_point(size = 4) +
  geom_text(
    data = correlations_lm,
    aes(x = -15, y = 25, label = label),
    inherit.aes = FALSE, hjust = 0, size = 5
  ) +
  scale_color_brewer(palette = "Set2") +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(x = "Model prediction", y = "Human data", color = "Motivation", shape = "Action") +
  facet_wrap(~model, nrow = 1) +
  coord_fixed(xlim = c(-20, 30), ylim = c(-20, 30))
```

### Summary: All model correlations

```{r}
bind_rows(
  correlations |> mutate(param_source = "Stipulated"),
  correlations_lm |> mutate(param_source = "LM")
) |>
  select(param_source, model, r, ci_lower, ci_upper) |>
  arrange(param_source, model) |>
  knitr::kable(
    digits = 3,
    caption = "Model-human correlations",
    col.names = c("Parameters", "Model", "r", "CI lower", "CI upper")
  )
```

## Scenario-level analysis

### Human belief updates by scenario

```{r}
df_belief_update_scenario <- df |>
  filter(stage == "posterior") |>
  group_by(scenario_label, motivation, action, model) |>
  tidyboot_mean(belief_update)
```

```{r fig.width=14, fig.height=10}
ggplot(df_belief_update_scenario, aes(x = action, y = empirical_stat, fill = motivation)) +
  geom_col(position = position_dodge(width = 0.9), alpha = 0.9) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.9), width = 0.1, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Belief updates by scenario",
    y = "Belief update",
    x = "Observed action",
    fill = "Motivation"
  ) +
  theme(legend.position = "bottom") +
  facet_wrap(~scenario_label, ncol = 4)
```

