---
title: Experiment 1 (forward planning) data analysis
author: Alicia Chen
date: 2026-01-03
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    page-layout: full
    search: false
execute:
  warning: false 
---

```{r setup}
library(here)
library(tidyverse)
library(tidyboot)
library(lme4)

theme_set(theme_classic(base_size = 18))
set.seed(67)
```

This experiment manipulates how much two people want the food (`low` vs. `high` reward), and the relationship between the two people (0, 50, 75, or 100 on a 100 point scale between 'intimate' and 'formal'). 

```{r}
# main_trials_long.csv is generated by analysis/json_to_csv.py
# It excludes participants who failed attention check or got 0 on memory check
df <- read_csv(here("data", "forw_plan", "main_trials_long.csv")) |>
  mutate(
    motivation = factor(motivation, levels = c("low", "high")),
    model = "humans"
  )
```

Calculate means

```{r}
df_summary <- df |>
  group_by(intimacy, motivation, action) |>
  tidyboot_mean(p_action)
```

Plot

```{r}
ggplot(df_summary, aes(x = action, y = empirical_stat, color = intimacy, group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), width = 0.2
  ) +
  geom_line(position = position_dodge(width = 0.2), size = 1) +
  scale_color_viridis_c(begin = 0, end = 0.85) +
  labs(y = "P(action)", x = "Action", color = "Intimacy") +
  theme(legend.position = "bottom") +
  facet_wrap(~motivation, labeller = label_both)
```

Plot each scenario separately
```{r}
df_summary_scenario <- df |>
  group_by(scenario_label, intimacy, motivation, action) |>
  tidyboot_mean(p_action)
```

```{r}
ggplot(df_summary_scenario, aes(x = action, y = empirical_stat, color = intimacy, group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), width = 0.2
  ) +
  geom_line(position = position_dodge(width = 0.2), size = 1) +
  scale_color_viridis_c(begin = 0, end = 0.85) +
  labs(y = "P(action)", x = "Action", color = "Intimacy") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid(motivation~scenario_label, labeller = labeller(motivation = label_both, scenario_label = label_wrap_gen(15)))
```



## Model comparison

### Model description

We fit three models to the human data. In all models, a joint actor chooses an action proportional to its total utility, following a softmax choice rule with inverse temperature $\alpha$.

**Full model**: The total utility of an action $a$ given motivational state $s$ and intimacy $I$ is:

$$U_{tot}(a|s, I) = w_r \cdot r(a|s, I) - w_d \cdot d(a|I) - w_c \cdot c(a)$$

where:

- $r(a|s, I)$ is the reward, scaled by the motivational state $s$ and the intimacy $I$. The more the actors want to eat the food, and the more intimate the relationship, the higher the reward of eating the food together (for Action 0, the reward is 0).
- $d(a|I) = (1 - I) \cdot \rho(a)$ is the discomfort from saliva transfer, where $\rho(a) = [0, 0, 1, 2]$ is the saliva-transfer risk for actions 0-3. Higher intimacy mitigates discomfort.
- $c(a) = [0, 1, 1, 1]$ is the sharing cost, capturing the baseline cost of coordination.

**Vanilla model**: Includes all utility terms but does NOT consider how intimacy scales either reward or cost. Reward and discomfort are not modulated by relationship closeness.

**Discomfort-only model**: Only considers how intimacy mitigates the discomfort of taking a risky action ($U = -w_d \cdot d(a|I)$), without a representation of reward or sharing cost.

**Social planning (LLM)**: Same structure as the full model, but uses LLM-derived scenario-specific values for risk $\rho(a)$, effort $c(a)$, and base reward $r_0$ instead of stipulated constants. The LLM (Llama-3.3-70B-Instruct) was prompted to rate each scenario's actions on scales of saliva-transfer risk, coordination effort, and sharing reward.

**LLM Direct**: Direct LLM predictions without any cognitive model structure. The LLM (Llama-3.3-70B-Instruct) was presented with the exact experiment instructions and stimuli, and asked to predict the probability distribution over actions for each scenario and condition. This tests whether an LLM can directly simulate human behavior without explicit utility-based reasoning.

### Load model predictions

```{r}
# Load LLM direct predictions and join with model fits
llm_direct <- read_csv(here("model", "llm_direct_predictions.csv")) |>
  select(scenario_label, intimacy, motivation, action, pred_llm = pred_llm_mean)

df_preds <- read_csv(here("model", "forward_planning_fits.csv")) |>
  left_join(llm_direct, by = c("scenario_label", "intimacy", "motivation", "action")) |>
  select(subject_id, scenario_label, intimacy, motivation, action, p_action,
         pred_full, pred_vanilla, pred_discomfort_only, pred_full_lm, pred_llm) |>
  pivot_longer(
    cols = c(p_action, pred_full, pred_vanilla, pred_discomfort_only, pred_full_lm, pred_llm),
    names_to = "model",
    values_to = "p_action"
  ) |>
  mutate(
    model = case_when(
      model == "p_action" ~ "Humans",
      model == "pred_full" ~ "Social planning",
      model == "pred_vanilla" ~ "Vanilla planning",
      model == "pred_discomfort_only" ~ "Discomfort only",
      model == "pred_full_lm" ~ "LLM hybrid",
      model == "pred_llm" ~ "LLM only"
    ),
    model = factor(model, levels = c("Vanilla planning", "Discomfort only", "Social planning", "LLM hybrid", "LLM only", "Humans")),
    motivation = factor(motivation, levels = c("low", "high"))
  )

df_preds_summary <- df_preds |>
  group_by(intimacy, motivation, action, model) |>
  tidyboot_mean(p_action)
```

### Model predictions vs human data

Cognitive models:
```{r fig.width=12, fig.height=5.2}
df_preds_summary |>
  filter(model %in% c("Vanilla planning", "Discomfort only", "Social planning", "Humans")) |>
  ggplot(aes(x = action, y = empirical_stat, color = intimacy, shape = factor(action), group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), width = 0.4
  ) +
  geom_line(position = position_dodge(width = 0.2)) +
  scale_color_viridis_c(begin = 0, end = 0.85) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(y = "P(action)", x = "Action", color = "Intimacy", shape = "Action") +
  facet_grid(motivation ~ model, labeller = labeller(motivation = label_both))
```

LLM-based models:
```{r fig.width=9, fig.height=5.2}
df_preds_summary |>
  filter(model %in% c("LLM hybrid", "LLM only", "Humans")) |>
  ggplot(aes(x = action, y = empirical_stat, color = intimacy, shape = factor(action), group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), width = 0.4
  ) +
  geom_line(position = position_dodge(width = 0.2)) +
  scale_color_viridis_c(begin = 0, end = 0.85) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(y = "P(action)", x = "Action", color = "Intimacy", shape = "Action") +
  facet_grid(motivation ~ model, labeller = labeller(motivation = label_both))
```

### Model-human correlation

```{r}
# Prepare data for correlation plots
df_corr <- df_preds_summary |>
  select(intimacy, motivation, action, model, empirical_stat, ci_lower, ci_upper) |>
  pivot_wider(
    names_from = model,
    values_from = c(empirical_stat, ci_lower, ci_upper)
  )

# Bootstrap function for correlation CI
boot_cor <- function(x, y, n_boot = 1000) {
  n <- length(x)
  boot_rs <- replicate(n_boot, {
    idx <- sample(n, replace = TRUE)
    cor(x[idx], y[idx])
  })
  c(
    r = cor(x, y),
    ci_lower = quantile(boot_rs, 0.025),
    ci_upper = quantile(boot_rs, 0.975)
  )
}

# Reshape for cognitive models
df_corr_cognitive <- df_corr |>
  pivot_longer(
    cols = c(`empirical_stat_Vanilla planning`, `empirical_stat_Discomfort only`, `empirical_stat_Social planning`),
    names_to = "model",
    names_prefix = "empirical_stat_",
    values_to = "model_pred"
  ) |>
  rename(
    Humans = `empirical_stat_Humans`,
    Humans_ci_lower = `ci_lower_Humans`,
    Humans_ci_upper = `ci_upper_Humans`
  ) |>
  mutate(
    model = factor(model, levels = c("Vanilla planning", "Discomfort only", "Social planning")),
    action = factor(action)
  )

# Reshape for LLM models
df_corr_llm <- df_corr |>
  pivot_longer(
    cols = c(`empirical_stat_LLM hybrid`, `empirical_stat_LLM only`),
    names_to = "model",
    names_prefix = "empirical_stat_",
    values_to = "model_pred"
  ) |>
  rename(
    Humans = `empirical_stat_Humans`,
    Humans_ci_lower = `ci_lower_Humans`,
    Humans_ci_upper = `ci_upper_Humans`
  ) |>
  mutate(
    model = factor(model, levels = c("LLM hybrid", "LLM only")),
    action = factor(action)
  )
```

Cognitive models:
```{r fig.width=12, fig.height=4}
# Calculate correlations for cognitive models
correlations_cognitive <- df_corr_cognitive |>
  group_by(model) |>
  summarize(
    boot_result = list(boot_cor(model_pred, Humans)),
    .groups = "drop"
  ) |>
  mutate(
    r = sapply(boot_result, `[`, "r"),
    ci_lower = sapply(boot_result, `[`, "ci_lower.2.5%"),
    ci_upper = sapply(boot_result, `[`, "ci_upper.97.5%"),
    label = paste0("r = ", round(r, 2), " [", round(ci_lower, 2), ", ", round(ci_upper, 2), "]")
  ) |>
  select(-boot_result)

ggplot(df_corr_cognitive, aes(x = model_pred, y = Humans, color = intimacy, shape = action)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
  geom_errorbar(aes(ymin = Humans_ci_lower, ymax = Humans_ci_upper), width = 0.02) +
  geom_point(size = 3) +
  geom_text(data = correlations_cognitive, aes(x = 0.05, y = 0.9, label = label),
            inherit.aes = FALSE, hjust = 0, size = 5) +
  scale_color_viridis_c(begin = 0, end = 0.85) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(x = "Model", y = "Humans", color = "Intimacy", shape = "Action") +
  facet_wrap(~model, nrow = 1) +
  coord_fixed(xlim = c(0, 1), ylim = c(0, 1))
```

LLM-based models:
```{r fig.width=8, fig.height=4}
# Calculate correlations for LLM models
correlations_llm <- df_corr_llm |>
  group_by(model) |>
  summarize(
    boot_result = list(boot_cor(model_pred, Humans)),
    .groups = "drop"
  ) |>
  mutate(
    r = sapply(boot_result, `[`, "r"),
    ci_lower = sapply(boot_result, `[`, "ci_lower.2.5%"),
    ci_upper = sapply(boot_result, `[`, "ci_upper.97.5%"),
    label = paste0("r = ", round(r, 2), " [", round(ci_lower, 2), ", ", round(ci_upper, 2), "]")
  ) |>
  select(-boot_result)

ggplot(df_corr_llm, aes(x = model_pred, y = Humans, color = intimacy, shape = action)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
  geom_errorbar(aes(ymin = Humans_ci_lower, ymax = Humans_ci_upper), width = 0.02) +
  geom_point(size = 3) +
  geom_text(data = correlations_llm, aes(x = 0.05, y = 0.9, label = label),
            inherit.aes = FALSE, hjust = 0, size = 5) +
  scale_color_viridis_c(begin = 0, end = 0.85) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(x = "Model", y = "Humans", color = "Intimacy", shape = "Action") +
  facet_wrap(~model, nrow = 1) +
  coord_fixed(xlim = c(0, 1), ylim = c(0, 1))
```

Plot motivation on separate panels

Cognitive models:
```{r fig.width=12, fig.height=6}
correlations_motiv_cognitive <- df_corr_cognitive |>
  group_by(model, motivation) |>
  summarize(
    boot_result = list(boot_cor(model_pred, Humans)),
    .groups = "drop"
  ) |>
  mutate(
    r = sapply(boot_result, `[`, "r"),
    ci_lower = sapply(boot_result, `[`, "ci_lower.2.5%"),
    ci_upper = sapply(boot_result, `[`, "ci_upper.97.5%"),
    label = paste0("r = ", round(r, 2), " [", round(ci_lower, 2), ", ", round(ci_upper, 2), "]")
  ) |>
  select(-boot_result)

ggplot(df_corr_cognitive, aes(x = model_pred, y = Humans, color = intimacy, shape = action)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
  geom_errorbar(aes(ymin = Humans_ci_lower, ymax = Humans_ci_upper), width = 0.02) +
  geom_point(size = 3) +
  geom_text(data = correlations_motiv_cognitive, aes(x = 0.05, y = 0.9, label = label),
            inherit.aes = FALSE, hjust = 0, size = 5) +
  scale_color_viridis_c(begin = 0, end = 0.85) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(x = "Model", y = "Humans", color = "Intimacy", shape = "Action") +
  facet_grid(motivation~model, labeller = labeller(motivation = label_both)) +
  coord_fixed(xlim = c(0, 1), ylim = c(0, 1))
```

LLM-based models:
```{r fig.width=8, fig.height=6}
correlations_motiv_llm <- df_corr_llm |>
  group_by(model, motivation) |>
  summarize(
    boot_result = list(boot_cor(model_pred, Humans)),
    .groups = "drop"
  ) |>
  mutate(
    r = sapply(boot_result, `[`, "r"),
    ci_lower = sapply(boot_result, `[`, "ci_lower.2.5%"),
    ci_upper = sapply(boot_result, `[`, "ci_upper.97.5%"),
    label = paste0("r = ", round(r, 2), " [", round(ci_lower, 2), ", ", round(ci_upper, 2), "]")
  ) |>
  select(-boot_result)

ggplot(df_corr_llm, aes(x = model_pred, y = Humans, color = intimacy, shape = action)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
  geom_errorbar(aes(ymin = Humans_ci_lower, ymax = Humans_ci_upper), width = 0.02) +
  geom_point(size = 3) +
  geom_text(data = correlations_motiv_llm, aes(x = 0.05, y = 0.9, label = label),
            inherit.aes = FALSE, hjust = 0, size = 5) +
  scale_color_viridis_c(begin = 0, end = 0.85) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(x = "Model", y = "Humans", color = "Intimacy", shape = "Action") +
  facet_grid(motivation~model, labeller = labeller(motivation = label_both)) +
  coord_fixed(xlim = c(0, 1), ylim = c(0, 1))
```

### Model fit results

```{r}
fit_results <- read_csv(here("model", "forward_planning_fit_results.csv"))
fit_results |>
  select(model, nll, n_params, starts_with("param_")) |>
  knitr::kable(digits = 3, caption = "Model fitting results")
```

The full model provides the best fit (lowest NLL), followed by the vanilla model and then the discomfort-only model.

### AIC/BIC comparison

```{r}
n_obs <- nrow(df)  # number of observations

fit_results <- fit_results |>
  mutate(
    AIC = 2 * n_params + 2 * nll,
    BIC = n_params * log(n_obs) + 2 * nll
  )

fit_results |>
  select(model, nll, n_params, AIC, BIC) |>
  arrange(AIC) |>
  knitr::kable(digits = 2, caption = "Model comparison (lower AIC/BIC is better)")
```

```{r}
# Compute differences from best model (full)
full_aic <- fit_results$AIC[fit_results$model == "full"]
full_bic <- fit_results$BIC[fit_results$model == "full"]

fit_results |>
  mutate(
    delta_AIC = AIC - full_aic,
    delta_BIC = BIC - full_bic
  ) |>
  select(model, delta_AIC, delta_BIC) |>
  knitr::kable(digits = 2, caption = "Difference from best model (Full)")
```

**Interpretation**: The full model (with stipulated parameters) has the lowest AIC and BIC, indicating the best trade-off between fit and complexity. For Full vs Vanilla (same number of parameters), the ΔAIC of `r round(fit_results$AIC[fit_results$model == "vanilla"] - full_aic, 2)` favors the full model, indicating that intimacy scaling of reward and discomfort improves model fit.

Comparing the stipulated Full model vs the LLM-derived Full model (both with 4 parameters), the ΔAIC of `r round(fit_results$AIC[fit_results$model == "full_lm"] - full_aic, 2)` favors the stipulated model. This suggests that the hand-crafted action parameters (risk = [0,0,1,2], effort = [0,1,1,1]) capture the relevant structure better than the LLM-derived scenario-specific values.

## LLM Models: Scenario-by-Scenario Predictions

We can examine how the LLM-based models (Social planning with LLM parameters and LLM Direct) capture variation across scenarios.

### LLM predictions by scenario

```{r}
df_preds_scenario <- df_preds |>
  filter(model %in% c("Humans", "LLM hybrid", "LLM only")) |>
  group_by(scenario_label, intimacy, motivation, action, model) |>
  tidyboot_mean(p_action)
```

```{r fig.width=14, fig.height=30}
ggplot(df_preds_scenario, aes(x = action, y = empirical_stat, color = intimacy, group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 2) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), width = 0.2
  ) +
  geom_line(position = position_dodge(width = 0.2)) +
  scale_color_viridis_c(begin = 0, end = 0.85) +
  labs(y = "P(action)", x = "Action", color = "Intimacy") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid(scenario_label ~ model + motivation,
             labeller = labeller(motivation = label_both))
```

### Model-human correlation by scenario (LLM models)

```{r}

# Compare LLM models' correlations per scenario
df_corr_scenario <- df_preds_scenario |>
  filter(model != "Humans") |>
  select(scenario_label, intimacy, motivation, action, model, model_pred = empirical_stat) |>
  left_join(
    df_preds_scenario |>
      filter(model == "Humans") |>
      select(scenario_label, intimacy, motivation, action,
             human_pred = empirical_stat, human_ci_lower = ci_lower, human_ci_upper = ci_upper),
    by = c("scenario_label", "intimacy", "motivation", "action")
  )

# Calculate correlations per scenario and model
scenario_correlations <- df_corr_scenario |>
  group_by(scenario_label, model) |>
  summarize(
    r = cor(model_pred, human_pred),
    .groups = "drop"
  ) |>
  mutate(label = paste0("r = ", round(r, 2)))

ggplot(df_corr_scenario, aes(x = model_pred, y = human_pred, color = intimacy)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
  geom_errorbar(aes(ymin = human_ci_lower, ymax = human_ci_upper), width = 0.02) +
  geom_point(size = 2, alpha = 0.7) +
  geom_text(data = scenario_correlations, aes(x = 0.1, y = 0.95, label = label),
            inherit.aes = FALSE, hjust = 0, size = 3) +
  scale_color_viridis_c(begin = 0, end = 0.85) +
  labs(x = "Model", y = "Humans", color = "Intimacy") +
  facet_grid(model ~ scenario_label) +
  coord_fixed(xlim = c(0, 1), ylim = c(0, 1)) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

### Summary: Correlation comparison across scenarios

```{r}
scenario_correlations |>
  pivot_wider(names_from = model, values_from = c(r, label)) |>
  select(scenario_label, `r_LLM hybrid`, `r_LLM only`) |>
  rename(
    Scenario = scenario_label,
    `LLM params r` = `r_LLM hybrid`,
    `LLM Direct r` = `r_LLM only`
  ) |>
  arrange(desc(`LLM Direct r`)) |>
  knitr::kable(digits = 3, caption = "Correlation with human data by scenario (LLM models)")
```

### Estimate stability: Bootstrap CIs on per-scenario correlations

With ~15 observations per cell, per-scenario correlations have substantial uncertainty. Here we bootstrap CIs for each scenario's model-human correlation and visualize them as a forest plot.

```{r}
# Bootstrap correlations per scenario and model
scenario_corr_boot <- df_corr_scenario |>
  group_by(scenario_label, model) |>
  summarize(
    boot_result = list(boot_cor(model_pred, human_pred, n_boot = 1000)),
    .groups = "drop"
  ) |>
  mutate(
    r = sapply(boot_result, `[`, "r"),
    ci_lower = sapply(boot_result, `[`, "ci_lower.2.5%"),
    ci_upper = sapply(boot_result, `[`, "ci_upper.97.5%")
  ) |>
  select(-boot_result)

# Add overall (pooled) correlation for comparison
overall_corr <- df_corr_scenario |>
  group_by(model) |>
  summarize(
    boot_result = list(boot_cor(model_pred, human_pred, n_boot = 1000)),
    .groups = "drop"
  ) |>
  mutate(
    scenario_label = "OVERALL",
    r = sapply(boot_result, `[`, "r"),
    ci_lower = sapply(boot_result, `[`, "ci_lower.2.5%"),
    ci_upper = sapply(boot_result, `[`, "ci_upper.97.5%")
  ) |>
  select(-boot_result)

scenario_corr_boot <- bind_rows(scenario_corr_boot, overall_corr) |>
  mutate(
    scenario_label = factor(scenario_label,
                            levels = c(setdiff(unique(scenario_label), "OVERALL"), "OVERALL")),
    is_overall = scenario_label == "OVERALL"
  )
```

```{r fig.width=10, fig.height=6}
# Forest plot
ggplot(scenario_corr_boot, aes(x = r, y = scenario_label, color = is_overall)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray70") +
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper), height = 0.3) +
  geom_point(aes(size = is_overall)) +
  scale_color_manual(values = c("FALSE" = "steelblue", "TRUE" = "firebrick"), guide = "none") +
  scale_size_manual(values = c("FALSE" = 2, "TRUE" = 4), guide = "none") +
  labs(x = "Correlation (r) with human data", y = "Scenario",
       title = "Per-scenario model-human correlations with 95% bootstrap CIs") +
  facet_wrap(~model) +
  theme(axis.text.y = element_text(size = 9))
```

```{r}
# Summary statistics on stability
stability_summary <- scenario_corr_boot |>
  filter(!is_overall) |>
  group_by(model) |>
  summarize(
    mean_r = mean(r),
    sd_r = sd(r),
    min_r = min(r),
    max_r = max(r),
    mean_ci_width = mean(ci_upper - ci_lower),
    .groups = "drop"
  )

stability_summary |>
  knitr::kable(digits = 3,
               caption = "Stability of per-scenario correlations",
               col.names = c("Model", "Mean r", "SD of r", "Min r", "Max r", "Mean CI width"))
```

**Interpretation**:
- **Mean CI width** indicates average uncertainty per scenario. With 15 obs/cell and ~32 points per scenario (4 intimacy × 2 motivation × 4 action), CIs are often wide.
- **SD of r** shows how much correlations vary across scenarios. If SD is large relative to mean CI width, there may be true scenario heterogeneity. If SD is similar to or smaller than what you'd expect from sampling noise alone, the variation is likely just noise.
- If most scenario CIs overlap substantially with the overall estimate (red), more data probably wouldn't change conclusions much. If some scenarios are clear outliers, that's worth investigating.