---
title: Experiment 1 (forward planning) data analysis
author: Alicia Chen
date: 2026-01-03
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    page-layout: full
    search: false
execute:
  warning: false 
---

```{r setup}
library(here)
library(tidyverse)
library(tidyboot)
library(lme4)

theme_set(theme_classic(base_size = 18))
```

This experiment manipulates how much two people want the food (`low` vs. `high` reward), and the relationship between the two people (0, 50, 75, or 100 on a 100 point scale between 'intimate' and 'formal'). 

```{r}
df <- read_csv(here("data", "forw_plan", "main_trials.csv")) |>
  rename(
    intimacy = intimacy_condition,
    motivation = reward_condition
  ) |>
  mutate(
    motivation = factor(motivation, levels = c("low", "high")),
    model = "humans"
  ) |>
  pivot_longer(
    cols = starts_with("action_"),
    names_to = "action",
    names_prefix = "action_",
    values_to = "p_action"
  )

write_csv(df, here("data", "forw_plan", "main_trials_long.csv"))
```

Calculate means

```{r}
df_summary <- df |>
  group_by(intimacy, motivation, action) |>
  tidyboot_mean(p_action)
```

Plot

```{r}
ggplot(df_summary, aes(x = action, y = empirical_stat, color = intimacy, group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), size = 1, width = 0.3
  ) +
  geom_line(position = position_dodge(width = 0.2), size = 1) +
  scale_color_viridis_c(begin = 0, end = 0.85) +
  labs(y = "P(action)", x = "Action", color = "Intimacy") +
  theme(legend.position = "bottom") +
  facet_wrap(~motivation, labeller = label_both)
```

## Model comparison

### Model description

We fit three models to the human data. In all models, a joint actor chooses an action proportional to its total utility, following a softmax choice rule with inverse temperature $\alpha$.

**Full model**: The total utility of an action $a$ given motivational state $s$ and intimacy $I$ is:

$$U_{tot}(a|s, I) = w_r \cdot r(a|s, I) - w_d \cdot d(a|I) - w_c \cdot c(a)$$

where:

- $r(a|s, I)$ is the reward, scaled by the motivational state $s$ and the intimacy $I$. The more the actors want to eat the food, and the more intimate the relationship, the higher the reward of eating the food together (for Action 0, the reward is 0).
- $d(a|I) = (1 - I) \cdot \rho(a)$ is the discomfort from saliva transfer, where $\rho(a) = [0, 0, 1, 2]$ is the saliva-transfer risk for actions 0-3. Higher intimacy mitigates discomfort.
- $c(a) = [0, 1, 1, 1]$ is the sharing cost, capturing the baseline cost of coordination.

**Vanilla model**: Includes all utility terms but does NOT consider how intimacy scales either reward or cost. Reward and discomfort are not modulated by relationship closeness.

**Discomfort-only model**: Only considers how intimacy mitigates the discomfort of taking a risky action ($U = -w_d \cdot d(a|I)$), without a representation of reward or sharing cost.

### Load model predictions

```{r}
df_preds <- read_csv(here("model", "forward_planning_fits.csv")) |>
  select(subject_id, scenario_label, intimacy, motivation, action, p_action,
         pred_full, pred_vanilla, pred_discomfort_only) |>
  pivot_longer(
    cols = c(p_action, pred_full, pred_vanilla, pred_discomfort_only),
    names_to = "model",
    values_to = "p_action"
  ) |>
  mutate(
    model = case_when(
      model == "p_action" ~ "Human",
      model == "pred_full" ~ "Full model",
      model == "pred_vanilla" ~ "Vanilla model",
      model == "pred_discomfort_only" ~ "Discomfort only"
    ),
    model = factor(model, levels = c("Vanilla model", "Discomfort only", "Full model", "Human")),
    motivation = factor(motivation, levels = c("low", "high"))
  )

df_preds_summary <- df_preds |>
  group_by(intimacy, motivation, action, model) |>
  tidyboot_mean(p_action)
```

### Model predictions vs human data

```{r fig.width=10, fig.height=6}
ggplot(df_preds_summary, aes(x = action, y = empirical_stat, color = intimacy, shape = factor(action), group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), width = 0.2
  ) +
  geom_line(position = position_dodge(width = 0.2)) +
  scale_color_viridis_c(begin = 0, end = 0.85) +
  labs(y = "P(action)", x = "Action", color = "Intimacy", shape = "Action") +
  facet_grid(motivation ~ model, labeller = labeller(motivation = label_both))
```

### Model-human correlation

```{r fig.width=10, fig.height=4}
# Prepare data for correlation plots
df_corr <- df_preds_summary |>
  select(intimacy, motivation, action, model, empirical_stat, ci_lower, ci_upper) |>
  pivot_wider(
    names_from = model,
    values_from = c(empirical_stat, ci_lower, ci_upper)
  )

# Reshape for faceted plotting
df_corr_long <- df_corr |>
  pivot_longer(
    cols = c(`empirical_stat_Vanilla model`, `empirical_stat_Discomfort only`, `empirical_stat_Full model`),
    names_to = "model",
    names_prefix = "empirical_stat_",
    values_to = "model_pred"
  ) |>
  rename(
    Human = `empirical_stat_Human`,
    Human_ci_lower = `ci_lower_Human`,
    Human_ci_upper = `ci_upper_Human`
  ) |>
  mutate(
    model = factor(model, levels = c("Vanilla model", "Discomfort only", "Full model")),
    action = factor(action)
  )

# Calculate correlations for annotations
correlations <- df_corr_long |>
  group_by(model) |>
  summarize(
    r = cor(model_pred, Human),
    label = paste0("r = ", round(r, 2))
  )

ggplot(df_corr_long, aes(x = model_pred, y = Human, color = intimacy, shape = action)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  geom_errorbar(aes(ymin = Human_ci_lower, ymax = Human_ci_upper), width = 0.02) +
  geom_point(size = 3) +
  geom_text(data = correlations, aes(x = 0.1, y = 0.9, label = label),
            inherit.aes = FALSE, hjust = 0, size = 5) +
  scale_color_viridis_c(begin = 0, end = 0.85) +
  labs(x = "Model prediction", y = "Human data", color = "Intimacy", shape = "Action") +
  facet_wrap(~model, nrow = 1) +
  coord_fixed(xlim = c(0, 1), ylim = c(0, 1))
```

### Model fit results

```{r}
fit_results <- read_csv(here("model", "forward_planning_fit_results.csv"))
fit_results |>
  select(model, nll, n_params, starts_with("param_")) |>
  knitr::kable(digits = 3, caption = "Model fitting results")
```

The full model provides the best fit (lowest NLL), followed by the vanilla model and then the discomfort-only model.

### AIC/BIC comparison

```{r}
n_obs <- nrow(df)  # number of observations

fit_results <- fit_results |>
  mutate(
    AIC = 2 * n_params + 2 * nll,
    BIC = n_params * log(n_obs) + 2 * nll
  )

fit_results |>
  select(model, nll, n_params, AIC, BIC) |>
  arrange(AIC) |>
  knitr::kable(digits = 2, caption = "Model comparison (lower AIC/BIC is better)")
```

```{r}
# Compute differences from best model (full)
full_aic <- fit_results$AIC[fit_results$model == "full"]
full_bic <- fit_results$BIC[fit_results$model == "full"]

fit_results |>
  mutate(
    delta_AIC = AIC - full_aic,
    delta_BIC = BIC - full_bic
  ) |>
  select(model, delta_AIC, delta_BIC) |>
  knitr::kable(digits = 2, caption = "Difference from best model (Full)")
```

**Interpretation**: The full model has the lowest AIC and BIC, indicating the best trade-off between fit and complexity. For Full vs Vanilla (same number of parameters), the ΔAIC of `r round(fit_results$AIC[fit_results$model == "vanilla"] - full_aic, 2)` favors the full model, indicating that intimacy scaling of reward and discomfort improves model fit.