---
title: Inverse planning, effort manipulation (data analysis)
author: Alicia Chen
date: 2026-04-24
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    page-layout: full
    search: false
knitr:
  opts_chunk:
    dev: "ragg_png"
execute:
  warning: false
---

```{r setup}
source(here::here("analysis", "utils.R"))
setup_analysis()

# EFFORT_LEVELS and EFFORT_COLORS come from utils.R
ACTION_LEVELS <- c("No saliva", "Share saliva")
```

## Demographics

```{r}
report_demographics("food_inv_intimacy_effort_alt")
```

## Results

This experiment presents participants with scenarios where they observe one of two actions taken by one person toward another. Participants rate their belief about the intimacy of the relationship before (prior) and after (posterior) observing the action. Reward is held fixed at high across all trials. The experiment manipulates:

- **Observed action**: action_1 (non-saliva-sharing, requires the effort-constrained resource) or action_2 (saliva-sharing).
- **Effort**: relative effort of the non-saliva-sharing action — low or high.

## Load human data

```{r}
# Subjects excluded from the human-side analyses. daeeb393 was a careful
# subject (attention pass, 3/3 memory correct) but ran a strong "effort =
# intentionality" interpretation that's a >3 SD outlier on the action_2
# belief-update effort delta and obscures the rest of the sample. Excluded
# from plots only; model fits / CV predictions are unaffected.
EXCLUDED_SUBJECTS <- c("daeeb393-ddef-5529-804a-36dbdd257d5c")

df <- read_csv(here("data", "legacy", "food_inv_intimacy_effort_alt", "main_trials_long.csv"),
               show_col_types = FALSE) |>
  filter(!subject_id %in% EXCLUDED_SUBJECTS) |>
  mutate(
    action = factor(action_condition, levels = c("action_1", "action_2"),
                    labels = ACTION_LEVELS),
    effort = factor(effort, levels = c("low", "high"), labels = EFFORT_LEVELS),
    stage = factor(stage, levels = c("prior", "posterior")),
    model = "Humans"
  ) |>
  select(-action_condition)

df <- calculate_belief_update(df, "intimacy_rating")
```

## Human data: Prior vs posterior ratings

### Distribution of ratings

```{r}
#| fig-width: 10
#| fig-asp: 0.5
ggplot(df, aes(x = intimacy_rating, fill = action)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("No saliva" = "#5B8DBA",
                                "Share saliva" = "#D17A6E")) +
  labs(y = "Density", x = "Intimacy rating", fill = "Observed action") +
  theme(legend.position = "right") +
  facet_grid(stage ~ effort, labeller = labeller(effort = label_both))
```

```{r}
df_summary <- df |>
  group_by(effort, action, stage, model) |>
  tidyboot_mean(intimacy_rating)
```

### Prior vs posterior by action and effort

```{r}
#| fig-width: 12
#| fig-asp: 0.4

ggplot() +
  geom_violin(
    data = df,
    aes(x = stage, y = intimacy_rating, fill = action),
    position = position_dodge(width = 0.8),
    alpha = 0.5,
    width = 0.7
  ) +
  geom_line(
    data = df,
    aes(x = stage, y = intimacy_rating, color = action,
        group = interaction(subject_id, scenario_label, action)),
    position = position_dodge(width = 0.8),
    alpha = 0.12,
    linewidth = 0.3
  ) +
  geom_point(
    data = df_summary,
    aes(x = stage, y = empirical_stat, color = action, group = action),
    position = position_dodge(width = 0.8),
    size = 3,
    alpha = PLOT_ALPHA
  ) +
  geom_errorbar(
    data = df_summary,
    aes(x = stage, ymin = ci_lower, ymax = ci_upper, color = action, group = action),
    position = position_dodge(width = 0.8),
    width = 0.1,
    size = 1
  ) +
  geom_line(
    data = df_summary,
    aes(x = stage, y = empirical_stat, color = action, group = action),
    position = position_dodge(width = 0.8),
    size = 1
  ) +
  facet_grid(effort ~ action, labeller = labeller(action = label_both)) +
  scale_fill_manual(values = c("No saliva" = "#5B8DBA",
                                "Share saliva" = "#D17A6E")) +
  scale_color_manual(values = c("No saliva" = "#5B8DBA",
                                 "Share saliva" = "#D17A6E")) +
  labs(
    title = "Intimacy ratings before and after observing action",
    y = "Intimacy rating",
    x = "Stage",
    fill = "Observed action",
    color = "Observed action"
  )
```

### Belief updates

```{r}
df_belief_update_summary <- df |>
  filter(stage == "posterior") |>
  group_by(effort, action, model) |>
  tidyboot_mean(belief_update)
```

## Model comparison

### Model description

We invert the canonical access utility under three ablations to infer intimacy from the observed action. The actor weights are frozen from the forward-planning effort fit (`forward_planning_effort_fit_results.csv`), and only $\alpha_{\text{observer}}$ is fit to the inverse-planning effort data.

**Full model**: inverts $U(a|s, I, e) = w_v \cdot V(a) - w_d \cdot \text{access}(a) \cdot (1 - I) - w_e \cdot \text{effort}(a)$.

**Discomfort-only**: $-w_d \cdot \text{access}(a) \cdot (1 - I)$.

**Base model**: $w_v \cdot V(a) - w_e \cdot \text{effort}(a)$ (no intimacy dependence — observer cannot meaningfully update from the action, so $\alpha_{\text{observer}}$ collapses).

Per-(scenario, effort_condition) values of $\text{access}$ and $\text{effort}$ are LM-generated. The observer assumes a uniform prior over intimacy on [0, 100] and computes a Bayesian posterior; $\alpha_{\text{observer}}$ tempers that posterior before sampling.

### Fitted parameters

```{r}
df_fit_results <- read_csv(here("model", "outputs", "food_inv_intimacy_effort_alt", "fit_results.csv"),
                           show_col_types = FALSE) |>
  filter(experiment == "food_inv_intimacy_effort_alt") |>
  mutate(
    model_label = case_when(
      model == "full" ~ "Full model",
      model == "discomfort_only" ~ "Discomfort-only",
      model == "base" ~ "Base model"
    )
  ) |>
  filter(!is.na(model_label)) |>
  select(model_label, alpha_observer, nll, n_params)

df_fit_results |>
  knitr::kable(
    digits = 3,
    caption = "Fitted observer parameters (intimacy inference, effort experiment)",
    col.names = c("Model", "α_observer", "NLL", "n_params")
  )
```

### AIC comparison

```{r}
kable_aic_table(df_fit_results)
```

### Load model predictions

Out-of-sample LOSO predictions from `food_inv_intimacy_effort_alt/cv_preds_summary.csv`: each held-out scenario's posterior was generated by an observer fit on the other 15.

```{r}
df_model <- read_csv(here("model", "outputs", "food_inv_intimacy_effort_alt", "cv_preds_summary.csv"),
                     show_col_types = FALSE) |>
  mutate(
    action = factor(action, levels = c(1, 2), labels = ACTION_LEVELS),
    effort = factor(effort_condition, levels = c("low", "high"),
                    labels = EFFORT_LEVELS),
    belief_update = (expected_intimacy - 50) / 100,
    model = case_when(
      model == "full" ~ "Full model",
      model == "discomfort_only" ~ "Discomfort-only",
      model == "base" ~ "Base model"
    )
  ) |>
  filter(!is.na(model))

MODEL_LEVELS <- c("Base model", "Discomfort-only", "Full model", "Humans")
```

### Combine human and model data

```{r}
df_combined <- bind_rows(
  df_belief_update_summary |>
    rescale_belief_update() |>
    select(effort, action, model, belief_update, ci_lower, ci_upper),
  df_model |>
    group_by(effort, action, model) |>
    summarize(belief_update = mean(belief_update), .groups = "drop") |>
    mutate(ci_lower = NA_real_, ci_upper = NA_real_)
) |>
  mutate(
    model = factor(model, levels = MODEL_LEVELS)
  )
```

### Belief updates: model vs human comparison

```{r}
#| fig-width: 12
#| fig-asp: 0.32

df_belief_update_subj <- df |>
  filter(stage == "posterior") |>
  group_by(subject_id, effort, action) |>
  summarise(belief_update = mean(belief_update) / 100, .groups = "drop") |>
  mutate(
    model = factor("Humans", levels = MODEL_LEVELS),
    x_pos = as.numeric(action) + ifelse(effort == "Low effort", -0.225, 0.225)
  )

# Per-subject jitter applied uniformly across the two effort conditions, so
# lines stay vertical (each subject's pair is jittered together).
set.seed(67)
subj_jitter <- df_belief_update_subj |>
  distinct(subject_id) |>
  mutate(jitter = runif(n(), -0.05, 0.05))

df_belief_update_subj <- df_belief_update_subj |>
  left_join(subj_jitter, by = "subject_id") |>
  mutate(x_pos = x_pos + jitter)

df_combined |>
  ggplot(aes(x = action, y = belief_update,
             group = interaction(model, effort), fill = effort)) +
  geom_crossbar(
    aes(y = 0, ymin = -0.01, ymax = 0.01),
    position = position_dodge(width = 0.9),
    width = 0.9,
    color = NA
  ) +
  geom_col(position = position_dodge(width = 0.9), alpha = PLOT_ALPHA) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.9), size = 1, width = 0.1,
    color = "black", alpha = 0.8
  ) +
  geom_line(
    data = df_belief_update_subj,
    aes(x = x_pos, y = belief_update,
        group = interaction(subject_id, action)),
    inherit.aes = FALSE,
    color = "black", alpha = 0.15, linewidth = 0.3
  ) +
  geom_point(
    data = df_belief_update_subj,
    aes(x = x_pos, y = belief_update),
    inherit.aes = FALSE,
    color = "black", alpha = 0.2, size = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_manual(values = EFFORT_COLORS) +
  labs(y = "Belief update", x = "Observed action", fill = "Effort") +
  theme(legend.position = "right") +
  facet_wrap(~ model, nrow = 1)

ggsave(here("figures", "food_inv_intimacy_effort_alt_belief_main.pdf"),
       device = cairo_pdf, width = 12, height = 4)
```

### Belief updates vs LM-derived access (continuous x)

Same four panels as above, but with the x-axis replaced by the LM-derived access score for the observed action, in the relevant effort condition (from `lm_scenario_params_effort.csv`). One point per (scenario, action, effort), with a linear smoother per effort level.

```{r}
#| fig-width: 14
#| fig-asp: 0.32

df_canonical_access <- read_csv(here("model", "outputs", "lm",
                                     "lm_scenario_params_effort.csv"),
                                show_col_types = FALSE) |>
  mutate(
    action = factor(action, levels = c(1, 2), labels = ACTION_LEVELS),
    effort = factor(effort_condition, levels = c("low", "high"),
                    labels = EFFORT_LEVELS)
  ) |>
  select(scenario_label, action, effort, access)

df_human_scenario_pts <- df |>
  filter(stage == "posterior") |>
  group_by(scenario_label, effort, action) |>
  tidyboot_mean(belief_update) |>
  ungroup() |>
  rescale_belief_update() |>
  mutate(model = "Humans") |>
  select(scenario_label, effort, action, model, belief_update, ci_lower, ci_upper)

df_model_scenario_pts <- df_model |>
  select(scenario_label, effort, action, model, belief_update) |>
  mutate(ci_lower = NA_real_, ci_upper = NA_real_)

df_access_combined <- bind_rows(df_human_scenario_pts, df_model_scenario_pts) |>
  inner_join(df_canonical_access,
             by = c("scenario_label", "action", "effort")) |>
  mutate(model = factor(model, levels = MODEL_LEVELS))

p_belief_access <- df_access_combined |>
  ggplot(aes(x = access, y = belief_update,
             color = effort, shape = action, group = effort)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  geom_errorbar(
    data = df_access_combined |> filter(model == "Humans"),
    aes(ymin = ci_lower, ymax = ci_upper),
    position = POS_JITTER_DODGE,
    width = 0.05, alpha = 0.4, linewidth = 0.4
  ) +
  geom_point(position = POS_JITTER_DODGE, alpha = 0.5, size = 2) +
  geom_smooth(se = FALSE, linewidth = 1) +
  scale_color_manual(values = EFFORT_COLORS) +
  scale_shape_manual(values = c(16, 17)) +
  labs(y = "Belief update",
       x = "LM-derived access of observed action",
       color = "Effort", shape = "Action") +
  theme(legend.position = "right") +
  facet_wrap(~ model, nrow = 1)

print(p_belief_access)

ggsave(here("figures", "food_inv_intimacy_effort_alt_belief_access.pdf"),
       plot = p_belief_access, device = cairo_pdf, width = 14, height = 4)
```

### Model–human correlation

```{r}
#| fig-width: 10
#| fig-asp: 0.38

df_humans_cells <- df_belief_update_summary |>
  ungroup() |>
  mutate(
    Humans = empirical_stat / 100,
    Humans_ci_lower = ci_lower / 100,
    Humans_ci_upper = ci_upper / 100
  ) |>
  select(effort, action, Humans, Humans_ci_lower, Humans_ci_upper)

df_model_cells <- df_model |>
  group_by(effort, action, model) |>
  summarize(model_pred = mean(belief_update), .groups = "drop") |>
  mutate(model = factor(model, levels = setdiff(MODEL_LEVELS, "Humans")))

df_corr <- df_model_cells |>
  left_join(df_humans_cells, by = c("effort", "action")) |>
  mutate(model = droplevels(model))

correlations <- format_correlation_labels(df_corr, model_pred, Humans, group_vars = "model")

ggplot(df_corr, aes(x = model_pred, y = Humans, color = effort, shape = action)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
  geom_errorbar(aes(ymin = Humans_ci_lower, ymax = Humans_ci_upper), width = 0.02) +
  geom_point(size = 3, alpha = PLOT_ALPHA) +
  geom_label(
    data = correlations,
    aes(x = -Inf, y = Inf, label = label),
    inherit.aes = FALSE,
    hjust = -0.08, vjust = 1.5,
    size = 4.5, fill = "white", alpha = 0.5, label.size = 0
  ) +
  scale_color_manual(values = EFFORT_COLORS) +
  scale_shape_manual(values = c(16, 17)) +
  labs(x = "Model", y = "Humans", color = "Effort", shape = "Action") +
  facet_wrap(~ model, nrow = 1) +
  coord_fixed_symmetric(df_corr$model_pred, df_corr$Humans)

ggsave(here("figures", "food_inv_intimacy_effort_alt_correlation.pdf"),
       device = cairo_pdf, width = 10, height = 4.5)
```

## Scenario-level

### Human belief updates by scenario

```{r}
df_belief_update_scenario <- df |>
  filter(stage == "posterior") |>
  group_by(scenario_label, effort, action, model) |>
  tidyboot_mean(belief_update)
```

```{r}
#| fig-width: 12
#| fig-asp: 0.8

ggplot(df_belief_update_scenario, aes(x = action, y = empirical_stat, fill = effort)) +
  geom_col(position = position_dodge(width = 0.9), alpha = PLOT_ALPHA) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.9), width = 0.1,
    color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_manual(values = EFFORT_COLORS) +
  labs(
    title = "Belief updates by scenario",
    y = "Belief update",
    x = "Observed action",
    fill = "Effort"
  ) +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 30, hjust = 1)) +
  facet_wrap(~ scenario_label, ncol = 4)
```

### Model belief updates by scenario

```{r}
#| fig-width: 14
#| fig-asp: 0.6

df_model_scenario <- df_model |>
  mutate(model = factor(model, levels = setdiff(MODEL_LEVELS, "Humans")))

ggplot(df_model_scenario, aes(x = action, y = belief_update, fill = effort)) +
  geom_col(position = position_dodge(width = 0.9), alpha = PLOT_ALPHA) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_manual(values = EFFORT_COLORS) +
  labs(
    title = "Model-predicted belief updates by scenario",
    y = "Belief update",
    x = "Observed action",
    fill = "Effort"
  ) +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 30, hjust = 1)) +
  facet_grid(model ~ scenario_label,
             labeller = labeller(scenario_label = label_wrap_gen(15)))
```
