---
title: "Study 1a — End-to-end LM baseline (exploratory)"
format:
  html:
    code-fold: true
    toc: true
    embed-resources: true
execute:
  echo: true
  warning: false
---

```{r setup, include=FALSE}
source(here::here("analysis", "utils.R"))
setup_analysis()
```

# Study 1a — end-to-end LM baseline (exploratory)

This is an exploratory comparison, **not** part of the main analysis. It asks how a
frontier LM prompted **end-to-end** — given the exact participant stimulus and asked
for the belief update directly, with no compositional decomposition — compares to
the structured inverse-planning model on the same Study 1a data.

The end-to-end task is run on two Together models so the result decomposes cleanly
rather than being confounded:

- **`meta-llama/Llama-3.3-70B-Instruct-Turbo`** — the *same* model the compositional
  pipeline uses for feature elicitation.
- **`moonshotai/Kimi-K2.6`** (instant mode) — a frontier model.

Reading the panels:

- **Structured (Llama-3.3-70B)** vs. **End-to-end (Llama-3.3-70B)** — model held
  fixed, so the difference isolates the contribution of the **compositional structure**.
- **End-to-end (Llama-3.3-70B)** vs. **End-to-end (Kimi-K2.6)** — approach held fixed,
  so the difference isolates the contribution of **raw model capability**.

The end-to-end predictions are *direct* (no fitting, no cross-validation); the
structured `full` predictions are the LOSO-CV out-of-sample ones from the main
pipeline. The primary comparison is the scale-invariant **cell-mean correlation**.

## Human belief updates (per cell)

```{r human}
data_path <- here::here("data", "food_inv_desire", "main_trials_long.csv")
have_human <- file.exists(data_path)
if (have_human) {
  df <- read_csv(data_path, show_col_types = FALSE)
  bu <- calculate_belief_update(df, "response") |>
    filter(stage == "posterior") |>
    mutate(
      action = factor_action(action_condition),
      intimacy = factor(intimacy, levels = INTIMACY_LEVELS),
      effort = factor_effort(effort)
    )
  human_agg <- bu |>
    group_by(action, intimacy, effort) |>
    tidyboot_mean(column = belief_update) |>
    ungroup()
  human_cells <- human_agg |>
    transmute(action, intimacy, effort, human = empirical_stat)
  cat("Human cells:", nrow(human_cells), "\n")
} else {
  cat("No human data at", data_path, "\n")
}
```

## Load predictions (structured full + end-to-end models)

```{r load-preds}
# Map each predictions file to a panel label. The structured `full` model comes
# from the main pipeline's cv_preds_summary.json; the two end-to-end files come
# from LM_evals. Panel labels/order are defined here (not in utils.R) to keep this
# exploratory artifact self-contained.
pred_sources <- list(
  list(
    path = here::here("model", "outputs", "food_inv_desire", "cv_preds_summary.json"),
    label = "Structured (Llama-3.3-70B)",
    filter_model = "full"
  ),
  list(
    path = here::here("LM_evals", "outputs", "food_inv_desire", "e2e_preds_e2e_llama_70b.json"),
    label = "End-to-end (Llama-3.3-70B)",
    filter_model = NULL
  ),
  list(
    path = here::here("LM_evals", "outputs", "food_inv_desire", "e2e_preds_e2e_kimi_k2.json"),
    label = "End-to-end (Kimi-K2.6)",
    filter_model = NULL
  )
)

read_pred_cells <- function(src) {
  if (!file.exists(src$path)) return(NULL)
  d <- read_model_json(src$path)
  if (!is.null(src$filter_model)) d <- d |> filter(model == src$filter_model)
  d |>
    mutate(
      action = factor(ACTION_LABELS[action + 1], levels = ACTION_LABELS),
      intimacy = factor(intimacy_condition, levels = INTIMACY_LEVELS),
      effort = factor_effort(effort_condition),
      source = src$label
    ) |>
    group_by(source, action, intimacy, effort) |>
    summarise(pred = mean(delta_desire, na.rm = TRUE), .groups = "drop")
}

preds <- purrr::map(pred_sources, read_pred_cells)
present_labels <- purrr::map_chr(pred_sources, "label")[!purrr::map_lgl(preds, is.null)]
preds_all <- dplyr::bind_rows(preds)
if (nrow(preds_all)) {
  preds_all <- preds_all |> mutate(source = factor(source, levels = present_labels))
  cat("Loaded prediction sources:", paste(present_labels, collapse = " | "), "\n")
} else {
  cat("No prediction files found yet — run the LM_evals runner and the CV.\n")
}
```

## Cell-level scatter (model vs. human), with per-panel r

Each point is one action × intimacy × effort cell: the source's predicted belief
update against the human mean. Dashed line is y = x; the per-panel r is the
cell-level correlation with a bootstrap 95% CI.

```{r scatter}
#| fig-width: 13
#| fig-height: 4.5
if (have_human && nrow(preds_all)) {
  scatter_df <- preds_all |> left_join(human_cells, by = c("action", "intimacy", "effort"))
  rlab <- format_correlation_labels(scatter_df, pred, human, group_vars = "source") |>
    mutate(panel_lab = sprintf("%s\n(r = %.2f [%.2f, %.2f])", source, r, ci_lower, ci_upper))
  panel_levels <- rlab |>
    mutate(source = factor(source, levels = present_labels)) |>
    arrange(source) |>
    pull(panel_lab)
  scatter_df <- scatter_df |>
    left_join(select(rlab, source, panel_lab), by = "source") |>
    mutate(panel_lab = factor(panel_lab, levels = panel_levels))
  lim <- range(c(scatter_df$human, scatter_df$pred), na.rm = TRUE)
  p_scatter <- ggplot(scatter_df, aes(pred, human, color = intimacy, shape = effort)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey70") +
    geom_point(size = 2.6) +
    facet_wrap(~ panel_lab, nrow = 1) +
    scale_color_intimacy() +
    coord_equal(xlim = lim, ylim = lim) +
    labs(x = "LM / model predicted belief update",
         y = "Human belief update", shape = "Effort of low-risk share")
  print(p_scatter)
  save_figure(p_scatter, "food_inv_desire_e2e_comparison.pdf")
} else {
  cat("Need human data and at least one prediction source.\n")
}
```

## By-condition bar overlay

Humans alongside each prediction source, faceted; bars are grouped by
relationship (fill) and effort of the low-risk share (pattern).

```{r bars}
#| fig-width: 14
#| fig-height: 5
if (have_human && nrow(preds_all)) {
  library(ggpattern)
  bar_levels <- c("Humans", present_labels)
  combined <- bind_rows(
    human_agg |>
      transmute(panel = "Humans", action, intimacy, effort,
                stat = empirical_stat, ci_lower, ci_upper),
    preds_all |>
      transmute(panel = as.character(source), action, intimacy, effort,
                stat = pred, ci_lower = NA_real_, ci_upper = NA_real_)
  ) |>
    mutate(panel = factor(panel, levels = bar_levels))

  pd <- position_dodge(width = 0.9)
  p_bar <- ggplot(combined, aes(action, stat, group = interaction(intimacy, effort))) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey70", linewidth = 1) +
    geom_col_pattern(
      aes(fill = intimacy, pattern = effort),
      position = pd, width = 0.85, linewidth = 0.2,
      pattern_fill = "white", pattern_colour = NA,
      pattern_density = 0.25, pattern_spacing = 0.02, pattern_angle = 45
    ) +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2,
                  position = pd, color = "grey10", na.rm = TRUE) +
    facet_wrap(~ panel, nrow = 1) +
    scale_fill_intimacy() +
    scale_pattern_effort() +
    scale_x_action() +
    labs(x = "Observed action", y = "Belief update") +
    guides(
      fill = guide_legend(nrow = 2, override.aes = list(pattern = "none")),
      pattern = guide_legend(nrow = 2, override.aes = list(fill = "grey10"))
    ) +
    theme(legend.position = "bottom")
  print(p_bar)
  save_figure(p_bar, "food_inv_desire_e2e_comparison_bars.pdf")
} else {
  cat("Need human data and at least one prediction source.\n")
}
```

## Summary — cell-mean correlation with humans

```{r summary}
if (have_human && nrow(preds_all)) {
  format_correlation_labels(
    preds_all |> left_join(human_cells, by = c("action", "intimacy", "effort")),
    pred, human, group_vars = "source"
  ) |>
    mutate(source = factor(source, levels = present_labels)) |>
    arrange(source) |>
    select(source, r, ci_lower, ci_upper) |>
    knitr::kable(digits = 3)
} else {
  cat("No summary — predictions or human data missing.\n")
}
```
