---
title: "Study 2a — Intimacy inference"
format:
  html:
    code-fold: true
    toc: true
    embed-resources: true
execute:
  echo: true
  warning: false
---

```{r setup, include=FALSE}
source(here::here("analysis", "utils.R"))
setup_analysis()
```

# Study 2a — Intimacy inference

Observer infers intimacy from a single observed action, given desire and effort.

**Status:** Data are in `data/food_inv_intimacy/`; the demographics and
data-loading chunks below run on them. The belief-update and model sections
are still disabled (`eval=FALSE`) and need enabling plus a render check.

## Demographics

```{r demographics}
if (file.exists(here::here("data", "food_inv_intimacy", "exit_survey.csv"))) {
  report_demographics("food_inv_intimacy")
} else {
  cat("No data yet for food_inv_intimacy.\n")
}
```

## Data

```{r load-data}
data_path <- here::here("data", "food_inv_intimacy", "main_trials_long.csv")
have_data <- file.exists(data_path)
if (have_data) {
  df <- read_csv(data_path, show_col_types = FALSE)
  cat("Trials:", nrow(df), "\n")
  glimpse(df)
} else {
  cat("No data yet for food_inv_intimacy.\n")
}
```

## Belief update

```{r belief-update, eval=FALSE}
# Compute belief update (posterior - prior) per (subject, scenario) via
# calculate_belief_update(df, "intimacy_rating") from utils.R, guarded by
# have_data. Data exist in data/food_inv_intimacy/; this chunk still needs to
# be written, enabled (drop eval=FALSE), and verified with a render.
```

## Model predictions

```{r preds, eval=FALSE}
preds_path <- here::here("model", "outputs", "food_inv_intimacy", "cv_preds_summary.json")
if (file.exists(preds_path)) {
  preds <- read_model_json(preds_path)
  glimpse(preds)
} else {
  cat("No CV predictions yet — run `make cv-food_inv_intimacy` once a fit + data exist.\n")
}
```

## Model-vs-human correlations

```{r model-vs-human, eval=FALSE}
# TODO: wire up once the model predictions exist (data are in).
# Standard pattern (cf. the cv-load / cv-scatter chunks in
# food-inv-desire-analysis.qmd):
#   1. Aggregate human belief updates by cell.
#   2. Aggregate model predictions by the same cell.
#   3. Plot scatter + report bootstrap Pearson via format_correlation_labels().
```

## Notes

See the manuscript Methods section for the exact design, plus
`experiments/food_inv_intimacy/README.md` for the trial structure spec.
