---
title: Forward planning, effort manipulation (data analysis)
author: Alicia Chen
date: 2026-04-24
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    page-layout: full
    search: false
knitr:
  opts_chunk:
    dev: "ragg_png"
execute:
  warning: false
---

```{r setup}
source(here::here("analysis", "utils.R"))
setup_analysis()
```

## Demographics

```{r}
report_demographics("food_forw_intimacy_effort")
```

## Results

This experiment manipulates the relationship between two people (intimacy: 0, 50, 75, 100 on a 100-point scale) and the relative effort of avoiding saliva-sharing (effort: `low` vs `high`). Reward is held fixed at high across all trials. Each scenario has two actions: action 1 is non-saliva-sharing (requires the effort-constrained resource) and action 2 is saliva-sharing.

Load the data (produced by `analysis/json_to_csv.py food_forw_intimacy_effort`, with attention/memory-check exclusions already applied):

```{r}
df <- read_csv(here("data", "food_forw_intimacy_effort", "main_trials_long.csv"), show_col_types = FALSE) |>
  mutate(
    intimacy = factor(intimacy, levels = INTIMACY_LEVELS),
    effort = factor(effort, levels = c("low", "high"), labels = c("Low effort", "High effort")),
    action = factor(action)
  )

df_summary <- df |>
  group_by(intimacy, effort, action) |>
  tidyboot_mean(p_action) |>
  ungroup()
```

### P(share saliva) by effort and intimacy

Since the two actions are complementary (they sum to 1), P(action 2) = P(share saliva) is a sufficient summary. x-axis shows effort of the non-saliva-sharing action; hue shows the relationship's intimacy level.

```{r}
#| fig-width: 6
#| fig-asp: 0.75

df_summary |>
  filter(action == 2) |>
  ggplot(aes(x = effort, y = empirical_stat, color = intimacy, group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 4, alpha = PLOT_ALPHA) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), linewidth = 1.2, width = 0.4
  ) +
  geom_line(position = position_dodge(width = 0.2), linewidth = 1.2) +
  scale_color_intimacy() +
  scale_y_continuous(limits = c(0, 1)) +
  labs(y = "P(share saliva)", x = "Effort of not sharing saliva", color = "Intimacy")

ggsave(here("figures", "food_forw_intimacy_effort_paction.pdf"), device = cairo_pdf, width = 6, height = 4.5)
```

### Per-scenario breakdown

```{r}
#| fig-width: 22
#| fig-asp: 0.25

df_summary_scenario <- df |>
  group_by(scenario_label, intimacy, effort, action) |>
  tidyboot_mean(p_action) |>
  ungroup()

ggplot(df_summary_scenario, aes(x = action, y = empirical_stat, color = intimacy, group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 3, alpha = PLOT_ALPHA) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), width = 0.4
  ) +
  geom_line(position = position_dodge(width = 0.2), linewidth = 1) +
  scale_color_intimacy() +
  labs(y = "P(action)", x = "Action", color = "Intimacy") +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 0)) +
  facet_grid(effort ~ scenario_label, labeller = labeller(scenario_label = label_wrap_gen(15)))
```

## Model comparison

### Model description

We fit three ablations of the canonical access utility to the human data. A joint actor chooses an action proportional to its total utility, following a softmax with inverse temperature $\alpha$. Per-(scenario, effort_condition, action) values of $\text{access}(a)$ and $\text{effort}(a)$ are LLM-generated (Llama-3.3-70B; see `model/lm/score_effort_features.py`). Reward is held fixed at high across all trials; both actions involve eating, so $V(a|s) = 1$ is stipulated for both. $w_v$ is therefore non-identified in the softmax and stays near its initialization during fitting.

**Full model**:

$$U(a|s, I, e) = w_v \cdot V(a) - w_d \cdot \text{access}(a) \cdot (1 - I) - w_e \cdot \text{effort}(a)$$

**Discomfort-only**: drops the food-reward and effort terms, keeping only $-w_d \cdot \text{access}(a) \cdot (1 - I)$.

**Base model**: drops the access term: $U = w_v \cdot V(a) - w_e \cdot \text{effort}(a)$.

### Load model predictions

Model predictions are the **out-of-sample LOSO predictions** from `model/outputs/food_forw_intimacy_effort/cv_preds.csv`: for each trial, the prediction was made by a model fit on the other 15 scenarios.

```{r}
df_cv <- read_csv(here("model", "outputs", "food_forw_intimacy_effort", "cv_preds.csv"),
                  show_col_types = FALSE)

df_humans_cv <- df_cv |>
  filter(variant == "full") |>
  transmute(
    subject_id, scenario_label = held_out_scenario,
    intimacy, effort, action,
    p_action, model = "Humans"
  )

df_model_cv <- df_cv |>
  mutate(
    model = case_when(
      variant == "full" ~ "Full model",
      variant == "discomfort_only" ~ "Discomfort-only",
      variant == "base" ~ "Base model"
    )
  ) |>
  transmute(
    subject_id, scenario_label = held_out_scenario,
    intimacy, effort, action,
    p_action = p_action_pred, model
  )

df_preds <- bind_rows(df_humans_cv, df_model_cv) |>
  mutate(
    intimacy = factor(intimacy, levels = INTIMACY_LEVELS),
    effort = factor(effort, levels = c("low", "high"),
                    labels = c("Low effort", "High effort")),
    model = factor(model, levels = c(
      "Base model", "Discomfort-only", "Full model", "Humans"
    ))
  )

df_preds_summary <- df_preds |>
  group_by(intimacy, effort, action, model) |>
  tidyboot_mean(p_action) |>
  ungroup()
```

### Model predictions vs human data

P(share saliva) = P(action 2) by effort × intimacy, one panel per model. Lines connect effort levels within an intimacy group; the model panels should reproduce the rise with intimacy and the rise with high effort visible in the Humans panel.

```{r}
#| fig-width: 14
#| fig-asp: 0.32

df_preds_summary |>
  filter(action == 2) |>
  ggplot(aes(x = effort, y = empirical_stat, color = intimacy, group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 4, alpha = PLOT_ALPHA) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), linewidth = 1.2, width = 0.4
  ) +
  geom_line(position = position_dodge(width = 0.2), linewidth = 1.2) +
  scale_color_intimacy() +
  scale_y_continuous(limits = c(0, 1)) +
  labs(y = "P(share saliva)", x = "Effort of not sharing saliva", color = "Intimacy") +
  facet_wrap(~ model, nrow = 1)

ggsave(here("figures", "food_forw_intimacy_effort_comparison.pdf"),
       device = cairo_pdf, width = 14, height = 4.5)
```

### Why the Base model is flat: a geometric view

The Base model has *one effective free parameter* in this experiment ($w_e$; $w_v$ is non-identified because $V$ is constant across actions). Sweeping $w_e$ traces out a 1D curve in $(P(\text{share} \mid \text{low effort}), P(\text{share} \mid \text{high effort}))$ space — that's the entire universe of predictions the Base model can produce. The human marginal sits in a region this curve cannot reach, no matter how $w_e$ is tuned.

Concretely: action 2 (share) has lower LM-rated effort than action 1 (not-share) in *both* effort conditions (the manipulation acts on action 1's effort). So:

- $w_e > 0 \Rightarrow$ predictions in the upper-right (above 0.5, with positive slope — curve is above the diagonal).
- $w_e < 0 \Rightarrow$ predictions in the lower-left (below 0.5, but with **negative** slope — curve is below the diagonal).
- $w_e = 0 \Rightarrow$ exactly $(0.5, 0.5)$.

Humans are in the lower-left *with positive slope* — above the diagonal but below 0.5 in both axes. The Base model cannot represent that combination because slope direction and intercept sign are coupled through the same parameter.

```{r}
#| fig-width: 6.5
#| fig-asp: 1
#| label: fig-base-model-achievable

# Base model achievable curve: sweep w_e and compute marginal P(share) per condition,
# averaged across the 16 scenarios using the LM-rated effort tables.
df_eff <- read_csv(here("model", "outputs", "lm", "lm_scenario_params_effort.csv"),
                   show_col_types = FALSE) |>
  select(scenario_label, effort_condition, action, effort) |>
  pivot_wider(names_from = action, values_from = effort, names_prefix = "a") |>
  mutate(gap = a1 - a2)  # effort(not_share) - effort(share); positive in both conditions

df_curve <- tidyr::expand_grid(
  w_e = seq(-5, 6, by = 0.05),
  df_eff |> select(scenario_label, effort_condition, gap)
) |>
  mutate(p_share = plogis(w_e * gap)) |>
  group_by(w_e, effort_condition) |>
  summarize(p_share = mean(p_share), .groups = "drop") |>
  pivot_wider(names_from = effort_condition, values_from = p_share) |>
  mutate(region = ifelse(w_e >= 0, "Allowed (w_e >= 0)", "Forbidden (w_e < 0)"))

# Marginal P(share) per condition for Humans + each fitted model
df_points <- df_preds |>
  filter(action == 2) |>
  mutate(effort_short = ifelse(effort == "Low effort", "low", "high")) |>
  group_by(model, effort_short) |>
  summarize(p_share = mean(p_action), .groups = "drop") |>
  pivot_wider(names_from = effort_short, values_from = p_share)

ggplot() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "grey60") +
  geom_path(data = df_curve, aes(x = low, y = high, color = region),
            linewidth = 1.3) +
  scale_color_manual(values = c("Allowed (w_e >= 0)" = "#7A4A5A",
                                "Forbidden (w_e < 0)" = "#cccccc")) +
  geom_point(data = df_points, aes(x = low, y = high, fill = model),
             shape = 21, size = 5, color = "black", stroke = 0.5) +
  scale_fill_manual(values = c("Humans" = "white",
                               "Full model" = "#2C5F8D",
                               "Discomfort-only" = "#7BA7C7",
                               "Base model" = "#D17A6E")) +
  ggrepel::geom_text_repel(
    data = df_points, aes(x = low, y = high, label = model),
    size = 3.5, point.padding = 0.5, box.padding = 0.5, seed = 1
  ) +
  coord_fixed(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(
    x = "P(share saliva) | low effort",
    y = "P(share saliva) | high effort",
    color = "Base model curve",
    fill = NULL,
    title = "Base model's achievable predictions (one curve, one parameter)",
    subtitle = "Humans live in a region the curve cannot reach"
  ) +
  theme(legend.position = "bottom", legend.box = "vertical")

ggsave(here("figures", "food_forw_intimacy_effort_base_geometric.pdf"),
       device = cairo_pdf, width = 6.5, height = 6.5)
```

The curve passes through $(0.5, 0.5)$ at $w_e = 0$ — that's the closest point to the human marginal, and it's the optimum the optimizer settles on. The Full model breaks out of this 1D constraint because the access × intimacy term provides a second free dimension that decouples the slope from the intercept.

### Model-human correlation

```{r}
#| label: fig-food-forw-intimacy-effort-correlation
#| fig-width: 12
#| fig-asp: 0.4

df_humans_cells <- df_preds_summary |>
  filter(model == "Humans") |>
  select(intimacy, effort, action,
         Humans = empirical_stat,
         Humans_ci_lower = ci_lower,
         Humans_ci_upper = ci_upper)

df_corr <- df_preds_summary |>
  filter(model != "Humans") |>
  select(intimacy, effort, action, model,
         model_pred = empirical_stat,
         model_ci_lower = ci_lower,
         model_ci_upper = ci_upper) |>
  left_join(df_humans_cells, by = c("intimacy", "effort", "action")) |>
  filter(!is.na(model_pred), !is.na(Humans)) |>
  mutate(
    action = factor(action, levels = c(1, 2),
                    labels = c("No saliva", "Share saliva")),
    model = droplevels(model)
  )

correlations <- format_correlation_labels(df_corr, model_pred, Humans, group_vars = "model")

ggplot(df_corr, aes(x = model_pred, y = Humans, color = intimacy, shape = action)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
  geom_errorbar(aes(ymin = Humans_ci_lower, ymax = Humans_ci_upper), width = 0.02) +
  geom_errorbarh(aes(xmin = model_ci_lower, xmax = model_ci_upper), height = 0.02) +
  geom_point(size = 3, alpha = PLOT_ALPHA) +
  geom_label(
    data = correlations,
    aes(x = -Inf, y = Inf, label = label),
    inherit.aes = FALSE,
    hjust = -0.08, vjust = 1.5,
    size = 4.5, fill = "white", alpha = 0.5, label.size = 0
  ) +
  scale_color_intimacy() +
  scale_shape_manual(values = c(16, 17)) +
  labs(x = "Model", y = "Humans", color = "Intimacy", shape = "Action") +
  facet_wrap(~ model, nrow = 1) +
  coord_fixed_symmetric(df_corr$model_pred, df_corr$Humans)

ggsave(here("figures", "food_forw_intimacy_effort_correlation.pdf"),
       device = cairo_pdf, width = 12, height = 4.8)
```

### Model fit results

```{r}
fit_results <- read_csv(here("model", "outputs",
                             "forward_planning_effort_fit_results.csv"),
                        show_col_types = FALSE)

fit_results |>
  select(model, nll, n_params, aic, bic, r, r_ci_lower, r_ci_upper,
         starts_with("param_")) |>
  knitr::kable(digits = 3, caption = "Model fitting results (in-sample)")
```

```{r}
n_obs <- nrow(df)

fit_results |>
  select(model, nll, n_params, aic, bic) |>
  arrange(aic) |>
  knitr::kable(digits = 2,
               caption = "Model comparison (lower AIC/BIC is better)")
```
