---
title: "Study 2a — Intimacy inference"
format:
  html:
    code-fold: true
    toc: true
    embed-resources: true
execute:
  echo: true
  warning: false
---

```{r setup, include=FALSE}
source(here::here("analysis", "utils.R"))
library(ggpattern)
setup_analysis()
```

# Study 2a — Intimacy inference

Observer infers intimacy from a single observed action, given desire and effort.

**Status:** Full sample is in `data/food_inv_intimacy/` (236 participants
retained of 237 recruited). The demographics and belief-update panels below run
on it. The model-prediction and model-vs-human sections are still disabled
(`eval=FALSE`) pending CV predictions — mirror the Study 1a qmd
(`food-inv-desire-analysis.qmd`) once `make cv-food_inv_intimacy` has written
`cv_preds_summary.json`.

## Demographics

```{r demographics}
if (file.exists(here::here("data", "food_inv_intimacy", "exit_survey.csv"))) {
  report_demographics("food_inv_intimacy")
} else {
  cat("No data yet for food_inv_intimacy.\n")
}
```

## Data

```{r load-data}
data_path <- here::here("data", "food_inv_intimacy", "main_trials_long.csv")
have_data <- file.exists(data_path)
if (have_data) {
  df <- read_csv(data_path, show_col_types = FALSE)
  cat("Trials:", nrow(df), "\n")
  glimpse(df)
} else {
  cat("No data yet for food_inv_intimacy.\n")
}
```

## Belief update

Belief update is the posterior − prior intimacy rating, collected on a 0–100
slider and stored on the 0–1 scale, so updates fall in [−1, 1]. Intimacy is the
inferred latent here; desire and effort are the given conditions crossed within
every trial (mirroring Study 1a, where desire is inferred and intimacy is
given). The panels below show the mean update per observed action × given
desire × given effort cell, with 95% cluster-bootstrap CIs (resampling
subjects, the project standard, since subjects contribute multiple rows per
cell).

```{r belief-update-prep}
if (have_data) {
  bu <- calculate_belief_update(df, "intimacy_rating") |>
    filter(stage == "posterior") |>
    mutate(
      action = factor_action(action_condition),
      desire = factor(desire, levels = c("low", "high"), labels = c("Low", "High")),
      effort = factor_effort(effort)
    )

  agg <- boot_cluster_means(bu, "belief_update",
                            c("action", "desire", "effort")) |>
    rename(empirical_stat = belief_update,
           ci_lower = belief_update_ci_lower,
           ci_upper = belief_update_ci_upper)
}
```

Grouped bars mirror the Study 1a house style: fill = given desire, pattern =
given effort of the low-risk share, one bar per action × desire × effort cell.

```{r belief-update-bars}
#| fig-width: 5
#| fig-height: 5
if (have_data) {
  pd_bar <- position_dodge(width = 0.9)
  p_bar <- ggplot(agg, aes(action, empirical_stat,
                           group = interaction(desire, effort))) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey70", linewidth = 1) +
    geom_col_pattern(
      aes(fill = desire, pattern = effort),
      position = pd_bar, width = 0.85, linewidth = 0.2,
      pattern_fill = "white", pattern_colour = NA,
      pattern_density = 0.25, pattern_spacing = 0.02, pattern_angle = 45
    ) +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2,
                  position = pd_bar, color = "grey10") +
    scale_fill_desire() +
    scale_pattern_effort() +
    scale_x_action() +
    labs(title = "Intimacy inference by observed action, given desire and effort",
         x = "Observed action", y = "Intimacy belief update", fill = "Desire") +
    guides(
      fill = guide_legend(nrow = 2, override.aes = list(pattern = "none")),
      pattern = guide_legend(nrow = 2, override.aes = list(fill = "grey10"))
    ) +
    theme(legend.position = "bottom")
  print(p_bar)
} else {
  cat("No data yet for food_inv_intimacy.\n")
}
```

The same cell means as lines: color = given desire, one panel per given effort,
so the graded pattern across observed actions is easy to read.

```{r belief-update-lines}
#| fig-width: 9
#| fig-height: 4.5
if (have_data) {
  pd <- position_dodge(width = 0.25)
  p_line <- ggplot(agg, aes(action, empirical_stat, color = desire, group = desire)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey70") +
    geom_line(position = pd, linewidth = 0.8) +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.15,
                  position = pd, alpha = 0.8) +
    geom_point(position = pd, size = 2.6) +
    facet_wrap(~ effort) +
    scale_color_desire() +
    scale_x_action() +
    labs(x = "Observed action", y = "Intimacy belief update", color = "Desire")
  print(p_line)
} else {
  cat("No data yet for food_inv_intimacy.\n")
}
```

## Model predictions

```{r preds, eval=FALSE}
preds_path <- here::here("model", "outputs", "food_inv_intimacy", "cv_preds_summary.json")
if (file.exists(preds_path)) {
  preds <- read_model_json(preds_path)
  glimpse(preds)
} else {
  cat("No CV predictions yet — run `make cv-food_inv_intimacy` once a fit + data exist.\n")
}
```

## Model-vs-human correlations

```{r model-vs-human, eval=FALSE}
# TODO: wire up once the model predictions exist (data are in).
# Standard pattern (cf. the cv-load / cv-scatter chunks in
# food-inv-desire-analysis.qmd):
#   1. Aggregate human belief updates by cell.
#   2. Aggregate model predictions by the same cell.
#   3. Plot scatter + report bootstrap Pearson via format_correlation_labels().
```

## Notes

See the manuscript Methods section for the exact design, plus
`experiments/food_inv_intimacy/README.md` for the trial structure spec.
