---
title: "Study 1b — Joint inference over desire and effort"
format:
  html:
    code-fold: true
    toc: true
    embed-resources: true
execute:
  echo: true
  warning: false
---

```{r setup, include=FALSE}
source(here::here("analysis", "utils.R"))
```

# Study 1b — Joint inference over desire and effort

Observer jointly infers desire and effort from a single observed action, given
intimacy. Each trial has two slider responses.

**Status:** Full sample is in `data/food_inv_joint_de/` (recruited and
retained counts are reported by the demographics chunk below). This document
reports demographics and data checks
only — all figures (the joint belief-update panels and the model-vs-human
panels) are generated in Python by `figures/results/figure_study1b.py` and
`figures/results/figure_model_scatter.py` (`make figures-results`), and the
model-comparison statistics by `model/cv/model_comparison.py`
(`make model-comparison`).

## Demographics

```{r demographics}
if (file.exists(here::here("data", "food_inv_joint_de", "exit_survey.csv"))) {
  report_demographics("food_inv_joint_de")
} else {
  cat("No data yet for food_inv_joint_de.\n")
}
```

## Data

```{r load-data}
data_path <- here::here("data", "food_inv_joint_de", "main_trials_long.csv")
have_data <- file.exists(data_path)
if (have_data) {
  df <- read_csv(data_path, show_col_types = FALSE)
  cat("Trials:", nrow(df), "\n")
  glimpse(df)
} else {
  cat("No data yet for food_inv_joint_de.\n")
}
```

## Notes

See the manuscript Methods section for the exact design, plus
`experiments/food_inv_joint_de/README.md` for the trial structure spec.
