---
title: Inverse planning — desire inference, no alternatives shown (data analysis)
author: Alicia Chen
date: 2026-04-26
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    page-layout: full
    search: false
knitr:
  opts_chunk:
    dev: "ragg_png"
execute:
  warning: false
---

```{r setup}
source(here::here("analysis", "utils.R"))
setup_analysis()
```

## Demographics

```{r}
report_demographics("food_inv_desire_intimacy_noalt")
```

## Results

This experiment is a variant of `food_inv_desire_intimacy_alt` in which participants observe a single action and infer the actor's desire **without seeing the other three alternative actions the actors could have taken**. The slider endpoints are the scenario's `reward_low` / `reward_high` paragraphs, so the response encodes P(high desire). The experiment manipulates:

- **Action**: Which action was taken (action 0–3, varying in saliva-transfer risk)
- **Intimacy**: The relationship closeness between the actors (0, 50, 75, or 100)

The cognitive model for this variant uses an LM-generated counterfactual alternative set as the actor's action space, padded to a fixed size so the memo program can enumerate a static slot dimension. Crucially, alternatives are conditioned on **(scenario, observed action, relationship_condition)** — *not* on motivation — because the desire-noalt observer sees relationship and infers motivation. Conditioning the action space on what the observer sees (rather than on the latent) keeps the counterfactual reasoning consistent with what a human observer entertains. Slot 0 holds the observed canonical action; slots 1..k hold the LM-generated alternatives for that (scenario, observed, relationship) cell; remaining slots are null-padded.

## Load human data

```{r}
# main_trials_long.csv is generated by analysis/json_to_csv.py
# It excludes participants who failed attention check or got 0 on memory check
df <- read_csv(here("data", "legacy", "food_inv_desire_intimacy_noalt", "main_trials_long.csv")) |>
  mutate(
    action = str_replace(action_condition, "action_", ""),
    intimacy = factor(intimacy),
    p_high_reward = response / 100,  # Convert to probability scale
    stage = factor(stage, levels = c("prior", "posterior")),
    model = "Humans"
  ) |>
  select(-action_condition, -response)

# Calculate belief updates
df <- calculate_belief_update(df, "p_high_reward")

cat("Trials per participant:", nrow(df) / n_distinct(df$subject_id), "\n")
cat("N participants (after exclusions):", n_distinct(df$subject_id), "\n")
cat("N scenarios:", n_distinct(df$scenario_label), "\n")
```

## Distribution of ratings

```{r}
#| fig-width: 12
#| fig-asp: 0.45

ggplot(df, aes(x = p_high_reward, fill = action)) +
  geom_density(alpha = 0.5) +
  scale_fill_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  labs(y = "Density", x = "P(high desire)", fill = "Action") +
  theme(legend.position = "right") +
  facet_grid(stage ~ intimacy, labeller = labeller(intimacy = label_both, stage = label_value))
```

## Summary statistics

```{r}
df_summary <- df |>
  group_by(intimacy, action, stage, model) |>
  tidyboot_mean(p_high_reward)
```

## Prior vs posterior by action and intimacy

```{r}
#| fig-width: 12
#| fig-asp: 0.7

ggplot() +
  geom_violin(
    data = df,
    aes(x = stage, y = p_high_reward, fill = action),
    position = position_dodge(width = 0.8),
    alpha = 0.5,
    width = 0.7
  ) +
  geom_point(
    data = df_summary,
    aes(x = stage, y = empirical_stat, color = action, group = action),
    position = position_dodge(width = 0.8),
    size = 3,
    alpha = PLOT_ALPHA
  ) +
  geom_errorbar(
    data = df_summary,
    aes(x = stage, ymin = ci_lower, ymax = ci_upper, color = action, group = action),
    position = position_dodge(width = 0.8),
    width = 0.1,
    size = 1
  ) +
  geom_line(
    data = df_summary,
    aes(x = stage, y = empirical_stat, color = action, group = action),
    position = position_dodge(width = 0.8),
    size = 1
  ) +
  facet_grid(intimacy ~ action, labeller = label_both) +
  scale_fill_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  scale_color_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  labs(
    title = "P(high desire) before and after observing action (no-alt)",
    y = "P(high desire)",
    x = "Stage",
    fill = "Observed action",
    color = "Observed action"
  )
```

## Belief updates

```{r}
df_belief_update_summary <- df |>
  filter(stage == "posterior") |>
  group_by(intimacy, action, model) |>
  tidyboot_mean(belief_update)
```

```{r}
#| fig-width: 10
#| fig-asp: 0.45

df_belief_update_summary |>
  ggplot(aes(x = action, y = empirical_stat, fill = intimacy, group = intimacy)) +
  geom_col(position = position_dodge(width = 0.9), alpha = PLOT_ALPHA) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.9), width = 0.1, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_intimacy() +
  labs(
    title = "Belief updates by action and intimacy (no-alt humans)",
    y = "Belief update (posterior − prior, on P scale)",
    x = "Observed action",
    fill = "Intimacy"
  ) +
  theme(legend.position = "right")
```

## Model comparison

### Model description

We compare three ablations of the canonical risk utility (Base model, Discomfort-only, Full model). The action space for each (scenario, observed action, relationship_condition) cell is the union of the observed canonical action (slot 0) and an LM-generated set of plausible counterfactual alternatives (slots 1..k). Risk, effort, and $V(a|s, m)$ for each alternative are all LM-scored (batched per scenario, 10 runs each). Because relationship is the conditioning axis (not motivation), V for each alternative is scored under both motivation states so the observer can integrate over motivation as the latent.

**Full model**: Inverts the full risk utility $U(a|s, I) = w_v \cdot V(a|s, m) - w_d \cdot \text{risk}(a) \cdot (1 - I) - w_e \cdot \text{effort}(a)$.

**Discomfort-only**: Drops V and effort, keeping only the risk-discomfort term $-w_d \cdot \text{risk}(a) \cdot (1 - I)$. Note: this ablation drops the only term that varies with motivation, so it is structurally incapable of inferring motivation from action — its predictions stay at the uniform prior P(high) = 0.5.

**Base model**: Drops the risk term: $U = w_v \cdot V(a|s, m) - w_e \cdot \text{effort}(a)$.

The actor's softmax is $P(a | I, m, s) \propto \pi(a|s, I) \cdot \exp(U(a|I, m, s))$, where $\pi$ is uniform over the valid slots in each padded cell. The observer assumes a uniform motivation prior P(high) = 0.5 and computes the posterior via Bayesian inference over the padded action space.

**Fitting strategy.** All actor weights are jointly fit with $\alpha_{\text{observer}}$ on no-alt desire data — the padded observer reasons over a variable-length LM-generated action set, so weights from the alt-shown experiments don't transplant cleanly. Per variant the free parameters are: Full = $\{w_v, w_d, w_e, \alpha_{\text{observer}}\}$, Discomfort-only = $\{w_d, \alpha_{\text{observer}}\}$, Base model = $\{w_v, w_e, \alpha_{\text{observer}}\}$. Actor's softmax temperature $\alpha$ is fixed at 1 for identifiability. Loss is binary cross-entropy between the human slider/100 and model P(high motivation). The fitted-parameters table below uses the all-data fit; predictions reported in the model-vs-human comparison plots are out-of-sample, from leave-one-scenario-out CV (`food_inv_desire_intimacy_noalt/cv_preds_summary.csv`).

### Fitted parameters

```{r}
df_fit_results <- read_csv(here("model", "outputs", "food_inv_desire_intimacy_noalt", "fit_results.csv")) |>
  mutate(
    utility = str_remove(model, "_padded$"),
    model_label = case_when(
      utility == "full" ~ "Full model",
      utility == "discomfort_only" ~ "Discomfort-only",
      utility == "base" ~ "Base model"
    )
  ) |>
  select(model_label, param_w_v, param_w_d, param_w_e, param_gamma, alpha_observer, nll, n_params) |>
  arrange(model_label)

df_fit_results |>
  knitr::kable(
    digits = 3,
    caption = "Joint-fit parameters for no-alt desire inference (relationship-keyed action space).",
    col.names = c("Utility", "w_v", "w_d", "w_e", "γ", "α_observer", "NLL", "n_params")
  )
```

### AIC comparison

Following Burnham and Anderson (2002), ΔAIC values of 0–2 indicate essentially equivalent models, 4–7 indicate considerably less support for the higher-AIC model, and values greater than 10 indicate essentially no support for the higher-AIC model.

```{r}
kable_aic_table(
  df_fit_results,
  caption = "Model comparison (3 utility variants; lower AIC is better)",
  model_col_label = "Utility"
)
```

### Load model predictions

```{r}
df_model <- read_csv(here("model", "outputs", "food_inv_desire_intimacy_noalt", "cv_preds_summary.csv")) |>
  mutate(
    action = as.character(observed_action),
    intimacy = factor(intimacy),
    p_high_reward = p_high,
    belief_update = p_high - 0.5,
    utility = str_remove(model, "_padded$"),
    model_label = case_when(
      utility == "full" ~ "Full model",
      utility == "discomfort_only" ~ "Discomfort-only",
      utility == "base" ~ "Base model"
    )
  )

MODEL_LEVELS <- c("Base model", "Discomfort-only", "Full model", "Humans")
```

### Combine human and model data

```{r}
df_human_long <- df_belief_update_summary |>
  mutate(belief_update = empirical_stat, model_label = "Humans") |>
  select(intimacy, action, model_label, belief_update, ci_lower, ci_upper)

df_model_long <- df_model |>
  group_by(intimacy, action, model_label) |>
  summarize(belief_update = mean(belief_update), .groups = "drop") |>
  mutate(ci_lower = NA_real_, ci_upper = NA_real_)
```

### Belief updates: Model vs human comparison

```{r}
#| fig-width: 12
#| fig-asp: 0.3

df_combined <- bind_rows(df_human_long, df_model_long) |>
  mutate(
    model_label = factor(model_label, levels = MODEL_LEVELS),
    action = factor(action)
  )

p_belief <- df_combined |>
  ggplot(aes(x = action, y = belief_update, group = interaction(model_label, intimacy), fill = intimacy)) +
  geom_crossbar(
    aes(y = 0, ymin = -0.01, ymax = 0.01),
    position = position_dodge(width = 0.9),
    width = 0.9,
    color = NA
  ) +
  geom_col(position = position_dodge(width = 0.9), alpha = PLOT_ALPHA) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.9), size = 1, width = 0.1, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_intimacy() +
  labs(y = "Belief update", x = "Observed action", fill = "Intimacy") +
  theme(legend.position = "right") +
  facet_wrap(~model_label, nrow = 1)

print(p_belief)

ggsave(here("figures", "food_inv_desire_intimacy_noalt_belief_main.pdf"), plot = p_belief, device = cairo_pdf)
```

### Belief updates vs LM-derived risk (continuous x)

X-axis is the LM-derived risk score for the observed canonical action (from `lm_scenario_params.csv`). One point per (scenario, action, intimacy), with a linear smoother per intimacy level.

```{r}
#| fig-width: 14
#| fig-asp: 0.3

df_canonical_risk <- read_csv(here("model", "outputs", "lm", "lm_scenario_params.csv"), show_col_types = FALSE) |>
  mutate(action = as.character(action)) |>
  select(scenario_label, action, access) |>
  rename(risk = access)

df_human_scenario_pts <- df |>
  filter(stage == "posterior") |>
  group_by(scenario_label, intimacy, action) |>
  tidyboot_mean(belief_update) |>
  ungroup() |>
  mutate(belief_update = empirical_stat, model_label = "Humans") |>
  select(scenario_label, intimacy, action, model_label, belief_update, ci_lower, ci_upper)

df_model_scenario_pts <- df_model |>
  select(scenario_label, intimacy, action, model_label, belief_update) |>
  mutate(ci_lower = NA_real_, ci_upper = NA_real_)

df_risk_combined <- bind_rows(df_human_scenario_pts, df_model_scenario_pts) |>
  inner_join(df_canonical_risk, by = c("scenario_label", "action")) |>
  mutate(model_label = factor(model_label, levels = MODEL_LEVELS))

p_belief_risk <- df_risk_combined |>
  ggplot(aes(x = risk, y = belief_update, color = intimacy, shape = factor(action), group = intimacy)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  geom_errorbar(
    data = df_risk_combined |> filter(model_label == "Humans"),
    aes(ymin = ci_lower, ymax = ci_upper),
    position = POS_JITTER_DODGE,
    width = 0.05, alpha = 0.4, linewidth = 0.4
  ) +
  geom_point(position = POS_JITTER_DODGE, alpha = 0.5, size = 2) +
  geom_smooth(se = FALSE, linewidth = 1) +
  scale_color_intimacy() +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(y = "Belief update", x = "LM-derived risk of observed action", color = "Intimacy", shape = "Action") +
  theme(legend.position = "right") +
  facet_wrap(~model_label, nrow = 1)

print(p_belief_risk)

ggsave(here("figures", "food_inv_desire_intimacy_noalt_belief_risk.pdf"),
       plot = p_belief_risk, device = cairo_pdf, width = 14, height = 4)
```

## Scenario-level

### Human belief updates by scenario

```{r}
df_belief_update_scenario <- df |>
  filter(stage == "posterior") |>
  group_by(scenario_label, intimacy, action, model) |>
  tidyboot_mean(belief_update)
```

```{r}
#| fig-width: 12
#| fig-asp: 0.85

ggplot(df_belief_update_scenario, aes(x = action, y = empirical_stat, fill = intimacy)) +
  geom_col(position = position_dodge(width = 0.9), alpha = PLOT_ALPHA) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.9), width = 0.1, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_intimacy() +
  labs(
    title = "Belief updates by scenario (no-alt humans)",
    y = "Belief update",
    x = "Observed action",
    fill = "Intimacy"
  ) +
  theme(legend.position = "right") +
  facet_wrap(~scenario_label, ncol = 4)
```

### Posterior P(high desire) by scenario

```{r}
df_posterior_scenario <- df |>
  filter(stage == "posterior") |>
  group_by(scenario_label, intimacy, action) |>
  tidyboot_mean(p_high_reward)
```

```{r}
#| fig-width: 12
#| fig-asp: 0.85

ggplot(df_posterior_scenario, aes(x = action, y = empirical_stat, fill = intimacy)) +
  geom_col(position = position_dodge(width = 0.9), alpha = PLOT_ALPHA) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.9), width = 0.1, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey") +
  scale_fill_intimacy() +
  labs(
    title = "Posterior P(high desire) by scenario (no-alt humans)",
    y = "Posterior P(high desire)",
    x = "Observed action",
    fill = "Intimacy"
  ) +
  theme(legend.position = "right") +
  facet_wrap(~scenario_label, ncol = 4)
```

### Model belief updates by scenario

```{r}
#| fig-width: 14
#| fig-asp: 0.9

df_model_scenario <- df_model |>
  mutate(model_label = factor(model_label, levels = c("Base model", "Discomfort-only", "Full model")))

ggplot(df_model_scenario, aes(x = action, y = belief_update, fill = intimacy)) +
  geom_col(position = position_dodge(width = 0.9), alpha = PLOT_ALPHA) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_intimacy() +
  labs(
    title = "Model-predicted belief updates by scenario",
    y = "Belief update",
    x = "Observed action",
    fill = "Intimacy"
  ) +
  theme(legend.position = "right") +
  facet_grid(model_label ~ scenario_label,
             labeller = labeller(scenario_label = label_wrap_gen(15)))
```
