---
title: "Study 2b — Joint inference over intimacy and effort"
format:
  html:
    code-fold: true
    toc: true
    embed-resources: true
execute:
  echo: true
  warning: false
---

```{r setup, include=FALSE}
source(here::here("analysis", "utils.R"))
setup_analysis()
```

# Study 2b — Joint inference over intimacy and effort

Observer jointly infers relationship intimacy and physical effort from a single observed action, given desire. Each trial has two slider responses: intimacy on a 0–100 numeric scale, and effort as a continuous rating between two states.

**Status:** Scaffolding stub. The plots below come online once pilot data lands in
`data/food_inv_joint_ie/`. Until then, this qmd is structure-only.

## Demographics

```{r demographics}
if (file.exists(here::here("data", "food_inv_joint_ie", "exit_survey.csv"))) {
  report_demographics("food_inv_joint_ie")
} else {
  cat("No data yet for food_inv_joint_ie.\n")
}
```

## Data

```{r load-data}
data_path <- here::here("data", "food_inv_joint_ie", "main_trials_long.csv")
if (file.exists(data_path)) {
  df <- read_csv(data_path, show_col_types = FALSE)
  cat("Trials:", nrow(df), "\n")
  glimpse(df)
} else {
  cat("No data yet for food_inv_joint_ie.\n")
}
```

## Belief update

Each trial yields two belief updates (posterior − prior): intimacy on the
numeric scale and effort on the paragraph-endpoint scale, where higher values
mean the low-risk share option takes more effort. Both ratings are collected on
0–100 sliders and stored on the 0–1 scale, so updates fall in [−1, 1]. The
joint plot below shows the mean update per observed action × given desire cell,
with crosshairs giving 95% cluster-bootstrap CIs (resampling subjects) on each
axis.

```{r belief-update}
#| fig-width: 11
#| fig-height: 7
if (exists("df")) {
  bu <- df |>
    calculate_belief_update("intimacy_rating") |>
    rename(intimacy_update = belief_update) |>
    calculate_belief_update("effort_rating") |>
    rename(effort_update = belief_update) |>
    filter(stage == "posterior") |>
    mutate(
      action = factor(action_condition,
                      levels = c("no_share", "low_risk_share", "high_risk_share"),
                      labels = c("No share", "Low-risk share", "High-risk share")),
      desire = factor(desire, levels = c("low", "high"), labels = c("Low", "High"))
    )

  agg <- boot_cluster_means(bu, c("intimacy_update", "effort_update"),
                            c("action", "desire"))

  # Symmetric, origin-centered, equal x/y limits (a vector plot needs equal units
  # on both axes), including the crosshair CI extents.
  lim <- max(abs(c(agg$intimacy_update, agg$effort_update,
                   agg$intimacy_update_ci_lower, agg$intimacy_update_ci_upper,
                   agg$effort_update_ci_lower, agg$effort_update_ci_upper)),
             na.rm = TRUE) * 1.08

  p <- ggplot(agg, aes(intimacy_update, effort_update,
                       color = desire, shape = action)) +
    geom_hline(yintercept = 0, linetype = "dotted", color = "grey70") +
    geom_vline(xintercept = 0, linetype = "dotted", color = "grey70") +
    geom_segment(aes(xend = intimacy_update, yend = effort_update), x = 0, y = 0,
                 alpha = 0.6, linewidth = 0.5) +
    geom_errorbar(aes(ymin = effort_update_ci_lower, ymax = effort_update_ci_upper),
                  width = 0, alpha = 0.7) +
    geom_errorbar(aes(xmin = intimacy_update_ci_lower, xmax = intimacy_update_ci_upper),
                  width = 0, alpha = 0.7, orientation = "y") +
    geom_point(size = 3) +
    scale_color_desire() +
    scale_shape_manual(values = ACTION_SHAPES) +
    coord_fixed() +
    labs(
      title = "Joint belief update by action and given desire",
      subtitle = "Color = given desire, shape = observed action; crosshairs 95% cluster-bootstrap CIs",
      x = "Intimacy belief update",
      y = "Effort belief update (low-risk share)",
      color = "Desire", shape = "Observed action"
    )
  print(p)
} else {
  cat("No data yet for food_inv_joint_ie.\n")
}
```

## Model predictions

```{r preds, eval=FALSE}
preds_path <- here::here("model", "outputs", "food_inv_joint_ie", "cv_preds_summary.json")
if (file.exists(preds_path)) {
  preds <- read_model_json(preds_path)
  glimpse(preds)
} else {
  cat("No CV predictions yet — run `make cv-food_inv_joint_ie` once a fit + data exist.\n")
}
```

## Model-vs-human correlations

```{r model-vs-human, eval=FALSE}
# TODO: wire up once both data and model predictions exist.
# Standard pattern (cf. food-inv-desire-analysis.qmd):
#   1. Aggregate human belief updates by cell.
#   2. Aggregate model predictions by the same cell.
#   3. Plot scatter + report bootstrap Pearson via format_correlation_labels().
```

## Notes

See the manuscript Methods section for the exact design, plus
`experiments/food_inv_joint_ie/README.md` for the trial structure spec.
