---
title: "Study 1a — Desire inference"
format:
  html:
    code-fold: true
    toc: true
    embed-resources: true
execute:
  echo: true
  warning: false
---

```{r setup, include=FALSE}
source(here::here("analysis", "utils.R"))
setup_analysis()
```

# Study 1a — Desire inference

Observer infers the actors' desire from a single observed action, given known effort and intimacy. Design: 2 (effort: low / high) × 4 (intimacy: 0 / 50 / 75 / 100) × 3 (observed action). The desire DV is a continuous rating ("how much would they like to eat the food?", from "not at all" to "extremely"), collected on a 0–100 slider and stored on the 0–1 scale. The model's observer reasons over an LM-generated alternative-action set per cell (see `model/lm/generate_alternatives.py`).

**Status:** Scaffolding stub. The original pilot — which used a 0–100
"probability of two states" desire slider — is archived at
`analysis/legacy/food-inv-desire-pilot-analysis.qmd` (data in
`data/legacy/food_inv_desire_pilot/`). The plots below come online once data on the
corrected direct 0–100 desire DV lands in `data/food_inv_desire/`. Until then, this
qmd is structure-only.

## Demographics

```{r demographics}
if (file.exists(here::here("data", "food_inv_desire", "exit_survey.csv"))) {
  report_demographics("food_inv_desire")
} else {
  cat("No data yet for food_inv_desire (active continuous 0–100 DV).\n")
}
```

## Data

```{r load-data}
data_path <- here::here("data", "food_inv_desire", "main_trials_long.csv")
if (file.exists(data_path)) {
  df <- read_csv(data_path, show_col_types = FALSE)
  cat("Trials:", nrow(df), "\n")
  glimpse(df)
} else {
  cat("No data yet for food_inv_desire (active continuous 0–100 DV).\n")
}
```

## Belief update

The desire DV is a continuous rating on the 0–1 scale; the belief update is
posterior − prior on that scale (so it falls in [−1, 1]), via
`calculate_belief_update()`, aggregated by observed action × intimacy × effort
with a bootstrap 95% CI.

```{r belief-update}
#| fig-width: 11
#| fig-height: 5.5
if (exists("df")) {
  bu <- calculate_belief_update(df, "response") |>
    filter(stage == "posterior") |>
    mutate(
      action = factor(action_condition,
                      levels = c("no_share", "low_risk_share", "high_risk_share"),
                      labels = c("No share", "Low-risk share", "High-risk share")),
      intimacy = factor(intimacy, levels = INTIMACY_LEVELS),
      effort = factor(effort, levels = c("low", "high"),
                      labels = c("Low effort", "High effort"))
    )

  agg <- bu |>
    group_by(action, intimacy, effort) |>
    tidyboot_mean(column = belief_update) |>
    ungroup()

  pd <- position_dodge(width = 0.3)
  p <- ggplot(agg, aes(action, empirical_stat, color = intimacy, linetype = effort,
                       group = interaction(intimacy, effort))) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey70") +
    geom_line(position = pd, linewidth = 0.8) +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0,
                  position = pd, alpha = 0.7) +
    geom_point(position = pd, size = 2.4) +
    scale_color_intimacy() +
    labs(
      title = "Desire belief update by action, intimacy, and effort",
      x = "Observed action", y = "Belief update",
      color = "Intimacy", linetype = "Effort of low-risk share"
    )
  print(p)
} else {
  cat("No data yet for food_inv_desire.\n")
}
```

## Model predictions

```{r preds, eval=FALSE}
preds_path <- here::here("model", "outputs", "food_inv_desire", "cv_preds_summary.json")
if (file.exists(preds_path)) {
  preds <- read_model_json(preds_path)
  glimpse(preds)
} else {
  cat("No CV predictions yet — run `make cv-food_inv_desire` once a fit + data exist.\n")
}
```

## Model-vs-human correlations

```{r model-vs-human, eval=FALSE}
# TODO: wire up once both data and model predictions exist.
# Model predicts a posterior over the 0–1 desire grid; the predicted rating is its
# mean (0–1 scale), matching the human DV.
# Standard pattern: aggregate human + model by cell, scatter, bootstrap Pearson.
```

## Notes

See the manuscript Methods section for the exact design, plus
`experiments/food_inv_desire/README.md` for the trial structure spec.
