---
title: "Study 1a — LM elicitation diagnostics"
format:
  html:
    code-fold: true
    toc: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
source(here::here("analysis", "utils.R"))
setup_analysis()
library(knitr)
```

# Study 1a — LM elicitation diagnostics

This notebook inspects the **LM elicitation** that feeds the Study 1a
(`food_inv_desire`) observer model — *not* the human data. The model's actor
softmaxes over `{observed action} ∪ LM-generated alternatives`, and the
reward / discomfort / effort terms are driven by LM-elicited per-action features.
The elicitation is a K-run simulated-observer mixture: each
(scenario × condition) cell is elicited `K` independent times, and the
run-to-run spread becomes part of the model's predicted distribution. This
notebook checks that the two elicitation stages produced sensible, well-covered,
internally-consistent output.

The two source files (under `model/outputs/lm/food_inv_desire/`):

- **`lm_alternatives.jsonl`** — stage 1. One record per generated counterfactual
  action (`generate_alternatives.py`): the action text plus its cell key
  (scenario, observed action, effort condition, intimacy condition) and `run_id`.
- **`lm_runs.jsonl`** — stage 2. One record per `(run, cell)`
  (`score_merged.py`), carrying that run's scored action list (slot 0 = the
  observed canonical action, slots 1..k = the run's alternatives) on risk,
  effort, and goal-satisfaction `g`, plus the run's given **intimacy** scalar for
  the cell's intimacy condition.

All three action features (risk / effort / g) are on a 0–1 scale (the absolute
scale is absorbed by the freely-fitted utility weights).

```{r load}
lm_dir <- here::here("model", "outputs", "lm", "food_inv_desire")
runs_path <- file.path(lm_dir, "lm_runs.jsonl")
alts_path <- file.path(lm_dir, "lm_alternatives.jsonl")

have_lm <- file.exists(runs_path) && file.exists(alts_path)

action_levels <- c("no_share", "low_risk_share", "high_risk_share")
action_labels <- c("No share", "Low-risk share", "High-risk share")
as_action <- function(x) factor(x, levels = action_levels, labels = action_labels)

if (have_lm) {
  # Stage 2: per-(run, cell) records with a nested `actions` list-column.
  runs_raw <- read_model_jsonl(runs_path)
  actions <- runs_raw |>
    select(run_id, scenario_label, observed_action, effort_condition,
           intimacy_condition, intimacy, actions) |>
    tidyr::unnest(actions)

  # Stage 1: flat per-alternative records.
  alts <- read_model_jsonl(alts_path)

  K <- n_distinct(runs_raw$run_id)
  n_scenarios <- n_distinct(runs_raw$scenario_label)
  # Cell grid for 1a = scenario × observed_action × effort × intimacy.
  n_cells <- n_scenarios * length(action_levels) *
    n_distinct(runs_raw$effort_condition) * n_distinct(runs_raw$intimacy_condition)

  cat("K (elicitation runs):", K, "\n")
  cat("Scenarios:", n_scenarios, "\n")
  cat("Cells (expected):", n_cells, " | lm_runs records:", nrow(runs_raw),
      " (expected", n_cells * K, ")\n")
  cat("Stage-1 alternative records:", nrow(alts), "\n")
} else {
  cat("No LM elicitation output found at", lm_dir,
      "\nRun `make lm-food_inv_desire` (or the generate + score scripts) first.\n")
}
```

## Coverage

Every `(scenario × observed-action × effort × intimacy)` cell should appear in
all `K` runs. The table below confirms each run covers the full cell grid (so the
mixture has the same support in every component).

```{r coverage}
if (have_lm) {
  per_run <- runs_raw |>
    count(run_id, name = "cells") |>
    summarise(min = min(cells), max = max(cells), runs = n())
  cat("Cells per run — min:", per_run$min, "max:", per_run$max,
      "across", per_run$runs, "runs (expected", n_cells, "each)\n")

  missing <- (n_cells * K) - nrow(runs_raw)
  cat("Missing (run, cell) records:", missing, "\n")
}
```

## Stage 1 — generated alternatives

### Set size per `(cell, run)`

The LM decides how many alternatives to propose per cell (no fixed quota), so set
size is itself a diagnostic — too few and the actor's choice set is degenerate,
implausibly many and the LM may be padding.

```{r alt-set-size}
#| fig-width: 8
#| fig-height: 4
if (have_lm) {
  set_sizes <- actions |>
    filter(!is_canonical) |>
    count(scenario_label, observed_action, effort_condition,
          intimacy_condition, run_id, name = "n_alts")

  cat("Alternatives per (cell, run) — min:", min(set_sizes$n_alts),
      "max:", max(set_sizes$n_alts),
      "mean:", round(mean(set_sizes$n_alts), 2),
      "median:", median(set_sizes$n_alts), "\n")

  ggplot(set_sizes, aes(n_alts)) +
    geom_bar(fill = "#4A7A4A", alpha = PLOT_ALPHA) +
    scale_x_continuous(breaks = scales::breaks_width(1)) +
    labs(title = "Alternatives generated per (cell, run)",
         x = "Number of alternatives", y = "Count of (cell, run) units")
}
```

### Run-to-run diversity

Cross-run diversity is the point of the K-run pipeline (dedup is *within* a run;
repetition *across* runs is preserved). For each cell, `unique / total` over all
runs measures how much genuinely new phrasing the runs explore — near 1 means
every run proposes distinct actions, near `1/K` means the runs collapse onto the
same set.

```{r alt-diversity}
#| fig-width: 8
#| fig-height: 4
if (have_lm) {
  diversity <- alts |>
    mutate(key = tolower(trimws(action_text))) |>
    group_by(scenario_label, observed_action, effort_condition, intimacy_condition) |>
    summarise(total = n(), unique = n_distinct(key), .groups = "drop") |>
    mutate(diversity = unique / total)

  cat("Per-cell unique/total — mean:", round(mean(diversity$diversity), 2),
      "median:", round(median(diversity$diversity), 2), "\n")

  ggplot(diversity, aes(diversity)) +
    geom_histogram(bins = 25, fill = "#4A7A4A", alpha = PLOT_ALPHA) +
    labs(title = "Run-to-run alternative diversity per cell",
         subtitle = "unique alternative texts / total across runs",
         x = "Unique / total", y = "Cells")
}
```

### Sample alternatives

A spot-check that the generated actions are on-topic and span a plausible
risk range for a high-stakes scenario (oysters, observed high-risk share). Shows
the distinct alternatives proposed across all runs of one cell.

```{r alt-sample}
if (have_lm) {
  alts |>
    filter(scenario_label == "oysters", observed_action == "high_risk_share",
           effort_condition == "low", intimacy_condition == "somewhat_formal") |>
    distinct(action_text) |>
    head(15) |>
    kable(col.names = "Alternative action (oysters / high-risk share / low effort / somewhat formal)")
}
```

## Stage 2 — canonical action features

Slot 0 of each scored run is the observed canonical action. Aggregating slot 0
across scenarios and runs by `observed_action` recovers each canonical action's
feature distribution. Two hard sanity checks:

- **risk** should be monotone: `no_share` ≈ 0 < `low_risk_share` < `high_risk_share`.
- **goal-satisfaction `g`** should be near 0 for `no_share` (it doesn't deliver
  the food) and high for both sharing actions (they do). `g` is desire-free.

`risk` and `g` are intimacy- and effort-independent (risk is effort-marginal; `g`
is desire-free), so they're collapsed over those axes here.

```{r canon-risk-g}
#| fig-width: 10
#| fig-height: 4.5
if (have_lm) {
  canon <- actions |> filter(is_canonical)

  canon_rg <- canon |>
    group_by(scenario_label, observed_action, run_id) |>
    summarise(risk = mean(risk, na.rm = TRUE),
              g = mean(g, na.rm = TRUE), .groups = "drop") |>
    pivot_longer(c(risk, g), names_to = "feature", values_to = "value") |>
    mutate(action = as_action(observed_action),
           feature = factor(feature, levels = c("risk", "g"),
                            labels = c("Risk", "Goal-satisfaction g")))

  ggplot(canon_rg, aes(action, value, color = action)) +
    geom_boxplot(outlier.alpha = 0.2, width = 0.6) +
    facet_wrap(~feature) +
    scale_color_manual(values = setNames(
      c("#7A9A4A", "#C9A85A", "#B05A5A"), action_labels), guide = "none") +
    labs(title = "Canonical action features (slot 0), pooled over scenarios × runs",
         x = "Observed action", y = "Feature value (0–1)") +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
}
```

### Effort by condition

Effort is scored *with* the low-risk-share effort paragraph appended, once per
`effort_condition`. The manipulation targets the low-risk share, so its effort
should rise from the low to the high condition; the other actions' effort is a
useful contrast.

```{r canon-effort}
#| fig-width: 10
#| fig-height: 4.5
if (have_lm) {
  canon_eff <- actions |>
    filter(is_canonical) |>
    group_by(scenario_label, observed_action, run_id, effort_condition) |>
    summarise(effort = mean(effort, na.rm = TRUE), .groups = "drop") |>
    mutate(action = as_action(observed_action),
           effort_cond = factor(effort_condition, levels = c("low", "high"),
                                labels = c("Low effort", "High effort")))

  pd <- position_dodge(width = 0.6)
  ggplot(canon_eff, aes(action, effort, color = effort_cond)) +
    geom_boxplot(outlier.alpha = 0.15, width = 0.6, position = pd) +
    scale_color_effort() +
    labs(title = "Effort of canonical actions by effort condition",
         x = "Observed action", y = "Effort (0–1)", color = "Effort condition") +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
}
```

### Run-to-run spread of features

The mixture's spread comes from run-to-run variation in the scored features.
Per-cell standard deviation of slot-0 risk and `g` across the `K` runs should be
non-trivially positive (else every mixture component is identical and the model
predicts a degenerate, zero-variance belief update).

```{r canon-spread}
#| fig-width: 8
#| fig-height: 4
if (have_lm) {
  spread <- actions |>
    filter(is_canonical) |>
    group_by(scenario_label, observed_action, effort_condition, intimacy_condition) |>
    summarise(risk = sd(risk, na.rm = TRUE),
              g = sd(g, na.rm = TRUE), .groups = "drop") |>
    pivot_longer(c(risk, g), names_to = "feature", values_to = "sd")

  spread |>
    group_by(feature) |>
    summarise(mean_sd = round(mean(sd, na.rm = TRUE), 3),
              max_sd = round(max(sd, na.rm = TRUE), 3), .groups = "drop") |>
    kable(caption = "Within-cell run-to-run SD of slot-0 features")

  ggplot(spread, aes(sd, fill = feature)) +
    geom_histogram(bins = 25, alpha = 0.7, position = "identity") +
    scale_fill_manual(values = c("risk" = "#B05A5A", "g" = "#4A7A4A")) +
    labs(title = "Run-to-run SD of canonical features (per cell)",
         x = "SD across runs", y = "Cells", fill = "Feature")
}
```

## Given intimacy scalar

Studies 1a/1b take intimacy as observed context. Its magnitude is LM-elicited
**per run** from the de-anchored relationship descriptors (4 levels, scenario-
independent). It should be **monotone** across `max_formal → max_intimate`, and —
because it is scored per run — should carry run-to-run spread like the action
features do.

```{r intimacy-scalar}
#| fig-width: 8
#| fig-height: 4.5
if (have_lm) {
  intim <- runs_raw |>
    distinct(intimacy_condition, run_id, intimacy) |>
    mutate(intimacy_condition = factor(intimacy_condition, levels = INTIMACY_LEVELS))

  intim_summ <- intim |>
    group_by(intimacy_condition) |>
    summarise(value = mean(intimacy),
              sd = sd(intimacy),
              n_runs = n_distinct(run_id), .groups = "drop")

  intim_summ |>
    mutate(across(c(value, sd), \(x) round(x, 3))) |>
    kable(caption = "Per-level intimacy scalar (mean over runs, run-to-run SD)")

  ggplot(intim_summ, aes(intimacy_condition, value, color = intimacy_condition)) +
    geom_errorbar(aes(ymin = value - sd, ymax = value + sd), width = 0.15) +
    geom_point(size = 3.5) +
    scale_color_intimacy() +
    ylim(0, 1) +
    labs(title = "Elicited intimacy magnitude by level",
         subtitle = "error bars = run-to-run SD",
         x = NULL, y = "Intimacy magnitude (0–1)", color = "Intimacy level") +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
}
```

::: {.callout-note}
Watch the run-to-run SD column. If it is ≈ 0 for every level, the intimacy axis
contributes no spread to the mixture (the 4 levels collapse to fixed constants
across runs) — scored at the low default rating temperature, the descriptors may
return the same value every run. Also check the *spacing*: if
`somewhat_intimate` sits very close to `somewhat_formal`, the two middle levels
are nearly non-identifiable for the model.
:::

## Do the alternatives differ by condition?

The alternatives are elicited *per cell* — `(scenario × observed-action × effort ×
intimacy)` — so each observed conditioning axis can in principle reshape the
imagined choice set. This matters in two ways. The load-bearing one is
**anchoring on the observed action**: the cognitive claim is that an observer
imagines counterfactuals to *what they saw*, so the alternative set should shift
with `observed_action`. The other two axes — effort and intimacy — are observed
context in Study 1a (the latent is desire), so conditioning on them doesn't leak
the latent; but it's worth seeing whether, and how much, they move the set. The
checks below look at the alternatives two ways: the composition of the **set**
itself (size, and run-to-run text overlap) and the scored **properties** of the
non-canonical actions (risk / effort / `g`).

```{r alt-condition-helpers}
if (have_lm) {
  # Order intimacy levels by their elicited magnitude (ascending) so the axis
  # reads formal -> intimate, without depending on a fixed level vector.
  intim_order <- runs_raw |>
    group_by(intimacy_condition) |>
    summarise(m = mean(intimacy), .groups = "drop") |>
    arrange(m) |> pull(intimacy_condition)

  cond_levels <- c(action_labels, "Low", "High", intim_order)

  # Melt the three observed conditioning axes into a long axis/level form so each
  # gets its own panel; levels differ per axis, so panels use a free x scale.
  to_cond_long <- function(df) {
    df |>
      mutate(`Observed action`    = as.character(as_action(observed_action)),
             `Effort condition`   = factor(effort_condition, c("low", "high"),
                                           c("Low", "High")),
             `Intimacy condition` = factor(intimacy_condition, intim_order)) |>
      pivot_longer(c(`Observed action`, `Effort condition`, `Intimacy condition`),
                   names_to = "axis", values_to = "level") |>
      mutate(axis  = factor(axis, c("Observed action", "Effort condition",
                                    "Intimacy condition")),
             level = factor(level, cond_levels))
  }

  # Long condition slugs crowd the narrow facet panels; wrap them onto two lines
  # (underscores -> spaces first) instead of rotating. Generic, so it survives
  # any renaming of the intimacy slugs.
  wrap_lab <- function(x) stringr::str_wrap(gsub("_", " ", x), width = 10)
}
```

### Set size by condition

Does the LM propose systematically more or fewer alternatives in some conditions?
Set size is pooled over the other axes within each panel.

```{r alt-setsize-condition}
#| fig-width: 11
#| fig-height: 4
if (have_lm) {
  set_sizes_cond <- actions |>
    filter(!is_canonical) |>
    count(scenario_label, observed_action, effort_condition,
          intimacy_condition, run_id, name = "n_alts") |>
    to_cond_long()

  ggplot(set_sizes_cond, aes(level, n_alts)) +
    geom_boxplot(fill = "#4A7A4A", alpha = PLOT_ALPHA, outlier.alpha = 0.1,
                 width = 0.6) +
    stat_summary(fun = mean, geom = "point", shape = 18, size = 2.5,
                 color = "#B05A5A") +
    facet_wrap(~axis, scales = "free_x") +
    scale_x_discrete(labels = wrap_lab) +
    labs(title = "Alternative set size by condition",
         subtitle = "Alternatives per (cell, run); red diamond = mean",
         x = NULL, y = "Number of alternatives") +
    theme(axis.text.x = element_text(size = 12))
}
```

### Properties by condition

The headline check. Each non-canonical action carries a scored risk, effort, and
`g`; this grid asks whether those distributions shift with each conditioning axis.
The strongest signal should be down the **observed-action** column (the anchoring
claim) — e.g. whether imagined alternatives skew lower-risk once a high-risk share
has already been observed, and whether they more often deliver the food (`g`)
after a `no_share`. Movement down the effort/intimacy columns shows how much those
framings bleed into the imagined set.

```{r alt-features-condition}
#| fig-width: 11
#| fig-height: 8
if (have_lm) {
  alt_feat <- actions |>
    filter(!is_canonical) |>
    select(scenario_label, observed_action, effort_condition, intimacy_condition,
           run_id, risk, effort, g) |>
    pivot_longer(c(risk, effort, g), names_to = "feature", values_to = "value") |>
    mutate(feature = factor(feature, c("risk", "effort", "g"),
                            c("Risk", "Effort", "Goal-satisfaction g"))) |>
    to_cond_long()

  ggplot(alt_feat, aes(level, value)) +
    geom_boxplot(fill = "#6B8E6B", alpha = PLOT_ALPHA, outlier.alpha = 0.04,
                 width = 0.6) +
    stat_summary(fun = mean, geom = "point", shape = 18, size = 2,
                 color = "#B05A5A") +
    facet_grid(feature ~ axis, scales = "free_x") +
    scale_x_discrete(labels = wrap_lab) +
    labs(title = "Alternative-action features by condition",
         subtitle = paste("Non-canonical (LM-generated) actions only, pooled over",
                          "scenarios × runs; red diamond = mean"),
         x = NULL, y = "Feature value (0–1)") +
    theme(axis.text.x = element_text(size = 12))
}
```

### Does the *set* of actions itself differ?

Set size and feature distributions can look stable while the actual actions
change. To test the set directly, we measure the **Jaccard overlap** of the
(normalized) alternative texts between cells that differ on one axis, and compare
it against a noise floor: how much two independent *runs of the same cell*
overlap. If conditioning genuinely reshapes the set, across-axis overlap should
sit *below* that run-to-run floor; if it sits at the floor, the axis adds nothing
beyond elicitation stochasticity. (Overlap is exact normalized-text match, so
paraphrases count as misses — these numbers are a lower bound on semantic
overlap.)

```{r alt-overlap-condition}
#| fig-width: 9
#| fig-height: 4.5
if (have_lm) {
  norm_txt <- function(x) {
    x <- tolower(trimws(x))
    x <- gsub("[[:punct:]]+$", "", x)
    gsub("\\s+", " ", x)
  }
  jacc <- function(a, b) {
    a <- unique(a); b <- unique(b)
    u <- length(union(a, b))
    if (u == 0) NA_real_ else length(intersect(a, b)) / u
  }
  mean_pairwise <- function(sets) {
    if (length(sets) < 2) return(NA_real_)
    pr <- combn(length(sets), 2)
    mean(vapply(seq_len(ncol(pr)),
                \(j) jacc(sets[[pr[1, j]]], sets[[pr[2, j]]]), numeric(1)),
         na.rm = TRUE)
  }
  set_by <- function(df, ...) {
    df |> group_by(...) |> summarise(set = list(unique(key)), .groups = "drop")
  }

  alt_keys <- alts |> mutate(key = norm_txt(action_text))

  # (1) Noise floor: same cell, across the K runs.
  base_ov <- set_by(alt_keys, scenario_label, observed_action,
                    effort_condition, intimacy_condition, run_id) |>
    group_by(scenario_label, observed_action, effort_condition, intimacy_condition) |>
    summarise(overlap = mean_pairwise(set), .groups = "drop") |>
    mutate(comparison = "Across runs\n(same cell)")

  # (2) Across effort, holding scenario / action / intimacy / run fixed.
  eff_ov <- set_by(alt_keys, scenario_label, observed_action,
                   intimacy_condition, run_id, effort_condition) |>
    group_by(scenario_label, observed_action, intimacy_condition, run_id) |>
    summarise(overlap = mean_pairwise(set), .groups = "drop") |>
    mutate(comparison = "Across effort\n(same run)")

  # (3) Across intimacy, holding scenario / action / effort / run fixed.
  int_ov <- set_by(alt_keys, scenario_label, observed_action,
                   effort_condition, run_id, intimacy_condition) |>
    group_by(scenario_label, observed_action, effort_condition, run_id) |>
    summarise(overlap = mean_pairwise(set), .groups = "drop") |>
    mutate(comparison = "Across intimacy\n(same run)")

  overlap <- bind_rows(base_ov, eff_ov, int_ov) |>
    mutate(comparison = factor(comparison, levels = c(
      "Across runs\n(same cell)", "Across effort\n(same run)",
      "Across intimacy\n(same run)")))

  overlap |>
    group_by(comparison) |>
    summarise(mean = round(mean(overlap, na.rm = TRUE), 3),
              median = round(median(overlap, na.rm = TRUE), 3), .groups = "drop") |>
    kable(caption = "Jaccard overlap of alternative-text sets (exact normalized match)")

  ggplot(overlap, aes(comparison, overlap, fill = comparison)) +
    geom_boxplot(alpha = PLOT_ALPHA, outlier.alpha = 0.1, width = 0.6) +
    scale_fill_manual(values = c("#8A8A8A", "#4A7A8A", "#8A5A8A"), guide = "none") +
    ylim(0, 1) +
    labs(title = "Do the alternative sets differ by condition?",
         subtitle = "Run-to-run overlap within a cell vs. across an experimental axis",
         x = NULL, y = "Jaccard overlap (alternative texts)")
}
```

## Failed ratings

A scoring call can fail to return a parseable value for an action; that feature
is stored as missing (`null` → `NA`) rather than dropped, so the loader can NaN-
mask it. The failure rate should be low.

```{r failed-ratings}
if (have_lm) {
  actions |>
    summarise(across(c(risk, effort, g),
                     list(n_missing = \(x) sum(is.na(x)),
                          pct = \(x) round(mean(is.na(x)) * 100, 2)))) |>
    pivot_longer(everything(),
                 names_to = c("feature", ".value"),
                 names_pattern = "(.*)_(n_missing|pct)") |>
    kable(caption = "Missing (failed) ratings per feature",
          col.names = c("Feature", "N missing", "% missing"))
}
```

## Takeaways

- **Coverage & set sizes** — confirm every run covers the full cell grid and that
  alternative set sizes are reasonable (a handful per cell, none empty).
- **Canonical features** — risk should be monotone across the three actions and
  `g` should separate `no_share` (≈ 0) from the sharing actions; this is the load-
  bearing sanity check for the whole pipeline.
- **Run-to-run spread** — the action features should carry positive per-cell SD;
  the intimacy scalar is the one to scrutinize, since it can come back constant
  across runs (see the callout above).
- **Alternatives by condition** — the clearest movement should be down the
  observed-action column (the anchoring claim: e.g. a high-risk share already
  observed yields lower-risk imagined alternatives, and a `no_share` yields
  higher-`g` ones); the effort/intimacy axes show how much those framings reshape
  the imagined set, with the Jaccard panel separating real reshaping from
  elicitation noise.

See `.claude/rules/model.md` for the full elicitation design and
`model/lm/score_merged.py` for the scoring procedure.
