---
title: "Study 2a — Intimacy inference"
format:
  html:
    code-fold: true
    toc: true
    embed-resources: true
execute:
  echo: true
  warning: false
---

```{r setup, include=FALSE}
source(here::here("analysis", "utils.R"))
```

# Study 2a — Intimacy inference

Observer infers intimacy from a single observed action, given desire and effort.

**Status:** Full sample is in `data/food_inv_intimacy/` (recruited and
retained counts are reported by the demographics chunk below). This document
reports demographics and data checks only — all figures (including the model-vs-human panels) are generated in
Python by `figures/results/figure_study2.py` and
`figures/results/figure_model_scatter.py` (`make figures-results`), and the
model-comparison statistics by `model/cv/model_comparison.py`
(`make model-comparison`).

## Demographics

```{r demographics}
if (file.exists(here::here("data", "food_inv_intimacy", "exit_survey.csv"))) {
  report_demographics("food_inv_intimacy")
} else {
  cat("No data yet for food_inv_intimacy.\n")
}
```

## Data

```{r load-data}
data_path <- here::here("data", "food_inv_intimacy", "main_trials_long.csv")
have_data <- file.exists(data_path)
if (have_data) {
  df <- read_csv(data_path, show_col_types = FALSE)
  cat("Trials:", nrow(df), "\n")
  glimpse(df)
} else {
  cat("No data yet for food_inv_intimacy.\n")
}
```

## Notes

See the manuscript Methods section for the exact design, plus
`experiments/food_inv_intimacy/README.md` for the trial structure spec.
