---
title: "Study 1a — Desire inference"
format:
  html:
    code-fold: true
    toc: true
    embed-resources: true
execute:
  echo: true
  warning: false
---

```{r setup, include=FALSE}
source(here::here("analysis", "utils.R"))
setup_analysis()
```

# Study 1a — Desire inference

Observer infers the actors' desire from a single observed action, given known effort and intimacy. Design: 2 (effort: low / high) × 4 (intimacy: 0 / 50 / 75 / 100) × 3 (observed action). The desire DV is a continuous 0–100 rating ("how much do they want to eat the food?", from "not at all" to "extremely"). The model's observer reasons over an LM-generated alternative-action set per cell (see `model/lm/generate_alternatives.py`).

**Status:** Scaffolding stub. The original pilot — which used a 0–100
"probability of two states" desire slider — is archived at
`analysis/legacy/food-inv-desire-pilot-analysis.qmd` (data in
`data/legacy/food_inv_desire_pilot/`). The plots below come online once data on the
corrected direct 0–100 desire DV lands in `data/food_inv_desire/`. Until then, this
qmd is structure-only.

## Demographics

```{r demographics}
if (file.exists(here::here("data", "food_inv_desire", "exit_survey.csv"))) {
  report_demographics("food_inv_desire")
} else {
  cat("No data yet for food_inv_desire (active continuous 0–100 DV).\n")
}
```

## Data

```{r load-data}
data_path <- here::here("data", "food_inv_desire", "main_trials_long.csv")
if (file.exists(data_path)) {
  df <- read_csv(data_path, show_col_types = FALSE)
  cat("Trials:", nrow(df), "\n")
  glimpse(df)
} else {
  cat("No data yet for food_inv_desire (active continuous 0–100 DV).\n")
}
```

## Belief update

```{r belief-update, eval=FALSE}
# Desire DV is a continuous 0-100 rating; belief update = posterior - prior on the 0-100 scale.
# Uses calculate_belief_update() from utils.R on the response column.
```

## Model predictions

```{r preds, eval=FALSE}
preds_path <- here::here("model", "outputs", "food_inv_desire", "cv_preds_summary.csv")
if (file.exists(preds_path)) {
  preds <- read_csv(preds_path, show_col_types = FALSE)
  glimpse(preds)
} else {
  cat("No CV predictions yet — run `make cv-food_inv_desire` once a fit + data exist.\n")
}
```

## Model-vs-human correlations

```{r model-vs-human, eval=FALSE}
# TODO: wire up once both data and model predictions exist.
# Model predicts P(high desire); map to the 0-100 rating via pred_desire = 100 * P(high).
# Standard pattern: aggregate human + model by cell, scatter, bootstrap Pearson.
```

## Notes

See the manuscript Methods section for the exact design, plus
`experiments/food_inv_desire/README.md` for the trial structure spec.
