---
title: Inverse planning — intimacy inference, alternatives shown (data analysis)
author: Alicia Chen
date: 2026-01-15
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    page-layout: full
    search: false
execute:
  warning: false
---

```{r setup}
source(here::here("analysis", "utils.R"))
setup_analysis()
```

## Demographics

```{r}
report_demographics("food_inv-intimacy_desire_alt")
```

## Results

This experiment presents participants with scenarios where they observe an action taken by one person toward another. Participants rate their belief about the intimacy of the relationship before (prior) and after (posterior) observing the action. The experiment manipulates:

- **Action**: Which action was taken (action 0-3, varying in saliva-transfer risk)
- **Motivation (desire)**: Whether both people want the food (high vs low)

## Load human data

```{r}
# main_trials_long.csv is generated by analysis/json_to_csv.py
# It excludes participants who failed attention check or got 0 on memory check
df <- read_csv(here("data", "food_inv-intimacy_desire_alt", "main_trials_long.csv")) |>
  mutate(
    action = str_replace(action_condition, "action_", ""),
    motivation = factor(motivation, levels = c("low", "high"), labels = MOTIVATION_LEVELS),
    stage = factor(stage, levels = c("prior", "posterior")),
    model = "Humans"
  ) |>
  select(-action_condition)

# Calculate belief updates
df <- calculate_belief_update(df, "intimacy_rating")
```

## Human data: Prior vs posterior ratings

First, look at summaries of human data

### Distribution of ratings

```{r}
ggplot(df, aes(x = intimacy_rating, fill = action)) +
  geom_density(alpha = 0.5) +
  scale_fill_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  labs(y = "Density", x = "Intimacy rating", fill = "Action") +
  theme(legend.position = "right") +
  facet_grid(stage ~ motivation, labeller = labeller(motivation = label_both))
```

Calculate summary statistics

```{r}
df_summary <- df |>
  group_by(motivation, action, stage, model) |>
  tidyboot_mean(intimacy_rating)
```

### Prior vs posterior by action and motivation

```{r}
#| fig-width: 12
#| fig-asp: 0.4

ggplot() +
  geom_violin(
    data = df,
    aes(x = stage, y = intimacy_rating, fill = action),
    position = position_dodge(width = 0.8),
    alpha = 0.5,
    width = 0.7
  ) +
  geom_point(
    data = df_summary,
    aes(x = stage, y = empirical_stat, color = action, group = action),
    position = position_dodge(width = 0.8),
    size = 3,
    alpha = PLOT_ALPHA
  ) +
  geom_errorbar(
    data = df_summary,
    aes(x = stage, ymin = ci_lower, ymax = ci_upper, color = action, group = action),
    position = position_dodge(width = 0.8),
    width = 0.1,
    size = 1
  ) +
  geom_line(
    data = df_summary,
    aes(x = stage, y = empirical_stat, color = action, group = action),
    position = position_dodge(width = 0.8),
    size = 1
  ) +
  facet_grid(motivation ~ action, labeller = labeller(action = label_both)) +
  scale_fill_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  scale_color_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  labs(
    title = "Intimacy ratings before and after observing action",
    y = "Intimacy rating",
    x = "Stage",
    fill = "Observed action",
    color = "Observed action"
  )
```

### Belief updates

```{r}
df_belief_update_summary <- df |>
  filter(stage == "posterior") |>
  group_by(motivation, action, model) |>
  tidyboot_mean(belief_update)
```

## Model comparison

### Model description

We compare three ablations of the canonical risk utility, performing inverse planning to infer intimacy from observed actions:

**Full model**: Inverts the full risk utility $U(a|s, I) = w_v \cdot V(a|s) - w_d \cdot \text{risk}(a) \cdot (1 - I) - w_e \cdot \text{effort}(a)$.

**Discomfort-only**: Drops the food-reward and effort terms, keeping only the risk-discomfort term $-w_d \cdot \text{risk}(a) \cdot (1 - I)$.

**Base model**: Drops the risk term: $U = w_v \cdot V(a|s) - w_e \cdot \text{effort}(a)$.

Per-scenario values of $\text{risk}(a)$ and $\text{effort}(a)$ are generated by an LLM; reward $V(a|s)$ is stipulated as a binary goal-satisfaction gate ($V=1$ iff the action satisfies the active goal — sharing under high motivation, not-sharing under low motivation). All models assume a uniform prior over intimacy (0-100 scale) and compute posteriors via Bayesian inference. The observer inverse temperature ($\alpha_{\text{observer}}$) is fitted to human data.

### Fitted parameters

```{r}
df_fit_results <- read_csv(here("model", "outputs", "food_inv-intimacy_desire_alt", "fit_results.csv")) |>
  filter(experiment == "food_inv-intimacy_desire_alt") |>
  mutate(
    model_label = case_when(
      model == "full" ~ "Full model",
      model == "discomfort_only" ~ "Discomfort-only",
      model == "base" ~ "Base model"
    )
  ) |>
  filter(!is.na(model_label)) |>
  select(model_label, alpha_observer, nll, n_params)

df_fit_results |>
  knitr::kable(
    digits = 3,
    caption = "Fitted observer parameters for intimacy inference",
    col.names = c("Model", "α_observer", "NLL", "n_params")
  )
```

### AIC comparison

Following Burnham and Anderson (2002), ΔAIC values of 0–2 indicate essentially equivalent models, 4–7 indicate considerably less support for the higher-AIC model, and values greater than 10 indicate essentially no support for the higher-AIC model.

```{r}
kable_aic_table(df_fit_results)
```

### Load model predictions

```{r}
df_model <- read_csv(here("model", "outputs", "food_inv-intimacy_desire_alt", "cv_preds_summary.csv")) |>
  rename(motivation = reward_condition) |>
  mutate(
    action = as.character(action),
    motivation = factor(motivation, levels = c("low", "high"), labels = MOTIVATION_LEVELS),
    # Belief update from uniform prior (mean = 50), scaled to proportion
    belief_update = (expected_intimacy - 50) / 100,
    model = case_when(
      model == "full" ~ "Full model",
      model == "discomfort_only" ~ "Discomfort-only",
      model == "base" ~ "Base model"
    )
  ) |>
  filter(!is.na(model))

MODEL_LEVELS <- c("Base model", "Discomfort-only", "Full model", "Humans")
```

### Combine human and model data

```{r}
df_combined <- bind_rows(
  df_belief_update_summary |>
    rescale_belief_update() |>
    select(motivation, action, model, belief_update, ci_lower, ci_upper),
  df_model |>
    group_by(motivation, action, model) |>
    summarize(belief_update = mean(belief_update), .groups = "drop") |>
    mutate(ci_lower = NA_real_, ci_upper = NA_real_)
) |>
  mutate(
    model = factor(model, levels = MODEL_LEVELS),
    action = factor(action)
  )
```

### Belief updates: Model vs human comparison

```{r}
#| fig-width: 12
#| fig-asp: 0.3

df_combined |>
  ggplot(aes(x = action, y = belief_update, group = interaction(model, motivation), fill = motivation)) +
  geom_crossbar(
    aes(y = 0, ymin = -0.01, ymax = 0.01),
    position = position_dodge(width = 0.9),
    width = 0.9,
    color = NA
  ) +
  geom_col(position = position_dodge(width = 0.9), alpha = PLOT_ALPHA) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.9), size = 1, width = 0.1, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_motivation() +
  labs(y = "Belief update", x = "Observed action", fill = "Desire") +
  theme(legend.position = "right") +
  facet_wrap(~model, nrow = 1)

ggsave(here("figures", "food_inv-intimacy_desire_alt_belief_main.pdf"), device = cairo_pdf)
```

### Belief updates vs LM-derived risk (continuous x)

Same four panels as above, but with the x-axis replaced by the LM-derived risk score for the observed action (from `lm_scenario_params.csv`). One point per (scenario, action, motivation), with a linear smoother per motivation.

```{r}
#| fig-width: 14
#| fig-asp: 0.3

df_canonical_risk <- read_csv(here("model", "outputs", "lm", "lm_scenario_params.csv"), show_col_types = FALSE) |>
  mutate(action = as.character(action)) |>
  select(scenario_label, action, access) |>
  rename(risk = access)

df_human_scenario_pts <- df |>
  filter(stage == "posterior") |>
  group_by(scenario_label, motivation, action) |>
  tidyboot_mean(belief_update) |>
  ungroup() |>
  rescale_belief_update() |>
  mutate(model = "Humans") |>
  select(scenario_label, motivation, action, model, belief_update, ci_lower, ci_upper)

df_model_scenario_pts <- df_model |>
  select(scenario_label, motivation, action, model, belief_update) |>
  mutate(ci_lower = NA_real_, ci_upper = NA_real_)

df_risk_combined <- bind_rows(df_human_scenario_pts, df_model_scenario_pts) |>
  inner_join(df_canonical_risk, by = c("scenario_label", "action")) |>
  mutate(model = factor(model, levels = MODEL_LEVELS))

p_belief_risk <- df_risk_combined |>
  ggplot(aes(x = risk, y = belief_update, color = motivation, shape = factor(action), group = motivation)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  geom_errorbar(
    data = df_risk_combined |> filter(model == "Humans"),
    aes(ymin = ci_lower, ymax = ci_upper),
    position = POS_JITTER_DODGE,
    width = 0.05, alpha = 0.4, linewidth = 0.4
  ) +
  geom_point(position = POS_JITTER_DODGE, alpha = 0.5, size = 2) +
  geom_smooth(se = FALSE, linewidth = 1) +
  scale_color_motivation() +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(y = "Belief update", x = "LM-derived risk of observed action", color = "Desire", shape = "Action") +
  theme(legend.position = "right") +
  facet_wrap(~model, nrow = 1)

print(p_belief_risk)

ggsave(here("figures", "food_inv-intimacy_desire_alt_belief_risk.pdf"),
       plot = p_belief_risk, device = cairo_pdf, width = 14, height = 4)
```


## Scenario-level

### Human belief updates by scenario

```{r}
df_belief_update_scenario <- df |>
  filter(stage == "posterior") |>
  group_by(scenario_label, motivation, action, model) |>
  tidyboot_mean(belief_update)
```

```{r}
#| fig-width: 12 
#| fig-asp: 0.8

ggplot(df_belief_update_scenario, aes(x = action, y = empirical_stat, fill = motivation)) +
  geom_col(position = position_dodge(width = 0.9), alpha = PLOT_ALPHA) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.9), width = 0.1, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_motivation() +
  labs(
    title = "Belief updates by scenario",
    y = "Belief update",
    x = "Observed action",
    fill = "Motivation"
  ) +
  theme(legend.position = "right") +
  facet_wrap(~scenario_label, ncol = 4)
```

### Model belief updates by scenario

```{r}
#| fig-width: 14
#| fig-asp: 0.9

df_model_scenario <- df_model |>
  mutate(model = factor(model, levels = setdiff(MODEL_LEVELS, "Humans")))

ggplot(df_model_scenario, aes(x = action, y = belief_update, fill = motivation)) +
  geom_col(position = position_dodge(width = 0.9), alpha = PLOT_ALPHA) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_motivation() +
  labs(
    title = "Model-predicted belief updates by scenario",
    y = "Belief update",
    x = "Observed action",
    fill = "Motivation"
  ) +
  theme(legend.position = "right") +
  facet_grid(model ~ scenario_label,
             labeller = labeller(scenario_label = label_wrap_gen(15)))
```
