---
title: Inverse planning — intimacy inference, no alternatives shown (data analysis)
author: Alicia Chen
date: 2026-04-19
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    page-layout: full
    search: false
execute:
  warning: false
---

```{r setup}
source(here::here("analysis", "utils.R"))
setup_analysis()
```

## Demographics

```{r}
report_demographics("food_inv_intimacy_desire_noalt")
```

## Results

This experiment is a variant of `food_inv_intimacy_desire_alt` in which participants observe a single action and infer the relationship **without seeing the other three alternative actions the actors could have taken**. The experiment manipulates the same factors as the alt-shown version:

- **Action**: Which action was taken (action 0-3, varying in saliva-transfer risk)
- **Motivation (desire)**: Whether both people want the food (high vs low)

The cognitive model for this variant uses an LM-generated counterfactual alternative set (conditioned on the observed action and motivation) as the actor's action space, padded to a fixed size so the memo program can enumerate a static slot dimension. Slot 0 holds the observed canonical action; slots 1..k hold the LM-generated alternatives; remaining slots are null-padded (access=effort=v=0).

## Load human data

```{r}
# main_trials_long.csv is generated by analysis/json_to_csv.py
# It excludes participants who failed attention check or got 0 on memory check
df <- read_csv(here("data", "food_inv_intimacy_desire_noalt", "main_trials_long.csv")) |>
  mutate(
    action = str_replace(action_condition, "action_", ""),
    motivation = factor(motivation, levels = c("low", "high"), labels = MOTIVATION_LEVELS),
    stage = factor(stage, levels = c("prior", "posterior")),
    model = "Humans"
  ) |>
  select(-action_condition)

# Calculate belief updates
df <- calculate_belief_update(df, "intimacy_rating")
```

## Human data: Prior vs posterior ratings

First, look at summaries of human data

### Distribution of ratings

```{r}
ggplot(df, aes(x = intimacy_rating, fill = action)) +
  geom_density(alpha = 0.5) +
  scale_fill_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  labs(y = "Density", x = "Intimacy rating", fill = "Action") +
  theme(legend.position = "right") +
  facet_grid(stage ~ motivation, labeller = labeller(motivation = label_both))
```

Calculate summary statistics

```{r}
df_summary <- df |>
  group_by(motivation, action, stage, model) |>
  tidyboot_mean(intimacy_rating)
```

### Prior vs posterior by action and motivation

```{r}
#| fig-width: 12
#| fig-asp: 0.4

ggplot() +
  geom_violin(
    data = df,
    aes(x = stage, y = intimacy_rating, fill = action),
    position = position_dodge(width = 0.8),
    alpha = 0.5,
    width = 0.7
  ) +
  geom_point(
    data = df_summary,
    aes(x = stage, y = empirical_stat, color = action, group = action),
    position = position_dodge(width = 0.8),
    size = 3,
    alpha = PLOT_ALPHA
  ) +
  geom_errorbar(
    data = df_summary,
    aes(x = stage, ymin = ci_lower, ymax = ci_upper, color = action, group = action),
    position = position_dodge(width = 0.8),
    width = 0.1,
    size = 1
  ) +
  geom_line(
    data = df_summary,
    aes(x = stage, y = empirical_stat, color = action, group = action),
    position = position_dodge(width = 0.8),
    size = 1
  ) +
  facet_grid(motivation ~ action, labeller = labeller(action = label_both)) +
  scale_fill_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  scale_color_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  labs(
    title = "Intimacy ratings before and after observing action (no-alt)",
    y = "Intimacy rating",
    x = "Stage",
    fill = "Observed action",
    color = "Observed action"
  )
```

### Belief updates

```{r}
df_belief_update_summary <- df |>
  filter(stage == "posterior") |>
  group_by(motivation, action, model) |>
  tidyboot_mean(belief_update)
```

## Model comparison

### Model description

We compare three ablations of the canonical risk utility (Base model, Discomfort-only, Full model). The action space for each (scenario, observed action, motivation) cell is the union of the observed canonical action and an LM-generated set of plausible counterfactual alternatives. Risk, effort, and $V(a|s, m)$ for each alternative are all LM-scored (batched per scenario, 10 runs each).

**Full model**: Inverts the full risk utility $U(a|s, I) = w_v \cdot V(a|s) - w_d \cdot \text{risk}(a) \cdot (1 - I) - w_e \cdot \text{effort}(a)$.

**Discomfort-only**: Drops the food-reward and effort terms, keeping only the risk-discomfort term $-w_d \cdot \text{risk}(a) \cdot (1 - I)$.

**Base model**: Drops the risk term: $U = w_v \cdot V(a|s) - w_e \cdot \text{effort}(a)$.

The softmax over actions is $P(a | I, s) \propto \pi(a|s, \text{mot}) \cdot \exp(U(a|I, s))$, where $\pi$ is uniform over the valid slots in each padded cell. All models assume a uniform intimacy prior and compute posteriors via Bayesian inference over the padded action space.

**Fitting strategy for the no-alt experiment.** Unlike the alt-shown experiments (which freeze the Exp 1 forward-plan actor weights and fit only $\alpha_{\text{observer}}$), the no-alt experiment refits **all** actor weights jointly with $\alpha_{\text{observer}}$ on the no-alt observer data. The padded observer reasons over a variable-length LM-generated action set, so the softmax competition structure differs from Exp 1's fixed four-action space — the Exp 1 weights don't transplant cleanly. Per variant the free parameters are: Full = $\{w_v, w_d, w_e, \alpha_{\text{observer}}\}$, Discomfort-only = $\{w_d, \alpha_{\text{observer}}\}$, Base model = $\{w_v, w_e, \alpha_{\text{observer}}\}$. Actor's softmax temperature $\alpha$ is fixed at 1 for identifiability. Predictions reported below are out-of-sample (leave-one-scenario-out) from `food_inv_intimacy_desire_noalt/cv_preds_summary.csv`.

### Fitted parameters

```{r}
df_fit_results <- read_csv(here("model", "outputs", "food_inv_intimacy_desire_noalt", "fit_results.csv")) |>
  mutate(
    utility = str_remove(model, "_padded$"),
    model_label = case_when(
      utility == "full" ~ "Full model",
      utility == "discomfort_only" ~ "Discomfort-only",
      utility == "base" ~ "Base model"
    )
  ) |>
  select(model_label, param_w_v, param_w_d, param_w_e, param_gamma, alpha_observer, nll, n_params) |>
  arrange(model_label)

df_fit_results |>
  knitr::kable(
    digits = 3,
    caption = "Joint-fit parameters for no-alt intimacy inference. All actor weights fit on no-alt data (not frozen from Exp 1).",
    col.names = c("Utility", "w_v", "w_d", "w_e", "γ", "α_observer", "NLL", "n_params")
  )
```

### AIC comparison

Following Burnham and Anderson (2002), ΔAIC values of 0–2 indicate essentially equivalent models, 4–7 indicate considerably less support for the higher-AIC model, and values greater than 10 indicate essentially no support for the higher-AIC model.

```{r}
kable_aic_table(
  df_fit_results,
  caption = "Model comparison (3 utility variants; lower AIC is better)",
  model_col_label = "Utility"
)
```

### Load model predictions

```{r}
df_model <- read_csv(here("model", "outputs", "food_inv_intimacy_desire_noalt", "cv_preds_summary.csv")) |>
  mutate(
    action = as.character(observed_action),
    motivation = factor(motivation, levels = c("low", "high"), labels = MOTIVATION_LEVELS),
    belief_update = (expected_intimacy - 50) / 100,
    utility = str_remove(model, "_padded$"),
    model_label = case_when(
      utility == "full" ~ "Full model",
      utility == "discomfort_only" ~ "Discomfort-only",
      utility == "base" ~ "Base model"
    )
  )
```

### Combine human and model data

```{r}
df_human_long <- df_belief_update_summary |>
  rescale_belief_update() |>
  mutate(model_label = "Humans") |>
  select(motivation, action, model_label, belief_update, ci_lower, ci_upper)

df_model_long <- df_model |>
  group_by(motivation, action, model_label) |>
  summarize(belief_update = mean(belief_update), .groups = "drop") |>
  mutate(ci_lower = NA_real_, ci_upper = NA_real_)
```

### Belief updates: Model vs human comparison

```{r}
#| fig-width: 12
#| fig-asp: 0.3

df_combined <- bind_rows(df_human_long, df_model_long) |>
  mutate(
    model_label = factor(model_label, levels = c("Base model", "Discomfort-only", "Full model", "Humans")),
    action = factor(action)
  )

p_belief <- df_combined |>
  ggplot(aes(x = action, y = belief_update, group = interaction(model_label, motivation), fill = motivation)) +
  geom_crossbar(
    aes(y = 0, ymin = -0.01, ymax = 0.01),
    position = position_dodge(width = 0.9),
    width = 0.9,
    color = NA
  ) +
  geom_col(position = position_dodge(width = 0.9), alpha = PLOT_ALPHA) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.9), size = 1, width = 0.1, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_motivation() +
  labs(y = "Belief update", x = "Observed action", fill = "Desire") +
  theme(legend.position = "right") +
  facet_wrap(~model_label, nrow = 1)

print(p_belief)

ggsave(here("figures", "food_inv_intimacy_desire_noalt_belief_main.pdf"), plot = p_belief, device = cairo_pdf)
```

### Belief updates vs LM-derived risk (continuous x)

Same four panels as above, but with the x-axis replaced by the LM-derived risk score for the observed canonical action (from `lm_scenario_params.csv`, the same source the model uses for slot 0 of its padded action space). One point per (scenario, action, motivation), with a linear smoother per motivation.

```{r}
#| fig-width: 14
#| fig-asp: 0.3

df_canonical_risk <- read_csv(here("model", "outputs", "lm", "lm_scenario_params.csv"), show_col_types = FALSE) |>
  mutate(action = as.character(action)) |>
  select(scenario_label, action, access) |>
  rename(risk = access)

df_human_scenario_pts <- df |>
  filter(stage == "posterior") |>
  group_by(scenario_label, motivation, action) |>
  tidyboot_mean(belief_update) |>
  ungroup() |>
  rescale_belief_update() |>
  mutate(model_label = "Humans") |>
  select(scenario_label, motivation, action, model_label, belief_update, ci_lower, ci_upper)

df_model_scenario_pts <- df_model |>
  select(scenario_label, motivation, action, model_label, belief_update) |>
  mutate(ci_lower = NA_real_, ci_upper = NA_real_)

df_risk_combined <- bind_rows(df_human_scenario_pts, df_model_scenario_pts) |>
  inner_join(df_canonical_risk, by = c("scenario_label", "action")) |>
  mutate(model_label = factor(model_label, levels = c("Base model", "Discomfort-only", "Full model", "Humans")))

p_belief_risk <- df_risk_combined |>
  ggplot(aes(x = risk, y = belief_update, color = motivation, shape = factor(action), group = motivation)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  geom_errorbar(
    data = df_risk_combined |> filter(model_label == "Humans"),
    aes(ymin = ci_lower, ymax = ci_upper),
    position = POS_JITTER_DODGE,
    width = 0.05, alpha = 0.4, linewidth = 0.4
  ) +
  geom_point(position = POS_JITTER_DODGE, alpha = 0.5, size = 2) +
  geom_smooth(se = FALSE, linewidth = 1) +
  scale_color_motivation() +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(y = "Belief update", x = "LM-derived risk of observed action", color = "Desire", shape = "Action") +
  theme(legend.position = "right") +
  facet_wrap(~model_label, nrow = 1)

print(p_belief_risk)

ggsave(here("figures", "food_inv_intimacy_desire_noalt_belief_risk.pdf"),
       plot = p_belief_risk, device = cairo_pdf, width = 14, height = 4)
```


## Scenario-level

### Human belief updates by scenario

```{r}
df_belief_update_scenario <- df |>
  filter(stage == "posterior") |>
  group_by(scenario_label, motivation, action, model) |>
  tidyboot_mean(belief_update)
```

```{r}
#| fig-width: 12
#| fig-asp: 0.8

ggplot(df_belief_update_scenario, aes(x = action, y = empirical_stat, fill = motivation)) +
  geom_col(position = position_dodge(width = 0.9), alpha = PLOT_ALPHA) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.9), width = 0.1, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_motivation() +
  labs(
    title = "Belief updates by scenario (no-alt humans)",
    y = "Belief update",
    x = "Observed action",
    fill = "Motivation"
  ) +
  theme(legend.position = "right") +
  facet_wrap(~scenario_label, ncol = 4)
```

### Model belief updates by scenario

```{r}
#| fig-width: 14
#| fig-asp: 0.9

df_model_scenario <- df_model |>
  mutate(model_label = factor(model_label, levels = c("Base model", "Discomfort-only", "Full model")))

ggplot(df_model_scenario, aes(x = action, y = belief_update, fill = motivation)) +
  geom_col(position = position_dodge(width = 0.9), alpha = PLOT_ALPHA) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_motivation() +
  labs(
    title = "Model-predicted belief updates by scenario",
    y = "Belief update",
    x = "Observed action",
    fill = "Motivation"
  ) +
  theme(legend.position = "right") +
  facet_grid(model_label ~ scenario_label,
             labeller = labeller(scenario_label = label_wrap_gen(15)))
```
