---
title: "Study 1a — Desire inference"
format:
  html:
    code-fold: true
    toc: true
    embed-resources: true
execute:
  echo: true
  warning: false
---

```{r setup, include=FALSE}
source(here::here("analysis", "utils.R"))
setup_analysis()
```

# Study 1a — Desire inference

## Data prep

```{r load-data}
data_path <- here::here("data", "food_inv_desire", "main_trials_long.csv")
if (file.exists(data_path)) {
  df <- read_csv(data_path, show_col_types = FALSE)
  cat("Trials:", nrow(df), "\n")
} else {
  cat("No data yet for food_inv_desire (active continuous 0–100 DV).\n")
}
```

```{r belief-update-prep}
if (exists("df")) {
  bu <- calculate_belief_update(df, "response") |>
    filter(stage == "posterior") |>
    mutate(
      action = factor_action(action_condition),
      intimacy = factor(intimacy, levels = INTIMACY_LEVELS),
      effort = factor_effort(effort)
    )

  agg <- bu |>
    group_by(action, intimacy, effort) |>
    tidyboot_mean(column = belief_update) |>
    ungroup()
}
```

```{r cv-load}
cv_path <- here::here("model", "outputs", "food_inv_desire", "cv_preds_summary.json")
if (file.exists(cv_path) && exists("agg")) {
  cv_preds <- read_model_json(cv_path) |>
    mutate(
      action = factor(ACTION_LABELS[action + 1], levels = ACTION_LABELS),
      intimacy = factor(intimacy_condition, levels = INTIMACY_LEVELS),
      effort = factor_effort(effort_condition),
      model_label = factor(MODEL_LABELS[model], levels = MODEL_LABELS)
    )
  cv_model_agg <- cv_preds |>
    group_by(model_label, action, intimacy, effort) |>
    summarise(empirical_stat = mean(delta_desire), .groups = "drop")
  cat("Loaded CV predictions for", nrow(cv_preds), "cells.\n")
} else {
  cat("No CV predictions / no human aggregate — run the CV and the belief-update chunk.\n")
}
```

## Out-of-sample predictions by condition

```{r cv-overlay-bars}
#| fig-width: 13
#| fig-height: 5

if (exists("cv_model_agg")) {
  library(ggpattern)
  combined <- bind_rows(
    cv_model_agg |>
      transmute(panel = as.character(model_label), action, intimacy, effort,
                empirical_stat, ci_lower = NA_real_, ci_upper = NA_real_),
    agg |>
      transmute(panel = "Humans", action, intimacy, effort,
                empirical_stat, ci_lower, ci_upper)
  ) |>
    mutate(panel = factor(panel, levels = PANEL_LEVELS)) |>
    # the discomfort-only model predicts ~0 belief update everywhere; flag those
    # cells so they can be drawn as a single small bar centered on 0 (a tile that
    # straddles the axis) instead of an invisible zero-height bar
    mutate(is_zero_stub = panel == "Discomfort-only" & abs(empirical_stat) < 0.012)

  pd_bar <- position_dodge(width = 0.9)
  p_bar <- ggplot(combined, aes(action, empirical_stat,
                                group = interaction(intimacy, effort))) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey70", linewidth=1) +
    geom_col_pattern(
      data = function(d) dplyr::filter(d, !is_zero_stub),
      aes(fill = intimacy, pattern = effort),
      position = pd_bar, width = 0.85, linewidth = 0.2,
      pattern_fill = "white", pattern_colour = NA,
      pattern_density = 0.25, pattern_spacing = 0.02, pattern_angle = 45
    ) +
    geom_tile_pattern(
      data = function(d) dplyr::filter(d, is_zero_stub),
      aes(y = 0, fill = intimacy, pattern = effort),
      height = 0.015, position = pd_bar, width = 0.85,
      linewidth = 0.2,
      pattern_fill = "white", pattern_colour = NA,
      pattern_density = 0.25, pattern_spacing = 0.02, pattern_angle = 45,
      show.legend = FALSE
    ) +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2,
                  position = pd_bar, color = "grey10", na.rm = TRUE) +
    facet_wrap(~ panel, nrow = 1) +
    scale_fill_intimacy() +
    scale_pattern_effort() +
    scale_x_action() +
    labs(x = "Observed action", y = "Belief update") +
    guides(
      fill = guide_legend(nrow = 2, override.aes = list(pattern = "none")),
      pattern = guide_legend(nrow = 2, override.aes = list(fill = "grey10"))
    ) +
    theme(legend.position = "bottom")
  print(p_bar)
  save_figure(p_bar, "food_inv_desire_cv_overlay_bars.pdf")
} else {
  cat("No CV predictions loaded.\n")
}
```

### Out-of-sample cell-level scatter

Each point is one action × intimacy × effort cell: the model's out-of-sample
predicted belief update (LOSO-CV) against the human mean. The dashed line is
y = x, and the per-panel r is the cell-level correlation (the discomfort-only
model predicts a constant ~0, so its correlation is undefined).

```{r cv-scatter}
#| fig-width: 12
#| fig-height: 4.5
if (exists("cv_model_agg")) {
  human_cells <- agg |> select(action, intimacy, effort, human = empirical_stat)
  scatter_df <- cv_model_agg |>
    rename(model = empirical_stat) |>
    left_join(human_cells, by = c("action", "intimacy", "effort"))
  rlab <- scatter_df |>
    group_by(model_label) |>
    summarise(r = if (sd(model) < 1e-3) NA_real_ else cor(model, human),
              .groups = "drop") |>
    mutate(panel_lab = sprintf("%s (r = %s)", model_label,
                               ifelse(is.na(r), "n/a", sprintf("%.2f", r))))
  panel_levels <- rlab |> arrange(model_label) |> pull(panel_lab)
  scatter_df <- scatter_df |>
    left_join(select(rlab, model_label, panel_lab), by = "model_label") |>
    mutate(panel_lab = factor(panel_lab, levels = panel_levels))
  lim <- range(c(scatter_df$human, scatter_df$model), na.rm = TRUE)
  p_scatter <- ggplot(scatter_df, aes(model, human, color = intimacy, shape = effort)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey70") +
    geom_point(size = 2.6) +
    facet_wrap(~ panel_lab, nrow = 1) +
    scale_color_intimacy() +
    coord_equal(xlim = lim, ylim = lim) +
    labs(x = "Model predicted belief update (out-of-sample)",
         y = "Human belief update", shape = "Effort of low-risk share")
  print(p_scatter)
} else {
  cat("No CV predictions loaded.\n")
}
```


