---
title: "Study 4a — Joint inference over desire and effort"
format:
  html:
    code-fold: true
    toc: true
    embed-resources: true
execute:
  echo: true
  warning: false
---

```{r setup, include=FALSE}
source(here::here("analysis", "utils.R"))
setup_analysis()
```

# Study 4a — Joint inference over desire and effort

Observer jointly infers desire and effort from a single observed action, given intimacy. Each trial has two slider responses.

**Status:** Scaffolding stub. The plots below come online once pilot data lands in
`data/food_inv_joint_de_3act/`. Until then, this qmd is structure-only.

## Demographics

```{r demographics}
if (file.exists(here::here("data", "food_inv_joint_de_3act", "exit_survey.csv"))) {
  report_demographics("food_inv_joint_de_3act")
} else {
  cat("No data yet for food_inv_joint_de_3act.\n")
}
```

## Data

```{r load-data}
data_path <- here::here("data", "food_inv_joint_de_3act", "main_trials_long.csv")
if (file.exists(data_path)) {
  df <- read_csv(data_path, show_col_types = FALSE)
  cat("Trials:", nrow(df), "\n")
  glimpse(df)
} else {
  cat("No data yet for food_inv_joint_de_3act.\n")
}
```

## Belief update

```{r belief-update, eval=FALSE}
# Compute belief update (posterior - prior) per (subject, scenario)
# Uses calculate_belief_update() from utils.R with the appropriate rating column.
# TODO: fill in the right rating column once data lands. Currently expected: p_high_reward_rating + p_effort_high_rating
```

## Model predictions

```{r preds, eval=FALSE}
preds_path <- here::here("model", "outputs", "food_inv_joint_de_3act", "cv_preds_summary.csv")
if (file.exists(preds_path)) {
  preds <- read_csv(preds_path, show_col_types = FALSE)
  glimpse(preds)
} else {
  cat("No CV predictions yet — run `make cv-food_inv_joint_de_3act` once a forward fit + data exist.\n")
}
```

## Model-vs-human correlations

```{r model-vs-human, eval=FALSE}
# TODO: wire up once both data and model predictions exist.
# Standard pattern (cf. food-inv-intimacy-desire-alt-analysis.qmd):
#   1. Aggregate human belief updates by cell.
#   2. Aggregate model predictions by the same cell.
#   3. Plot scatter + report bootstrap Pearson via format_correlation_labels().
```

## Notes

See the manuscript Methods section for the exact design, plus
`experiments/food_inv_joint_de_3act/README.md` for the trial structure spec.
