---
title: "Study 1a — Desire inference"
format:
  html:
    code-fold: true
    toc: true
    embed-resources: true
execute:
  echo: true
  warning: false
---

```{r setup, include=FALSE}
source(here::here("analysis", "utils.R"))
setup_analysis()
```

# Study 1a — Desire inference

Observer infers the actors' desire from a single observed action, given known effort and intimacy. Design: 2 (effort: low / high) × 4 (intimacy: 0 / 50 / 75 / 100) × 3 (observed action). The desire DV is a continuous rating ("how much would they like to eat the food?", from "not at all" to "extremely"), collected on a 0–100 slider and stored on the 0–1 scale. The model's observer reasons over an LM-generated alternative-action set per cell (see `model/lm/generate_alternatives.py`).

**Status:** Scaffolding stub. The original pilot — which used a 0–100
"probability of two states" desire slider — is archived at
`analysis/legacy/food-inv-desire-pilot-analysis.qmd` (data in
`data/legacy/food_inv_desire_pilot/`). The plots below come online once data on the
corrected direct 0–100 desire DV lands in `data/food_inv_desire/`. Until then, this
qmd is structure-only.

## Demographics

```{r demographics}
if (file.exists(here::here("data", "food_inv_desire", "exit_survey.csv"))) {
  report_demographics("food_inv_desire")
} else {
  cat("No data yet for food_inv_desire (active continuous 0–100 DV).\n")
}
```

## Data

```{r load-data}
data_path <- here::here("data", "food_inv_desire", "main_trials_long.csv")
if (file.exists(data_path)) {
  df <- read_csv(data_path, show_col_types = FALSE)
  cat("Trials:", nrow(df), "\n")
  glimpse(df)
} else {
  cat("No data yet for food_inv_desire (active continuous 0–100 DV).\n")
}
```

## Belief update

The desire DV is a continuous rating on the 0–1 scale; the belief update is
posterior − prior on that scale (so it falls in [−1, 1]), via
`calculate_belief_update()`, aggregated by observed action × intimacy × effort
with a bootstrap 95% CI.

```{r belief-update}
#| fig-width: 11
#| fig-height: 5.5
if (exists("df")) {
  bu <- calculate_belief_update(df, "response") |>
    filter(stage == "posterior") |>
    mutate(
      action = factor(action_condition,
                      levels = c("no_share", "low_risk_share", "high_risk_share"),
                      labels = c("No share", "Low-risk share", "High-risk share")),
      intimacy = factor(intimacy, levels = INTIMACY_LEVELS),
      effort = factor(effort, levels = c("low", "high"),
                      labels = c("Low effort", "High effort"))
    )

  agg <- bu |>
    group_by(action, intimacy, effort) |>
    tidyboot_mean(column = belief_update) |>
    ungroup()

  pd <- position_dodge(width = 0.3)
  p <- ggplot(agg, aes(action, empirical_stat, color = intimacy, linetype = effort,
                       group = interaction(intimacy, effort))) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey70") +
    geom_line(position = pd, linewidth = 0.8) +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0,
                  position = pd, alpha = 0.7) +
    geom_point(position = pd, size = 2.4) +
    scale_color_intimacy() +
    labs(
      title = "Desire belief update by action, intimacy, and effort",
      x = "Observed action", y = "Belief update",
      color = "Intimacy", linetype = "Effort of low-risk share"
    )
  print(p)
} else {
  cat("No data yet for food_inv_desire.\n")
}
```

### Bar-plot version

Same aggregates as above, drawn as bars so the magnitude and sign of the update
from 0 are easier to read. Effort stays on the same panel: within each observed
action, bars are grouped into effort blocks — a solid low-effort block of four
intimacy bars, then a striped high-effort block — so the four intimacy bars sit
next to each other within each effort category. Colour reads intimacy and the
stripe pattern reads effort (via `ggpattern`).

```{r belief-update-bars}
#| fig-width: 11
#| fig-height: 5.5
if (exists("agg")) {
  library(ggpattern)
  pd_bar <- position_dodge(width = 0.9)
  p_bar <- ggplot(agg, aes(action, empirical_stat,
                           group = interaction(intimacy, effort))) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey70") +
    geom_col_pattern(
      aes(fill = intimacy, pattern = effort),
      position = pd_bar, width = 0.85, color = "grey30", linewidth = 0.2,
      pattern_fill = "white", pattern_colour = NA,
      pattern_density = 0.25, pattern_spacing = 0.025, pattern_angle = 45
    ) +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2,
                  position = pd_bar, alpha = 0.7) +
    scale_fill_intimacy() +
    scale_pattern_manual(values = c("Low effort" = "none", "High effort" = "stripe"),
                         breaks = c("Low effort", "High effort")) +
    labs(
      title = "Desire belief update by action, intimacy, and effort",
      x = "Observed action", y = "Belief update",
      fill = "Intimacy", pattern = "Effort of low-risk share"
    ) +
    guides(
      fill = guide_legend(override.aes = list(pattern = "none")),
      pattern = guide_legend(override.aes = list(fill = "grey70"))
    )
  print(p_bar)
} else {
  cat("No data yet for food_inv_desire.\n")
}
```

## Out-of-sample model comparison (LOSO-CV)

The confirmatory analysis. Predictions come from leave-one-scenario-out CV
(`cv_food_inv_desire.py`): for each held-out scenario the weights are refit on the
other 15, then used to predict it — so every model-vs-human number here is
genuinely out-of-sample (CV is the sole prediction source). The **primary** metric
is the held-out per-trial log-likelihood; the cell-mean correlation is
secondary/descriptive.

### Held-out log-likelihood (primary)

```{r cv-heldout-ll}
#| fig-width: 7
#| fig-height: 4.5
MODEL_LABELS <- c(base = "Base", discomfort_only = "Discomfort-only", full = "Full")
ll_path <- here::here("model", "outputs", "food_inv_desire", "cv_trial_ll.jsonl")
if (file.exists(ll_path)) {
  ll <- read_model_jsonl(ll_path)
  summ <- ll |>
    group_by(model) |>
    summarise(mean_ll = mean(held_out_ll), total_nll = -sum(held_out_ll),
              n = n(), .groups = "drop") |>
    mutate(model = factor(MODEL_LABELS[model], levels = MODEL_LABELS)) |>
    arrange(model)

  # subject cluster-bootstrap of the paired per-trial LL differences
  wide <- ll |>
    select(subject_id, scenario_label, model, held_out_ll) |>
    pivot_wider(names_from = model, values_from = held_out_ll) |>
    mutate(d_fb = full - base, d_fd = full - discomfort_only)
  boot_ci <- function(col, nboot = 4000) {
    bs <- split(wide[[col]], wide$subject_id); subs <- names(bs); set.seed(1)
    m <- replicate(nboot, mean(unlist(bs[sample(subs, length(subs), replace = TRUE)])))
    list(point = mean(wide[[col]]), lo = unname(quantile(m, .025)),
         hi = unname(quantile(m, .975)), pgt = mean(m > 0))
  }
  fb <- boot_ci("d_fb"); fd <- boot_ci("d_fd")

  print(summ)
  cat(sprintf("\nfull − base:            ΔLL/trial = %+.4f  95%% CI [%+.4f, %+.4f]  P(Δ>0)=%.3f\n",
              fb$point, fb$lo, fb$hi, fb$pgt))
  cat(sprintf("full − discomfort_only: ΔLL/trial = %+.4f  95%% CI [%+.4f, %+.4f]\n",
              fd$point, fd$lo, fd$hi))

  p <- ggplot(summ, aes(model, mean_ll, fill = model)) +
    geom_col(width = 0.6) +
    geom_text(aes(label = sprintf("%.3f", mean_ll)), vjust = 1.4, color = "white", size = 4) +
    labs(
      title = "Out-of-sample mean held-out log-likelihood per trial (higher = better)",
      subtitle = sprintf("full − base ΔLL/trial = %+.4f  [%+.4f, %+.4f]", fb$point, fb$lo, fb$hi),
      x = NULL, y = "Mean held-out LL / trial"
    ) +
    guides(fill = "none")
  print(p)
} else {
  cat("No CV results yet — run `make cv-food_inv_desire` (or `CV_WORKERS=4 ...`).\n")
}
```

### Out-of-sample predictions by condition

```{r cv-load}
cv_path <- here::here("model", "outputs", "food_inv_desire", "cv_preds_summary.json")
ACT_LABELS <- c("No share", "Low-risk share", "High-risk share")
if (file.exists(cv_path) && exists("agg")) {
  cv_preds <- read_model_json(cv_path) |>
    mutate(
      action = factor(ACT_LABELS[action + 1], levels = ACT_LABELS),
      intimacy = factor(intimacy_condition, levels = INTIMACY_LEVELS),
      effort = factor(effort_condition, levels = c("low", "high"),
                      labels = c("Low effort", "High effort")),
      model_label = factor(MODEL_LABELS[model], levels = MODEL_LABELS)
    )
  cv_model_agg <- cv_preds |>
    group_by(model_label, action, intimacy, effort) |>
    summarise(empirical_stat = mean(delta_desire), .groups = "drop")
  cat("Loaded CV predictions for", nrow(cv_preds), "cells.\n")
} else {
  cat("No CV predictions / no human aggregate — run the CV and the belief-update chunk.\n")
}
```

```{r cv-overlay}
#| fig-width: 15
#| fig-height: 5
if (exists("cv_model_agg")) {
  PANEL_LEVELS <- c("Base", "Discomfort-only", "Full", "Human")
  combined <- bind_rows(
    cv_model_agg |>
      transmute(panel = as.character(model_label), action, intimacy, effort,
                empirical_stat, ci_lower = NA_real_, ci_upper = NA_real_),
    agg |>
      transmute(panel = "Human", action, intimacy, effort,
                empirical_stat, ci_lower, ci_upper)
  ) |>
    mutate(panel = factor(panel, levels = PANEL_LEVELS))
  pd <- position_dodge(width = 0.3)
  p <- ggplot(combined, aes(action, empirical_stat, color = intimacy,
                            linetype = effort, group = interaction(intimacy, effort))) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey70") +
    geom_line(position = pd, linewidth = 0.8) +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0,
                  position = pd, alpha = 0.7, na.rm = TRUE) +
    geom_point(position = pd, size = 2.2) +
    facet_wrap(~ panel, nrow = 1) +
    scale_color_intimacy() +
    labs(title = "Desire belief update — ablations vs human (out-of-sample LOSO-CV)",
         x = "Observed action", y = "Belief update",
         color = "Intimacy", linetype = "Effort of low-risk share") +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
  print(p)
} else {
  cat("No CV predictions loaded.\n")
}
```

Same out-of-sample panels as bars (fill = intimacy, stripe = effort).

```{r cv-overlay-bars}
#| fig-width: 16
#| fig-height: 5.5
if (exists("cv_model_agg")) {
  library(ggpattern)
  PANEL_LEVELS <- c("Base", "Discomfort-only", "Full", "Human")
  combined <- bind_rows(
    cv_model_agg |>
      transmute(panel = as.character(model_label), action, intimacy, effort,
                empirical_stat, ci_lower = NA_real_, ci_upper = NA_real_),
    agg |>
      transmute(panel = "Human", action, intimacy, effort,
                empirical_stat, ci_lower, ci_upper)
  ) |>
    mutate(panel = factor(panel, levels = PANEL_LEVELS))
  pd_bar <- position_dodge(width = 0.9)
  p_bar <- ggplot(combined, aes(action, empirical_stat,
                                group = interaction(intimacy, effort))) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey70") +
    geom_col_pattern(
      aes(fill = intimacy, pattern = effort),
      position = pd_bar, width = 0.85, color = "grey30", linewidth = 0.2,
      pattern_fill = "white", pattern_colour = NA,
      pattern_density = 0.25, pattern_spacing = 0.02, pattern_angle = 45
    ) +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2,
                  position = pd_bar, alpha = 0.7, na.rm = TRUE) +
    facet_wrap(~ panel, nrow = 1) +
    scale_fill_intimacy() +
    scale_pattern_manual(values = c("Low effort" = "none", "High effort" = "stripe"),
                         breaks = c("Low effort", "High effort")) +
    labs(
      title = "Desire belief update — ablations vs human, bars (out-of-sample LOSO-CV)",
      x = "Observed action", y = "Belief update",
      fill = "Intimacy", pattern = "Effort of low-risk share"
    ) +
    guides(
      fill = guide_legend(override.aes = list(pattern = "none")),
      pattern = guide_legend(override.aes = list(fill = "grey70"))
    ) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
  print(p_bar)
} else {
  cat("No CV predictions loaded.\n")
}
```

### Out-of-sample cell-level scatter

```{r cv-scatter}
#| fig-width: 13
#| fig-height: 5
if (exists("cv_model_agg")) {
  human_cells <- agg |> select(action, intimacy, effort, human = empirical_stat)
  scatter_df <- cv_model_agg |>
    rename(model = empirical_stat) |>
    left_join(human_cells, by = c("action", "intimacy", "effort"))
  rlab <- scatter_df |>
    group_by(model_label) |>
    summarise(r = if (sd(model) < 1e-3) NA_real_ else cor(model, human),
              .groups = "drop") |>
    mutate(panel_lab = sprintf("%s (r = %s)", model_label,
                               ifelse(is.na(r), "n/a", sprintf("%.2f", r))))
  panel_levels <- rlab |> arrange(model_label) |> pull(panel_lab)
  scatter_df <- scatter_df |>
    left_join(select(rlab, model_label, panel_lab), by = "model_label") |>
    mutate(panel_lab = factor(panel_lab, levels = panel_levels))
  lim <- range(c(scatter_df$human, scatter_df$model), na.rm = TRUE)
  p <- ggplot(scatter_df, aes(model, human, color = intimacy, shape = effort)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey70") +
    geom_point(size = 2.6) +
    facet_wrap(~ panel_lab, nrow = 1) +
    scale_color_intimacy() +
    coord_equal(xlim = lim, ylim = lim) +
    labs(title = "Out-of-sample cell-level model vs human (24 cells each)",
         subtitle = "LOSO-CV; each point = one action × intimacy × effort cell",
         x = "Model predicted belief update (CV)", y = "Human belief update")
  print(p)
} else {
  cat("No CV predictions loaded.\n")
}
```

## Manuscript figure drafts

Two candidate main-text figures (drafts).

### Prior → posterior slope plot

Open point = prior desire (before the action); filled point = posterior (after);
the arrow is the belief update. Effort is collapsed. Reads off three things at
once: the **prior rises with intimacy**, **posteriors converge by action**
(no-share pulls low, sharing pulls high), and the update is the arrow length —
including the deeper no-share drop at max intimacy.

```{r fig-slope}
#| fig-width: 8.5
#| fig-height: 5.5
if (exists("df")) {
  sl <- df |>
    mutate(
      action = factor(action_condition,
                      levels = c("no_share", "low_risk_share", "high_risk_share"),
                      labels = c("No share", "Low-risk share", "High-risk share")),
      intimacy = factor(intimacy, levels = INTIMACY_LEVELS)
    ) |>
    group_by(action, intimacy, stage) |>
    tidyboot_mean(column = response) |>
    ungroup() |>
    mutate(across(c(empirical_stat, ci_lower, ci_upper), ~ .x * 100))
  off <- tibble(
    intimacy = factor(INTIMACY_LEVELS, levels = INTIMACY_LEVELS),
    dx = seq(-0.27, 0.27, length.out = length(INTIMACY_LEVELS))
  )
  sl <- sl |> left_join(off, by = "intimacy") |>
    mutate(xpos = as.integer(action) + dx)
  wide <- sl |>
    select(action, intimacy, xpos, stage, m = empirical_stat) |>
    pivot_wider(names_from = stage, values_from = m)

  p_slope <- ggplot() +
    geom_segment(
      data = wide,
      aes(x = xpos, xend = xpos, y = prior, yend = posterior, color = intimacy),
      arrow = arrow(length = unit(0.10, "in"), type = "closed"), linewidth = 0.8
    ) +
    geom_errorbar(
      data = sl, aes(x = xpos, ymin = ci_lower, ymax = ci_upper, color = intimacy),
      width = 0, alpha = 0.45, linewidth = 0.4
    ) +
    geom_point(
      data = filter(sl, stage == "prior"),
      aes(xpos, empirical_stat, color = intimacy), shape = 21, fill = "white", size = 1.9
    ) +
    geom_point(
      data = filter(sl, stage == "posterior"),
      aes(xpos, empirical_stat, color = intimacy), size = 2.1
    ) +
    scale_color_intimacy() +
    scale_x_continuous(breaks = 1:3, labels = levels(sl$action)) +
    labs(
      title = "Prior → posterior desire by action and intimacy",
      subtitle = "Open = prior, filled = posterior, arrow = belief update (effort collapsed)",
      x = "Observed action", y = "Desire rating (0–100)", color = "Intimacy"
    )
  print(p_slope)
} else {
  cat("No data loaded.\n")
}
```

### Held-out ΔLL interval plot (confirmatory)

Out-of-sample model improvement: Δ held-out log-likelihood per trial, with the
subject cluster-bootstrap 95% CI; > 0 favors the full model. Free x-scales so the
small (but reliably positive) full − base contrast is legible next to the large
full − discomfort-only floor.

```{r fig-dll}
#| fig-width: 8
#| fig-height: 3.2
ll_path <- here::here("model", "outputs", "food_inv_desire", "cv_trial_ll.jsonl")
if (file.exists(ll_path)) {
  llw <- read_model_jsonl(ll_path) |>
    select(subject_id, scenario_label, model, held_out_ll) |>
    pivot_wider(names_from = model, values_from = held_out_ll) |>
    mutate(d_fb = full - base, d_fd = full - discomfort_only)
  boot_ci <- function(col, nboot = 4000) {
    bs <- split(llw[[col]], llw$subject_id); subs <- names(bs); set.seed(1)
    m <- replicate(nboot, mean(unlist(bs[sample(subs, length(subs), replace = TRUE)])))
    tibble(point = mean(llw[[col]]), lo = unname(quantile(m, .025)),
           hi = unname(quantile(m, .975)))
  }
  comps <- bind_rows(
    boot_ci("d_fb") |> mutate(comparison = "Full − Base"),
    boot_ci("d_fd") |> mutate(comparison = "Full − Discomfort-only")
  ) |>
    mutate(comparison = factor(comparison, levels = c("Full − Base", "Full − Discomfort-only")))

  p_dll <- ggplot(comps, aes(point, y = 0)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
    geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0.25, linewidth = 0.7) +
    geom_point(size = 3) +
    geom_text(aes(label = sprintf("Δ = %.3f  [%.3f, %.3f]", point, lo, hi)),
              vjust = -1.6, size = 3.3) +
    facet_wrap(~ comparison, scales = "free_x", ncol = 1) +
    scale_y_continuous(limits = c(-1, 1), breaks = NULL) +
    labs(
      title = "Out-of-sample model improvement (LOSO-CV)",
      subtitle = "Δ held-out log-likelihood / trial; > 0 favors full. Bars = subject-bootstrap 95% CI.",
      x = "Δ held-out LL / trial", y = NULL
    )
  print(p_dll)
} else {
  cat("No CV results — run the CV first.\n")
}
```

## Notes

See the manuscript Methods section for the exact design, plus
`experiments/food_inv_desire/README.md` for the trial structure spec.
