---
title: "Study 1b — Joint inference over desire and effort"
format:
  html:
    code-fold: true
    toc: true
    embed-resources: true
execute:
  echo: true
  warning: false
---

```{r setup, include=FALSE}
source(here::here("analysis", "utils.R"))
setup_analysis()
```

# Study 1b — Joint inference over desire and effort

Observer jointly infers desire and effort from a single observed action, given intimacy. Each trial has two slider responses.

**Status:** Data are in `data/food_inv_joint_de/`; the demographics and
belief-update panels below run on them. The model-prediction and
model-vs-human sections are still disabled (`eval=FALSE`) pending CV
predictions.

## Demographics

```{r demographics}
if (file.exists(here::here("data", "food_inv_joint_de", "exit_survey.csv"))) {
  report_demographics("food_inv_joint_de")
} else {
  cat("No data yet for food_inv_joint_de.\n")
}
```

## Data

```{r load-data}
data_path <- here::here("data", "food_inv_joint_de", "main_trials_long.csv")
have_data <- file.exists(data_path)
if (have_data) {
  df <- read_csv(data_path, show_col_types = FALSE)
  cat("Trials:", nrow(df), "\n")
  glimpse(df)
} else {
  cat("No data yet for food_inv_joint_de.\n")
}
```

## Belief update

Each trial yields two belief updates (posterior − prior): desire on the direct
rating scale and effort on the paragraph-endpoint scale, where higher values
mean the low-risk share option takes more effort. Both ratings are collected on
0–100 sliders and stored on the 0–1 scale, so updates fall in [−1, 1]. The
joint plot below shows the mean update per observed action × given intimacy
cell, with crosshairs giving 95% cluster-bootstrap CIs (resampling subjects) on
each axis.

```{r belief-update}
#| fig-width: 11
#| fig-height: 7
if (have_data) {
  bu <- df |>
    calculate_belief_update("desire_rating") |>
    rename(desire_update = belief_update) |>
    calculate_belief_update("effort_rating") |>
    rename(effort_update = belief_update) |>
    filter(stage == "posterior") |>
    mutate(
      action = factor_action(action_condition),
      intimacy = factor(intimacy, levels = INTIMACY_LEVELS)
    )

  agg <- boot_cluster_means(bu, c("desire_update", "effort_update"),
                            c("action", "intimacy"))

  # Symmetric, origin-centered, equal x/y limits (a vector plot needs equal units
  # on both axes), including the crosshair CI extents.
  lim <- max(abs(c(agg$desire_update, agg$effort_update,
                   agg$desire_update_ci_lower, agg$desire_update_ci_upper,
                   agg$effort_update_ci_lower, agg$effort_update_ci_upper)),
             na.rm = TRUE) * 1.08

  p <- ggplot(agg, aes(desire_update, effort_update,
                       color = intimacy, shape = action)) +
    geom_hline(yintercept = 0, linetype = "dotted", color = "grey70") +
    geom_vline(xintercept = 0, linetype = "dotted", color = "grey70") +
    geom_segment(aes(xend = desire_update, yend = effort_update), x = 0, y = 0,
                 alpha = 0.6, linewidth = 0.5) +
    geom_errorbar(aes(ymin = effort_update_ci_lower, ymax = effort_update_ci_upper),
                  width = 0, alpha = 0.7) +
    geom_errorbar(aes(xmin = desire_update_ci_lower, xmax = desire_update_ci_upper),
                  width = 0, alpha = 0.7, orientation = "y") +
    geom_point(size = 3) +
    scale_color_intimacy() +
    scale_shape_manual(values = ACTION_SHAPES) +
    coord_fixed() +
    labs(
      title = "Joint belief update by action and given intimacy",
      subtitle = "Color = given intimacy, shape = observed action; crosshairs 95% cluster-bootstrap CIs",
      x = "Desire belief update",
      y = "Effort belief update (low-risk share)",
      color = "Intimacy", shape = "Observed action"
    )
  print(p)
} else {
  cat("No data yet for food_inv_joint_de.\n")
}
```

The same cell means split by DV: one panel per inferred variable, with lines
tracking each given-intimacy level across observed actions.

```{r belief-update-by-dv}
#| fig-width: 12
#| fig-height: 5
if (have_data) {
  agg_long <- agg |>
    rename_with(\(x) paste0(x, "_mean"), c(desire_update, effort_update)) |>
    pivot_longer(
      -c(action, intimacy),
      names_to = c("measure", ".value"),
      names_pattern = "(desire|effort)_update_(mean|ci_lower|ci_upper)"
    ) |>
    mutate(measure = factor(measure, levels = c("desire", "effort"),
                            labels = c("Desire", "Effort of low-risk share")))

  pd <- position_dodge(width = 0.25)
  p <- ggplot(agg_long, aes(action, mean, color = intimacy, group = intimacy)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey70") +
    geom_line(position = pd, linewidth = 0.8) +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.15,
                  position = pd, alpha = 0.8) +
    geom_point(position = pd, size = 2.6) +
    facet_wrap(~ measure) +
    scale_color_intimacy() +
    scale_x_discrete(labels = \(x) sub(" ", "\n", x)) +
    labs(
      title = "Belief updates by action and given intimacy",
      x = "Observed action", y = "Belief update",
      color = "Intimacy"
    )
  print(p)
} else {
  cat("No data yet for food_inv_joint_de.\n")
}
```

## Model predictions

```{r preds, eval=FALSE}
preds_path <- here::here("model", "outputs", "food_inv_joint_de", "cv_preds_summary.json")
if (file.exists(preds_path)) {
  preds <- read_model_json(preds_path)
  glimpse(preds)
} else {
  cat("No CV predictions yet — run `make cv-food_inv_joint_de` once a fit + data exist.\n")
}
```

## Model-vs-human correlations

```{r model-vs-human, eval=FALSE}
# TODO: wire up once the model predictions exist (data are in).
# Standard pattern (cf. the cv-load / cv-scatter chunks in
# food-inv-desire-analysis.qmd):
#   1. Aggregate human belief updates by cell.
#   2. Aggregate model predictions by the same cell.
#   3. Plot scatter + report bootstrap Pearson via format_correlation_labels().
```

## Notes

See the manuscript Methods section for the exact design, plus
`experiments/food_inv_joint_de/README.md` for the trial structure spec.
