---
title: "Study 1a — LM elicitation diagnostics"
format:
  html:
    code-fold: true
    toc: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
source(here::here("analysis", "utils.R"))
setup_analysis()
library(knitr)
```

# Study 1a — LM elicitation diagnostics

This notebook inspects the **LM elicitation** that feeds the Study 1a
(`food_inv_desire`) observer model — *not* the human data. The model's actor
softmaxes over `{observed action} ∪ LM-generated alternatives`, and the
reward / discomfort / effort terms are driven by LM-elicited per-action features.
The elicitation is a K-run simulated-observer mixture: each
(scenario × condition) cell is elicited `K` independent times, and the
run-to-run spread becomes part of the model's predicted distribution. This
notebook checks that the two elicitation stages produced sensible, well-covered,
internally-consistent output.

The two source files (under `model/outputs/lm/food_inv_desire/`):

- **`lm_alternatives.jsonl`** — stage 1. One record per generated counterfactual
  action (`generate_alternatives.py`): the action text plus its cell key
  (scenario, observed action, effort condition, intimacy condition) and `run_id`.
- **`lm_runs.jsonl`** — stage 2. One record per `(run, cell)`
  (`score_merged.py`), carrying that run's scored action list (slot 0 = the
  observed canonical action, slots 1..k = the run's alternatives) on risk,
  effort, and goal-satisfaction `g`, plus the run's given **intimacy** scalar for
  the cell's intimacy condition.

All three action features (risk / effort / g) are on a 0–1 scale (the absolute
scale is absorbed by the freely-fitted utility weights).

```{r load}
lm_dir <- here::here("model", "outputs", "lm", "food_inv_desire")
runs_path <- file.path(lm_dir, "lm_runs.jsonl")
alts_path <- file.path(lm_dir, "lm_alternatives.jsonl")

have_lm <- file.exists(runs_path) && file.exists(alts_path)

action_levels <- c("no_share", "low_risk_share", "high_risk_share")
action_labels <- c("No share", "Low-risk share", "High-risk share")
as_action <- function(x) factor(x, levels = action_levels, labels = action_labels)

if (have_lm) {
  # Stage 2: per-(run, cell) records with a nested `actions` list-column.
  runs_raw <- read_model_jsonl(runs_path)
  actions <- runs_raw |>
    select(run_id, scenario_label, observed_action, effort_condition,
           intimacy_condition, intimacy, actions) |>
    tidyr::unnest(actions)

  # Stage 1: flat per-alternative records.
  alts <- read_model_jsonl(alts_path)

  K <- n_distinct(runs_raw$run_id)
  n_scenarios <- n_distinct(runs_raw$scenario_label)
  # Cell grid for 1a = scenario × observed_action × effort × intimacy.
  n_cells <- n_scenarios * length(action_levels) *
    n_distinct(runs_raw$effort_condition) * n_distinct(runs_raw$intimacy_condition)

  cat("K (elicitation runs):", K, "\n")
  cat("Scenarios:", n_scenarios, "\n")
  cat("Cells (expected):", n_cells, " | lm_runs records:", nrow(runs_raw),
      " (expected", n_cells * K, ")\n")
  cat("Stage-1 alternative records:", nrow(alts), "\n")
} else {
  cat("No LM elicitation output found at", lm_dir,
      "\nRun `make lm-food_inv_desire` (or the generate + score scripts) first.\n")
}
```

## Coverage

Every `(scenario × observed-action × effort × intimacy)` cell should appear in
all `K` runs. The table below confirms each run covers the full cell grid (so the
mixture has the same support in every component).

```{r coverage}
if (have_lm) {
  per_run <- runs_raw |>
    count(run_id, name = "cells") |>
    summarise(min = min(cells), max = max(cells), runs = n())
  cat("Cells per run — min:", per_run$min, "max:", per_run$max,
      "across", per_run$runs, "runs (expected", n_cells, "each)\n")

  missing <- (n_cells * K) - nrow(runs_raw)
  cat("Missing (run, cell) records:", missing, "\n")
}
```

## Stage 1 — generated alternatives

### Set size per `(cell, run)`

The LM decides how many alternatives to propose per cell (no fixed quota), so set
size is itself a diagnostic — too few and the actor's choice set is degenerate,
implausibly many and the LM may be padding.

```{r alt-set-size}
#| fig-width: 8
#| fig-height: 4
if (have_lm) {
  set_sizes <- actions |>
    filter(!is_canonical) |>
    count(scenario_label, observed_action, effort_condition,
          intimacy_condition, run_id, name = "n_alts")

  cat("Alternatives per (cell, run) — min:", min(set_sizes$n_alts),
      "max:", max(set_sizes$n_alts),
      "mean:", round(mean(set_sizes$n_alts), 2),
      "median:", median(set_sizes$n_alts), "\n")

  ggplot(set_sizes, aes(n_alts)) +
    geom_bar(fill = "#4A7A4A", alpha = PLOT_ALPHA) +
    scale_x_continuous(breaks = scales::breaks_width(1)) +
    labs(title = "Alternatives generated per (cell, run)",
         x = "Number of alternatives", y = "Count of (cell, run) units")
}
```

### Truncation at the padding cap

The observer's actor evaluates a fixed-width action space of `MAX_ACTIONS = 12`
slots — slot 0 is the observed action, so at most 11 alternatives per cell survive;
any beyond that are silently dropped by the padded-table loaders. If many cells run
over the cap, the model is reasoning over a truncated choice set, so it's worth
knowing how often that happens and how many alternatives it costs.

```{r alt-truncation}
#| fig-width: 8
#| fig-height: 4
if (have_lm) {
  # The actor softmaxes over at most MAX_ACTIONS slots (slot 0 = the observed
  # action), so only the first MAX_ACTIONS - 1 alternatives per cell are kept; the
  # loaders in model/tables.py drop the rest. Constant mirrors MAX_ACTIONS there.
  MAX_ACTIONS <- 12L
  keep <- MAX_ACTIONS - 1L

  n_over <- sum(set_sizes$n_alts > keep)
  dropped <- sum(pmax(set_sizes$n_alts - keep, 0))
  cat("(cell, run) units over the cap (", keep, " kept): ", n_over, " of ",
      nrow(set_sizes), " (", round(100 * n_over / nrow(set_sizes), 1), "%)\n",
      sep = "")
  cat("Alternatives dropped to padding: ", dropped, " of ", sum(set_sizes$n_alts),
      " (", round(100 * dropped / sum(set_sizes$n_alts), 1), "%)\n", sep = "")

  ggplot(set_sizes, aes(n_alts, fill = n_alts > keep)) +
    geom_bar(alpha = PLOT_ALPHA) +
    geom_vline(xintercept = keep + 0.5, linetype = "dashed", color = "grey25") +
    annotate("text", x = keep + 0.5, y = Inf, hjust = -0.1, vjust = 1.5,
             label = paste0("cap = ", keep), color = "grey25", size = 3.5) +
    scale_fill_manual(values = c(`FALSE` = "#4A7A4A", `TRUE` = "#B05A5A"),
                      guide = "none") +
    scale_x_continuous(breaks = scales::breaks_width(1)) +
    labs(title = "Alternatives per (cell, run) vs. the padding cap",
         subtitle = paste0("Red = units whose extra alternatives are dropped at ",
                           "MAX_ACTIONS = ", MAX_ACTIONS),
         x = "Number of alternatives", y = "Count of (cell, run) units")
}
```

### Run-to-run diversity

Cross-run diversity is the point of the K-run pipeline (dedup is *within* a run;
repetition *across* runs is preserved). For each cell, `unique / total` over all
runs measures how much genuinely new phrasing the runs explore — near 1 means
every run proposes distinct actions, near `1/K` means the runs collapse onto the
same set.

```{r alt-diversity}
#| fig-width: 8
#| fig-height: 4
if (have_lm) {
  diversity <- alts |>
    mutate(key = tolower(trimws(action_text))) |>
    group_by(scenario_label, observed_action, effort_condition, intimacy_condition) |>
    summarise(total = n(), unique = n_distinct(key), .groups = "drop") |>
    mutate(diversity = unique / total)

  cat("Per-cell unique/total — mean:", round(mean(diversity$diversity), 2),
      "median:", round(median(diversity$diversity), 2), "\n")

  ggplot(diversity, aes(diversity)) +
    geom_histogram(bins = 25, fill = "#4A7A4A", alpha = PLOT_ALPHA) +
    labs(title = "Run-to-run alternative diversity per cell",
         subtitle = "unique alternative texts / total across runs",
         x = "Unique / total", y = "Cells")
}
```

### Sample alternatives

A spot-check that the generated actions are on-topic and span a plausible
risk range for a high-stakes scenario (oysters, observed high-risk share). Shows
the distinct alternatives proposed across all runs of one cell.

```{r alt-sample}
if (have_lm) {
  alts |>
    filter(scenario_label == "oysters", observed_action == "high_risk_share",
           effort_condition == "low", intimacy_condition == "somewhat_formal") |>
    distinct(action_text) |>
    head(15) |>
    kable(col.names = "Alternative action (oysters / high-risk share / low effort / somewhat formal)")
}
```

## Stage 2 — canonical action features

Slot 0 of each scored run is the observed canonical action. Aggregating slot 0
across scenarios and runs by `observed_action` recovers each canonical action's
feature distribution. Two hard sanity checks:

- **risk** should be monotone: `no_share` ≈ 0 < `low_risk_share` < `high_risk_share`.
- **goal-satisfaction `g`** should be near 0 for `no_share` (it doesn't deliver
  the food) and high for both sharing actions (they do). `g` is desire-free.

`risk` and `g` are intimacy- and effort-independent (risk is effort-marginal; `g`
is desire-free), so they're collapsed over those axes here.

```{r canon-risk-g}
#| fig-width: 10
#| fig-height: 4.5
if (have_lm) {
  canon <- actions |> filter(is_canonical)

  canon_rg <- canon |>
    group_by(scenario_label, observed_action, run_id) |>
    summarise(risk = mean(risk, na.rm = TRUE),
              g = mean(g, na.rm = TRUE), .groups = "drop") |>
    pivot_longer(c(risk, g), names_to = "feature", values_to = "value") |>
    mutate(action = as_action(observed_action),
           feature = factor(feature, levels = c("risk", "g"),
                            labels = c("Risk", "Goal-satisfaction g")))

  # Per-scenario mean trajectory across the three actions, kept faint as
  # background texture; `canon_rg_mean` is the bold overall mean trajectory.
  canon_rg_line <- canon_rg |>
    group_by(scenario_label, feature, action) |>
    summarise(value = mean(value, na.rm = TRUE), .groups = "drop")
  canon_rg_mean <- canon_rg |>
    group_by(feature, action) |>
    summarise(value = mean(value, na.rm = TRUE), .groups = "drop")

  ggplot(canon_rg, aes(action, value, color = action)) +
    geom_line(data = canon_rg_line, aes(group = scenario_label),
              color = "grey55", linewidth = 0.15, alpha = 0.12) +
    geom_boxplot(width = 0.55, fill = "white", alpha = 0.55) +
    geom_point(position = position_jitter(width = 0.16, height = 0.04, seed = 67),
               alpha = 0.22, size = 0.75) +
    # geom_line(data = canon_rg_mean, aes(group = 1),
    #           color = "grey15", linewidth = 0.9) +
    # geom_point(data = canon_rg_mean, color = "grey15", size = 2.1) +
    facet_wrap(~feature) +
    scale_color_manual(values = setNames(
      c("#7A9A4A", "#C9A85A", "#B05A5A"), action_labels)) +
    scale_y_continuous(limits = c(-0.05, 1.05), breaks = seq(0, 1, 0.25),
                       expand = expansion(mult = 0)) +
    labs(title = "Observed action features",
         x = "Observed action", y = "Feature value (0–1)") +
    theme(plot.title = element_text(size = 15, hjust = 0),
          axis.text.x = element_text(angle = 20, hjust = 1))
}
```

### Effort by condition

Effort is scored *with* the low-risk-share effort paragraph appended, once per
`effort_condition`. The manipulation targets the low-risk share, so its effort
should rise from the low to the high condition; the other actions' effort is a
useful contrast.

```{r canon-effort}
#| fig-width: 8
#| fig-height: 4.5
if (have_lm) {
  canon_eff <- actions |>
    filter(is_canonical) |>
    group_by(scenario_label, observed_action, run_id, effort_condition) |>
    summarise(effort = mean(effort, na.rm = TRUE), .groups = "drop") |>
    mutate(action = as_action(observed_action),
           effort_cond = factor(effort_condition, levels = c("low", "high"),
                                labels = c("Low effort", "High effort")))

  # Mean effort trajectory per effort condition across the three actions; x offset
  # by ±0.15 so each line sits under its dodged boxes (low left, high right —
  # matching position_dodge(width = 0.6) over 2 groups).
  canon_eff_mean <- canon_eff |>
    group_by(effort_cond, action) |>
    summarise(effort = mean(effort, na.rm = TRUE), .groups = "drop") |>
    mutate(xpos = as.integer(action) +
             ifelse(effort_cond == "Low effort", -0.15, 0.15))

  pd <- position_dodge(width = 0.7)
  ggplot(canon_eff, aes(action, effort, color = effort_cond)) +
    geom_boxplot(outlier.shape = NA, width = 0.6, position = pd, fill = NA) +
    geom_point(position = position_jitterdodge(jitter.width = 0.25, jitter.height = 0.03,
                                               dodge.width = 0.6, seed = 67),
               alpha = 0.2, size = 0.75) +
    geom_line(data = canon_eff_mean,
              aes(x = xpos, y = effort, color = effort_cond, group = effort_cond),
              linewidth = 1, inherit.aes = FALSE, show.legend = FALSE) +
    geom_point(data = canon_eff_mean,
               aes(x = xpos, y = effort, color = effort_cond),
               size = 2.1, inherit.aes = FALSE, show.legend = FALSE) +
    scale_color_effort() +
    scale_y_continuous(limits = c(-0.03, 1.03), breaks = seq(0, 1, 0.25),
                       expand = expansion(mult = 0)) +
    labs(title = "Effort of observed actions by effort condition",
         x = "Observed action", y = "Effort (0–1)", color = "Effort condition") +
    theme(plot.title = element_text(size = 15, hjust = 0),
          axis.text.x = element_text(angle = 20, hjust = 1))
}
```

### Effort manipulation specificity

The effort manipulation targets the low-risk share, so the high vs. low effort
condition should move *its* effort and leave the other two actions' effort roughly
unchanged. Plotting the per-`(scenario, run)` effort delta (high − low) by action
makes that specificity explicit, rather than asking the reader to compare two
boxplots by eye.

```{r canon-effort-delta}
#| fig-width: 7
#| fig-height: 4.5
if (have_lm) {
  eff_delta <- actions |>
    filter(is_canonical) |>
    group_by(scenario_label, observed_action, run_id, effort_condition) |>
    summarise(effort = mean(effort, na.rm = TRUE), .groups = "drop") |>
    pivot_wider(names_from = effort_condition, values_from = effort) |>
    mutate(delta = high - low, action = as_action(observed_action))

  eff_delta |>
    group_by(action) |>
    summarise(mean_delta = round(mean(delta, na.rm = TRUE), 3), .groups = "drop") |>
    kable(caption = "Effort delta (high − low condition) by observed action")

  ggplot(eff_delta, aes(action, delta, color = action)) +
    geom_hline(yintercept = 0, linewidth = 0.3, color = "grey50") +
    geom_boxplot(width = 0.5, fill = "white", alpha = 0.55, outlier.shape = NA) +
    geom_point(position = position_jitter(width = 0.16, height = 0, seed = 67),
               alpha = 0.18, size = 0.7) +
    scale_color_manual(values = setNames(
      c("#7A9A4A", "#C9A85A", "#B05A5A"), action_labels), guide = "none") +
    labs(title = "Effort manipulation is specific to the low-risk share",
         subtitle = "Per (scenario, run): effort(high condition) − effort(low condition)",
         x = "Observed action", y = "Effort delta (high − low)") +
    theme(plot.title = element_text(size = 15, hjust = 0),
          axis.text.x = element_text(angle = 20, hjust = 1))
}
```

### Run-to-run spread of features

The mixture's spread comes from run-to-run variation in the scored features.
Per-cell standard deviation of slot-0 risk and `g` across the `K` runs should be
non-trivially positive (else every mixture component is identical and the model
predicts a degenerate, zero-variance belief update).

```{r canon-spread}
#| fig-width: 8
#| fig-height: 4
if (have_lm) {
  spread <- actions |>
    filter(is_canonical) |>
    group_by(scenario_label, observed_action, effort_condition, intimacy_condition) |>
    summarise(risk = sd(risk, na.rm = TRUE),
              g = sd(g, na.rm = TRUE), .groups = "drop") |>
    pivot_longer(c(risk, g), names_to = "feature", values_to = "sd")

  spread |>
    group_by(feature) |>
    summarise(mean_sd = round(mean(sd, na.rm = TRUE), 3),
              max_sd = round(max(sd, na.rm = TRUE), 3), .groups = "drop") |>
    kable(caption = "Within-cell run-to-run SD of slot-0 features")

  ggplot(spread, aes(sd, fill = feature)) +
    geom_histogram(bins = 25, alpha = 0.7, position = "identity") +
    scale_fill_manual(values = c("risk" = "#B05A5A", "g" = "#4A7A4A")) +
    labs(title = "Run-to-run SD of canonical features (per cell)",
         x = "SD across runs", y = "Cells", fill = "Feature")
}
```

### Per-scenario monotonicity

The canonical risk/`g` checks above pool over all 16 scenarios, which can hide a
single scenario whose elicited features are out of order (e.g. a low-risk share
rated riskier than the high-risk share). This breaks the pooled check out by
scenario and flags any violations of the two orderings the pipeline relies on:
risk should rise `no_share → low_risk_share → high_risk_share`, and `g` for
`no_share` should sit below both sharing actions.

```{r canon-monotonicity}
#| fig-width: 9
#| fig-height: 4.5
if (have_lm) {
  canon_by_scen <- actions |>
    filter(is_canonical) |>
    group_by(scenario_label, observed_action) |>
    summarise(risk = mean(risk, na.rm = TRUE), g = mean(g, na.rm = TRUE),
              .groups = "drop")

  risk_wide <- canon_by_scen |>
    select(scenario_label, observed_action, risk) |>
    pivot_wider(names_from = observed_action, values_from = risk)
  bad_risk <- risk_wide |>
    filter(!(no_share <= low_risk_share & low_risk_share <= high_risk_share))
  cat("Scenarios with non-monotone canonical risk:", nrow(bad_risk), "of",
      nrow(risk_wide), "\n")
  if (nrow(bad_risk)) cat("  ", paste(bad_risk$scenario_label, collapse = ", "), "\n")

  g_wide <- canon_by_scen |>
    select(scenario_label, observed_action, g) |>
    pivot_wider(names_from = observed_action, values_from = g)
  bad_g <- g_wide |>
    filter(no_share >= pmin(low_risk_share, high_risk_share))
  cat("Scenarios where no_share g is not below both sharing actions:",
      nrow(bad_g), "\n")
  if (nrow(bad_g)) cat("  ", paste(bad_g$scenario_label, collapse = ", "), "\n")

  prof <- canon_by_scen |>
    pivot_longer(c(risk, g), names_to = "feature", values_to = "value") |>
    mutate(action = as_action(observed_action),
           feature = factor(feature, c("risk", "g"),
                            c("Risk", "Goal-satisfaction g")))
  prof_mean <- prof |>
    group_by(feature, action) |>
    summarise(value = mean(value, na.rm = TRUE), .groups = "drop")

  ggplot(prof, aes(action, value, group = scenario_label)) +
    geom_line(color = "grey60", linewidth = 0.3, alpha = 0.5) +
    geom_line(data = prof_mean, aes(group = 1), color = "#B05A5A", linewidth = 1.1) +
    geom_point(data = prof_mean, aes(group = 1), color = "#B05A5A", size = 2.2) +
    facet_wrap(~feature) +
    scale_y_continuous(limits = c(-0.05, 1.05), breaks = seq(0, 1, 0.25)) +
    labs(title = "Per-scenario canonical feature profiles",
         subtitle = "Grey = each scenario; red = pooled mean (risk should rise; no_share g should be lowest)",
         x = "Observed action", y = "Feature value (0–1)") +
    theme(plot.title = element_text(size = 15, hjust = 0),
          axis.text.x = element_text(angle = 20, hjust = 1))
}
```

## Given intimacy scalar

Studies 1a/1b take intimacy as observed context. Its magnitude is LM-elicited
**per run** from the de-anchored relationship descriptors (4 levels, scenario-
independent). It should be **monotone** across `max_formal → max_intimate`, and —
because it is scored per run — should carry run-to-run spread like the action
features do.

```{r intimacy-scalar}
#| fig-width: 8
#| fig-height: 4.5
if (have_lm) {
  intim <- runs_raw |>
    distinct(intimacy_condition, run_id, intimacy) |>
    mutate(intimacy_condition = factor(intimacy_condition, levels = INTIMACY_LEVELS))

  intim_summ <- intim |>
    group_by(intimacy_condition) |>
    summarise(value = mean(intimacy),
              sd = sd(intimacy),
              n_runs = n_distinct(run_id), .groups = "drop")

  intim_summ |>
    mutate(across(c(value, sd), \(x) round(x, 3))) |>
    kable(caption = "Per-level intimacy scalar (mean over runs, run-to-run SD)")

  ggplot(intim_summ, aes(intimacy_condition, value, color = intimacy_condition)) +
    geom_errorbar(aes(ymin = value - sd, ymax = value + sd), width = 0.15) +
    geom_point(size = 3.5) +
    scale_color_intimacy() +
    ylim(0, 1) +
    labs(title = "Elicited intimacy magnitude by level",
         subtitle = "error bars = run-to-run SD",
         x = NULL, y = "Intimacy magnitude (0–1)", color = "Intimacy level") +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
}
```

::: {.callout-note}
Watch the run-to-run SD column. If it is ≈ 0 for every level, the intimacy axis
contributes no spread to the mixture (the 4 levels collapse to fixed constants
across runs) — scored at the low default rating temperature, the descriptors may
return the same value every run. Also check the *spacing*: if
`somewhat_intimate` sits very close to `somewhat_formal`, the two middle levels
are nearly non-identifiable for the model.
:::

## Do the alternatives differ by condition?

The alternatives are elicited *per cell* — `(scenario × observed-action × effort ×
intimacy)` — so each observed conditioning axis can in principle reshape the
imagined choice set. This matters in two ways. The load-bearing one is
**anchoring on the observed action**: the cognitive claim is that an observer
imagines counterfactuals to *what they saw*, so the alternative set should shift
with `observed_action`. The other two axes — effort and intimacy — are observed
context in Study 1a (the latent is desire), so conditioning on them doesn't leak
the latent; but it's worth seeing whether, and how much, they move the set. The
checks below look at the alternatives two ways: the composition of the **set**
itself (size, and run-to-run text overlap) and the scored **properties** of the
non-canonical actions (risk / effort / `g`).

```{r alt-condition-helpers}
if (have_lm) {
  # Order intimacy levels by their elicited magnitude (ascending) so the axis
  # reads formal -> intimate, without depending on a fixed level vector.
  intim_order <- runs_raw |>
    group_by(intimacy_condition) |>
    summarise(m = mean(intimacy), .groups = "drop") |>
    arrange(m) |> pull(intimacy_condition)

  cond_levels <- c(action_labels, "Low", "High", intim_order)

  # Melt the three observed conditioning axes into a long axis/level form so each
  # gets its own panel; levels differ per axis, so panels use a free x scale.
  to_cond_long <- function(df) {
    df |>
      mutate(`Observed action`    = as.character(as_action(observed_action)),
             `Effort condition`   = factor(effort_condition, c("low", "high"),
                                           c("Low", "High")),
             `Intimacy condition` = factor(intimacy_condition, intim_order)) |>
      pivot_longer(c(`Observed action`, `Effort condition`, `Intimacy condition`),
                   names_to = "axis", values_to = "level") |>
      mutate(axis  = factor(axis, c("Observed action", "Effort condition",
                                    "Intimacy condition")),
             level = factor(level, cond_levels))
  }

  # Long condition slugs crowd the narrow facet panels; wrap them onto two lines
  # (underscores -> spaces first) instead of rotating. Generic, so it survives
  # any renaming of the intimacy slugs.
  wrap_lab <- function(x) stringr::str_wrap(gsub("_", " ", x), width = 10)
}
```

### Set size by condition

Does the LM propose systematically more or fewer alternatives in some conditions?
Set size is pooled over the other axes within each panel.

```{r alt-setsize-condition}
#| fig-width: 11
#| fig-height: 4
if (have_lm) {
  set_sizes_cond <- actions |>
    filter(!is_canonical) |>
    count(scenario_label, observed_action, effort_condition,
          intimacy_condition, run_id, name = "n_alts") |>
    to_cond_long()

  ggplot(set_sizes_cond, aes(level, n_alts)) +
    geom_boxplot(fill = "#4A7A4A", alpha = PLOT_ALPHA, outlier.shape = NA,
                 width = 0.6) +
    geom_point(position = position_jitter(width = 0.18, height = 0.12, seed = 67),
               alpha = 0.12, size = 0.5, color = "#1F3A1F") +
    stat_summary(fun = mean, geom = "point", shape = 18, size = 2.5,
                 color = "#B05A5A") +
    facet_wrap(~axis, scales = "free_x") +
    scale_x_discrete(labels = wrap_lab) +
    labs(title = "Alternative set size by condition",
         subtitle = "Alternatives per (cell, run); red diamond = mean",
         x = NULL, y = "Number of alternatives") +
    theme(axis.text.x = element_text(size = 12))
}
```

### Properties by condition

The headline check. Each non-canonical action carries a scored risk, effort, and
`g`; this grid asks whether those distributions shift with each conditioning axis.
The strongest signal should be down the **observed-action** column (the anchoring
claim) — e.g. whether imagined alternatives skew lower-risk once a high-risk share
has already been observed, and whether they more often deliver the food (`g`)
after a `no_share`. Movement down the effort/intimacy columns shows how much those
framings bleed into the imagined set.

```{r alt-features-condition}
#| fig-width: 11
#| fig-height: 8
if (have_lm) {
  alt_feat <- actions |>
    filter(!is_canonical) |>
    select(scenario_label, observed_action, effort_condition, intimacy_condition,
           run_id, risk, effort, g) |>
    pivot_longer(c(risk, effort, g), names_to = "feature", values_to = "value") |>
    mutate(feature = factor(feature, c("risk", "effort", "g"),
                            c("Risk", "Effort", "Goal-satisfaction g"))) |>
    to_cond_long()

  ggplot(alt_feat, aes(level, value)) +
    geom_boxplot(fill = "#6B8E6B", alpha = PLOT_ALPHA, outlier.shape = NA,
                 width = 0.6) +
    geom_point(position = position_jitter(width = 0.18, height = 0.01, seed = 67),
               alpha = 0.05, size = 0.4, color = "#1F3A1F") +
    stat_summary(fun = mean, geom = "point", shape = 18, size = 2,
                 color = "#B05A5A") +
    facet_grid(feature ~ axis, scales = "free_x") +
    scale_x_discrete(labels = wrap_lab) +
    labs(title = "Alternative-action features by condition",
         subtitle = paste("Non-canonical (LM-generated) actions only, pooled over",
                          "scenarios × runs; red diamond = mean"),
         x = NULL, y = "Feature value (0–1)") +
    theme(axis.text.x = element_text(size = 12))
}
```

### Does the *set* of actions itself differ?

Set size and feature distributions can look stable while the actual actions
change. To test the set directly, we measure the **Jaccard overlap** of the
(normalized) alternative texts between cells that differ on one axis, and compare
it against a noise floor: how much two independent *runs of the same cell*
overlap. If conditioning genuinely reshapes the set, across-axis overlap should
sit *below* that run-to-run floor; if it sits at the floor, the axis adds nothing
beyond elicitation stochasticity. (Overlap is exact normalized-text match, so
paraphrases count as misses — these numbers are a lower bound on semantic
overlap.)

```{r alt-overlap-condition}
#| fig-width: 9
#| fig-height: 4.5
if (have_lm) {
  norm_txt <- function(x) {
    x <- tolower(trimws(x))
    x <- gsub("[[:punct:]]+$", "", x)
    gsub("\\s+", " ", x)
  }
  jacc <- function(a, b) {
    a <- unique(a); b <- unique(b)
    u <- length(union(a, b))
    if (u == 0) NA_real_ else length(intersect(a, b)) / u
  }
  mean_pairwise <- function(sets) {
    if (length(sets) < 2) return(NA_real_)
    pr <- combn(length(sets), 2)
    mean(vapply(seq_len(ncol(pr)),
                \(j) jacc(sets[[pr[1, j]]], sets[[pr[2, j]]]), numeric(1)),
         na.rm = TRUE)
  }
  set_by <- function(df, ...) {
    df |> group_by(...) |> summarise(set = list(unique(key)), .groups = "drop")
  }

  alt_keys <- alts |> mutate(key = norm_txt(action_text))

  # (1) Noise floor: same cell, across the K runs.
  base_ov <- set_by(alt_keys, scenario_label, observed_action,
                    effort_condition, intimacy_condition, run_id) |>
    group_by(scenario_label, observed_action, effort_condition, intimacy_condition) |>
    summarise(overlap = mean_pairwise(set), .groups = "drop") |>
    mutate(comparison = "Across runs\n(same cell)")

  # (2) Across effort, holding scenario / action / intimacy / run fixed.
  eff_ov <- set_by(alt_keys, scenario_label, observed_action,
                   intimacy_condition, run_id, effort_condition) |>
    group_by(scenario_label, observed_action, intimacy_condition, run_id) |>
    summarise(overlap = mean_pairwise(set), .groups = "drop") |>
    mutate(comparison = "Across effort\n(same run)")

  # (3) Across intimacy, holding scenario / action / effort / run fixed.
  int_ov <- set_by(alt_keys, scenario_label, observed_action,
                   effort_condition, run_id, intimacy_condition) |>
    group_by(scenario_label, observed_action, effort_condition, run_id) |>
    summarise(overlap = mean_pairwise(set), .groups = "drop") |>
    mutate(comparison = "Across intimacy\n(same run)")

  overlap <- bind_rows(base_ov, eff_ov, int_ov) |>
    mutate(comparison = factor(comparison, levels = c(
      "Across runs\n(same cell)", "Across effort\n(same run)",
      "Across intimacy\n(same run)")))

  overlap |>
    group_by(comparison) |>
    summarise(mean = round(mean(overlap, na.rm = TRUE), 3),
              median = round(median(overlap, na.rm = TRUE), 3), .groups = "drop") |>
    kable(caption = "Jaccard overlap of alternative-text sets (exact normalized match)")

  ggplot(overlap, aes(comparison, overlap, fill = comparison)) +
    geom_boxplot(alpha = PLOT_ALPHA, outlier.shape = NA, width = 0.6) +
    geom_point(position = position_jitter(width = 0.15, height = 0.01, seed = 67),
               alpha = 0.15, size = 0.6, color = "#333333") +
    scale_fill_manual(values = c("#8A8A8A", "#4A7A8A", "#8A5A8A"), guide = "none") +
    ylim(0, 1) +
    labs(title = "Do the alternative sets differ by condition?",
         subtitle = "Run-to-run overlap within a cell vs. across an experimental axis",
         x = NULL, y = "Jaccard overlap (alternative texts)")
}
```

## The observed action in its choice set

The alternatives matter only through their contrast with the observed action: the
actor softmaxes over `{observed action} ∪ alternatives`, so an action is diagnostic
exactly to the extent its features stand out from the counterfactuals the observer
imagined. The sections above plot the observed action and the alternatives
separately; here we join each observed action to *its own* alternatives within the
same `(cell, run)` and ask three questions — how far the observed action's features
sit from its alternatives', whether the alternatives bracket it, and whether the
joint feature space supports the competing-explanation trade-offs the model relies
on.

### Feature gap: observed action vs. its alternatives

For each `(cell, run)`, the observed action (slot 0) minus the mean of *its* scored
alternatives, per feature. This is the diagnosticity quantity in raw form: a
positive risk gap for a high-risk share means the observer imagined safer options
the actors passed up (so the action is evidence about the relational cost), and a
negative `g` gap for `no_share` means the imagined alternatives would have delivered
the food better (so not sharing is evidence about desire).

```{r alt-gap}
#| fig-width: 10
#| fig-height: 4.5
if (have_lm) {
  # Within each (cell, run): slot-0 feature minus the mean over that cell's
  # alternatives. Cells with no alternatives are dropped (no contrast to measure).
  gap <- actions |>
    group_by(scenario_label, observed_action, effort_condition,
             intimacy_condition, run_id) |>
    filter(any(is_canonical), any(!is_canonical)) |>
    summarise(
      risk   = risk[is_canonical][1]   - mean(risk[!is_canonical],   na.rm = TRUE),
      effort = effort[is_canonical][1] - mean(effort[!is_canonical], na.rm = TRUE),
      g      = g[is_canonical][1]      - mean(g[!is_canonical],      na.rm = TRUE),
      .groups = "drop") |>
    pivot_longer(c(risk, effort, g), names_to = "feature", values_to = "gap") |>
    mutate(action = as_action(observed_action),
           feature = factor(feature, c("risk", "effort", "g"),
                            c("Risk", "Effort", "Goal-satisfaction g")))

  gap |>
    group_by(action, feature) |>
    summarise(median = round(median(gap, na.rm = TRUE), 2), .groups = "drop") |>
    pivot_wider(names_from = feature, values_from = median) |>
    kable(caption = "Median feature gap (observed action − its alternatives)")

  ggplot(gap, aes(action, gap, color = action)) +
    geom_hline(yintercept = 0, linewidth = 0.3, color = "grey50") +
    geom_boxplot(width = 0.55, fill = "white", alpha = 0.55, outlier.shape = NA) +
    geom_point(position = position_jitter(width = 0.16, height = 0, seed = 67),
               alpha = 0.12, size = 0.6) +
    facet_wrap(~feature) +
    scale_color_manual(values = setNames(
      c("#7A9A4A", "#C9A85A", "#B05A5A"), action_labels), guide = "none") +
    labs(title = "Observed action minus its alternatives",
         subtitle = "Per (cell, run): slot-0 feature − mean over that cell's alternatives",
         x = "Observed action", y = "Feature gap (observed − alternatives)") +
    theme(plot.title = element_text(size = 15, hjust = 0),
          axis.text.x = element_text(angle = 20, hjust = 1))
}
```

### Does the alternative set bracket the observed action?

A gap says how far apart the observed action and its alternatives sit on average;
bracketing says whether the alternatives fall on *both sides* of it. When the
observed action is the extreme of its set on a feature (no imagined option is
riskier, or none delivers the food better), that one-sidedness is what drives the
inference; when it is bracketed, the alternatives offer competing explanations in
both directions.

```{r alt-bracket}
#| fig-width: 10
#| fig-height: 4.5
if (have_lm) {
  pos_of <- function(canon, alt) {
    alt <- alt[!is.na(alt)]
    if (!length(alt) || is.na(canon)) return(NA_character_)
    if (canon > max(alt)) "Above all alternatives"
    else if (canon < min(alt)) "Below all alternatives"
    else "Bracketed"
  }
  bracket <- actions |>
    group_by(scenario_label, observed_action, effort_condition,
             intimacy_condition, run_id) |>
    filter(any(is_canonical), any(!is_canonical)) |>
    summarise(
      Risk   = pos_of(risk[is_canonical][1],   risk[!is_canonical]),
      Effort = pos_of(effort[is_canonical][1], effort[!is_canonical]),
      `Goal-satisfaction g` = pos_of(g[is_canonical][1], g[!is_canonical]),
      .groups = "drop") |>
    pivot_longer(c(Risk, Effort, `Goal-satisfaction g`),
                 names_to = "feature", values_to = "position") |>
    filter(!is.na(position)) |>
    mutate(action = as_action(observed_action),
           feature = factor(feature, c("Risk", "Effort", "Goal-satisfaction g")),
           position = factor(position, c("Below all alternatives", "Bracketed",
                                         "Above all alternatives")))

  ggplot(bracket, aes(action, fill = position)) +
    geom_bar(position = "fill", alpha = PLOT_ALPHA, width = 0.7) +
    facet_wrap(~feature) +
    scale_fill_manual(values = c("Below all alternatives" = "#7AA0B5",
                                 "Bracketed" = "#C9C0A8",
                                 "Above all alternatives" = "#B05A5A")) +
    scale_y_continuous(labels = scales::percent) +
    labs(title = "Where the observed action sits in its alternative set",
         subtitle = "Share of (cell, run) units by the observed action's position on each feature",
         x = "Observed action", y = "Share of cells", fill = NULL) +
    theme(plot.title = element_text(size = 15, hjust = 0),
          axis.text.x = element_text(angle = 20, hjust = 1),
          legend.position = "top")
}
```

### Joint feature structure

Every diagnosticity argument in the model is a *trade-off* — high-risk sharing can
be explained by strong desire or by a costly low-risk option, so the same action is
weaker evidence for intimacy when desire is high. That reasoning is only possible if
the elicited features are not collinear: there must be actions that are high-`g` at
low risk *and* high-`g` at high risk, and the low-risk options must actually cost
more effort. This plots the joint geometry of the feature space (each point an
action, alternatives in grey) and reports the pairwise feature correlations.

```{r alt-joint-structure}
#| fig-width: 10
#| fig-height: 5
if (have_lm) {
  # One row per distinct action within a (scenario, run): risk and g don't vary by
  # effort/intimacy cell, and effort is averaged over its two conditions, so each
  # action contributes a single point to the feature cloud.
  cloud <- actions |>
    group_by(scenario_label, run_id, action_text, is_canonical) |>
    summarise(risk = mean(risk, na.rm = TRUE),
              effort = mean(effort, na.rm = TRUE),
              g = mean(g, na.rm = TRUE),
              observed_action = first(observed_action), .groups = "drop") |>
    mutate(category = if_else(is_canonical,
                              as.character(as_action(observed_action)), "Alternative"))

  cloud |>
    summarise(`Risk–Effort` = cor(risk, effort, use = "complete.obs"),
              `Risk–g` = cor(risk, g, use = "complete.obs"),
              `Effort–g` = cor(effort, g, use = "complete.obs")) |>
    pivot_longer(everything(), names_to = "Feature pair", values_to = "Pearson r") |>
    mutate(`Pearson r` = round(`Pearson r`, 2)) |>
    kable(caption = "Feature correlations across all elicited actions")

  cloud_long <- cloud |>
    pivot_longer(c(effort, g), names_to = "yvar", values_to = "yval") |>
    mutate(yvar = factor(yvar, c("effort", "g"),
                         c("Effort", "Goal-satisfaction g")))
  cloud_alt <- filter(cloud_long, !is_canonical)
  cloud_can <- filter(cloud_long, is_canonical)

  ggplot(mapping = aes(risk, yval)) +
    geom_point(data = cloud_alt, color = "grey75", alpha = 0.08, size = 0.5) +
    geom_point(data = cloud_can, aes(color = category), alpha = 0.5, size = 1.1) +
    facet_wrap(~yvar) +
    scale_color_manual(values = setNames(
      c("#7A9A4A", "#C9A85A", "#B05A5A"), action_labels), name = "Observed action") +
    scale_x_continuous(limits = c(-0.03, 1.03), breaks = seq(0, 1, 0.25)) +
    scale_y_continuous(limits = c(-0.03, 1.03), breaks = seq(0, 1, 0.25)) +
    labs(title = "Joint feature structure of the action set",
         subtitle = "Each point an action (grey = LM alternatives); risk vs. the other two features",
         x = "Risk (0–1)", y = "Feature value (0–1)") +
    theme(plot.title = element_text(size = 15, hjust = 0),
          legend.position = "top")
}
```

### Dominated alternatives

An alternative that is weakly worse than the observed action on *every* utility
term — no higher `g`, no lower effort, no lower risk — is Pareto-dominated by what
the actors actually did, so it adds nothing to the softmax contrast regardless of
the fitted weights. A high dominated fraction means part of the generated set is
inert padding. (The reverse — the observed action itself dominated by an
alternative — would be a surprising choice and is reported alongside.)

```{r alt-dominated}
#| fig-width: 7
#| fig-height: 4.5
if (have_lm) {
  dom <- actions |>
    group_by(scenario_label, observed_action, effort_condition,
             intimacy_condition, run_id) |>
    filter(any(is_canonical), any(!is_canonical)) |>
    mutate(c_risk = risk[is_canonical][1], c_eff = effort[is_canonical][1],
           c_g = g[is_canonical][1]) |>
    filter(!is_canonical) |>
    mutate(dom_by_canon = c_g >= g & c_eff <= effort & c_risk <= risk &
             (c_g > g | c_eff < effort | c_risk < risk),
           dominates_canon = g >= c_g & effort <= c_eff & risk <= c_risk &
             (g > c_g | effort < c_eff | risk < c_risk)) |>
    ungroup()

  cat("Alternatives Pareto-dominated by the observed action:",
      round(100 * mean(dom$dom_by_canon, na.rm = TRUE), 1), "%\n")
  cat("Alternatives that dominate the observed action:",
      round(100 * mean(dom$dominates_canon, na.rm = TRUE), 1), "%\n")

  dom_by_action <- dom |>
    mutate(action = as_action(observed_action)) |>
    group_by(action) |>
    summarise(`Dominated by observed` = 100 * mean(dom_by_canon, na.rm = TRUE),
              `Dominates observed` = 100 * mean(dominates_canon, na.rm = TRUE),
              .groups = "drop") |>
    pivot_longer(c(`Dominated by observed`, `Dominates observed`),
                 names_to = "direction", values_to = "pct")

  ggplot(dom_by_action, aes(action, pct, fill = direction)) +
    geom_col(position = position_dodge(width = 0.7), width = 0.65,
             alpha = PLOT_ALPHA) +
    scale_fill_manual(values = c("Dominated by observed" = "#6B8E6B",
                                 "Dominates observed" = "#B05A5A")) +
    labs(title = "Pareto dominance between alternatives and the observed action",
         subtitle = "Share of alternatives weakly worse / better on all of g, effort, risk",
         x = "Observed action", y = "% of alternatives", fill = NULL) +
    theme(plot.title = element_text(size = 15, hjust = 0),
          axis.text.x = element_text(angle = 20, hjust = 1),
          legend.position = "top")
}
```

## Semantic structure of the alternatives

The exact-text diversity and Jaccard panels above count surface forms, so
paraphrases ("tear it in half" vs. "rip it in two") read as distinct. This section
reads an optional artifact from `model/lm/embed_alternatives.py`, which embeds each
unique alternative two ways. For **diversity**, it clusters the alternatives *within
each scenario* into action types — per scenario because the dominant axis of
embedding variation is the scenario itself (its names, food, setting), so a single
global clustering would just recover the 16 scenarios rather than abstract action
types. For the cross-scenario **anchoring** view, it instead labels each alternative
by its nearest canonical action (no-share / low-risk / high-risk share), which is
scenario-invariant by construction since every scenario shares that three-action
frame; the cosine to that nearest canonical also catches near-paraphrases that add
little contrast. Generate it with
`uv run python model/lm/embed_alternatives.py --study food_inv_desire`; until then
this section reports the artifact is missing and renders nothing.

```{r semantic-load}
sem_path <- file.path(lm_dir, "lm_alternatives_semantic.jsonl")
clusters_path <- file.path(lm_dir, "lm_clusters.json")
have_sem <- have_lm && file.exists(sem_path)

if (have_sem) {
  sem <- read_model_jsonl(sem_path)
  alts_sem <- alts |>
    left_join(sem, by = c("scenario_label", "action_text"))
  cat("Semantic artifact:", nrow(sem), "unique (scenario, alt) pairs across",
      n_distinct(sem$scenario_label), "scenarios (per-scenario clusters)\n")
} else {
  cat("No semantic artifact at", sem_path,
      "\nRun `uv run python model/lm/embed_alternatives.py --study food_inv_desire`",
      "to generate it.\n")
}
```

### Action types (one example scenario)

Clusters are per scenario, so there is no single global type list to show. For one
illustrative scenario, each within-scenario cluster with a representative
alternative shows the kinds of counterfactual the LM proposes there — a healthy set
spans several distinct types rather than collapsing onto one.

```{r semantic-exemplars}
if (have_sem && file.exists(clusters_path)) {
  cl <- jsonlite::fromJSON(clusters_path, simplifyMatrix = FALSE)
  ex_scen <- "oysters"
  cl$clusters |>
    filter(scenario == ex_scen) |>
    transmute(cluster, size,
              exemplar = vapply(exemplars, \(x) x[[1]], character(1))) |>
    arrange(desc(size)) |>
    kable(caption = paste0("Within-scenario action types for '", ex_scen,
                           "' (one example scenario)"))
}
```

### Semantic diversity per cell

The semantic analog of the exact-text diversity panel: distinct action *types* per
cell divided by the number of alternatives. Near 1 means each alternative is a
different kind of action; low values mean the set is many paraphrases of a few
types.

```{r semantic-diversity}
#| fig-width: 8
#| fig-height: 4
if (have_sem) {
  sem_div <- alts_sem |>
    filter(!is.na(cluster)) |>
    group_by(scenario_label, observed_action, effort_condition, intimacy_condition) |>
    summarise(n = n(), types = n_distinct(cluster), .groups = "drop") |>
    mutate(type_ratio = types / n)

  cat("Distinct types per cell — mean:", round(mean(sem_div$types), 2),
      "| types/alts — mean:", round(mean(sem_div$type_ratio), 2), "\n")

  ggplot(sem_div, aes(type_ratio)) +
    geom_histogram(bins = 25, fill = "#4A7A4A", alpha = PLOT_ALPHA) +
    labs(title = "Semantic diversity per cell",
         subtitle = "Distinct per-scenario action types / total alternatives, across runs",
         x = "Distinct types / total", y = "Cells")
}
```

### Type mix by observed action (semantic anchoring)

The semantic version of the anchoring check: if observers imagine counterfactuals
to *what they saw*, the imagined set should shift with the observed action — e.g.
skew toward lower-risk actions once a high-risk share has been observed. Each
alternative is labeled by its nearest canonical action (a scenario-invariant
type, so it aggregates across scenarios), and each row is that distribution for one
observed action (rows sum to 100%).

```{r semantic-anchoring}
#| fig-width: 8
#| fig-height: 4.5
if (have_sem) {
  type_by_action <- alts_sem |>
    filter(!is.na(nearest_canonical)) |>
    count(observed_action, nearest_canonical) |>
    group_by(observed_action) |>
    mutate(prop = n / sum(n)) |>
    ungroup() |>
    mutate(action = as_action(observed_action),
           nearest = as_action(nearest_canonical))

  ggplot(type_by_action, aes(nearest, action, fill = prop)) +
    geom_tile(color = "white", linewidth = 0.4) +
    geom_text(aes(label = scales::percent(prop, accuracy = 1)),
              size = 3.6, color = "grey15") +
    scale_fill_viridis_c(option = "mako", direction = -1) +
    labs(title = "Imagined alternatives by resemblance to a canonical action",
         subtitle = "Each alternative labeled by its nearest canonical action; rows sum to 100%",
         x = "Alternative most resembles…", y = "Observed action", fill = "Proportion") +
    theme(plot.title = element_text(size = 15, hjust = 0),
          axis.text.x = element_text(angle = 20, hjust = 1))
}
```

### Near-paraphrases of the canonical actions

Each alternative's cosine similarity to the nearest canonical action of its
scenario. Alternatives near 1 are essentially restatements of one of the three
canonical actions and add little contrast to the actor's choice set; a long right
tail near the dashed threshold flags how much of the generated set is redundant
with what was already on the menu.

```{r semantic-neardup}
#| fig-width: 8
#| fig-height: 4
if (have_sem) {
  dup_thresh <- 0.85
  cat("Alternatives with cosine ≥", dup_thresh, "to a canonical action:",
      round(100 * mean(sem$sim_to_canonical >= dup_thresh, na.rm = TRUE), 1), "%\n")

  ggplot(sem, aes(sim_to_canonical)) +
    geom_histogram(bins = 30, fill = "#6B8E6B", alpha = PLOT_ALPHA) +
    geom_vline(xintercept = dup_thresh, linetype = "dashed", color = "#B05A5A") +
    labs(title = "Alternative similarity to the nearest canonical action",
         subtitle = "Embedding cosine; high = near-paraphrase of an already-listed action",
         x = "Max cosine similarity to a canonical action", y = "Unique alternatives")
}
```

## Failed ratings

A scoring call can fail to return a parseable value for an action; that feature
is stored as missing (`null` → `NA`) rather than dropped, so the loader can NaN-
mask it. The failure rate should be low.

```{r failed-ratings}
if (have_lm) {
  actions |>
    summarise(across(c(risk, effort, g),
                     list(n_missing = \(x) sum(is.na(x)),
                          pct = \(x) round(mean(is.na(x)) * 100, 2)))) |>
    pivot_longer(everything(),
                 names_to = c("feature", ".value"),
                 names_pattern = "(.*)_(n_missing|pct)") |>
    kable(caption = "Missing (failed) ratings per feature",
          col.names = c("Feature", "N missing", "% missing"))
}
```

## Takeaways

- **Coverage & set sizes** — confirm every run covers the full cell grid and that
  alternative set sizes are reasonable (a handful per cell, none empty); the
  truncation panel flags how often a cell exceeds the `MAX_ACTIONS` cap and loses
  alternatives the model never sees.
- **Canonical features** — risk should be monotone across the three actions and
  `g` should separate `no_share` (≈ 0) from the sharing actions; this is the load-
  bearing sanity check for the whole pipeline. The per-scenario panel flags any
  scenario that breaks those orderings, and the effort-delta panel confirms the
  effort manipulation moves only the low-risk share.
- **Run-to-run spread** — the action features should carry positive per-cell SD;
  the intimacy scalar is the one to scrutinize, since it can come back constant
  across runs (see the callout above).
- **Alternatives by condition** — the clearest movement should be down the
  observed-action column (the anchoring claim: e.g. a high-risk share already
  observed yields lower-risk imagined alternatives, and a `no_share` yields
  higher-`g` ones); the effort/intimacy axes show how much those framings reshape
  the imagined set, with the Jaccard panel separating real reshaping from
  elicitation noise.
- **The observed action in its choice set** — the diagnosticity view: the feature
  gap (observed − alternatives) should separate by observed action (a high-risk
  share above its imagined alternatives on risk, a `no_share` below them on `g`),
  the bracketing panel shows whether alternatives straddle the observed action, and
  the joint-structure panel confirms risk / effort / `g` are not collinear so the
  model's competing-explanation trade-offs are identifiable; the dominance panel
  flags inert alternatives (Pareto-worse than the observed action on every term).
- **Semantic structure** (optional; needs `embed_alternatives.py`) — per-scenario
  clustering gives a meaning-level diversity metric, the nearest-canonical mix tests
  the anchoring claim in a scenario-invariant frame (does the imagined set skew
  lower-risk once a high-risk share is observed?), and the canonical-similarity
  histogram catches near-paraphrase padding the exact-match panels miss.

See `.claude/rules/model.md` for the full elicitation design and
`model/lm/score_merged.py` for the scoring procedure.
