---
title: Forward planning, effort manipulation (data analysis)
author: Alicia Chen
date: 2026-04-24
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    page-layout: full
    search: false
knitr:
  opts_chunk:
    dev: "ragg_png"
execute:
  warning: false
---

```{r setup}
source(here::here("analysis", "utils.R"))
setup_analysis()
```

## Demographics

```{r}
df_exit <- read_csv(here("data", "forw_plan_effort", "exit_survey.csv"), show_col_types = FALSE)

n_total <- nrow(df_exit)
n_passed <- df_exit |>
  filter(attention_passed == TRUE, memory_correct_count > 0) |>
  nrow()

cat("Total participants recruited:", n_total, "\n")
cat("Passed attention + memory checks:", n_passed, "\n")
cat("Mean age:", round(mean(df_exit$age, na.rm = TRUE), 1),
    "SD age:", round(sd(df_exit$age, na.rm = TRUE), 1),
    "Min age:", min(df_exit$age, na.rm = TRUE),
    "Max age:", max(df_exit$age, na.rm = TRUE))
cat("\nGender:\n")
print(table(df_exit$gender))
```

## Results

This experiment manipulates the relationship between two people (intimacy: 0, 50, 75, 100 on a 100-point scale) and the relative effort of avoiding saliva-sharing (effort: `low` vs `high`). Reward is held fixed at high across all trials. Each scenario has two actions: action 1 is non-saliva-sharing (requires the effort-constrained resource) and action 2 is saliva-sharing.

Load the data (produced by `analysis/json_to_csv.py forw_plan_effort`, with attention/memory-check exclusions already applied):

```{r}
df <- read_csv(here("data", "forw_plan_effort", "main_trials_long.csv"), show_col_types = FALSE) |>
  mutate(
    intimacy = factor(intimacy, levels = INTIMACY_LEVELS),
    effort = factor(effort, levels = c("low", "high"), labels = c("Low effort", "High effort")),
    action = factor(action)
  )

df_summary <- df |>
  group_by(intimacy, effort, action) |>
  tidyboot_mean(p_action) |>
  ungroup()
```

### P(share saliva) by effort and intimacy

Since the two actions are complementary (they sum to 1), P(action 2) = P(share saliva) is a sufficient summary. x-axis shows effort of the non-saliva-sharing action; hue shows the relationship's intimacy level.

```{r}
#| fig-width: 6
#| fig-asp: 0.75

df_summary |>
  filter(action == 2) |>
  ggplot(aes(x = effort, y = empirical_stat, color = intimacy, group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 4, alpha = PLOT_ALPHA) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), linewidth = 1.2, width = 0.4
  ) +
  geom_line(position = position_dodge(width = 0.2), linewidth = 1.2) +
  scale_color_intimacy() +
  scale_y_continuous(limits = c(0, 1)) +
  labs(y = "P(share saliva)", x = "Effort of not sharing saliva", color = "Intimacy")

ggsave(here("figures", "forw_plan_effort_paction.pdf"), device = cairo_pdf, width = 6, height = 4.5)
```

### Per-scenario breakdown

```{r}
#| fig-width: 22
#| fig-asp: 0.25

df_summary_scenario <- df |>
  group_by(scenario_label, intimacy, effort, action) |>
  tidyboot_mean(p_action) |>
  ungroup()

ggplot(df_summary_scenario, aes(x = action, y = empirical_stat, color = intimacy, group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 3, alpha = PLOT_ALPHA) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), width = 0.4
  ) +
  geom_line(position = position_dodge(width = 0.2), linewidth = 1) +
  scale_color_intimacy() +
  labs(y = "P(action)", x = "Action", color = "Intimacy") +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 0)) +
  facet_grid(effort ~ scenario_label, labeller = labeller(scenario_label = label_wrap_gen(15)))
```

## Model comparison

### Model description

We fit three ablations of the canonical access utility to the human data. A joint actor chooses an action proportional to its total utility, following a softmax with inverse temperature $\alpha$. Per-(scenario, effort_condition, action) values of $\text{access}(a)$ and $\text{effort}(a)$ are LLM-generated (Llama-3.3-70B; see `model/lm_scenario_params_effort.py`). Reward is held fixed at high across all trials; both actions involve eating, so $V(a|s) = 1$ is stipulated for both. $w_v$ is therefore non-identified in the softmax and stays near its initialization during fitting.

**Full model**:

$$U(a|s, I, e) = w_v \cdot V(a) - w_d \cdot \text{access}(a) \cdot (1 - I) - w_e \cdot \text{effort}(a)$$

**Discomfort-only**: drops the food-reward and effort terms, keeping only $-w_d \cdot \text{access}(a) \cdot (1 - I)$.

**Base model**: drops the access term: $U = w_v \cdot V(a) - w_e \cdot \text{effort}(a)$.

**Why the Base model is flat across effort conditions.** In this experiment $V(a) = 1$ for both actions, so $w_v \cdot V$ is constant in the softmax and the only differentiator is $-w_e \cdot \text{effort}(a)$. The fitted $w_e$ lands at the optimizer's lower clip ($\approx 10^{-6}$) on every LOSO fold, which makes the utility effectively constant and the predictions $\approx 0.5$ in every cell — independent of effort condition or intimacy. The reason $w_e$ collapses to zero is that effort and access are negatively correlated in the stimuli (the higher-access action is usually the lower-effort one). With only an effort term to work with, a positive $w_e$ would predict a uniform preference for the lower-effort = higher-access action across all intimacy levels, which contradicts the low-intimacy human data; setting $w_e \approx 0$ gives uniform predictions, which is a less-bad fit. This also explains the negative pooled $r$ for the Base model: the access $\times$ intimacy interaction is what drives behavior, and effort alone cannot substitute for it.

### Load model predictions

Model predictions are the **out-of-sample LOSO predictions** from `model/outputs/cv_loso_preds_effort.csv`: for each trial, the prediction was made by a model fit on the other 15 scenarios.

```{r}
df_cv <- read_csv(here("model", "outputs", "cv_loso_preds_effort.csv"),
                  show_col_types = FALSE)

df_humans_cv <- df_cv |>
  filter(variant == "access_full") |>
  transmute(
    subject_id, scenario_label = held_out_scenario,
    intimacy, effort, action,
    p_action, model = "Humans"
  )

df_model_cv <- df_cv |>
  mutate(
    model = case_when(
      variant == "access_full" ~ "Full model",
      variant == "access_only" ~ "Discomfort-only",
      variant == "no_access" ~ "Base model"
    )
  ) |>
  transmute(
    subject_id, scenario_label = held_out_scenario,
    intimacy, effort, action,
    p_action = p_action_pred, model
  )

df_preds <- bind_rows(df_humans_cv, df_model_cv) |>
  mutate(
    intimacy = factor(intimacy, levels = INTIMACY_LEVELS),
    effort = factor(effort, levels = c("low", "high"),
                    labels = c("Low effort", "High effort")),
    model = factor(model, levels = c(
      "Base model", "Discomfort-only", "Full model", "Humans"
    ))
  )

df_preds_summary <- df_preds |>
  group_by(intimacy, effort, action, model) |>
  tidyboot_mean(p_action) |>
  ungroup()
```

### Model predictions vs human data

P(share saliva) = P(action 2) by effort × intimacy, one panel per model. Lines connect effort levels within an intimacy group; the model panels should reproduce the rise with intimacy and the rise with high effort visible in the Humans panel.

```{r}
#| fig-width: 14
#| fig-asp: 0.32

df_preds_summary |>
  filter(action == 2) |>
  ggplot(aes(x = effort, y = empirical_stat, color = intimacy, group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 4, alpha = PLOT_ALPHA) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), linewidth = 1.2, width = 0.4
  ) +
  geom_line(position = position_dodge(width = 0.2), linewidth = 1.2) +
  scale_color_intimacy() +
  scale_y_continuous(limits = c(0, 1)) +
  labs(y = "P(share saliva)", x = "Effort of not sharing saliva", color = "Intimacy") +
  facet_wrap(~ model, nrow = 1)

ggsave(here("figures", "forw_plan_effort_comparison.pdf"),
       device = cairo_pdf, width = 14, height = 4.5)
```

### Model-human correlation

```{r}
#| label: fig-forw-plan-effort-correlation
#| fig-width: 12
#| fig-asp: 0.4

df_humans_cells <- df_preds_summary |>
  filter(model == "Humans") |>
  select(intimacy, effort, action,
         Humans = empirical_stat,
         Humans_ci_lower = ci_lower,
         Humans_ci_upper = ci_upper)

df_corr <- df_preds_summary |>
  filter(model != "Humans") |>
  select(intimacy, effort, action, model,
         model_pred = empirical_stat,
         model_ci_lower = ci_lower,
         model_ci_upper = ci_upper) |>
  left_join(df_humans_cells, by = c("intimacy", "effort", "action")) |>
  filter(!is.na(model_pred), !is.na(Humans)) |>
  mutate(
    action = factor(action, levels = c(1, 2),
                    labels = c("No saliva", "Share saliva")),
    model = droplevels(model)
  )

correlations <- df_corr |>
  group_by(model) |>
  summarize(boot_result = list(boot_cor(model_pred, Humans)), .groups = "drop") |>
  mutate(
    r = sapply(boot_result, function(x) x$r),
    ci_lower = sapply(boot_result, function(x) x$ci_lower),
    ci_upper = sapply(boot_result, function(x) x$ci_upper),
    label = paste0(
      "r = ", round(r, 2),
      " (", round(ci_lower, 2), ", ", round(ci_upper, 2), ")"
    )
  ) |>
  select(-boot_result)

ggplot(df_corr, aes(x = model_pred, y = Humans, color = intimacy, shape = action)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
  geom_errorbar(aes(ymin = Humans_ci_lower, ymax = Humans_ci_upper), width = 0.02) +
  geom_errorbarh(aes(xmin = model_ci_lower, xmax = model_ci_upper), height = 0.02) +
  geom_point(size = 3, alpha = PLOT_ALPHA) +
  geom_label(
    data = correlations,
    aes(x = -Inf, y = Inf, label = label),
    inherit.aes = FALSE,
    hjust = -0.08, vjust = 1.5,
    size = 4.5, fill = "white", alpha = 0.5, label.size = 0
  ) +
  scale_color_intimacy() +
  scale_shape_manual(values = c(16, 17)) +
  labs(x = "Model", y = "Humans", color = "Intimacy", shape = "Action") +
  facet_wrap(~ model, nrow = 1) +
  coord_fixed_symmetric(df_corr$model_pred, df_corr$Humans)

ggsave(here("figures", "forw_plan_effort_correlation.pdf"),
       device = cairo_pdf, width = 12, height = 4.8)
```

### Model fit results

```{r}
fit_results <- read_csv(here("model", "outputs",
                             "forward_planning_effort_fit_results.csv"),
                        show_col_types = FALSE)

fit_results |>
  select(model, nll, n_params, aic, bic, r, r_ci_lower, r_ci_upper,
         starts_with("param_")) |>
  knitr::kable(digits = 3, caption = "Model fitting results (in-sample)")
```

```{r}
n_obs <- nrow(df)

fit_results |>
  select(model, nll, n_params, aic, bic) |>
  arrange(aic) |>
  knitr::kable(digits = 2,
               caption = "Model comparison (lower AIC/BIC is better)")
```
