---
title: Combined model-human correlation plots, per-scenario points
author: Alicia Chen
date: today
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
knitr:
  opts_chunk:
    dev: "ragg_png"
execute:
  warning: false
---

```{r setup}
source(here::here("analysis", "utils.R"))
setup_analysis()
library(ggnewscale)

INTIMACY_INFERENCE_CONDITIONS <- c("Low", "High")
DESIRE_INFERENCE_CONDITIONS <- c("Intimacy: 0", "Intimacy: 50", "Intimacy: 75", "Intimacy: 100")

DESIRE_INFERENCE_COLORS <- INTIMACY_COLORS
names(DESIRE_INFERENCE_COLORS) <- paste0("Intimacy: ", names(INTIMACY_COLORS))
```

This notebook parallels `inv-plan-combined-correlation.qmd` (and adds forward planning), but each point is one scenario × condition × action cell instead of an average across scenarios. With 16 food scenarios, every panel has roughly 16× the points of the aggregated version. Error bars are dropped since the number of points makes them visually unhelpful; correlations are still computed with bootstrap CIs over the full point cloud.

# Forward planning

## Canonical forward planning (`food_forw_intimacy_desire`)

```{r}
df_fp_cv <- read_csv(here("model", "outputs", "food_forw_intimacy_desire", "cv_preds.csv"), show_col_types = FALSE)

df_fp_human <- df_fp_cv |>
  filter(variant == "full") |>
  group_by(scenario_label = held_out_scenario, intimacy, motivation, action) |>
  summarize(human_p_action = mean(p_action), .groups = "drop")

df_fp_model <- df_fp_cv |>
  mutate(model = case_when(
    variant == "full" ~ "Full model",
    variant == "discomfort_only" ~ "Discomfort-only",
    variant == "base" ~ "Base model"
  )) |>
  group_by(scenario_label = held_out_scenario, intimacy, motivation, action, model) |>
  summarize(model_p_action = mean(p_action_pred), .groups = "drop")

df_fp_corr <- df_fp_model |>
  left_join(df_fp_human, by = c("scenario_label", "intimacy", "motivation", "action")) |>
  mutate(
    intimacy = factor(intimacy, levels = c(0, 50, 75, 100)),
    motivation = factor(motivation, levels = c("low", "high"), labels = MOTIVATION_LEVELS),
    action = factor(action),
    model = factor(model, levels = c("Base model", "Discomfort-only", "Full model"))
  )
```

```{r}
correlations_fp <- format_correlation_labels(
  df_fp_corr, model_p_action, human_p_action, group_vars = "model"
)
```

```{r}
#| fig-width: 11
#| fig-asp: 0.4

ggplot(df_fp_corr, aes(x = model_p_action, y = human_p_action,
                       color = intimacy, shape = action)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
  geom_point(size = 1.5, alpha = 0.55) +
  geom_label(
    data = correlations_fp,
    aes(x = -Inf, y = Inf, label = label),
    inherit.aes = FALSE, hjust = -0.08, vjust = 1.5, size = 5,
    fill = "white", alpha = 0.5, label.size = 0
  ) +
  scale_color_intimacy() +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(x = "Model P(action)", y = "Human P(action)",
       color = "Intimacy", shape = "Action") +
  facet_wrap(~model, nrow = 1) +
  coord_fixed_symmetric(df_fp_corr$model_p_action, df_fp_corr$human_p_action)

ggsave(here("figures", "food_forw_intimacy_desire_correlation_by_scenario.pdf"), device = cairo_pdf)
```

```{r}
correlations_fp |>
  select(model, r, ci_lower, ci_upper) |>
  knitr::kable(
    digits = 3,
    caption = "Per-scenario model-human correlations (canonical forward planning)",
    col.names = c("Model", "r", "CI lower", "CI upper")
  )
```

## Forward planning, effort (`food_forw_intimacy_effort`)

```{r}
df_fpe_cv <- read_csv(here("model", "outputs", "food_forw_intimacy_effort", "cv_preds.csv"), show_col_types = FALSE)

df_fpe_human <- df_fpe_cv |>
  filter(variant == "full") |>
  group_by(scenario_label = held_out_scenario, intimacy, effort, action) |>
  summarize(human_p_action = mean(p_action), .groups = "drop")

df_fpe_model <- df_fpe_cv |>
  mutate(model = case_when(
    variant == "full" ~ "Full model",
    variant == "discomfort_only" ~ "Discomfort-only",
    variant == "base" ~ "Base model"
  )) |>
  group_by(scenario_label = held_out_scenario, intimacy, effort, action, model) |>
  summarize(model_p_action = mean(p_action_pred), .groups = "drop")

EFFORT_ACTION_LEVELS <- c("No saliva", "Share saliva")

df_fpe_corr <- df_fpe_model |>
  left_join(df_fpe_human, by = c("scenario_label", "intimacy", "effort", "action")) |>
  mutate(
    intimacy = factor(intimacy, levels = c(0, 50, 75, 100)),
    effort = factor(effort, levels = c("low", "high"), labels = EFFORT_LEVELS),
    action = factor(action, levels = c(1, 2), labels = EFFORT_ACTION_LEVELS),
    model = factor(model, levels = c("Base model", "Discomfort-only", "Full model"))
  )
```

```{r}
correlations_fpe <- format_correlation_labels(
  df_fpe_corr, model_p_action, human_p_action, group_vars = "model"
)
```

```{r}
#| fig-width: 11
#| fig-asp: 0.4

ggplot(df_fpe_corr, aes(x = model_p_action, y = human_p_action,
                        color = intimacy, shape = action)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
  geom_point(size = 1.5, alpha = 0.55) +
  geom_label(
    data = correlations_fpe,
    aes(x = -Inf, y = Inf, label = label),
    inherit.aes = FALSE, hjust = -0.08, vjust = 1.5, size = 5,
    fill = "white", alpha = 0.5, label.size = 0
  ) +
  scale_color_intimacy() +
  scale_shape_manual(values = c(4, 8)) +
  labs(x = "Model P(action)", y = "Human P(action)",
       color = "Intimacy", shape = "Action") +
  facet_wrap(~model, nrow = 1) +
  coord_fixed_symmetric(df_fpe_corr$model_p_action, df_fpe_corr$human_p_action)

ggsave(here("figures", "food_forw_intimacy_effort_correlation_by_scenario.pdf"), device = cairo_pdf)
```

```{r}
correlations_fpe |>
  select(model, r, ci_lower, ci_upper) |>
  knitr::kable(
    digits = 3,
    caption = "Per-scenario model-human correlations (forward planning, effort)",
    col.names = c("Model", "r", "CI lower", "CI upper")
  )
```

# Inverse planning

## Load `food_inv_intimacy_desire_alt` (intimacy inference) data

### Human data

```{r}
df_intimacy <- read_csv(here("data", "legacy", "food_inv_intimacy_desire_alt", "main_trials_long.csv")) |>
  mutate(
    action = str_replace(action_condition, "action_", ""),
    motivation = factor(motivation, levels = c("low", "high"), labels = MOTIVATION_LEVELS),
    stage = factor(stage, levels = c("prior", "posterior"))
  ) |>
  group_by(subject_id, scenario_label) |>
  mutate(
    belief_update = ifelse(
      stage == "posterior",
      intimacy_rating[stage == "posterior"] - intimacy_rating[stage == "prior"],
      NA
    )
  ) |>
  ungroup()

df_intimacy_summary <- df_intimacy |>
  filter(stage == "posterior") |>
  group_by(scenario_label, motivation, action) |>
  summarize(human_belief_update = mean(belief_update, na.rm = TRUE), .groups = "drop")
```

### Model predictions

```{r}
df_model_intimacy <- read_csv(here("model", "outputs", "food_inv_intimacy_desire_alt", "cv_preds_summary.csv")) |>
  rename(motivation = reward_condition) |>
  mutate(
    action = as.character(action),
    motivation = factor(motivation, levels = c("low", "high"), labels = MOTIVATION_LEVELS),
    model_belief_update = expected_intimacy - 50,
    model = case_when(
      model == "full" ~ "Full model",
      model == "discomfort_only" ~ "Discomfort-only",
      model == "base" ~ "Base model"
    )
  ) |>
  filter(!is.na(model)) |>
  select(scenario_label, motivation, action, model, model_belief_update)
```

## Load `food_inv_desire_intimacy_alt` (desire inference) data

### Human data

```{r}
df_desire <- read_csv(here("data", "legacy", "food_inv_desire_intimacy_alt", "main_trials_long.csv")) |>
  mutate(
    action = str_replace(action_condition, "action_", ""),
    intimacy = factor(intimacy),
    p_high_reward = response / 100,
    stage = factor(stage, levels = c("prior", "posterior"))
  ) |>
  group_by(subject_id, scenario_label) |>
  mutate(
    belief_update = ifelse(
      stage == "posterior",
      p_high_reward[stage == "posterior"] - p_high_reward[stage == "prior"],
      NA
    )
  ) |>
  ungroup()

df_desire_summary <- df_desire |>
  filter(stage == "posterior") |>
  group_by(scenario_label, intimacy, action) |>
  summarize(human_belief_update = mean(belief_update, na.rm = TRUE), .groups = "drop")
```

### Model predictions

```{r}
df_model_desire <- read_csv(here("model", "outputs", "food_inv_desire_intimacy_alt", "cv_preds_summary.csv")) |>
  rename(intimacy = intimacy_condition) |>
  mutate(
    action = as.character(action),
    intimacy = factor(intimacy),
    model_belief_update = (p_high_reward / 100) - 0.5,
    model = case_when(
      model == "full" ~ "Full model",
      model == "discomfort_only" ~ "Discomfort-only",
      model == "base" ~ "Base model"
    )
  ) |>
  filter(!is.na(model)) |>
  select(scenario_label, intimacy, action, model, model_belief_update)
```

## Combine and rescale

```{r}
df_corr_intimacy <- df_intimacy_summary |>
  left_join(df_model_intimacy, by = c("scenario_label", "motivation", "action")) |>
  mutate(
    experiment = "Intimacy inference",
    human_belief_update = human_belief_update / 100,
    model_belief_update = model_belief_update / 100
  )

df_corr_desire <- df_desire_summary |>
  left_join(df_model_desire, by = c("scenario_label", "intimacy", "action")) |>
  mutate(experiment = "Desire inference")

df_combined <- bind_rows(
  df_corr_intimacy |>
    mutate(condition = as.character(motivation)) |>
    select(scenario_label, action, model, experiment, condition, human_belief_update, model_belief_update),
  df_corr_desire |>
    mutate(condition = paste0("Intimacy: ", intimacy)) |>
    select(scenario_label, action, model, experiment, condition, human_belief_update, model_belief_update)
) |>
  mutate(
    model = factor(model, levels = c("Base model", "Discomfort-only", "Full model")),
    action = factor(action),
    condition = factor(condition, levels = c(
      "Low", "High",
      "Intimacy: 0", "Intimacy: 50", "Intimacy: 75", "Intimacy: 100"
    ))
  )
```

## Compute correlations

```{r}
correlations_combined <- format_correlation_labels(
  df_combined, model_belief_update, human_belief_update, group_vars = "model"
)
```

## Per-scenario correlation plot

```{r}
#| fig-width: 11
#| fig-asp: 0.4

df_plot <- df_combined

ggplot(df_plot, aes(x = model_belief_update, y = human_belief_update, shape = action)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
  geom_point(
    data = \(d) filter(d, condition %in% INTIMACY_INFERENCE_CONDITIONS),
    aes(color = condition), size = 1.5, alpha = 0.55
  ) +
  scale_color_manual(name = "Intimacy inference", values = MOTIVATION_COLORS, guide = guide_legend(order = 2)) +
  new_scale_color() +
  geom_point(
    data = \(d) filter(d, condition %in% DESIRE_INFERENCE_CONDITIONS),
    aes(color = condition), size = 1.5, alpha = 0.55
  ) +
  scale_color_manual(
    name = "Desire inference",
    values = DESIRE_INFERENCE_COLORS,
    labels = \(x) gsub("Intimacy: (\\d+)", "\\1 intimacy", x),
    guide = guide_legend(order = 3)
  ) +
  geom_label(
    data = correlations_combined,
    aes(x = -Inf, y = Inf, label = label),
    inherit.aes = FALSE, hjust = -0.08, vjust = 1.5, size = 5,
    fill = "white", alpha = 0.5, label.size = 0
  ) +
  scale_shape_manual(values = c(16, 17, 15, 18), guide = guide_legend(order = 1)) +
  labs(x = "Model belief update", y = "Human belief update", shape = "Action") +
  theme(
    legend.position = "right",
    legend.spacing.y = unit(0, "lines"),
  ) +
  facet_wrap(~model, nrow = 1) +
  coord_fixed_symmetric(df_plot$model_belief_update, df_plot$human_belief_update)

ggsave(here("figures", "inv_plan_combined_correlation_by_scenario_main.pdf"), device = cairo_pdf)
```

## Summary table

```{r}
correlations_combined |>
  select(model, r, ci_lower, ci_upper) |>
  knitr::kable(
    digits = 3,
    caption = "Per-scenario model-human correlations (intimacy + desire inference, alt-shown)",
    col.names = c("Model", "r", "CI lower", "CI upper")
  )
```

# Effort experiments combined

```{r}
EFFORT_ACTION_LEVELS <- c("No saliva", "Share saliva")
EXCLUDED_SUBJECTS_EFFORT <- c("daeeb393-ddef-5529-804a-36dbdd257d5c")
```

## Load `food_inv_intimacy_effort_alt` (intimacy inference, effort) data

### Human data

```{r}
df_intimacy_effort <- read_csv(here("data", "legacy", "food_inv_intimacy_effort_alt", "main_trials_long.csv")) |>
  filter(!subject_id %in% EXCLUDED_SUBJECTS_EFFORT) |>
  mutate(
    action = factor(action_condition, levels = c("action_1", "action_2"),
                    labels = EFFORT_ACTION_LEVELS),
    effort = factor(effort, levels = c("low", "high"), labels = EFFORT_LEVELS),
    stage = factor(stage, levels = c("prior", "posterior"))
  ) |>
  group_by(subject_id, scenario_label) |>
  mutate(
    belief_update = ifelse(
      stage == "posterior",
      intimacy_rating[stage == "posterior"] - intimacy_rating[stage == "prior"],
      NA
    )
  ) |>
  ungroup()

df_intimacy_effort_summary <- df_intimacy_effort |>
  filter(stage == "posterior") |>
  group_by(scenario_label, effort, action) |>
  summarize(human_belief_update = mean(belief_update, na.rm = TRUE), .groups = "drop")
```

### Model predictions

```{r}
df_model_intimacy_effort <- read_csv(here("model", "outputs", "food_inv_intimacy_effort_alt", "cv_preds_summary.csv")) |>
  rename(effort = effort_condition) |>
  mutate(
    action = factor(action, levels = c(1, 2), labels = EFFORT_ACTION_LEVELS),
    effort = factor(effort, levels = c("low", "high"), labels = EFFORT_LEVELS),
    model_belief_update = expected_intimacy - 50,
    model = case_when(
      model == "full" ~ "Full model",
      model == "discomfort_only" ~ "Discomfort-only",
      model == "base" ~ "Base model"
    )
  ) |>
  filter(!is.na(model)) |>
  select(scenario_label, effort, action, model, model_belief_update)
```

## Load `food_inv_effort_intimacy_alt` (effort inference) data

### Human data

```{r}
df_effort_intimacy <- read_csv(here("data", "legacy", "food_inv_effort_intimacy_alt", "main_trials_long.csv")) |>
  mutate(
    action = factor(action_condition, levels = c("action_1", "action_2"),
                    labels = EFFORT_ACTION_LEVELS),
    intimacy = factor(intimacy, levels = c(0, 50, 75, 100)),
    p_effort_high = response / 100,
    stage = factor(stage, levels = c("prior", "posterior"))
  ) |>
  group_by(subject_id, scenario_label) |>
  mutate(
    belief_update = ifelse(
      stage == "posterior",
      p_effort_high[stage == "posterior"] - p_effort_high[stage == "prior"],
      NA
    )
  ) |>
  ungroup()

df_effort_intimacy_summary <- df_effort_intimacy |>
  filter(stage == "posterior") |>
  group_by(scenario_label, intimacy, action) |>
  summarize(human_belief_update = mean(belief_update, na.rm = TRUE), .groups = "drop")
```

### Model predictions

```{r}
df_model_effort_intimacy <- read_csv(here("model", "outputs", "food_inv_effort_intimacy_alt", "cv_preds_summary.csv")) |>
  mutate(
    action = factor(action, levels = c(1, 2), labels = EFFORT_ACTION_LEVELS),
    intimacy = factor(intimacy, levels = c(0, 50, 75, 100)),
    model_belief_update = p_effort_high - 0.5,
    model = case_when(
      model == "full" ~ "Full model",
      model == "discomfort_only" ~ "Discomfort-only",
      model == "base" ~ "Base model"
    )
  ) |>
  filter(!is.na(model)) |>
  select(scenario_label, intimacy, action, model, model_belief_update)
```

## Combine and rescale

```{r}
df_corr_intimacy_effort <- df_intimacy_effort_summary |>
  left_join(df_model_intimacy_effort, by = c("scenario_label", "effort", "action")) |>
  mutate(
    experiment = "Inv plan effort",
    human_belief_update = human_belief_update / 100,
    model_belief_update = model_belief_update / 100
  )

df_corr_effort_intimacy <- df_effort_intimacy_summary |>
  left_join(df_model_effort_intimacy, by = c("scenario_label", "intimacy", "action")) |>
  mutate(experiment = "Inv plan effort inferred")

df_combined_effort <- bind_rows(
  df_corr_intimacy_effort |>
    mutate(condition = as.character(effort)) |>
    select(scenario_label, action, model, experiment, condition, human_belief_update, model_belief_update),
  df_corr_effort_intimacy |>
    mutate(condition = paste0("Intimacy: ", intimacy)) |>
    select(scenario_label, action, model, experiment, condition, human_belief_update, model_belief_update)
) |>
  mutate(
    model = factor(model, levels = c("Base model", "Discomfort-only", "Full model")),
    action = factor(action, levels = EFFORT_ACTION_LEVELS),
    condition = factor(condition, levels = c(
      "Low effort", "High effort",
      "Intimacy: 0", "Intimacy: 50", "Intimacy: 75", "Intimacy: 100"
    ))
  )
```

## Compute correlations

```{r}
correlations_combined_effort <- format_correlation_labels(
  df_combined_effort, model_belief_update, human_belief_update, group_vars = "model"
)
```

## Per-scenario correlation plot

```{r}
#| fig-width: 11
#| fig-asp: 0.4

df_plot_effort <- df_combined_effort

ggplot(df_plot_effort, aes(x = model_belief_update, y = human_belief_update, shape = action)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
  geom_point(
    data = \(d) filter(d, condition %in% EFFORT_LEVELS),
    aes(color = condition), size = 1.5, alpha = 0.55
  ) +
  scale_color_manual(name = "Effort intimacy inference", values = EFFORT_COLORS,
                     guide = guide_legend(order = 2)) +
  new_scale_color() +
  geom_point(
    data = \(d) filter(d, condition %in% DESIRE_INFERENCE_CONDITIONS),
    aes(color = condition), size = 1.5, alpha = 0.55
  ) +
  scale_color_manual(
    name = "Effort inference",
    values = DESIRE_INFERENCE_COLORS,
    labels = \(x) gsub("Intimacy: (\\d+)", "\\1 intimacy", x),
    guide = guide_legend(order = 3)
  ) +
  geom_label(
    data = correlations_combined_effort,
    aes(x = -Inf, y = Inf, label = label),
    inherit.aes = FALSE, hjust = -0.08, vjust = 1.5, size = 5,
    fill = "white", alpha = 0.5, label.size = 0
  ) +
  scale_shape_manual(values = c(16, 17), guide = guide_legend(order = 1)) +
  labs(x = "Model belief update", y = "Human belief update", shape = "Action") +
  theme(
    legend.position = "right",
    legend.spacing.y = unit(0, "lines"),
  ) +
  facet_wrap(~model, nrow = 1) +
  coord_fixed_symmetric(df_plot_effort$model_belief_update, df_plot_effort$human_belief_update)

ggsave(here("figures", "food_inv_intimacy_effort_alt_combined_correlation_by_scenario_main.pdf"), device = cairo_pdf)
```

## Summary table

```{r}
correlations_combined_effort |>
  select(model, r, ci_lower, ci_upper) |>
  knitr::kable(
    digits = 3,
    caption = "Per-scenario model-human correlations (both effort experiments combined)",
    col.names = c("Model", "r", "CI lower", "CI upper")
  )
```

# All inverse planning experiments combined

```{r}
df_combined_all <- bind_rows(
  df_corr_intimacy |>
    mutate(condition = as.character(motivation), action = as.character(action)) |>
    select(scenario_label, action, model, experiment, condition, human_belief_update, model_belief_update),
  df_corr_desire |>
    mutate(condition = paste0("Intimacy: ", intimacy), action = as.character(action)) |>
    select(scenario_label, action, model, experiment, condition, human_belief_update, model_belief_update),
  df_corr_intimacy_effort |>
    mutate(condition = as.character(effort), action = as.character(action)) |>
    select(scenario_label, action, model, experiment, condition, human_belief_update, model_belief_update),
  df_corr_effort_intimacy |>
    mutate(condition = paste0("Intimacy: ", intimacy), action = as.character(action)) |>
    select(scenario_label, action, model, experiment, condition, human_belief_update, model_belief_update)
) |>
  mutate(
    model = factor(model, levels = c("Base model", "Discomfort-only", "Full model")),
    action = factor(action, levels = c("0", "1", "2", "3", "No saliva", "Share saliva")),
    condition = factor(condition, levels = c(
      "Low", "High",
      "Intimacy: 0", "Intimacy: 50", "Intimacy: 75", "Intimacy: 100",
      "Low effort", "High effort"
    ))
  )
```

## Compute correlations

```{r}
correlations_combined_all <- format_correlation_labels(
  df_combined_all, model_belief_update, human_belief_update, group_vars = "model"
)
```

## Per-scenario correlation plot

```{r}
#| fig-width: 13
#| fig-asp: 0.45

df_plot_all <- df_combined_all

ggplot(df_plot_all, aes(x = model_belief_update, y = human_belief_update, shape = action)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
  geom_point(
    data = \(d) filter(d, condition %in% INTIMACY_INFERENCE_CONDITIONS),
    aes(color = condition), size = 1.2, alpha = 0.5
  ) +
  scale_color_manual(name = "Intimacy inference", values = MOTIVATION_COLORS,
                     guide = guide_legend(order = 2)) +
  new_scale_color() +
  geom_point(
    data = \(d) filter(d, condition %in% DESIRE_INFERENCE_CONDITIONS),
    aes(color = condition), size = 1.2, alpha = 0.5
  ) +
  scale_color_manual(
    name = "Desire / effort inference",
    values = DESIRE_INFERENCE_COLORS,
    labels = \(x) gsub("Intimacy: (\\d+)", "\\1 intimacy", x),
    guide = guide_legend(order = 3)
  ) +
  new_scale_color() +
  geom_point(
    data = \(d) filter(d, condition %in% EFFORT_LEVELS),
    aes(color = condition), size = 1.2, alpha = 0.5
  ) +
  scale_color_manual(name = "Effort intimacy inference", values = EFFORT_COLORS,
                     guide = guide_legend(order = 4)) +
  geom_label(
    data = correlations_combined_all,
    aes(x = -Inf, y = Inf, label = label),
    inherit.aes = FALSE, hjust = -0.08, vjust = 1.5, size = 5,
    fill = "white", alpha = 0.5, label.size = 0
  ) +
  scale_shape_manual(
    values = c("0" = 16, "1" = 17, "2" = 15, "3" = 18,
               "No saliva" = 4, "Share saliva" = 8),
    guide = guide_legend(order = 1)
  ) +
  labs(x = "Model belief update", y = "Human belief update", shape = "Action") +
  theme(
    legend.position = "right",
    legend.spacing.y = unit(0, "lines"),
  ) +
  facet_wrap(~model, nrow = 1) +
  coord_fixed_symmetric(df_plot_all$model_belief_update, df_plot_all$human_belief_update)

ggsave(here("figures", "inv_plan_all_combined_correlation_by_scenario_main.pdf"), device = cairo_pdf)
```

## Summary table

```{r}
correlations_combined_all |>
  select(model, r, ci_lower, ci_upper) |>
  knitr::kable(
    digits = 3,
    caption = "Per-scenario model-human correlations (all four inverse planning experiments)",
    col.names = c("Model", "r", "CI lower", "CI upper")
  )
```
