---
title: Combined alt-shown inverse planning correlation plots (intimacy + desire)
author: Alicia Chen
date: today
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
knitr:
  opts_chunk:
    dev: "ragg_png"
execute:
  warning: false
---

```{r setup}
source(here::here("analysis", "utils.R"))
setup_analysis()
library(ggnewscale)

# Define condition sets for each experiment
INTIMACY_INFERENCE_CONDITIONS <- c("Low", "High")
DESIRE_INFERENCE_CONDITIONS <- c("Intimacy: 0", "Intimacy: 50", "Intimacy: 75", "Intimacy: 100")

# Create intimacy colors with correct names for this analysis
DESIRE_INFERENCE_COLORS <- INTIMACY_COLORS
names(DESIRE_INFERENCE_COLORS) <- paste0("Intimacy: ", names(INTIMACY_COLORS))
```

This notebook aggregates model-human correlations across both alt-shown inverse planning experiments:
- **`inv_plan_intimacy_alt`**: Intimacy inference (infer relationship closeness from observed action)
- **`inv_plan_desire_alt`**: Desire inference (infer motivation from observed action)

## Load `inv_plan_intimacy_alt` (Intimacy Inference) Data

### Human data

```{r}
df_intimacy <- read_csv(here("data", "inv_plan_intimacy_alt", "main_trials_long.csv")) |>
  mutate(
    action = str_replace(action_condition, "action_", ""),
    motivation = factor(motivation, levels = c("low", "high"), labels = MOTIVATION_LEVELS),
    stage = factor(stage, levels = c("prior", "posterior"))
  ) |>
  group_by(subject_id, scenario_label) |>
  mutate(
    belief_update = ifelse(
      stage == "posterior",
      intimacy_rating[stage == "posterior"] - intimacy_rating[stage == "prior"],
      NA
    )
  ) |>
  ungroup()

df_intimacy_summary <- df_intimacy |>
  filter(stage == "posterior") |>
  group_by(motivation, action) |>
  tidyboot_mean(belief_update) |>
  rename(human_belief_update = empirical_stat)
```

### Model predictions

```{r}
df_model_intimacy <- read_csv(here("model", "outputs", "inv_plan_intimacy_alt_preds_summary.csv")) |>
  rename(motivation = reward_condition) |>
  mutate(
    action = as.character(action),
    motivation = factor(motivation, levels = c("low", "high"), labels = MOTIVATION_LEVELS),
    model_belief_update = expected_intimacy - 50,
    model = case_when(
      model == "access_full" ~ "Full model",
      model == "access_only" ~ "Discomfort-only",
      model == "no_access" ~ "Base model"
    )
  ) |>
  group_by(motivation, action, model) |>
  summarize(model_belief_update = mean(model_belief_update), .groups = "drop")
```

## Load `inv_plan_desire_alt` (Desire Inference) Data

### Human data

```{r}
df_desire <- read_csv(here("data", "inv_plan_desire_alt", "main_trials_long.csv")) |>
  mutate(
    action = str_replace(action_condition, "action_", ""),
    intimacy = factor(intimacy),
    p_high_reward = response / 100,
    stage = factor(stage, levels = c("prior", "posterior"))
  ) |>
  group_by(subject_id, scenario_label) |>
  mutate(
    belief_update = ifelse(
      stage == "posterior",
      p_high_reward[stage == "posterior"] - p_high_reward[stage == "prior"],
      NA
    )
  ) |>
  ungroup()

df_desire_summary <- df_desire |>
  filter(stage == "posterior") |>
  group_by(intimacy, action) |>
  tidyboot_mean(belief_update) |>
  rename(human_belief_update = empirical_stat)
```

### Model predictions

```{r}
df_model_desire <- read_csv(here("model", "outputs", "inv_plan_desire_alt_preds_summary.csv")) |>
  rename(intimacy = intimacy_condition) |>
  mutate(
    action = as.character(action),
    intimacy = factor(intimacy),
    model_belief_update = (p_high_reward / 100) - 0.5,
    model = case_when(
      model == "access_full" ~ "Full model",
      model == "access_only" ~ "Discomfort-only",
      model == "no_access" ~ "Base model"
    )
  ) |>
  group_by(intimacy, action, model) |>
  summarize(model_belief_update = mean(model_belief_update), .groups = "drop")
```

## Combine and Rescale

We rescale the intimacy-inference belief updates by dividing by 100 to put both experiments on a -1 to 1 scale:
- Intimacy inference: intimacy ratings (0-100), so belief updates are on ~-50 to +50 scale → ÷100 gives -0.5 to +0.5
- Desire inference: probabilities (0-1), so belief updates are already on -0.5 to +0.5 scale

```{r}
# Combine human and model for intimacy inference (rescale from 0-100 to 0-1)
df_corr_intimacy <- df_intimacy_summary |>
  left_join(df_model_intimacy, by = c("motivation", "action")) |>
  mutate(
    experiment = "Intimacy inference",
    # Rescale from percentage scale to proportion scale
    human_belief_update = human_belief_update / 100,
    ci_lower = ci_lower / 100,
    ci_upper = ci_upper / 100,
    model_belief_update = model_belief_update / 100
  )

# Combine human and model for desire inference (already on 0-1 scale)
df_corr_desire <- df_desire_summary |>
  left_join(df_model_desire, by = c("intimacy", "action")) |>
  mutate(experiment = "Desire inference")

# Combine both experiments
df_combined <- bind_rows(
  df_corr_intimacy |>
    mutate(condition = as.character(motivation)) |>
    select(action, model, experiment, condition, human_belief_update, ci_lower, ci_upper, model_belief_update),
  df_corr_desire |>
    mutate(condition = paste0("Intimacy: ", intimacy)) |>
    select(action, model, experiment, condition, human_belief_update, ci_lower, ci_upper, model_belief_update)
) |>
  mutate(
    model = factor(model, levels = c("Base model", "Discomfort-only", "Full model")),
    action = factor(action),
    condition = factor(condition, levels = c(
      "Low", "High",
      "Intimacy: 0", "Intimacy: 50", "Intimacy: 75", "Intimacy: 100"
    ))
  )
```

## Compute Correlations

```{r}
correlations_combined <- df_combined |>
  group_by(model) |>
  summarize(
    boot_result = list(boot_cor(model_belief_update, human_belief_update)),
    .groups = "drop"
  ) |>
  mutate(
    r = sapply(boot_result, function(x) x$r),
    ci_lower = sapply(boot_result, function(x) x$ci_lower),
    ci_upper = sapply(boot_result, function(x) x$ci_upper),
    label = paste0("r = ", round(r, 2), " (", round(ci_lower, 2), ", ", round(ci_upper, 2), ")")
  ) |>
  select(-boot_result)
```

## Aggregated Correlation Plot

```{r}
#| fig-width: 11
#| fig-asp: 0.4

df_plot <- df_combined

ggplot(df_plot, aes(x = model_belief_update, y = human_belief_update, shape = action)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
  # Intimacy-inference points (motivation)
  geom_errorbar(
    data = \(d) filter(d, condition %in% INTIMACY_INFERENCE_CONDITIONS),
    aes(ymin = ci_lower, ymax = ci_upper, color = condition), width = 0.02
  ) +
  geom_point(
    data = \(d) filter(d, condition %in% INTIMACY_INFERENCE_CONDITIONS),
    aes(color = condition), size = 3, alpha = PLOT_ALPHA
  ) +
  scale_color_manual(name = "Intimacy inference", values = MOTIVATION_COLORS, guide = guide_legend(order = 2)) +
  new_scale_color() +
  # Desire-inference points (intimacy)
  geom_errorbar(
    data = \(d) filter(d, condition %in% DESIRE_INFERENCE_CONDITIONS),
    aes(ymin = ci_lower, ymax = ci_upper, color = condition), width = 0.02
  ) +
  geom_point(
    data = \(d) filter(d, condition %in% DESIRE_INFERENCE_CONDITIONS),
    aes(color = condition), size = 3, alpha = PLOT_ALPHA
  ) +
  scale_color_manual(
    name = "Desire inference",
    values = DESIRE_INFERENCE_COLORS,
    labels = \(x) gsub("Intimacy: (\\d+)", "\\1 intimacy", x),
    guide = guide_legend(order = 3)
  ) +
  # Correlation labels
  geom_label(
    data = correlations_combined,
    aes(x = -Inf, y = Inf, label = label),
    inherit.aes = FALSE, hjust = -0.08, vjust = 1.5, size = 5,
    fill = "white", alpha = 0.5, label.size = 0
  ) +
  scale_shape_manual(values = c(16, 17, 15, 18), guide = guide_legend(order = 1)) +
  labs(x = "Model belief update", y = "Human belief update", shape = "Action") +

  theme(
    legend.position = "right",
    legend.spacing.y = unit(0, "lines"),
  ) +
  facet_wrap(~model, nrow = 1) +
  coord_fixed_symmetric(df_plot$model_belief_update, df_plot$human_belief_update)

ggsave(here("figures", "inv_plan_combined_correlation_main.pdf"), device = cairo_pdf)
```

## Summary Table

```{r}
correlations_combined |>
  select(model, r, ci_lower, ci_upper) |>
  knitr::kable(
    digits = 3,
    caption = "Aggregated model-human correlations (intimacy + desire inference, alt-shown)",
    col.names = c("Model", "r", "CI lower", "CI upper")
  )
```
