---
title: Inverse planning — desire inference, no alternatives shown (data analysis)
author: Alicia Chen
date: 2026-04-26
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    page-layout: full
    search: false
knitr:
  opts_chunk:
    dev: "ragg_png"
execute:
  warning: false
---

```{r setup}
source(here::here("analysis", "utils.R"))
setup_analysis()
```

## Demographics

```{r}
report_demographics("inv_plan_desire_noalt")
```

## Results

This experiment is a variant of `inv_plan_desire_alt` in which participants observe a single action and infer the actor's desire **without seeing the other three alternative actions the actors could have taken**. The slider endpoints are the scenario's `reward_low` / `reward_high` paragraphs, so the response encodes P(high desire). The experiment manipulates:

- **Action**: Which action was taken (action 0–3, varying in saliva-transfer risk)
- **Intimacy**: The relationship closeness between the actors (0, 50, 75, or 100)

This notebook reports the raw human data only — modeling is not included.

## Load human data

```{r}
# main_trials_long.csv is generated by analysis/json_to_csv.py
# It excludes participants who failed attention check or got 0 on memory check
df <- read_csv(here("data", "inv_plan_desire_noalt", "main_trials_long.csv")) |>
  mutate(
    action = str_replace(action_condition, "action_", ""),
    intimacy = factor(intimacy),
    p_high_reward = response / 100,  # Convert to probability scale
    stage = factor(stage, levels = c("prior", "posterior")),
    model = "Humans"
  ) |>
  select(-action_condition, -response)

# Calculate belief updates
df <- calculate_belief_update(df, "p_high_reward")

cat("Trials per participant:", nrow(df) / n_distinct(df$subject_id), "\n")
cat("N participants (after exclusions):", n_distinct(df$subject_id), "\n")
cat("N scenarios:", n_distinct(df$scenario_label), "\n")
```

## Distribution of ratings

```{r}
#| fig-width: 12
#| fig-asp: 0.45

ggplot(df, aes(x = p_high_reward, fill = action)) +
  geom_density(alpha = 0.5) +
  scale_fill_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  labs(y = "Density", x = "P(high desire)", fill = "Action") +
  theme(legend.position = "right") +
  facet_grid(stage ~ intimacy, labeller = labeller(intimacy = label_both, stage = label_value))
```

## Summary statistics

```{r}
df_summary <- df |>
  group_by(intimacy, action, stage, model) |>
  tidyboot_mean(p_high_reward)
```

## Prior vs posterior by action and intimacy

```{r}
#| fig-width: 12
#| fig-asp: 0.7

ggplot() +
  geom_violin(
    data = df,
    aes(x = stage, y = p_high_reward, fill = action),
    position = position_dodge(width = 0.8),
    alpha = 0.5,
    width = 0.7
  ) +
  geom_point(
    data = df_summary,
    aes(x = stage, y = empirical_stat, color = action, group = action),
    position = position_dodge(width = 0.8),
    size = 3,
    alpha = PLOT_ALPHA
  ) +
  geom_errorbar(
    data = df_summary,
    aes(x = stage, ymin = ci_lower, ymax = ci_upper, color = action, group = action),
    position = position_dodge(width = 0.8),
    width = 0.1,
    size = 1
  ) +
  geom_line(
    data = df_summary,
    aes(x = stage, y = empirical_stat, color = action, group = action),
    position = position_dodge(width = 0.8),
    size = 1
  ) +
  facet_grid(intimacy ~ action, labeller = label_both) +
  scale_fill_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  scale_color_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  labs(
    title = "P(high desire) before and after observing action (no-alt)",
    y = "P(high desire)",
    x = "Stage",
    fill = "Observed action",
    color = "Observed action"
  )
```

## Belief updates

```{r}
df_belief_update_summary <- df |>
  filter(stage == "posterior") |>
  group_by(intimacy, action, model) |>
  tidyboot_mean(belief_update)
```

```{r}
#| fig-width: 10
#| fig-asp: 0.45

df_belief_update_summary |>
  ggplot(aes(x = action, y = empirical_stat, fill = intimacy, group = intimacy)) +
  geom_col(position = position_dodge(width = 0.9), alpha = PLOT_ALPHA) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.9), width = 0.1, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_intimacy() +
  labs(
    title = "Belief updates by action and intimacy (no-alt humans)",
    y = "Belief update (posterior − prior, on P scale)",
    x = "Observed action",
    fill = "Intimacy"
  ) +
  theme(legend.position = "right")
```

## Belief updates vs LM-derived risk

X-axis is the LM-derived risk score for the observed action (from `lm_scenario_params.csv`). One point per (scenario, action, intimacy), with a linear smoother per intimacy level.

```{r}
#| fig-width: 9
#| fig-asp: 0.55

df_canonical_risk <- read_csv(here("model", "outputs", "lm_scenario_params.csv"), show_col_types = FALSE) |>
  mutate(action = as.character(action)) |>
  select(scenario_label, action, access) |>
  rename(risk = access)

df_human_scenario_pts <- df |>
  filter(stage == "posterior") |>
  group_by(scenario_label, intimacy, action) |>
  tidyboot_mean(belief_update) |>
  ungroup() |>
  mutate(belief_update = empirical_stat) |>
  select(scenario_label, intimacy, action, belief_update, ci_lower, ci_upper) |>
  inner_join(df_canonical_risk, by = c("scenario_label", "action"))

df_human_scenario_pts |>
  ggplot(aes(x = risk, y = belief_update, color = intimacy, shape = factor(action), group = intimacy)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = POS_JITTER_DODGE,
    width = 0.05, alpha = 0.4, linewidth = 0.4
  ) +
  geom_point(position = POS_JITTER_DODGE, alpha = 0.7, size = 2) +
  geom_smooth(se = FALSE, linewidth = 1) +
  scale_color_intimacy() +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(y = "Belief update", x = "LM-derived risk of observed action",
       color = "Intimacy", shape = "Action") +
  theme(legend.position = "right")
```

## Scenario-level

### Human belief updates by scenario

```{r}
df_belief_update_scenario <- df |>
  filter(stage == "posterior") |>
  group_by(scenario_label, intimacy, action, model) |>
  tidyboot_mean(belief_update)
```

```{r}
#| fig-width: 12
#| fig-asp: 0.85

ggplot(df_belief_update_scenario, aes(x = action, y = empirical_stat, fill = intimacy)) +
  geom_col(position = position_dodge(width = 0.9), alpha = PLOT_ALPHA) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.9), width = 0.1, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_intimacy() +
  labs(
    title = "Belief updates by scenario (no-alt humans)",
    y = "Belief update",
    x = "Observed action",
    fill = "Intimacy"
  ) +
  theme(legend.position = "right") +
  facet_wrap(~scenario_label, ncol = 4)
```

### Posterior P(high desire) by scenario

```{r}
df_posterior_scenario <- df |>
  filter(stage == "posterior") |>
  group_by(scenario_label, intimacy, action) |>
  tidyboot_mean(p_high_reward)
```

```{r}
#| fig-width: 12
#| fig-asp: 0.85

ggplot(df_posterior_scenario, aes(x = action, y = empirical_stat, fill = intimacy)) +
  geom_col(position = position_dodge(width = 0.9), alpha = PLOT_ALPHA) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.9), width = 0.1, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey") +
  scale_fill_intimacy() +
  labs(
    title = "Posterior P(high desire) by scenario (no-alt humans)",
    y = "Posterior P(high desire)",
    x = "Observed action",
    fill = "Intimacy"
  ) +
  theme(legend.position = "right") +
  facet_wrap(~scenario_label, ncol = 4)
```
