---
title: Inverse planning — desire inference, alternatives shown (data analysis)
author: Alicia Chen
date: 2026-01-15
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    page-layout: full
    search: false
knitr:
  opts_chunk:
    dev: "ragg_png"
execute:
  warning: false
---

```{r setup}
source(here::here("analysis", "utils.R"))
setup_analysis()
```

## Demographics

```{r}
df_exit <- read_csv(here("data", "inv_plan_desire_alt", "exit_survey.csv"), show_col_types = FALSE)

n_total <- nrow(df_exit)
n_passed <- df_exit |>
  filter(attention_passed == TRUE, memory_correct_count > 0) |>
  nrow()

cat("Total participants recruited:", n_total, "\n")
cat("Passed attention checks:", n_passed, "\n")
cat("Mean age:", round(mean(df_exit$age, na.rm = TRUE), 1),
    "SD age:", round(sd(df_exit$age, na.rm = TRUE), 1),
    "Min age:", min(df_exit$age, na.rm = TRUE),
    "Max age:", max(df_exit$age, na.rm = TRUE))
cat("\nGender:\n")
print(table(df_exit$gender))
```

## Results

This experiment presents participants with scenarios where they observe an action taken by one person toward another. Participants rate their belief about the probability that the actors have high motivation (want the food) before (prior) and after (posterior) observing the action. The experiment manipulates:
- **Action**: Which action was taken (action 0-3, varying in saliva-transfer risk)
- **Intimacy**: The relationship closeness between the actors (0, 50, 75, or 100)

## Load human data

```{r}
# main_trials_long.csv is generated by analysis/json_to_csv.py
# It excludes participants who failed attention check or got 0 on memory check
df <- read_csv(here("data", "inv_plan_desire_alt", "main_trials_long.csv")) |>
  mutate(
    action = str_replace(action_condition, "action_", ""),
    intimacy = factor(intimacy),
    p_high_reward = response / 100,  # Convert to probability scale
    stage = factor(stage, levels = c("prior", "posterior")),
    model = "Humans"
  ) |>
  select(-action_condition, -response)

# Calculate belief updates
df <- calculate_belief_update(df, "p_high_reward")
```

## Human data: Prior vs posterior ratings

### Distribution of ratings

```{r}
#| fig-width: 12
#| fig-asp: 0.45

ggplot(df, aes(x = p_high_reward, fill = action)) +
  geom_density(alpha = 0.5) +
  scale_fill_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  labs(y = "Density", x = "P(high motivation)", fill = "Action") +
  theme(legend.position = "right") +
  facet_grid(stage ~ intimacy, labeller = labeller(intimacy = label_both, stage = label_value))
```

### Summary statistics

```{r}
df_summary <- df |>
  group_by(intimacy, action, stage, model) |>
  tidyboot_mean(p_high_reward)
```

### Prior vs posterior by action and intimacy

```{r}
#| fig-width: 12
#| fig-asp: 0.7

ggplot() +
  geom_violin(
    data = df,
    aes(x = stage, y = p_high_reward, fill = action),
    position = position_dodge(width = 0.8),
    alpha = 0.5,
    width = 0.7
  ) +
  geom_point(
    data = df_summary,
    aes(x = stage, y = empirical_stat, color = action, group = action),
    position = position_dodge(width = 0.8),
    size = 3,
    alpha = PLOT_ALPHA
  ) +
  geom_errorbar(
    data = df_summary,
    aes(x = stage, ymin = ci_lower, ymax = ci_upper, color = action, group = action),
    position = position_dodge(width = 0.8),
    width = 0.1,
    size = 1
  ) +
  geom_line(
    data = df_summary,
    aes(x = stage, y = empirical_stat, color = action, group = action),
    position = position_dodge(width = 0.8),
    size = 1
  ) +
  facet_grid(intimacy ~ action, labeller = label_both) +
  scale_fill_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  scale_color_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  labs(
    title = "P(high motivation) before and after observing action",
    y = "P(high motivation)",
    x = "Stage",
    fill = "Observed action",
    color = "Observed action"
  )
```

### Belief updates

```{r}
df_belief_update_summary <- df |>
  filter(stage == "posterior") |>
  group_by(intimacy, action, model) |>
  tidyboot_mean(belief_update)
```

## Model comparison

### Model description

We compare three ablations of the canonical access utility, performing inverse planning to infer desire/motivation from observed actions:

**Full model**: Inverts the full access utility $U(a|s, I) = w_v \cdot V(a|s) - w_d \cdot \text{access}(a) \cdot (1 - I) - w_e \cdot \text{effort}(a)$.

**Access only**: Drops the food-reward and effort terms, keeping only the access-discomfort term $-w_d \cdot \text{access}(a) \cdot (1 - I)$.

**Base (no access)**: Drops the access term: $U = w_v \cdot V(a|s) - w_e \cdot \text{effort}(a)$.

Per-scenario values of $\text{access}(a)$ and $\text{effort}(a)$ are generated by an LLM; reward $V(a|s)$ is stipulated as a binary goal-satisfaction gate ($V=1$ iff the action satisfies the active goal — sharing under high motivation, not-sharing under low motivation). All models assume a uniform prior over reward (P(high motivation) = 0.5) and compute posteriors via Bayesian inference. The observer inverse temperature ($\alpha_{\text{observer}}$) is fitted to human data.

### Fitted parameters

```{r}
df_fit_results <- read_csv(here("model", "outputs", "inverse_planning_fit_results.csv")) |>
  filter(experiment == "reward") |>
  mutate(
    model_label = case_when(
      model == "access_full" ~ "Full model",
      model == "access_only" ~ "Access only",
      model == "no_access" ~ "Base"
    )
  ) |>
  select(model_label, alpha_observer, nll, n_params)

df_fit_results |>
  knitr::kable(
    digits = 3,
    caption = "Fitted observer parameters for desire inference",
    col.names = c("Model", "α_observer", "NLL", "n_params")
  )
```

### AIC comparison

Following Burnham and Anderson (2002), ΔAIC values of 0–2 indicate essentially equivalent models, 4–7 indicate considerably less support for the higher-AIC model, and values greater than 10 indicate essentially no support for the higher-AIC model.

```{r}
# Calculate AIC
df_fit_results <- df_fit_results |>
  mutate(AIC = 2 * n_params + 2 * nll)

# Find best model
best_aic <- min(df_fit_results$AIC)

df_fit_results |>
  mutate(delta_AIC = AIC - best_aic) |>
  select(model_label, nll, n_params, AIC, delta_AIC) |>
  arrange(AIC) |>
  knitr::kable(
    digits = 2,
    caption = "Model comparison (lower AIC is better)",
    col.names = c("Model", "NLL", "n_params", "AIC", "ΔAIC")
  )
```

### Load model predictions

```{r}
df_model <- read_csv(here("model", "outputs", "inv_plan_desire_alt_preds_summary.csv")) |>
  rename(intimacy = intimacy_condition) |>
  mutate(
    action = as.character(action),
    intimacy = factor(intimacy),
    # Model predictions are P(high desire) in percentage
    p_high_reward = p_high_reward / 100,
    # Belief update from uniform prior (50%)
    belief_update = p_high_reward - 0.5,
    model = case_when(
      model == "access_full" ~ "Full model",
      model == "access_only" ~ "Access only",
      model == "no_access" ~ "Base"
    )
  )
```

### Combine human and model data

```{r}
df_combined <- bind_rows(
  df_belief_update_summary |>
    mutate(belief_update = empirical_stat) |>
    select(intimacy, action, model, belief_update, ci_lower, ci_upper),
  df_model |>
    group_by(intimacy, action, model) |>
    summarize(belief_update = mean(belief_update), .groups = "drop") |>
    mutate(ci_lower = NA_real_, ci_upper = NA_real_)
) |>
  mutate(
    model = factor(model, levels = c("Base", "Access only", "Full model", "Humans")),
    action = factor(action)
  )
```

### Belief updates: Model vs human comparison

```{r}
#| fig-width: 12
#| fig-asp: 0.3

df_combined |>
  ggplot(aes(x = action, y = belief_update, group = interaction(model, intimacy), fill = intimacy)) +
  geom_crossbar(
    aes(y = 0, ymin = -0.015, ymax = 0.015),
    position = position_dodge(width = 0.9),
    width = 0.9,
    color = NA
  ) +
  geom_col(position = position_dodge(width = 0.9), alpha = PLOT_ALPHA) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.9), size = 1, width = 0, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_intimacy() +
  labs(y = "Belief update", x = "Observed action", fill = "Intimacy") +
  theme(legend.position = "right") +
  facet_wrap(~model, nrow = 1)

ggsave(here("figures", "inv_plan_desire_alt_belief_main.pdf"), device = cairo_pdf)
```

## Scenario-level

### Human belief updates by scenario

```{r}
df_belief_update_scenario <- df |>
  filter(stage == "posterior") |>
  group_by(scenario_label, intimacy, action, model) |>
  tidyboot_mean(belief_update)
```
