---
title: Experiment 2a no-alt (inverse planning - intimacy inference, no alternatives shown) data analysis
author: Alicia Chen
date: 2026-04-19
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    page-layout: full
    search: false
execute:
  warning: false
---

```{r setup}
source(here::here("analysis", "utils.R"))
setup_analysis()
```

## Demographics

```{r}
df_exit <- read_csv(here("data", "inv_plan_intimacy_noalt", "exit_survey.csv"), show_col_types = FALSE)

n_total <- nrow(df_exit)
n_passed <- df_exit |>
  filter(attention_passed == TRUE, memory_correct_count > 0) |>
  nrow()

cat("Total participants recruited:", n_total, "\n")
cat("Passed attention checks:", n_passed, "\n")
cat("Mean age:", round(mean(df_exit$age, na.rm = TRUE), 1),
    "SD age:", round(sd(df_exit$age, na.rm = TRUE), 1),
    "Min age:", min(df_exit$age, na.rm = TRUE),
    "Max age:", max(df_exit$age, na.rm = TRUE))
cat("\nGender:\n")
print(table(df_exit$gender))
```

## Results

This experiment is a variant of Exp 2a in which participants observe a single action and infer the relationship **without seeing the other three alternative actions the actors could have taken**. The experiment manipulates the same factors as Exp 2a:

- **Action**: Which action was taken (action 0-3, varying in saliva-transfer risk)
- **Motivation (desire)**: Whether both people want the food (high vs low)

The cognitive model for this variant uses an LM-generated counterfactual alternative set (conditioned on the observed action and motivation) as the actor's action space, padded to a fixed size so the memo program can enumerate a static slot dimension. Slot 0 holds the observed canonical action; slots 1..k hold the LM-generated alternatives; remaining slots are null-padded (access=effort=is_share=0).

## Load human data

```{r}
# main_trials_long.csv is generated by analysis/json_to_csv.py
# It excludes participants who failed attention check or got 0 on memory check
df <- read_csv(here("data", "inv_plan_intimacy_noalt", "main_trials_long.csv")) |>
  mutate(
    action = str_replace(action_condition, "action_", ""),
    motivation = factor(motivation, levels = c("low", "high"), labels = MOTIVATION_LEVELS),
    stage = factor(stage, levels = c("prior", "posterior")),
    model = "Humans"
  ) |>
  select(-action_condition)

# Calculate belief updates
df <- calculate_belief_update(df, "intimacy_rating")
```

## Human data: Prior vs posterior ratings

First, look at summaries of human data

### Distribution of ratings

```{r}
ggplot(df, aes(x = intimacy_rating, fill = action)) +
  geom_density(alpha = 0.5) +
  scale_fill_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  labs(y = "Density", x = "Intimacy rating", fill = "Action") +
  theme(legend.position = "right") +
  facet_grid(stage ~ motivation, labeller = labeller(motivation = label_both))
```

Calculate summary statistics

```{r}
df_summary <- df |>
  group_by(motivation, action, stage, model) |>
  tidyboot_mean(intimacy_rating)
```

### Prior vs posterior by action and motivation

```{r}
#| fig-width: 12
#| fig-asp: 0.4

ggplot() +
  geom_violin(
    data = df,
    aes(x = stage, y = intimacy_rating, fill = action),
    position = position_dodge(width = 0.8),
    alpha = 0.5,
    width = 0.7
  ) +
  geom_point(
    data = df_summary,
    aes(x = stage, y = empirical_stat, color = action, group = action),
    position = position_dodge(width = 0.8),
    size = 3,
    alpha = PLOT_ALPHA
  ) +
  geom_errorbar(
    data = df_summary,
    aes(x = stage, ymin = ci_lower, ymax = ci_upper, color = action, group = action),
    position = position_dodge(width = 0.8),
    width = 0.1,
    size = 1
  ) +
  geom_line(
    data = df_summary,
    aes(x = stage, y = empirical_stat, color = action, group = action),
    position = position_dodge(width = 0.8),
    size = 1
  ) +
  facet_grid(motivation ~ action, labeller = labeller(action = label_both)) +
  scale_fill_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  scale_color_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  labs(
    title = "Intimacy ratings before and after observing action (no-alt)",
    y = "Intimacy rating",
    x = "Stage",
    fill = "Observed action",
    color = "Observed action"
  )
```

### Belief updates

```{r}
df_belief_update_summary <- df |>
  filter(stage == "posterior") |>
  group_by(motivation, action, model) |>
  tidyboot_mean(belief_update)
```

## Model comparison

### Model description

We compare three ablations of the canonical access utility (Base, Access only, Full). The action space for each (scenario, observed action, motivation) cell is the union of the observed canonical action and an LM-generated set of plausible counterfactual alternatives. Access and effort for each alternative are scored by an LLM (batched per scenario, 10 runs each); $V(a|s)$ uses an `is_share` tag elicited at generation time.

**Full model**: Inverts the full access utility $U(a|s, I) = w_v \cdot V(a|s) - w_d \cdot \text{access}(a) \cdot (1 - I) - w_e \cdot \text{effort}(a)$.

**Access only**: Drops the food-reward and effort terms, keeping only the access-discomfort term $-w_d \cdot \text{access}(a) \cdot (1 - I)$.

**Base (no access)**: Drops the access term: $U = w_v \cdot V(a|s) - w_e \cdot \text{effort}(a)$.

The softmax over actions is $P(a | I, s) \propto \pi(a|s, \text{mot}) \cdot \exp(U(a|I, s))$, where $\pi$ is uniform over the valid slots in each padded cell. All models assume a uniform intimacy prior and compute posteriors via Bayesian inference over the padded action space. Actor parameters are frozen per-variant from the forward-planning fit. Only the observer inverse temperature ($\alpha_{\text{observer}}$) is fitted to the no-alt human data.

### Fitted parameters

```{r}
df_fit_results <- read_csv(here("model", "outputs", "inverse_planning_noalt_fit_results.csv")) |>
  mutate(
    utility = str_remove(model, "_padded$"),
    model_label = case_when(
      utility == "access_full" ~ "Full model",
      utility == "access_only" ~ "Access only",
      utility == "no_access" ~ "Base"
    )
  ) |>
  select(model_label, alpha_observer, nll, n_params) |>
  arrange(model_label)

df_fit_results |>
  knitr::kable(
    digits = 3,
    caption = "Fitted observer parameters for no-alt intimacy inference (3 utility variants)",
    col.names = c("Utility", "α_observer", "NLL", "n_params")
  )
```

### AIC comparison

Following Burnham and Anderson (2002), ΔAIC values of 0–2 indicate essentially equivalent models, 4–7 indicate considerably less support for the higher-AIC model, and values greater than 10 indicate essentially no support for the higher-AIC model.

```{r}
# Calculate AIC
df_fit_results <- df_fit_results |>
  mutate(AIC = 2 * n_params + 2 * nll)

# Find best model
best_aic <- min(df_fit_results$AIC)

df_fit_results |>
  mutate(delta_AIC = AIC - best_aic) |>
  select(model_label, nll, n_params, AIC, delta_AIC) |>
  arrange(AIC) |>
  knitr::kable(
    digits = 2,
    caption = "Model comparison (3 utility variants; lower AIC is better)",
    col.names = c("Utility", "NLL", "n_params", "AIC", "ΔAIC")
  )
```

### Load model predictions

```{r}
df_model <- read_csv(here("model", "outputs", "inv_plan_intimacy_noalt_preds_summary.csv")) |>
  mutate(
    action = as.character(observed_action),
    motivation = factor(motivation, levels = c("low", "high"), labels = MOTIVATION_LEVELS),
    belief_update = (expected_intimacy - 50) / 100,
    utility = str_remove(model, "_padded$"),
    model_label = case_when(
      utility == "access_full" ~ "Full model",
      utility == "access_only" ~ "Access only",
      utility == "no_access" ~ "Base"
    )
  )
```

### Combine human and model data

```{r}
df_human_long <- df_belief_update_summary |>
  mutate(
    belief_update = empirical_stat / 100,
    ci_lower = ci_lower / 100,
    ci_upper = ci_upper / 100,
    model_label = "Humans"
  ) |>
  select(motivation, action, model_label, belief_update, ci_lower, ci_upper)

df_model_long <- df_model |>
  group_by(motivation, action, model_label) |>
  summarize(belief_update = mean(belief_update), .groups = "drop") |>
  mutate(ci_lower = NA_real_, ci_upper = NA_real_)
```

### Belief updates: Model vs human comparison

```{r}
#| fig-width: 12
#| fig-asp: 0.3

df_combined <- bind_rows(df_human_long, df_model_long) |>
  mutate(
    model_label = factor(model_label, levels = c("Base", "Access only", "Full model", "Humans")),
    action = factor(action)
  )

p_belief <- df_combined |>
  ggplot(aes(x = action, y = belief_update, group = interaction(model_label, motivation), fill = motivation)) +
  geom_crossbar(
    aes(y = 0, ymin = -0.01, ymax = 0.01),
    position = position_dodge(width = 0.9),
    width = 0.9,
    color = NA
  ) +
  geom_col(position = position_dodge(width = 0.9), alpha = PLOT_ALPHA) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.9), size = 1, width = 0.1, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_motivation() +
  labs(y = "Belief update", x = "Observed action", fill = "Desire") +
  theme(legend.position = "right") +
  facet_wrap(~model_label, nrow = 1)

print(p_belief)

ggsave(here("figures", "exp2a_noalt_belief_main.pdf"), plot = p_belief, device = cairo_pdf)
```


## Scenario-level

### Human belief updates by scenario

```{r}
df_belief_update_scenario <- df |>
  filter(stage == "posterior") |>
  group_by(scenario_label, motivation, action, model) |>
  tidyboot_mean(belief_update)
```

```{r}
#| fig-width: 12
#| fig-asp: 0.8

ggplot(df_belief_update_scenario, aes(x = action, y = empirical_stat, fill = motivation)) +
  geom_col(position = position_dodge(width = 0.9), alpha = PLOT_ALPHA) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.9), width = 0.1, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_motivation() +
  labs(
    title = "Belief updates by scenario (no-alt humans)",
    y = "Belief update",
    x = "Observed action",
    fill = "Motivation"
  ) +
  theme(legend.position = "right") +
  facet_wrap(~scenario_label, ncol = 4)
```

### Model belief updates by scenario

```{r}
#| fig-width: 14
#| fig-asp: 0.9

df_model_scenario <- df_model |>
  mutate(model_label = factor(model_label, levels = c("Base", "Access only", "Full model")))

ggplot(df_model_scenario, aes(x = action, y = belief_update, fill = motivation)) +
  geom_col(position = position_dodge(width = 0.9), alpha = PLOT_ALPHA) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_motivation() +
  labs(
    title = "Model-predicted belief updates by scenario",
    y = "Belief update",
    x = "Observed action",
    fill = "Motivation"
  ) +
  theme(legend.position = "right") +
  facet_grid(model_label ~ scenario_label,
             labeller = labeller(scenario_label = label_wrap_gen(15)))
```
