---
title: "LM elicitation overview — across the four studies"
format:
  html:
    code-fold: true
    toc: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
source(here::here("analysis", "utils.R"))
setup_analysis()
library(knitr)
library(ggrepel)
```

# LM elicitation overview

This notebook checks the **LM elicitation** that feeds all four inverse-planning
observer models — pooling across the four studies rather than diagnosing one. The
per-study diagnostic notebook for Study 1a
(`food-inv-desire-lm-elicitation.qmd`) drills into a single elicitation; this one
asks the cross-study questions: does the elicited feature map recover the intended
action structure everywhere, do the desire and intimacy manipulations come out
right, and do the imagined alternatives cover the space and shift with what was
observed.

Each study's actor softmaxes over `{observed action} ∪ LM-generated alternatives`;
the reward / discomfort / effort terms are driven by LM-elicited per-action
features (risk / effort / goal-satisfaction `g`, all on 0–1). The given latents are
LM-elicited scalars: **intimacy** per relationship level in the given-relationship
studies (1a/1b), **desire** per desire condition in the given-desire studies
(2a/2b).

```{r roster}
# Study roster. Slug is the stable identifier; label is the manuscript number.
STUDIES <- tibble::tribble(
  ~slug,                ~label, ~infers,            ~given_latent,
  "food_inv_desire",    "1a",   "desire",           "intimacy",
  "food_inv_joint_de",  "1b",   "desire + effort",  "intimacy",
  "food_inv_intimacy",  "2a",   "intimacy",         "desire",
  "food_inv_joint_ie",  "2b",   "intimacy + effort","desire"
)

action_levels <- c("no_share", "low_risk_share", "high_risk_share")
action_labels <- c("No share", "Low-risk share", "High-risk share")
ACTION_COLORS <- setNames(c("#7A9A4A", "#C9A85A", "#B05A5A"), action_labels)
as_action <- function(x) factor(x, levels = action_levels, labels = action_labels)

lm_path <- function(slug, file) here::here("model", "outputs", "lm", slug, file)
```

```{r load}
# Add any columns a study's records lack (1a/1b carry intimacy*, 2a/2b carry
# desire*) so all four can be row-bound into one long action table.
ensure_cols <- function(df, cols) {
  for (c in cols) if (!c %in% names(df)) df[[c]] <- NA
  df
}

load_study_actions <- function(slug, label) {
  path <- lm_path(slug, "lm_runs.jsonl")
  if (!file.exists(path)) return(NULL)
  raw <- read_model_jsonl(path) |>
    ensure_cols(c("effort_condition", "intimacy_condition", "desire_condition",
                  "intimacy", "desire"))
  raw |>
    select(run_id, scenario_label, observed_action, effort_condition,
           intimacy_condition, desire_condition, intimacy, desire, actions) |>
    tidyr::unnest(actions) |>
    mutate(slug = slug, study = label)
}

actions_all <- purrr::pmap(STUDIES[c("slug", "label")],
                           \(slug, label) load_study_actions(slug, label)) |>
  purrr::compact() |>
  bind_rows() |>
  mutate(study = factor(study, levels = STUDIES$label))

have_data <- nrow(actions_all) > 0

if (have_data) {
  obs_all <- filter(actions_all, is_observed)
  cat("Loaded studies:", paste(levels(droplevels(actions_all$study)), collapse = ", "), "\n")
  actions_all |>
    group_by(study) |>
    summarise(records = n_distinct(run_id, scenario_label, observed_action,
                                   effort_condition, intimacy_condition, desire_condition),
              action_rows = n(),
              observed = sum(is_observed),
              alternatives = sum(!is_observed), .groups = "drop") |>
    kable(caption = "Elicitation volume per study (after unnesting actions)")
} else {
  cat("No lm_runs.jsonl found for any study — run the LM elicitation first.\n")
}
```

# Does the feature map recover the intended action structure?

The load-bearing sanity check for the whole pipeline. Across **all four
elicitations pooled**, the observed actions should show: **risk**
monotone (`no_share` ≈ 0 < `low_risk_share` < `high_risk_share`), **goal-satisfaction
`g`** near 0 for `no_share` and high for both sharing actions (it is desire-free),
and **effort** lifted by the effort manipulation on the low-risk share. Each thin
grey line is one scenario (pooled over studies and runs); the bold line is the
grand mean.

```{r obs-structure}
#| fig-width: 12
#| fig-height: 4.5
if (have_data) {
  # risk and g are effort/intimacy/desire-independent; effort is pooled over its
  # two conditions here for the uniform overview (the manipulation detail follows).
  obs_feat <- obs_all |>
    group_by(scenario_label, observed_action) |>
    summarise(risk = mean(risk, na.rm = TRUE),
              effort = mean(effort, na.rm = TRUE),
              g = mean(g, na.rm = TRUE), .groups = "drop") |>
    pivot_longer(c(risk, effort, g), names_to = "feature", values_to = "value") |>
    mutate(action = as_action(observed_action),
           feature = factor(feature, c("risk", "effort", "g"),
                            c("Risk", "Effort", "Goal-satisfaction g")))

  obs_mean <- obs_feat |>
    group_by(feature, action) |>
    summarise(value = mean(value, na.rm = TRUE), .groups = "drop")

  ggplot(obs_feat, aes(action, value)) +
    geom_line(aes(group = scenario_label), color = "grey65",
              linewidth = 0.25, alpha = 0.4) +
    geom_line(data = obs_mean, aes(group = 1), color = "grey15", linewidth = 1.1) +
    geom_point(data = obs_mean, aes(color = action), size = 3) +
    facet_wrap(~feature) +
    scale_color_manual(values = ACTION_COLORS, guide = "none") +
    scale_y_continuous(limits = c(-0.05, 1.05), breaks = seq(0, 1, 0.25)) +
    labs(title = "LM feature map recovers the intended action structure",
         subtitle = "Pooled over all four elicitations; thin lines = scenarios, bold = grand mean",
         x = "Observed action", y = "Feature value (0–1)") +
    theme(plot.title = element_text(size = 15, hjust = 0),
          axis.text.x = element_text(angle = 20, hjust = 1))
}
```

```{r obs-monotone-check}
if (have_data) {
  wide <- obs_all |>
    group_by(scenario_label, observed_action) |>
    summarise(risk = mean(risk, na.rm = TRUE), g = mean(g, na.rm = TRUE),
              .groups = "drop")
  risk_w <- wide |> select(-g) |> pivot_wider(names_from = observed_action, values_from = risk)
  g_w <- wide |> select(-risk) |> pivot_wider(names_from = observed_action, values_from = g)
  bad_risk <- risk_w |> filter(!(no_share <= low_risk_share & low_risk_share <= high_risk_share))
  bad_g <- g_w |> filter(no_share >= pmin(low_risk_share, high_risk_share))
  cat("Scenarios with non-monotone observed-action risk:", nrow(bad_risk), "of", nrow(risk_w),
      if (nrow(bad_risk)) paste0(" (", paste(bad_risk$scenario_label, collapse = ", "), ")") else "", "\n")
  cat("Scenarios where no_share g is not below both sharing actions:", nrow(bad_g),
      if (nrow(bad_g)) paste0(" (", paste(bad_g$scenario_label, collapse = ", "), ")") else "", "\n")
}
```

## Is the effort manipulation correct?

The effort manipulation targets the **low-risk share** (its preparation paragraph
is appended in the high condition), so high vs. low effort should lift *its* effort
and leave the other actions roughly unchanged. Pooled across all four elicitations.

```{r effort-manip}
#| fig-width: 8
#| fig-height: 4.5
if (have_data) {
  eff <- obs_all |>
    filter(effort_condition %in% c("low", "high")) |>
    group_by(scenario_label, observed_action, effort_condition) |>
    summarise(effort = mean(effort, na.rm = TRUE), .groups = "drop") |>
    mutate(action = as_action(observed_action),
           effort_cond = factor(effort_condition, c("low", "high"),
                                c("Low effort", "High effort")))

  eff_mean <- eff |>
    group_by(action, effort_cond) |>
    summarise(effort = mean(effort, na.rm = TRUE), .groups = "drop") |>
    mutate(xpos = as.integer(action) + ifelse(effort_cond == "Low effort", -0.15, 0.15))

  ggplot(eff, aes(action, effort, color = effort_cond)) +
    geom_boxplot(outlier.shape = NA, width = 0.6,
                 position = position_dodge(width = 0.7), fill = NA) +
    geom_line(data = eff_mean,
              aes(x = xpos, y = effort, group = effort_cond, color = effort_cond),
              linewidth = 1, inherit.aes = FALSE, show.legend = FALSE) +
    geom_point(data = eff_mean, aes(x = xpos, y = effort, color = effort_cond),
               size = 2.5, inherit.aes = FALSE, show.legend = FALSE) +
    scale_color_effort() +
    scale_y_continuous(limits = c(-0.03, 1.03), breaks = seq(0, 1, 0.25)) +
    labs(title = "Effort manipulation lifts the low-risk share specifically",
         subtitle = "Pooled over all four elicitations; per (scenario) means",
         x = "Observed action", y = "Effort (0–1)", color = "Effort condition") +
    theme(plot.title = element_text(size = 15, hjust = 0),
          axis.text.x = element_text(angle = 20, hjust = 1))
}
```

## Is the desire manipulation correct?

Studies 2a/2b take **desire** as given context, LM-elicited per run as a scalar for
each `desire_condition` (low / high). It should be cleanly ordered low < high, and —
because it is scored per run — should carry run-to-run spread. Each thin line is one
scenario; pooled across 2a + 2b.

```{r desire-manip}
#| fig-width: 7
#| fig-height: 4.5
if (have_data) {
  # desire is non-NA only for the given-desire studies (2a/2b).
  desire_scalar <- actions_all |>
    filter(!is.na(desire_condition), !is.na(desire)) |>
    distinct(study, scenario_label, desire_condition, run_id, desire)

  have_desire <- nrow(desire_scalar) > 0
  if (have_desire) {
    dsr_scen <- desire_scalar |>
      group_by(scenario_label, desire_condition) |>
      summarise(desire = mean(desire, na.rm = TRUE), .groups = "drop") |>
      mutate(cond = factor(desire_condition, c("low", "high"), c("Low", "High")))
    dsr_mean <- dsr_scen |> group_by(cond) |>
      summarise(desire = mean(desire), .groups = "drop")

    print(
      ggplot(dsr_scen, aes(cond, desire)) +
        geom_line(aes(group = scenario_label), color = "grey65", linewidth = 0.3, alpha = 0.5) +
        geom_jitter(data = desire_scalar |>
                      mutate(cond = factor(desire_condition, c("low","high"), c("Low","High"))),
                    aes(cond, desire), width = 0.08, height = 0, alpha = 0.06, size = 0.5,
                    color = "#7A4A5A") +
        geom_line(data = dsr_mean, aes(group = 1), color = "#7A4A5A", linewidth = 1.2) +
        geom_point(data = dsr_mean, aes(color = cond), size = 4) +
        scale_color_desire() +
        scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25)) +
        labs(title = "Desire manipulation (2a/2b): elicited scalar by condition",
             subtitle = "Thin lines = scenarios; points = per-run values (run-to-run spread)",
             x = "Desire condition", y = "Elicited desire magnitude (0–1)", color = NULL) +
        theme(plot.title = element_text(size = 15, hjust = 0))
    )
  }
}
```

## Is the intimacy manipulation correct?

Studies 1a/1b take **intimacy** as given context, LM-elicited per run from the
de-anchored relationship descriptors (4 levels, scenario-independent). It should be
monotone `max_formal → max_intimate`. Pooled across 1a + 1b.

```{r intimacy-manip}
#| fig-width: 8
#| fig-height: 4.5
if (have_data) {
  intim_scalar <- actions_all |>
    filter(!is.na(intimacy_condition), !is.na(intimacy)) |>
    distinct(study, intimacy_condition, run_id, intimacy) |>
    mutate(intimacy_condition = factor(intimacy_condition, INTIMACY_LEVELS))

  have_intim <- nrow(intim_scalar) > 0
  if (have_intim) {
    intim_summ <- intim_scalar |>
      group_by(intimacy_condition) |>
      summarise(value = mean(intimacy, na.rm = TRUE), sd = sd(intimacy, na.rm = TRUE),
                .groups = "drop")
    intim_summ |>
      mutate(across(c(value, sd), \(x) round(x, 3))) |>
      kable(caption = "Pooled intimacy scalar (mean over 1a+1b runs, run-to-run SD)")

    print(
      ggplot(intim_summ, aes(intimacy_condition, value, color = intimacy_condition)) +
        geom_errorbar(aes(ymin = value - sd, ymax = value + sd), width = 0.15) +
        geom_point(size = 4) +
        scale_color_intimacy() +
        scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25)) +
        labs(title = "Intimacy manipulation (1a/1b): elicited magnitude by level",
             subtitle = "Error bars = run-to-run SD (note: near zero — constant across runs)",
             x = NULL, y = "Elicited intimacy magnitude (0–1)", color = "Intimacy level") +
        theme(plot.title = element_text(size = 15, hjust = 0),
              axis.text.x = element_text(angle = 20, hjust = 1))
    )
  }
}
```

# Where do the observed actions sit in risk × effort space?

Each point is an observed action for one scenario (features averaged over runs),
pooled across all four elicitations. Shape marks the action. Because the effort
manipulation only changes the cost of the **low-risk share** action, only that
action is split by effort condition (low vs high); no-share and high-risk share are
effort-invariant, so each is shown once (their two effort-condition rows collapsed,
marked "n/a"). The three actions separate horizontally by risk, and the effort
manipulation shifts the low-risk share vertically.

```{r risk-effort-2x2}
#| fig-width: 8.5
#| fig-height: 6
if (have_data) {
  set.seed(67)
  obs_re <- obs_all |>
    filter(effort_condition %in% c("low", "high")) |>
    group_by(study, scenario_label, observed_action, effort_condition) |>
    summarise(risk = mean(risk, na.rm = TRUE), effort = mean(effort, na.rm = TRUE),
              .groups = "drop") |>
    mutate(action = as_action(observed_action))

  # Effort condition only modulates the low-risk share, so keep both effort points
  # for that action; for no-share and high-risk share (effort-invariant) collapse
  # the two effort-condition rows into a single point each, labelled "n/a".
  low_risk <- obs_re |>
    filter(observed_action == "low_risk_share") |>
    mutate(effort_cond = factor(effort_condition, c("low", "high"),
                                c("Low effort", "High effort")))
  invariant <- obs_re |>
    filter(observed_action != "low_risk_share") |>
    group_by(study, scenario_label, observed_action, action) |>
    summarise(risk = mean(risk), effort = mean(effort), .groups = "drop") |>
    mutate(effort_cond = "n/a")
  obs_re <- bind_rows(low_risk, invariant) |>
    mutate(effort_cond = factor(effort_cond,
                                levels = c("Low effort", "High effort", "n/a")))

  ggplot(obs_re, aes(risk, effort, color = effort_cond, shape = action)) +
    geom_jitter(width = 0.015, height = 0.015, size = 2, alpha = 0.7) +
    scale_color_manual(values = c("Low effort" = "#B5C9A8",
                                  "High effort" = "#4A7A4A", "n/a" = "grey70")) +
    scale_shape_manual(values = c("No share" = 16, "Low-risk share" = 17,
                                  "High-risk share" = 15)) +
    scale_x_continuous(limits = c(-0.05, 1.05), breaks = seq(0, 1, 0.25)) +
    scale_y_continuous(limits = c(-0.05, 1.05), breaks = seq(0, 1, 0.25)) +
    labs(title = "Observed actions in risk × effort space",
         subtitle = "Effort condition shown only for low-risk share (the action it modulates); other actions collapsed",
         x = "Risk (0–1)", y = "Effort (0–1)",
         color = "Effort condition", shape = "Observed action") +
    theme(plot.title = element_text(size = 15, hjust = 0))
}
```

# Alternatives

The remaining sections look at the **LM-generated alternatives** — the imagined
counterfactual choice set. The deep-dive into one scenario's set uses Study 1a
(its embeddings are available for the semantic view); the coverage and
condition-shift checks pool across studies.

```{r alt-params}
# Featured scenario for the single-scenario deep-dive. Change `ex_scenario` to any
# of the 16 (apples, basketball, birthday, brunch, cooking, dip, drinks, driving,
# fair, gala, hike, oysters, social, soup, takeout, wedding).
ex_study <- "food_inv_desire"
ex_label <- "1a"
ex_scenario <- "oysters"
```

## One scenario in feature space, colored by goal-satisfaction

Every distinct action in `r ex_scenario`'s (`r ex_label`) elicited choice set,
placed in the risk × effort plane the model reasons over and colored by its
goal-satisfaction `g`. The three observed actions are starred and labeled; the
grey-to-yellow alternatives show how the imagined options trade risk and effort off
against delivering the food.

```{r alt-feature-scatter}
#| fig-width: 8.5
#| fig-height: 6
if (have_data) {
  ex <- actions_all |>
    filter(slug == ex_study, scenario_label == ex_scenario) |>
    group_by(action_text, is_observed) |>
    summarise(risk = mean(risk, na.rm = TRUE), effort = mean(effort, na.rm = TRUE),
              g = mean(g, na.rm = TRUE),
              observed_action = first(observed_action), .groups = "drop") |>
    filter(!is.na(risk), !is.na(effort), !is.na(g))

  ex_alt <- filter(ex, !is_observed)
  ex_obs <- filter(ex, is_observed) |> mutate(action = as_action(observed_action))

  set.seed(67)
  print(
    ggplot(mapping = aes(risk, effort)) +
      geom_jitter(data = ex_alt, aes(fill = g), width = 0.012, height = 0.012,
                  shape = 21, color = "white", stroke = 0.1, size = 2.6, alpha = 0.85) +
      geom_point(data = ex_obs, shape = 23, aes(fill = g), color = "black",
                 size = 5, stroke = 0.9) +
      ggrepel::geom_label_repel(data = ex_obs, aes(label = action), size = 3.2,
                                fill = "white", alpha = 0.85, box.padding = 0.6,
                                segment.color = "grey50", min.segment.length = 0) +
      scale_fill_viridis_c(option = "viridis", limits = c(0, 1)) +
      scale_x_continuous(limits = c(-0.05, 1.05), breaks = seq(0, 1, 0.25)) +
      scale_y_continuous(limits = c(-0.05, 1.05), breaks = seq(0, 1, 0.25)) +
      labs(title = paste0("Elicited choice set for '", ex_scenario, "' (", ex_label, ")"),
           subtitle = "Each point a distinct action; ◆ = observed actions, colored by goal-satisfaction g",
           x = "Risk (0–1)", y = "Effort (0–1)", fill = "Goal-sat. g") +
      theme(plot.title = element_text(size = 15, hjust = 0))
  )
}
```

## One scenario in semantic space

A UMAP projection of the same scenario's alternatives (embedded text), so the
*kinds* of counterfactual the LM proposes are visible as clusters. Points are
colored by their per-scenario action-type cluster; the observed actions are
starred. Each cluster is labeled on the figure with a representative alternative,
and the full list of types is tabled below.

```{r semantic-load}
proj_path <- lm_path(ex_study, "lm_alternatives_projection.jsonl")
clusters_path <- lm_path(ex_study, "lm_clusters.json")
have_proj <- file.exists(proj_path)
if (!have_proj) {
  cat("No projection artifact at", proj_path,
      "\nRun: uv run python model/lm/project_alternatives.py --study", ex_study, "\n")
}
```

```{r semantic-space}
#| fig-width: 10
#| fig-height: 7
if (have_proj) {
  proj <- read_model_jsonl(proj_path) |> filter(scenario_label == ex_scenario)
  proj_alt <- filter(proj, !is_observed) |> mutate(cluster = factor(cluster))
  proj_obs <- filter(proj, is_observed) |> mutate(action = as_action(observed_action))

  # cluster exemplars (one representative alternative per cluster) from the
  # embed_alternatives.py artifact, placed at each cluster's centroid.
  cl_labels <- NULL
  if (file.exists(clusters_path)) {
    cl <- jsonlite::fromJSON(clusters_path, simplifyDataFrame = TRUE)
    ex_cl <- cl$clusters |> filter(scenario == ex_scenario) |>
      transmute(cluster = factor(cluster), size,
                exemplar = vapply(exemplars, \(x) x[[1]], character(1)))
    centroids <- proj_alt |> group_by(cluster) |>
      summarise(dim1 = mean(dim1), dim2 = mean(dim2), .groups = "drop")
    cl_labels <- centroids |> left_join(ex_cl, by = "cluster") |>
      mutate(lab = paste0("[", cluster, "] ",
                          stringr::str_trunc(exemplar, 42)))
  }

  p <- ggplot(proj_alt, aes(dim1, dim2)) +
    geom_point(aes(color = cluster), size = 1.4, alpha = 0.55) +
    geom_point(data = proj_obs, shape = 8, size = 4, stroke = 1.1, color = "black") +
    ggrepel::geom_text_repel(data = proj_obs, aes(label = action), size = 3.2,
                             fontface = "bold", box.padding = 0.5,
                             min.segment.length = 0, color = "grey15") +
    scale_color_viridis_d(option = "turbo", guide = "none") +
    labs(title = paste0("Semantic space of alternatives for '", ex_scenario, "' (", ex_label, ")"),
         subtitle = "UMAP of alternative texts; color = per-scenario action-type cluster; ✳ = observed actions",
         x = "UMAP-1", y = "UMAP-2") +
    theme(plot.title = element_text(size = 15, hjust = 0),
          axis.text = element_blank(), axis.ticks = element_blank())

  if (!is.null(cl_labels)) {
    p <- p + ggrepel::geom_label_repel(
      data = cl_labels, aes(dim1, dim2, label = lab, color = cluster),
      size = 2.9, fill = "white", alpha = 0.9, box.padding = 0.7,
      max.overlaps = Inf, segment.color = "grey60", show.legend = FALSE)
  }
  print(p)
}
```

```{r semantic-table}
if (have_proj && file.exists(clusters_path)) {
  cl <- jsonlite::fromJSON(clusters_path, simplifyDataFrame = TRUE)
  cl$clusters |>
    filter(scenario == ex_scenario) |>
    transmute(cluster, size, exemplar = vapply(exemplars, \(x) x[[1]], character(1))) |>
    arrange(desc(size)) |>
    kable(caption = paste0("Action types (clusters) for '", ex_scenario,
                           "' with a representative alternative"))
}
```

## Do the alternatives differ by observed action?

The load-bearing cognitive claim is that observers imagine counterfactuals to *what
they saw*, so the alternative set should shift with the **observed action** and
cover the feature space — that is what makes the observed action diagnostic. Pooled
across all four studies, the alternatives (grey) are shown in the feature planes the
model reasons over, faceted by observed action, with the observed actions starred.

```{r alt-by-observed}
#| fig-width: 11
#| fig-height: 7
if (have_data) {
  set.seed(67)
  by_obs <- actions_all |>
    select(observed_action, is_observed, risk, effort, g) |>
    filter(!is.na(risk)) |>
    mutate(action = as_action(observed_action))

  alt_pts <- filter(by_obs, !is_observed) |>
    pivot_longer(c(g, effort), names_to = "yvar", values_to = "yval") |>
    mutate(yvar = factor(yvar, c("g", "effort"), c("Goal-satisfaction g", "Effort")))
  obs_pts <- by_obs |> filter(is_observed) |>
    group_by(action, observed_action) |>
    summarise(risk = mean(risk), g = mean(g), effort = mean(effort), .groups = "drop") |>
    pivot_longer(c(g, effort), names_to = "yvar", values_to = "yval") |>
    mutate(yvar = factor(yvar, c("g", "effort"), c("Goal-satisfaction g", "Effort")))

  print(
    ggplot(mapping = aes(risk, yval)) +
      geom_jitter(data = alt_pts, width = 0.015, height = 0.015,
                  color = "grey70", alpha = 0.04, size = 0.4) +
      geom_point(data = obs_pts, aes(color = action), shape = 18, size = 3) +
      facet_grid(yvar ~ action) +
      scale_color_manual(values = ACTION_COLORS, guide = "none") +
      scale_x_continuous(limits = c(-0.05, 1.05), breaks = seq(0, 1, 0.5)) +
      scale_y_continuous(limits = c(-0.05, 1.05), breaks = seq(0, 1, 0.5)) +
      labs(title = "Alternatives cover the feature space, faceted by observed action",
           subtitle = "Grey = LM alternatives (all studies pooled); ◆ = mean for that observed action",
           x = "Risk (0–1)", y = "Feature value (0–1)") +
      theme(plot.title = element_text(size = 15, hjust = 0))
  )
}
```

For Study 1a, the semantic embeddings let us label each alternative by the
observed action it most resembles (scenario-invariant), which tests the anchoring
claim directly: does the imagined set skew toward the contrasting region given what
was observed? Rows sum to 100%.

```{r alt-anchoring}
#| fig-width: 8
#| fig-height: 4.5
sem_path <- lm_path("food_inv_desire", "lm_alternatives_semantic.jsonl")
alts_path <- lm_path("food_inv_desire", "lm_alternatives.jsonl")
if (file.exists(sem_path) && file.exists(alts_path)) {
  sem <- read_model_jsonl(sem_path)
  alts1a <- read_model_jsonl(alts_path) |>
    left_join(sem, by = c("scenario_label", "action_text"))

  anchor <- alts1a |>
    filter(!is.na(nearest_observed_action)) |>
    count(observed_action, nearest_observed_action) |>
    group_by(observed_action) |>
    mutate(prop = n / sum(n)) |>
    ungroup() |>
    mutate(observed = as_action(observed_action),
           nearest = as_action(nearest_observed_action))

  print(
    ggplot(anchor, aes(nearest, observed, fill = prop)) +
      geom_tile(color = "white", linewidth = 0.4) +
      geom_text(aes(label = scales::percent(prop, accuracy = 1)), size = 3.6, color = "grey15") +
      scale_fill_viridis_c(option = "mako", direction = -1) +
      labs(title = "Anchoring (1a): imagined alternatives by nearest observed action",
           subtitle = "Each alternative labeled by its nearest observed action; rows sum to 100%",
           x = "Alternative most resembles…", y = "Observed action", fill = "Proportion") +
      theme(plot.title = element_text(size = 15, hjust = 0),
            axis.text.x = element_text(angle = 20, hjust = 1))
  )
}
```

## Beyond the observed action — do the alternatives shift with condition?

The alternatives are elicited per cell, so the other observed axes could also
reshape the imagined set. Intimacy (1a/1b) and desire (2a/2b) are *given context* in
their studies, so conditioning on them doesn't leak the inferred latent — but it is
worth seeing whether the imagined set moves with them. Below, the alternatives'
scored features are shown by intimacy condition (1a/1b) and by desire
condition (2a/2b); a clear shift means the framing bleeds into what observers
imagine.

```{r alt-by-intimacy}
#| fig-width: 11
#| fig-height: 4.5
if (have_data) {
  alt_int <- actions_all |>
    filter(!is_observed, !is.na(intimacy_condition), !is.na(risk)) |>
    mutate(intimacy_condition = factor(intimacy_condition, INTIMACY_LEVELS)) |>
    select(intimacy_condition, risk, effort, g) |>
    pivot_longer(c(risk, effort, g), names_to = "feature", values_to = "value") |>
    mutate(feature = factor(feature, c("risk", "effort", "g"),
                            c("Risk", "Effort", "Goal-satisfaction g")))

  if (nrow(alt_int) > 0) {
    print(
      ggplot(alt_int, aes(intimacy_condition, value, fill = intimacy_condition)) +
        geom_boxplot(outlier.shape = NA, width = 0.65, alpha = PLOT_ALPHA) +
        stat_summary(fun = mean, geom = "point", shape = 18, size = 2.5, color = "#B05A5A") +
        facet_wrap(~feature) +
        scale_fill_intimacy() +
        scale_x_discrete(labels = \(x) stringr::str_wrap(gsub("_", " ", x), 10)) +
        labs(title = "Alternative features by intimacy condition (1a/1b)",
             subtitle = "Alternative actions pooled; red diamond = mean",
             x = NULL, y = "Feature value (0–1)", fill = "Intimacy") +
        theme(plot.title = element_text(size = 15, hjust = 0),
              axis.text.x = element_text(size = 11),
              legend.position = "none")
    )
  }
}
```

```{r alt-by-desire}
#| fig-width: 11
#| fig-height: 4.5
if (have_data) {
  alt_des <- actions_all |>
    filter(!is_observed, !is.na(desire_condition), !is.na(risk)) |>
    mutate(cond = factor(desire_condition, c("low", "high"), c("Low", "High"))) |>
    select(cond, risk, effort, g) |>
    pivot_longer(c(risk, effort, g), names_to = "feature", values_to = "value") |>
    mutate(feature = factor(feature, c("risk", "effort", "g"),
                            c("Risk", "Effort", "Goal-satisfaction g")))

  if (nrow(alt_des) > 0) {
    print(
      ggplot(alt_des, aes(cond, value, fill = cond)) +
        geom_boxplot(outlier.shape = NA, width = 0.6, alpha = PLOT_ALPHA) +
        stat_summary(fun = mean, geom = "point", shape = 18, size = 2.5, color = "#B05A5A") +
        facet_wrap(~feature) +
        scale_fill_desire() +
        labs(title = "Alternative features by desire condition (2a/2b)",
             subtitle = "Alternative actions pooled; red diamond = mean (contrast to intimacy)",
             x = "Desire condition", y = "Feature value (0–1)", fill = "Desire") +
        theme(plot.title = element_text(size = 15, hjust = 0),
              legend.position = "none")
    )
  }
}
```

# Takeaways

- **Feature map** — risk should be monotone across the three actions, `g` should
  separate `no_share` from the sharing actions, and the effort manipulation should
  lift only the low-risk share. These hold pooled across all four elicitations
  (violations, if any, are flagged above).
- **Manipulation checks** — the desire scalar (2a/2b) should be cleanly ordered
  low < high with run-to-run spread; the intimacy scalar (1a/1b) should be monotone
  across the four levels (it is constant across runs by construction — no spread).
- **Risk × effort geometry** — the observed actions separate by risk and the
  effort manipulation shifts the low-risk share, so the feature space is not
  collinear.
- **Alternatives** — the imagined set should cover the feature space and shift with
  the observed action (the anchoring claim); the intimacy/desire panels show how
  much the given-context framings additionally reshape the set.
```

