---
title: Non-food forward planning (data analysis)
author: Alicia Chen
date: 2026-04-26
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    page-layout: full
    search: false
knitr:
  opts_chunk:
    dev: "ragg_png"
execute:
  warning: false
---

```{r setup}
source(here::here("analysis", "utils.R"))
setup_analysis()
```

## Demographics

```{r}
df_exit <- report_demographics("nonfood_forw_plan")
```

## Results

This experiment is the non-food analogue of `forw_plan`, run on the 16 scenarios in `experiments/scenarios_nonfood.csv` (substance sharing, shared physical space, and informational/situational privacy). It manipulates motivation (`low` vs. `high`) and intimacy (0, 50, 75, 100), and asks participants to distribute probability across four candidate actions.

Load csv, calculate means

```{r}
# main_trials_long.csv is generated by analysis/json_to_csv.py
# It excludes participants who failed attention check or got 0 on memory check
df <- read_csv(here("data", "nonfood_forw_plan", "main_trials_long.csv")) |>
  mutate(
    intimacy = factor(intimacy),
    motivation = factor(motivation, levels = c("low", "high"), labels = MOTIVATION_LEVELS),
    model = "humans"
  )

df_summary <- df |>
  group_by(intimacy, motivation, action) |>
  tidyboot_mean(p_action)
```

### P(action) by intimacy and motivation

```{r}
#| fig-width: 8
#| fig-asp: 0.5

df_summary |>
  ggplot(aes(x = action, y = empirical_stat, color = intimacy, shape = factor(action), group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 4, alpha = PLOT_ALPHA) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), size = 1.2, width = 0.5
  ) +
  geom_line(position = position_dodge(width = 0.2), linewidth = 1.2) +
  scale_color_intimacy() +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(y = "P(Action)", x = "Action", color = "Intimacy", shape = "Action") +
  facet_wrap(~motivation)
```

### Per-scenario summary

```{r}
#| label: fig-nonfood-forw-plan-scenario-summary
#| fig-width: 23
#| fig-asp: 0.18

df_summary_scenario <- df |>
  group_by(scenario_label, intimacy, motivation, action) |>
  tidyboot_mean(p_action)

ggplot(df_summary_scenario, aes(x = action, y = empirical_stat, color = intimacy, group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 3, alpha = PLOT_ALPHA) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), width = 0.4
  ) +
  geom_line(position = position_dodge(width = 0.2), linewidth = 1) +
  scale_color_intimacy() +
  labs(y = "P(action)", x = "Action", color = "Intimacy") +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid(motivation ~ scenario_label, labeller = labeller(scenario_label = label_wrap_gen(15)))
```

### Per-scenario summary, grouped by scenario type

`scenarios_nonfood.csv` tags each scenario as `substance` (chapstick, towel, hat, hairbrush, harmonica, sunscreen), `space` (blanket, sleeping-bag, bed, locker-room, sauna), or `privacy` (breakup, payment, gossip, home, navigation).

```{r}
#| fig-width: 23
#| fig-asp: 0.18

df_scenario_types <- read_csv(here("experiments", "scenarios_nonfood.csv"), show_col_types = FALSE) |>
  distinct(scenario_label, scenario_type)

df_summary_typed <- df_summary_scenario |>
  left_join(df_scenario_types, by = "scenario_label") |>
  mutate(scenario_label = paste0(scenario_label, "\n(", scenario_type, ")"))

ggplot(df_summary_typed, aes(x = action, y = empirical_stat, color = intimacy, group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 3, alpha = PLOT_ALPHA) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), width = 0.4
  ) +
  geom_line(position = position_dodge(width = 0.2), linewidth = 1) +
  scale_color_intimacy() +
  labs(y = "P(action)", x = "Action", color = "Intimacy") +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid(motivation ~ scenario_type + scenario_label,
             labeller = labeller(scenario_label = label_wrap_gen(15)),
             scales = "free_x", space = "free_x")
```

## Model comparison

Sample size is small (n = `r df_exit |> nrow()` recruited; LOSO CV refits on 15 scenarios per fold). Read effect sizes with that caveat.

### Model description

Three ablations of the canonical access utility, identical to the food forward-planning model — the only difference is that `access(a)` and `effort(a)` are now LM-elicited from `experiments/scenarios_nonfood.csv` rather than `experiments/scenarios.csv`. The same domain-general LM prompt set is used for both (`model/lm_prompts.py`).

**Full model**: $U(a|s, I) = w_v \cdot V(a|s) - w_d \cdot \text{access}(a) \cdot (1 - I) - w_e \cdot \text{effort}(a)$

**Discomfort-only**: only $-w_d \cdot \text{access}(a) \cdot (1 - I)$.

**Base model**: $U = w_v \cdot V(a|s) - w_e \cdot \text{effort}(a)$ (no access/intimacy term).

### Load LOSO predictions

Out-of-sample predictions from `model/outputs/cv_loso_preds_nonfood.csv`: each trial's prediction is from a model fit on the other 15 scenarios.

```{r}
df_cv <- read_csv(here("model", "outputs", "cv_loso_preds_nonfood.csv"), show_col_types = FALSE)

df_humans_cv <- df_cv |>
  filter(variant == "access_full") |>
  transmute(
    subject_id, scenario_label = held_out_scenario,
    intimacy, motivation, action,
    p_action, model = "Humans"
  )

df_model_cv <- df_cv |>
  mutate(
    model = case_when(
      variant == "access_full" ~ "Full model",
      variant == "access_only" ~ "Discomfort-only",
      variant == "no_access" ~ "Base model"
    )
  ) |>
  transmute(
    subject_id, scenario_label = held_out_scenario,
    intimacy, motivation, action,
    p_action = p_action_pred, model
  )

df_preds <- bind_rows(df_humans_cv, df_model_cv) |>
  mutate(
    intimacy = factor(intimacy),
    model = factor(model, levels = c(
      "Base model", "Discomfort-only", "Full model", "Humans"
    )),
    motivation = factor(motivation, levels = c("low", "high"), labels = MOTIVATION_LEVELS)
  )

df_preds_summary <- df_preds |>
  group_by(intimacy, motivation, action, model) |>
  tidyboot_mean(p_action) |>
  ungroup()
```

### Model predictions vs human data

```{r}
#| fig-width: 12
#| fig-asp: 0.43

df_preds_summary |>
  ggplot(aes(x = action, y = empirical_stat, color = intimacy, shape = factor(action), group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 4, alpha = PLOT_ALPHA) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), size = 1.2, width = 0.5
  ) +
  geom_line(position = position_dodge(width = 0.2), linewidth = 1.2) +
  scale_color_intimacy() +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(y = "P(Action)", x = "Action", color = "Intimacy", shape = "Action") +
  facet_grid(motivation ~ model)

ggsave(here("figures", "nonfood_forw_plan_comparison.pdf"), device = cairo_pdf)
```

### P(action) vs LM-derived access

```{r}
#| fig-width: 14
#| fig-asp: 0.45

df_canonical_access <- read_csv(here("model", "outputs", "lm_scenario_params_nonfood.csv"),
                                show_col_types = FALSE) |>
  select(scenario_label, action, access)

df_human_pts <- df |>
  group_by(scenario_label, intimacy, motivation, action) |>
  tidyboot_mean(p_action) |>
  ungroup() |>
  mutate(p_action = empirical_stat, model = "Humans") |>
  select(scenario_label, intimacy, motivation, action, model, p_action, ci_lower, ci_upper)

df_model_pts <- df_preds |>
  filter(model != "Humans") |>
  group_by(scenario_label, intimacy, motivation, action, model) |>
  summarize(p_action = mean(p_action), .groups = "drop") |>
  mutate(ci_lower = NA_real_, ci_upper = NA_real_)

df_access_combined <- bind_rows(df_human_pts, df_model_pts) |>
  inner_join(df_canonical_access, by = c("scenario_label", "action")) |>
  mutate(model = factor(model, levels = c("Base model", "Discomfort-only", "Full model", "Humans")))

p_paction_access <- df_access_combined |>
  ggplot(aes(x = access, y = p_action, color = intimacy, shape = factor(action), group = intimacy)) +
  geom_errorbar(
    data = df_access_combined |> filter(model == "Humans"),
    aes(ymin = ci_lower, ymax = ci_upper),
    position = POS_JITTER_DODGE,
    width = 0.05, alpha = 0.4, linewidth = 0.4
  ) +
  geom_point(position = POS_JITTER_DODGE, alpha = 0.5, size = 1.8) +
  geom_smooth(se = FALSE, linewidth = 1) +
  scale_color_intimacy() +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(y = "P(action)", x = "LM-derived access of action", color = "Intimacy", shape = "Action") +
  theme(legend.position = "right") +
  facet_grid(motivation ~ model)

print(p_paction_access)

ggsave(here("figures", "nonfood_forw_plan_paction_access.pdf"),
       plot = p_paction_access, device = cairo_pdf, width = 14, height = 6)
```

### Model–human correlation

```{r}
boot_cor <- function(x, y, n_boot = 1000) {
  complete <- complete.cases(x, y)
  x <- x[complete]; y <- y[complete]
  n <- length(x)
  if (n < 3) return(list(r = NA_real_, ci_lower = NA_real_, ci_upper = NA_real_))
  boot_rs <- replicate(n_boot, {
    idx <- sample(n, replace = TRUE)
    cor(x[idx], y[idx])
  })
  list(
    r = cor(x, y),
    ci_lower = as.numeric(quantile(boot_rs, 0.025, na.rm = TRUE)),
    ci_upper = as.numeric(quantile(boot_rs, 0.975, na.rm = TRUE))
  )
}

df_humans_summary <- df_preds_summary |>
  filter(model == "Humans") |>
  select(intimacy, motivation, action,
         Humans = empirical_stat,
         Humans_ci_lower = ci_lower,
         Humans_ci_upper = ci_upper)

df_corr <- df_preds_summary |>
  filter(model != "Humans") |>
  select(intimacy, motivation, action, model, model_pred = empirical_stat) |>
  left_join(df_humans_summary, by = c("intimacy", "motivation", "action")) |>
  filter(!is.na(model_pred), !is.na(Humans)) |>
  mutate(action = factor(action), model = droplevels(model))
```

```{r}
#| label: fig-nonfood-forw-plan-correlation
#| fig-width: 10
#| fig-asp: 0.4

correlations <- format_correlation_labels(df_corr, model_pred, Humans, group_vars = "model")

ggplot(df_corr, aes(x = model_pred, y = Humans, color = intimacy, shape = action)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
  geom_errorbar(aes(ymin = Humans_ci_lower, ymax = Humans_ci_upper), width = 0.02) +
  geom_point(size = 3, alpha = PLOT_ALPHA) +
  geom_label(
    data = correlations,
    aes(x = -Inf, y = Inf, label = label),
    inherit.aes = FALSE,
    hjust = -0.08, vjust = 1.5, size = 4.5,
    fill = "white", alpha = 0.5, label.size = 0
  ) +
  scale_color_intimacy() +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(x = "Model", y = "Humans", color = "Intimacy", shape = "Action") +
  facet_wrap(~ model, nrow = 1) +
  coord_fixed_symmetric(df_corr$model_pred, df_corr$Humans)

ggsave(here("figures", "nonfood_forw_plan_correlation.pdf"), device = cairo_pdf)
```

### All-data fit results (AIC / BIC)

```{r}
fit_results <- read_csv(here("model", "outputs", "forward_planning_fit_results_nonfood.csv"),
                        show_col_types = FALSE)

fit_results |>
  select(model, nll, n_params, starts_with("param_")) |>
  knitr::kable(digits = 3, caption = "Model fitting results (nonfood, all-data fit)")
```

```{r}
fit_results <- add_aic_bic(fit_results, n_obs = nrow(df))

fit_results |>
  select(model, nll, n_params, AIC, BIC) |>
  arrange(AIC) |>
  knitr::kable(digits = 2, caption = "Model comparison (lower AIC/BIC is better)")

best_aic <- fit_results$AIC[fit_results$model == "access_full"]
best_bic <- fit_results$BIC[fit_results$model == "access_full"]

fit_results |>
  mutate(delta_AIC = AIC - best_aic, delta_BIC = BIC - best_bic) |>
  select(model, delta_AIC, delta_BIC) |>
  knitr::kable(digits = 2, caption = "Difference from Full model")
```

## Residual diagnostics

How well does the canonical model capture nonfood scenario-level structure? The `scenario_type` heatmap below is the place to look for systematic miss-fits — the linear `access · (1 − I)` shape may struggle with privacy scenarios where norms are categorical.

### Setup

```{r}
df_cell <- df_preds |>
  filter(model %in% c("Humans", "Full model")) |>
  group_by(scenario_label, intimacy, motivation, action, model) |>
  summarize(p_action = mean(p_action), .groups = "drop") |>
  pivot_wider(names_from = model, values_from = p_action) |>
  rename(full_model = `Full model`, humans = Humans) |>
  mutate(residual = humans - full_model) |>
  inner_join(df_canonical_access, by = c("scenario_label", "action")) |>
  left_join(df_scenario_types, by = "scenario_label")
```

### Cross-scenario dispersion: humans vs Full model

```{r}
#| fig-width: 12
#| fig-asp: 0.45

df_dispersion <- df_cell |>
  group_by(intimacy, motivation, action) |>
  summarize(
    Humans = sd(humans),
    `Full model` = sd(full_model),
    .groups = "drop"
  ) |>
  pivot_longer(c(Humans, `Full model`), names_to = "source", values_to = "sd") |>
  mutate(source = factor(source, levels = c("Full model", "Humans")))

ggplot(df_dispersion, aes(x = factor(action), y = sd, fill = source)) +
  geom_col(position = position_dodge(width = 0.8), alpha = PLOT_ALPHA, width = 0.75) +
  scale_fill_manual(values = c("Full model" = "#6b8cae", "Humans" = "#b4584d")) +
  labs(x = "Action", y = "SD across scenarios of P(action)", fill = NULL) +
  facet_grid(motivation ~ intimacy, labeller = labeller(intimacy = label_both)) +
  theme(legend.position = "top")
```

### Per-scenario fit quality

```{r}
#| fig-width: 8
#| fig-asp: 0.7

df_scenario_fit <- df_cell |>
  group_by(scenario_label, scenario_type) |>
  summarize(
    r = cor(humans, full_model),
    rmse = sqrt(mean(residual^2)),
    .groups = "drop"
  ) |>
  arrange(r)

ggplot(df_scenario_fit, aes(x = reorder(scenario_label, r), y = r, fill = scenario_type)) +
  geom_col(alpha = PLOT_ALPHA) +
  geom_text(aes(label = sprintf("%.2f", r)), hjust = -0.15, size = 3.5) +
  coord_flip(ylim = c(min(df_scenario_fit$r) - 0.05, 1.05)) +
  labs(x = "Scenario", y = "Pearson r (Humans vs Full model)", fill = "Type")
```

### Mean residual per (scenario, action), grouped by scenario type

```{r}
#| fig-width: 7
#| fig-asp: 1.1

df_cell_action <- df_cell |>
  group_by(scenario_label, scenario_type, action) |>
  summarize(mean_residual = mean(residual), .groups = "drop")

scenario_order <- df_cell_action |>
  group_by(scenario_label, scenario_type) |>
  summarize(mabs = mean(abs(mean_residual)), .groups = "drop") |>
  arrange(scenario_type, mabs) |>
  pull(scenario_label)

df_cell_action |>
  mutate(
    scenario_label = factor(scenario_label, levels = scenario_order),
    action = factor(action)
  ) |>
  ggplot(aes(x = action, y = scenario_label, fill = mean_residual)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%+.2f", mean_residual)), size = 3) +
  scale_fill_gradient2(low = "#3b4cc0", mid = "white", high = "#b40426",
                       midpoint = 0,
                       limits = c(-1, 1) * max(abs(df_cell_action$mean_residual))) +
  labs(x = "Action", y = "Scenario", fill = "Residual\n(H − M)") +
  facet_grid(scenario_type ~ ., scales = "free_y", space = "free_y") +
  theme(panel.grid = element_blank())
```

### Residual vs access, by scenario type

If privacy scenarios systematically deviate from the linear `access · (1 − I)` shape, expect a non-flat smoother for that facet.

```{r}
#| fig-width: 12
#| fig-asp: 0.4

ggplot(df_cell, aes(x = access, y = residual, color = scenario_label, shape = factor(action),
                    group = scenario_label)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  geom_point(alpha = 0.3, size = 1.3) +
  geom_smooth(se = FALSE, linewidth = 0.7, method = "loess") +
  scale_color_viridis_d(option = "turbo") +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(x = "LM-derived access of action", y = "Residual (Humans − Full model)",
       color = "Scenario", shape = "Action") +
  theme(legend.position = "right") +
  facet_wrap(~ scenario_type)
```

## LM-elicited V — validity

V is now LM-elicited rather than stipulated. Two quick sanity checks below before reading the model fit. (Validity is at the LM-table level, not at the fit level — these don't depend on the fit at all.)

```{r}
df_lmv <- read_csv(here("model", "outputs", "lm_scenario_v_nonfood.csv"), show_col_types = FALSE) |>
  left_join(df_scenario_types, by = "scenario_label")
```

### Sign reversal: refrain should be ≥ 0 under LOW motivation, ≤ 0 under HIGH

```{r}
#| fig-width: 9
#| fig-asp: 0.45

df_refrain <- df_lmv |> filter(action == 0)

ggplot(df_refrain, aes(x = scenario_label, y = v, fill = motivation)) +
  geom_col(position = "dodge", alpha = 0.85) +
  geom_hline(yintercept = 0, color = "grey40") +
  scale_fill_manual(values = c(high = "#b4584d", low = "#6b8cae")) +
  labs(x = NULL, y = "V (refrain, action 0)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid(~ scenario_type, scales = "free_x", space = "free_x")
```

### V vs access collinearity (w_v vs w_d identifiability)

If V and access correlate strongly (|r| > 0.85), w_v and w_d won't be cleanly separable in the fit.

```{r}
#| fig-width: 8
#| fig-asp: 0.45

cor_v_acc <- df_v_vs_access <- df_lmv |>
  inner_join(df_canonical_access, by = c("scenario_label", "action"))

cor_v_acc <- df_v_vs_access |>
  group_by(motivation) |>
  summarize(r = cor(v, access), .groups = "drop")

ggplot(df_v_vs_access, aes(x = access, y = v, color = scenario_type, shape = factor(action))) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "grey40", linewidth = 0.5) +
  geom_label(data = cor_v_acc,
             aes(x = -Inf, y = Inf, label = sprintf("r = %.2f", r)),
             inherit.aes = FALSE,
             hjust = -0.1, vjust = 1.4, size = 4) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(x = "LM-derived access", y = "LM-V (signed valence)") +
  facet_wrap(~ motivation)
```
