---
title: "Study 3b — Joint inference over intimacy and effort (non-food)"
format:
  html:
    code-fold: true
    toc: true
    embed-resources: true
execute:
  echo: true
  warning: false
---

```{r setup, include=FALSE}
source(here::here("analysis", "utils.R"))
```

# Study 3b — Joint inference over intimacy and effort (non-food)

Study 2b's design on the non-food stimulus set: the observer jointly infers
relationship intimacy and effort from a single observed action, given desire.
Each trial has two slider responses: intimacy on a 0–100 numeric scale, and
effort as a continuous rating between two states.

**Status:** Full sample is in `data/nonfood_inv_joint_ie/` (recruited and
retained counts are reported by the demographics chunk below). This document
reports demographics and data checks only — all figures (the joint
belief-update panels and the model-vs-human panels) are generated in Python by
`figures/scripts/figure_study3.py` and `figures/scripts/figure_model_scatter.py`
(`make figures-results`), and the model-comparison statistics by
`model/cv/model_comparison.py` (`make model-comparison`).

## Demographics

```{r demographics}
if (file.exists(here::here("data", "nonfood_inv_joint_ie", "exit_survey.csv"))) {
  report_demographics("nonfood_inv_joint_ie")
} else {
  cat("No data yet for nonfood_inv_joint_ie.\n")
}
```

## Data

```{r load-data}
data_path <- here::here("data", "nonfood_inv_joint_ie", "main_trials_long.csv")
have_data <- file.exists(data_path)
if (have_data) {
  df <- read_csv(data_path, show_col_types = FALSE)
  cat("Trials:", nrow(df), "\n")
  glimpse(df)
} else {
  cat("No data yet for nonfood_inv_joint_ie.\n")
}
```

## Notes

See the manuscript Methods section for the exact design, plus
`experiments/nonfood_inv_joint_ie/README.md` for the trial structure spec.
